---
title: "Investigate ALStructure 2"
author: "Matthew Stephens"
date: 2018-04-05
output: workflowr::wflow_html
---




## Load package

```{r}
#library("devtools")
#install_github("storeylab/alstructure", build_vignettes=TRUE)
 library("alstructure")
```


Define a version that can run the factorization directly on `X`/2 instead of `F_hat`, and it also returns `F_hat` and `run_time` in the list for my convenience and allows a seed to be set (as the initialization for factorizing is random); otherwise this function is the same as 
`alstructure`.
```{r}
my_alstructure = function (X, d_hat = NULL, svd_method = "base", tol = 1e-05, 
    max_iters = 1000, order_method = "ave_admixture", P_init, 
    Q_init, useX = FALSE, seed = 1) 
{
    set.seed(seed)
    run_time <- proc.time()
    X <- alstructure:::impute_mean(X)
    if (is.null(d_hat)) {
        d_hat <- estimate_d(X)
    }
    F_obj <- estimate_F(X, d = d_hat, svd_method = svd_method)
    F_hat <- F_obj$F_hat
    rowspace <- F_obj$rowspace
    if(useX){
      factors <- factor_F(F = X/2, d = d_hat, tol = tol, max_iters = max_iters)} else {
      factors <- factor_F(F = F_hat, d = d_hat, tol = tol, max_iters = max_iters)  }#changed Fhat to X here!
    if (order_method == "ave_admixture") {
        ordering <- order_pops(factors$P_hat, factors$Q_hat, 
            method = "ave_admixture", Q_space = NULL)
        factors$Q_hat <- ordering$Q_ordered
        factors$P_hat <- ordering$P_ordered
    }
    else if (order_method == "var_explained") {
        ordering <- order_pops(factors$P_hat, factors$Q_hat, 
            method = "var_explained", Q_space = rowspace)
        factors$Q_hat <- ordering$Q_ordered
        factors$P_hat <- ordering$P_ordered
    }
    run_time <- proc.time() - run_time
    vals = list(P_hat = factors$P_hat, Q_hat = factors$Q_hat, 
        rowspace = rowspace, iter = factors$iter, tol = factors$tol, F_hat = F_hat, run_time=run_time)
    return(vals)
}
```

Simulation as from the vignette, but 30 populations and 300 individuals:
```{r}
m = 10000 # number of SNPS
n = 300 # number of individuals
d = 30 # dimension of latent subspace
alpha = rep(0.1,30)
seed = 12345

sim_data <- simulate_admixture(m, n, d, alpha, BN_params = NA, seed = seed)
```

See what happens if we used d=10 in fitting
```{r}
fit <- my_alstructure(X = sim_data$X, d_hat = 10)
fit2 <- my_alstructure(X = sim_data$X, d_hat = 10,useX=TRUE)

mean((sim_data$F - fit$P_hat %*% fit$Q_hat)^2)
mean((sim_data$F - fit2$P_hat %*% fit2$Q_hat)^2)

mean((sim_data$X - fit$P_hat %*% fit$Q_hat)^2)
mean((sim_data$X - fit2$P_hat %*% fit2$Q_hat)^2)

fit$run_time
fit2$run_time
```


So now things are getting more interesting...

One issue is that Fhat itself requires an estimate of $d$.
I'm currently giving it the true value, which might be an unfair advantage. I wonder if we can estimate $d$ instead.
```{r}
d_est = estimate_d(sim_data$X)
```

OK, so that didn't work!
Try flash:
```{r}
sim.f = flashr::flash(sim_data$X)
```


Questions: 
- can we make things faster by exploiting Fhat low rank?



