---
title: "Investigate ALStructure"
author: "Matthew Stephens"
date: 2018-04-05
output: html_document
---

<!-- The file analysis/chunks.R contains chunks that define default settings
shared across the workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

## Load package

```{r}
#library("devtools")
#install_github("storeylab/alstructure", build_vignettes=TRUE)
library("alstructure")
set.seed("12345")
```


Define a version that can run the factorization directly on `X`/2 instead of `F_hat`, and it also returns `F_hat` and `run_time` in the list for my convenience and allows a seed to be set (as the initialization for factorizing is random); otherwise this function is the same as 
`alstructure`.
```{r}
my_alstructure = function (X, d_hat = NULL, svd_method = "base", tol = 1e-05, 
    max_iters = 1000, order_method = "ave_admixture", P_init, 
    Q_init, useX = FALSE, seed = 1) 
{
    set.seed(seed)
    run_time <- proc.time()
    X <- alstructure:::impute_mean(X)
    if (is.null(d_hat)) {
        d_hat <- estimate_d(X)
    }
    F_obj <- estimate_F(X, d = d_hat, svd_method = svd_method)
    F_hat <- F_obj$F_hat
    rowspace <- F_obj$rowspace
    if(useX){
      factors <- factor_F(F = X/2, d = d_hat, tol = tol, max_iters = max_iters)} else {
      factors <- factor_F(F = F_hat, d = d_hat, tol = tol, max_iters = max_iters)  }
    if (order_method == "ave_admixture") {
        ordering <- order_pops(factors$P_hat, factors$Q_hat, 
            method = "ave_admixture", Q_space = NULL)
        factors$Q_hat <- ordering$Q_ordered
        factors$P_hat <- ordering$P_ordered
    }
    else if (order_method == "var_explained") {
        ordering <- order_pops(factors$P_hat, factors$Q_hat, 
            method = "var_explained", Q_space = rowspace)
        factors$Q_hat <- ordering$Q_ordered
        factors$P_hat <- ordering$P_ordered
    }
    run_time <- proc.time() - run_time
    vals = list(P_hat = factors$P_hat, Q_hat = factors$Q_hat, 
        rowspace = rowspace, iter = factors$iter, tol = factors$tol, F_hat = F_hat, run_time=run_time)
    return(vals)
}
```

## Simulations

Simulation as from the vignette:
```{r}
m = 10000 # number of SNPS
n = 100 # number of individuals
d = 3 # dimension of latent subspace
alpha = c(0.1, 0.1, 0.1) # dirichlet parameters
seed = 12345

sim_data <- simulate_admixture(m, n, d, alpha, BN_params = NA, seed = seed)

names(sim_data)
```


```{r}
fit <- my_alstructure(X = sim_data$X, d_hat = d)
fit2 <- my_alstructure(X = sim_data$X, d_hat = d, useX=TRUE)
```

The two lead to almost identical errors and similar run-times:
```{r}
mean((sim_data$F - fit$P_hat %*% fit$Q_hat)^2)
mean((sim_data$F - fit2$P_hat %*% fit2$Q_hat)^2)
fit$run_time
fit2$run_time
```

Now try a harder example with 10 populations:
```{r}
m = 10000 # number of SNPS
n = 100 # number of individuals
d = 10 # dimension of latent subspace
alpha = rep(0.1, d) # dirichlet parameters
seed = 12345

sim_data <- simulate_admixture(m, n, d, alpha, BN_params = NA, seed = seed)
fit <- my_alstructure(X = sim_data$X, d_hat = d)
fit2 <- my_alstructure(X = sim_data$X, d_hat = d,useX=TRUE)

mean((sim_data$F - fit$P_hat %*% fit$Q_hat)^2)
mean((sim_data$F - fit2$P_hat %*% fit2$Q_hat)^2)
fit$run_time
fit2$run_time
```


Now try a harder example with 30 populations:
```{r}
d = 30 # dimension of latent subspace
n=300
alpha = rep(0.1, d) # dirichlet parameters
seed = 12345

sim_data <- simulate_admixture(m, n, d, alpha, BN_params = NA, seed = seed)

fit <- my_alstructure(X = sim_data$X, d_hat = d)
fit2 <- my_alstructure(X = sim_data$X, d_hat = d,useX=TRUE)

mean((sim_data$F - fit$P_hat %*% fit$Q_hat)^2)
mean((sim_data$F - fit2$P_hat %*% fit2$Q_hat)^2)

mean((sim_data$X - fit$P_hat %*% fit$Q_hat)^2)
mean((sim_data$X - fit2$P_hat %*% fit2$Q_hat)^2)

fit$run_time
fit2$run_time
```


So now things are getting more interesting... the fits are
similar but the dimension reduction has reduced run time.


One issue is that Fhat itself requires an estimate of $d$.
I'm currently giving it the true value, which might be an unfair advantage. I wonder if we can estimate $d$ instead.
```{r}
d_est = estimate_d(sim_data$X)
```

OK, so that didn't work! 
Let's just see what happens with 16.


```{r}
fit <- my_alstructure(X = sim_data$X, d_hat = 16)
fit2 <- my_alstructure(X = sim_data$X, d_hat = 16,useX=TRUE)
mean((sim_data$F - fit$P_hat %*% fit$Q_hat)^2)
mean((sim_data$F - fit2$P_hat %*% fit2$Q_hat)^2)

mean((sim_data$X - fit$P_hat %*% fit$Q_hat)^2)
mean((sim_data$X - fit2$P_hat %*% fit2$Q_hat)^2)

fit$run_time
fit2$run_time
```

So I guess more investigation is required to see if
the run time is consistently improved....


## Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
