---
title: "ash_delta_operator"
author: "Matthew Stephens"
date: "2019-10-07"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

Consider the posterior mean of ash in the case with constant standard errors (say 1) as a shrinkage operator $s(x;g)$ where the operator
obviously depends on $g$. 

Now define $\delta(x)$ to be the value (or a value, if not uniquely defined) such that
$s(x+\delta(x)) = x$. The motivation is that if you add $\delta(x)$
to $x$ and shrink, you get back to x.

That is, $\delta(x)= s^{-1}(x)-x$. 

For soft-thresholding $s(x) = (x-c)_+$, and $\delta(x) = 0$ for $x<c$
and $\delta(x) = c$ for $x>c$ works. 

Here I look at what the form of $\delta(x)$ looks like for ash.
I'm particularly interested in whether $\delta$ is increasing
because if it is I think I can approximate ash solutions to linear regression using a linear combination of lasso solutions.

```{r}
library(ashr)
k=10 # number of mixture components in ash prior, g
maxsd = 10 # maximum sd in g
g = normalmix(rep(1/k,k),rep(0,k),(maxsd/(2^k)) * 2^(1:k))
x = seq(0,10,length=100)
x.a = ashr::ash.workhorse(x,1,g=g,fixg=TRUE)
plot(x,get_pm(x.a))
abline(a=0,b=1)
```

Now plot delta directly.
```{r}
z = get_pm(x.a) #so s^{-1}(z)=x, so delta(z) = x-z
plot(z,x-z)
```

So delta is not increasing. 

