---
title: "Comparing fastICA and alternating projection for binary group recovery"
author: "Matthew Stephens"
date: "2026-08-15"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

We compare two approaches for finding binary group structure in whitened data:

1. **Non-centered fastICA** (from `fastica_centered_02.Rmd`): maximizes the log-cosh objective over unit-norm weight vectors on whitened data with a row of 1s prepended as an intercept.

2. **Alternating projection**: iteratively projects a binary vector onto the subspace spanned by the data, then rounds back to ±1.  Specifically, each update is `v <- sign(U %*% (t(U) %*% v))` where U has orthonormal columns spanning the data subspace.

Both methods receive the same preprocessed data: centered and whitened `X1` with a row of 1s prepended to form `X1_aug`.  For the alternating projection, `U = t(X1_aug) / sqrt(n)`, which has orthonormal columns because `X1_aug %*% t(X1_aug) = n * I` (whitening plus centering makes the intercept row orthogonal to all whitened rows).

For each simulation setting (taken from `fastica_centered_02.Rmd`) we run each method from **100 random starting points** and report the fraction of starts that produce a solution correlating >0.95 with at least one true group column.

## Helper functions

```{r setup}
# X is (n.comp+1) x n.  W is (n.comp+1) x n_starts (one weight vector per column).
# P = t(X) %*% W is n x n_starts (source estimates).
fastica_update = function(X, W) {
  P  <- t(X) %*% W          # n x n_starts: source estimates
  G  <- tanh(P)
  G2 <- 1 - G^2
  W  <- X %*% G - sweep(W, 2, colSums(G2), "*")
  sweep(W, 2, sqrt(colSums(W^2)), "/")
}

# X is (n.comp+1) x n.  V is n x n_starts (one binary vector per column).
# P = X %*% V / n is (n.comp+1) x n_starts (subspace coefficients).
alt_proj_update = function(X, V) {
  n <- ncol(X)
  P <- X %*% V / n           # (n.comp+1) x n_starts: subspace coefficients
  V <- sign(t(X) %*% P)     # n x n_starts: project back and round
  V[V == 0] <- 1L
  V
}

# Center columns of X, whiten to n.comp dimensions.
preprocess = function(X, n.comp = 10) {
  X <- scale(X, scale = FALSE)
  sqrt(nrow(X)) * t(svd(X)$u[, 1:n.comp])
}
```

## Comparison functions

```{r compare-fns}
max_cor_with_truth = function(v, L) max(abs(cor(v, L)))

run_fastica = function(X1_aug, L, n_starts = 100, n_iter = 50, threshold = 0.95) {
  set.seed(1)
  W <- matrix(rnorm(nrow(X1_aug) * n_starts), nrow(X1_aug), n_starts)
  W <- sweep(W, 2, sqrt(colSums(W^2)), "/")
  for (i in seq_len(n_iter))
    W <- fastica_update(X1_aug, W)
  Lhat <- t(X1_aug) %*% W  # n x n_starts
  success <- sum(apply(Lhat, 2, function(lhat) {
    mc <- max_cor_with_truth(lhat, L)
    !is.na(mc) && mc > threshold
  }))
  success / n_starts
}

run_alt_proj = function(X1_aug, L, n_starts = 100, n_iter = 50, threshold = 0.95) {
  n <- ncol(X1_aug)
  set.seed(1)
  V <- sign(matrix(rnorm(n * n_starts), n, n_starts))
  V[V == 0] <- 1L
  for (i in seq_len(n_iter))
    V <- alt_proj_update(X1_aug, V)
  success <- sum(apply(V, 2, function(v) {
    mc <- max_cor_with_truth(v, L)
    !is.na(mc) && mc > threshold
  }))
  success / n_starts
}

run_comparison = function(X, L, n.comp, label, n_starts = 100, n_iter = 50) {
  X1     <- preprocess(X, n.comp = n.comp)
  X1_aug <- rbind(rep(1, ncol(X1)), X1)

  cat("Running fastICA for:", label, "\n")
  ica_time <- system.time(ica_rate <- run_fastica(X1_aug, L, n_starts = n_starts, n_iter = n_iter))

  cat("Running alt-proj for:", label, "\n")
  ap_time  <- system.time(ap_rate  <- run_alt_proj(X1_aug, L, n_starts = n_starts, n_iter = n_iter))

  cat(sprintf("  fastICA: %.0f%% (%.0fms)  AltProj: %.0f%% (%.0fms)\n",
              100 * ica_rate, 1000 * ica_time["elapsed"],
              100 * ap_rate,  1000 * ap_time["elapsed"]))
  c(fastICA = ica_rate, AltProj = ap_rate,
    fastICA_ms = 1000 * ica_time["elapsed"], AltProj_ms = 1000 * ap_time["elapsed"])
}
```

## Simulation 1: 3 overlapping groups

Three groups, 20 members each out of n=100; groups may overlap. Background is -1, group members are +1.

```{r sim1-3groups}
K <- 3; p <- 1000; n <- 100
set.seed(1)
L <- matrix(-1, nrow = n, ncol = K)
for (i in 1:K) { L[sample(1:n, 20), i] <- 1 }
FF <- matrix(rnorm(p * K), nrow = p, ncol = K)
X  <- L %*% t(FF) + rnorm(n * p, 0, 0.01)

res1 <- run_comparison(X, L, n.comp = 10, label = "3 overlapping groups (n=100, K=3)")
```

## Simulation 2: 9 overlapping groups

Nine groups, 20 members each out of n=100. Background is 0, group members are +1.

```{r sim2-9groups}
K <- 9; p <- 1000; n <- 100
set.seed(1)
L <- matrix(0, nrow = n, ncol = K)
for (i in 1:K) { L[sample(1:n, 20), i] <- 1 }
FF <- matrix(rnorm(p * K), nrow = p, ncol = K)
X  <- L %*% t(FF) + rnorm(n * p, 0, 0.01)

res2 <- run_comparison(X, L, n.comp = 10, label = "9 overlapping groups (n=100, K=9)")
```

## Simulation 3: 4 non-overlapping groups

Four equal non-overlapping groups of 25 (n=100). Using n.comp=10.


```{r sim3-4nonoverlap}
n <- 100; p <- 1000
set.seed(1)
L <- matrix(0, nrow = n, ncol = 4)
L[1:25,   1] <- 1
L[26:50,  2] <- 1
L[51:75,  3] <- 1
L[76:100, 4] <- 1
FF <- matrix(rnorm(p * 4), nrow = p)
X  <- L %*% t(FF) + rnorm(n * p, 0, 0.01)

res3 <- run_comparison(X, L, n.comp = 10, label = "4 non-overlapping groups (n=100)")
```

## Simulation 4: 9 non-overlapping groups

Nine equal non-overlapping groups of 25 (n=225). Using n.comp=20.

```{r sim4-9nonoverlap}
n <- 225; p <- 1000
set.seed(1)
L <- matrix(0, nrow = n, ncol = 9)
for (i in 1:9) { L[((i - 1) * 25 + 1):(i * 25), i] <- 1 }
FF <- matrix(rnorm(p * 9), nrow = p)
X  <- L %*% t(FF) + rnorm(n * p, 0, 0.01)

res4 <- run_comparison(X, L, n.comp = 20, label = "9 non-overlapping groups (n=225)")
```

## Simulation 5: Bifurcating tree

A 4-leaf bifurcating tree with 6 binary branches (n=100). Using n.comp=10 (fewer PCs reduces noise, as found in `fastica_centered_02.Rmd`).

```{r sim5-tree}
n <- 100; p <- 1000
set.seed(1)
L <- matrix(0, nrow = n, ncol = 6)
L[1:50,   1] <- 1
L[51:100, 2] <- 1
L[1:25,   3] <- 1
L[26:50,  4] <- 1
L[51:75,  5] <- 1
L[76:100, 6] <- 1
FF <- matrix(rnorm(p * 6), nrow = p)
X  <- L %*% t(FF) + rnorm(n * p, 0, 0.01)

res5 <- run_comparison(X, L, n.comp = 10, label = "Bifurcating tree (n=100, 6 branches)")
```

## Diagnosing the 9 non-overlapping groups case

Is alternating projection finding combinations of multiple groups rather than single groups? We re-run both methods on sim 4, store all 100 result vectors, and examine the absolute correlation of each result with each of the 9 true group columns.

```{r diag-9nonoverlap}
n <- 225; p <- 1000; n_starts <- 100; n_iter <- 50
set.seed(1)
L9 <- matrix(0, nrow = n, ncol = 9)
for (i in 1:9) L9[((i-1)*25+1):(i*25), i] <- 1
FF <- matrix(rnorm(p * 9), nrow = p)
X9 <- L9 %*% t(FF) + rnorm(n * p, 0, 0.01)

X1     <- preprocess(X9, n.comp = 20)
X1_aug <- rbind(rep(1, ncol(X1)), X1)

# fastICA: store all weight vectors
set.seed(1)
W <- matrix(rnorm(nrow(X1_aug) * n_starts), nrow(X1_aug), n_starts)
W <- sweep(W, 2, sqrt(colSums(W^2)), "/")
for (i in seq_len(n_iter)) W <- fastica_update(X1_aug, W)
Lhat_ica <- t(X1_aug) %*% W   # n x n_starts

# alt_proj: store all binary vectors
set.seed(1)
V <- sign(matrix(rnorm(n * n_starts), n, n_starts))
V[V == 0] <- 1L
for (i in seq_len(n_iter)) V <- alt_proj_update(X1_aug, V)
Lhat_ap <- V                   # n x n_starts

# Absolute correlation of each result with each true group: n_starts x 9
cor_ica <- abs(cor(Lhat_ica, L9))
cor_ap  <- abs(cor(Lhat_ap,  L9))
```

For each start, we look at how many groups it correlates with above 0.5 — a single-group solution scores 1, a combination scores 2 or more.

```{r diag-ngroups}
n_groups_hit = function(cor_mat, thresh = 0.5) rowSums(cor_mat > thresh)

tab_ica <- table(n_groups_hit(cor_ica))
tab_ap  <- table(n_groups_hit(cor_ap))

cat("fastICA — number of groups with |cor| > 0.5:\n"); print(tab_ica)
cat("AltProj — number of groups with |cor| > 0.5:\n"); print(tab_ap)
```

We can also check whether the fastICA objective is higher for single-group solutions than for combos, using the weight vectors W already computed.

```{r diag-objective}
# fastICA objective for each start
P <- t(X1_aug) %*% W                         # n x n_starts
obj_ica <- colMeans(log(cosh(P)))

n_hit <- n_groups_hit(cor_ica)
boxplot(obj_ica ~ n_hit,
        xlab = "Number of groups with |cor| > 0.5",
        ylab = "fastICA objective  mean(log cosh(Xw))",
        main = "fastICA objective by solution type (9 non-overlapping groups)")
```

Heatmap of absolute correlations for each start (rows = starts sorted by best single-group correlation, columns = true groups):

```{r diag-heatmap, fig.width=8, fig.height=5}
par(mfrow = c(1, 2))

order_ica <- order(apply(cor_ica, 1, max), decreasing = TRUE)
image(t(cor_ica[order_ica, ]), zlim = c(0, 1), axes = FALSE,
      main = "fastICA", xlab = "True group", ylab = "Start (sorted)")
axis(1, at = seq(0, 1, length.out = 9), labels = 1:9)

order_ap <- order(apply(cor_ap, 1, max), decreasing = TRUE)
image(t(cor_ap[order_ap, ]), zlim = c(0, 1), axes = FALSE,
      main = "AltProj", xlab = "True group", ylab = "Start (sorted)")
axis(1, at = seq(0, 1, length.out = 9), labels = 1:9)

par(mfrow = c(1, 1))
```

## Summary

```{r summary-table}
results <- data.frame(
  Simulation = c(
    "3 overlapping groups (n=100)",
    "9 overlapping groups (n=100)",
    "4 non-overlapping groups (n=100)",
    "9 non-overlapping groups (n=225)",
    "Bifurcating tree (n=100, 6 branches)"
  ),
  n.comp = c(10, 10, 10, 20, 10),
  fastICA_pct = round(100 * c(res1["fastICA"], res2["fastICA"], res3["fastICA"],
                               res4["fastICA"], res5["fastICA"])),
  fastICA_ms  = round(c(res1["fastICA_ms.elapsed"], res2["fastICA_ms.elapsed"],
                         res3["fastICA_ms.elapsed"], res4["fastICA_ms.elapsed"],
                         res5["fastICA_ms.elapsed"])),
  AltProj_pct = round(100 * c(res1["AltProj"], res2["AltProj"], res3["AltProj"],
                               res4["AltProj"], res5["AltProj"])),
  AltProj_ms  = round(c(res1["AltProj_ms.elapsed"], res2["AltProj_ms.elapsed"],
                         res3["AltProj_ms.elapsed"], res4["AltProj_ms.elapsed"],
                         res5["AltProj_ms.elapsed"]))
)

knitr::kable(results,
  col.names = c("Simulation", "n.comp",
                "fastICA (% success)", "fastICA (ms)",
                "AltProj (% success)", "AltProj (ms)"),
  caption   = "Fraction of 100 random starts recovering a solution with max correlation > 0.95 with any true group column, with wall-clock time in milliseconds."
)
```
