---
title: "Comparing fastICA and alternating projection for binary group recovery"
author: "Matthew Stephens"
date: "2026-08-15"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

We compare two approaches for finding binary group structure in whitened data:

1. **Non-centered fastICA** (from `fastica_centered_02.Rmd`): maximizes the log-cosh objective over unit-norm weight vectors on whitened data with a row of 1s prepended as an intercept.

2. **Alternating projection**: iteratively projects a binary vector onto the subspace spanned by the data, then rounds back to ±1.  Specifically, each update is `v <- sign(U %*% (t(U) %*% v))` where U has orthonormal columns spanning the data subspace.

Both methods receive the same preprocessed data: centered and whitened `X1` with a row of 1s prepended to form `X1_aug`.  For the alternating projection, `U = t(X1_aug) / sqrt(n)`, which has orthonormal columns because `X1_aug %*% t(X1_aug) = n * I` (whitening plus centering makes the intercept row orthogonal to all whitened rows).

For each simulation setting (taken from `fastica_centered_02.Rmd`) we run each method from **100 random starting points** and report the fraction of starts that produce a solution correlating >0.99 with at least one true group column.

## Helper functions

```{r setup}
# X is (n.comp+1) x n.  W is (n.comp+1) x n_starts (one weight vector per column).
# P = t(X) %*% W is n x n_starts (source estimates).
fastica_update = function(X, W) {
  P  <- t(X) %*% W          # n x n_starts: source estimates
  G  <- tanh(P)
  G2 <- 1 - G^2
  W  <- X %*% G - sweep(W, 2, colSums(G2), "*")
  sweep(W, 2, sqrt(colSums(W^2)), "/")
}

# X is (n.comp+1) x n.  V is n x n_starts (one binary vector per column).
# P = X %*% V / n is (n.comp+1) x n_starts (subspace coefficients).
alt_proj_update = function(X, V) {
  n <- ncol(X)
  P <- X %*% V / n           # (n.comp+1) x n_starts: subspace coefficients
  V <- sign(t(X) %*% P)     # n x n_starts: project back and round
  V[V == 0] <- 1L
  V
}

# Center columns of X, whiten to n.comp dimensions.
preprocess = function(X, n.comp = 10) {
  X <- scale(X, scale = FALSE)
  sqrt(nrow(X)) * t(svd(X)$u[, 1:n.comp])
}
```

## Comparison functions

```{r compare-fns}
max_cor_with_truth = function(v, L) max(abs(cor(v, L)))

vPv_scores = function(X1_aug, V) {
  # v'Pv = ||X1_aug v||^2 / n  for each column of V (binary +-1 vectors)
  colSums((X1_aug %*% V)^2) / ncol(X1_aug)
}

# Gradient-only step: omits the Hessian correction, so each step is gradient
# ascent on the log-cosh objective rather than a Newton step.
fastica_gradient_update = function(X, W) {
  P <- t(X) %*% W
  G <- tanh(P)
  W <- X %*% G   # gradient only — no colSums(G2) * W term
  sweep(W, 2, sqrt(colSums(W^2)), "/")
}

run_fastica = function(X1_aug, L, n_starts = 100, n_iter = 50) {
  set.seed(1)
  W <- matrix(rnorm(nrow(X1_aug) * n_starts), nrow(X1_aug), n_starts)
  W <- sweep(W, 2, sqrt(colSums(W^2)), "/")
  for (i in seq_len(n_iter))
    W <- fastica_update(X1_aug, W)
  Lhat <- t(X1_aug) %*% W  # n x n_starts
  # Centre each column before binarising: handles asymmetric (0/1) group encoding.
  # Also flag near-degenerate columns (tiny sd) before centering amplifies noise.
  near_const <- apply(Lhat, 2, sd) < 1e-3
  V <- sign(sweep(Lhat, 2, colMeans(Lhat))); V[V == 0] <- 1L
  degenerate        <- near_const | apply(V, 2, function(v) all(v == v[1]))
  max_cors          <- apply(V, 2, max_cor_with_truth, L = L)
  max_cors[degenerate] <- 0  # degenerate starts count as failures
  list(max_cors = max_cors, vPv = vPv_scores(X1_aug, V),
       prop_degenerate = mean(degenerate))
}

# Hybrid: n_grad gradient steps (ascent) then Newton steps for convergence.
run_fastica_hybrid = function(X1_aug, L, n_starts = 100, n_iter = 50, n_grad = 5) {
  set.seed(1)
  W <- matrix(rnorm(nrow(X1_aug) * n_starts), nrow(X1_aug), n_starts)
  W <- sweep(W, 2, sqrt(colSums(W^2)), "/")
  for (i in seq_len(n_grad))
    W <- fastica_gradient_update(X1_aug, W)
  for (i in seq_len(n_iter - n_grad))
    W <- fastica_update(X1_aug, W)
  Lhat <- t(X1_aug) %*% W
  near_const <- apply(Lhat, 2, sd) < 1e-3
  V <- sign(sweep(Lhat, 2, colMeans(Lhat))); V[V == 0] <- 1L
  degenerate        <- near_const | apply(V, 2, function(v) all(v == v[1]))
  max_cors          <- apply(V, 2, max_cor_with_truth, L = L)
  max_cors[degenerate] <- 0
  list(max_cors = max_cors, vPv = vPv_scores(X1_aug, V),
       prop_degenerate = mean(degenerate))
}

run_alt_proj = function(X1_aug, L, n_starts = 100, n_iter = 50) {
  n <- ncol(X1_aug)
  set.seed(1)
  V <- sign(matrix(rnorm(n * n_starts), n, n_starts))
  V[V == 0] <- 1L
  for (i in seq_len(n_iter))
    V <- alt_proj_update(X1_aug, V)
  degenerate        <- apply(V, 2, function(v) all(v == v[1]))
  max_cors          <- apply(V, 2, max_cor_with_truth, L = L)
  max_cors[degenerate] <- 0
  list(max_cors = max_cors, vPv = vPv_scores(X1_aug, V),
       prop_degenerate = mean(degenerate))
}

run_comparison = function(X, L, n.comp, label, n_starts = 100, n_iter = 50) {
  X1     <- preprocess(X, n.comp = n.comp)
  X1_aug <- rbind(rep(1, ncol(X1)), X1)
  # Convert each L column to +-1 by thresholding at the column mean
  L_bin  <- apply(L, 2, function(col) { v <- sign(col - mean(col)); v[v == 0] <- 1L; v })
  true_vPv <- min(vPv_scores(X1_aug, L_bin))

  cat("Running fastICA for:", label, "\n")
  ica_time <- system.time(ica <- run_fastica(X1_aug, L, n_starts = n_starts, n_iter = n_iter))

  cat("Running fastICA-hybrid for:", label, "\n")
  hyb_time <- system.time(hyb <- run_fastica_hybrid(X1_aug, L, n_starts = n_starts, n_iter = n_iter))

  cat("Running alt-proj for:", label, "\n")
  ap_time  <- system.time(ap  <- run_alt_proj(X1_aug, L, n_starts = n_starts, n_iter = n_iter))

  ica_approx <- mean(ica$max_cors > 0.95)
  ica_exact  <- mean(ica$max_cors > 0.99)
  hyb_approx <- mean(hyb$max_cors > 0.95)
  hyb_exact  <- mean(hyb$max_cors > 0.99)
  ap_approx  <- mean(ap$max_cors  > 0.95)
  ap_exact   <- mean(ap$max_cors  > 0.99)
  ica_beat   <- mean(ica$vPv >= true_vPv - 1e-10)
  hyb_beat   <- mean(hyb$vPv >= true_vPv - 1e-10)
  ap_beat    <- mean(ap$vPv  >= true_vPv - 1e-10)
  cat(sprintf("  ICA: approx=%.0f%% exact=%.0f%% degen=%.0f%%  Hybrid: approx=%.0f%% exact=%.0f%% degen=%.0f%%  AP: approx=%.0f%% exact=%.0f%% degen=%.0f%%\n",
              100*ica_approx, 100*ica_exact, 100*ica$prop_degenerate,
              100*hyb_approx, 100*hyb_exact, 100*hyb$prop_degenerate,
              100*ap_approx,  100*ap_exact,  100*ap$prop_degenerate))
  c(fastICA_approx = ica_approx, fastICA_exact = ica_exact,
    fastICA_ms = 1000 * ica_time[["elapsed"]], fastICA_beat = ica_beat,
    fastICA_degen = ica$prop_degenerate,
    Hybrid_approx = hyb_approx,   Hybrid_exact = hyb_exact,
    Hybrid_ms = 1000 * hyb_time[["elapsed"]], Hybrid_beat = hyb_beat,
    Hybrid_degen = hyb$prop_degenerate,
    AltProj_approx = ap_approx,  AltProj_exact = ap_exact,
    AltProj_ms = 1000 * ap_time[["elapsed"]],  AltProj_beat = ap_beat,
    AltProj_degen = ap$prop_degenerate,
    true_vPv = true_vPv)
}
```

# Low-noise simulations (noise_sd = 0.01)

## Simulation 1: 3 overlapping groups

Three groups, 20 members each out of n=100; groups may overlap. Background is -1, group members are +1.

```{r sim1-3groups}
K <- 3; p <- 1000; n <- 100
set.seed(1)
L <- matrix(-1, nrow = n, ncol = K)
for (i in 1:K) { L[sample(1:n, 20), i] <- 1 }
FF <- matrix(rnorm(p * K), nrow = p, ncol = K)
X  <- L %*% t(FF) + rnorm(n * p, 0, 0.01)

res1 <- run_comparison(X, L, n.comp = 10, label = "3 overlapping groups (n=100, K=3)")
```

## Simulation 2: 9 overlapping groups

Nine groups, 20 members each out of n=100. Background is 0, group members are +1.

```{r sim2-9groups}
K <- 9; p <- 1000; n <- 100
set.seed(1)
L <- matrix(0, nrow = n, ncol = K)
for (i in 1:K) { L[sample(1:n, 20), i] <- 1 }
FF <- matrix(rnorm(p * K), nrow = p, ncol = K)
X  <- L %*% t(FF) + rnorm(n * p, 0, 0.01)

res2 <- run_comparison(X, L, n.comp = 10, label = "9 overlapping groups (n=100, K=9)")
```

## Simulation 3: 4 non-overlapping groups

Four equal non-overlapping groups of 25 (n=100). Using n.comp=10.


```{r sim3-4nonoverlap}
n <- 100; p <- 1000
set.seed(1)
L <- matrix(0, nrow = n, ncol = 4)
L[1:25,   1] <- 1
L[26:50,  2] <- 1
L[51:75,  3] <- 1
L[76:100, 4] <- 1
FF <- matrix(rnorm(p * 4), nrow = p)
X  <- L %*% t(FF) + rnorm(n * p, 0, 0.01)

res3 <- run_comparison(X, L, n.comp = 10, label = "4 non-overlapping groups (n=100)")
```

## Simulation 4: 9 non-overlapping groups

Nine equal non-overlapping groups of 25 (n=225). Using n.comp=20.

```{r sim4-9nonoverlap}
n <- 225; p <- 1000
set.seed(1)
L <- matrix(0, nrow = n, ncol = 9)
for (i in 1:9) { L[((i - 1) * 25 + 1):(i * 25), i] <- 1 }
FF <- matrix(rnorm(p * 9), nrow = p)
X  <- L %*% t(FF) + rnorm(n * p, 0, 0.01)

res4 <- run_comparison(X, L, n.comp = 20, label = "9 non-overlapping groups (n=225)")
```

## Simulation 5: Bifurcating tree

A 4-leaf bifurcating tree with 6 binary branches (n=100). Using n.comp=10 (fewer PCs reduces noise, as found in `fastica_centered_02.Rmd`).

```{r sim5-tree}
n <- 100; p <- 1000
set.seed(1)
L <- matrix(0, nrow = n, ncol = 6)
L[1:50,   1] <- 1
L[51:100, 2] <- 1
L[1:25,   3] <- 1
L[26:50,  4] <- 1
L[51:75,  5] <- 1
L[76:100, 6] <- 1
FF <- matrix(rnorm(p * 6), nrow = p)
X  <- L %*% t(FF) + rnorm(n * p, 0, 0.01)

res5 <- run_comparison(X, L, n.comp = 10, label = "Bifurcating tree (n=100, 6 branches)")
```

## Diagnosing the 9 non-overlapping groups case

Is alternating projection finding combinations of multiple groups rather than single groups? We re-run both methods on sim 4, store all 100 result vectors, and examine the absolute correlation of each result with each of the 9 true group columns.

```{r diag-9nonoverlap}
n <- 225; p <- 1000; n_starts <- 100; n_iter <- 50
set.seed(1)
L9 <- matrix(0, nrow = n, ncol = 9)
for (i in 1:9) L9[((i-1)*25+1):(i*25), i] <- 1
FF <- matrix(rnorm(p * 9), nrow = p)
X9 <- L9 %*% t(FF) + rnorm(n * p, 0, 0.01)

X1     <- preprocess(X9, n.comp = 20)
X1_aug <- rbind(rep(1, ncol(X1)), X1)

# fastICA: store all weight vectors
set.seed(1)
W <- matrix(rnorm(nrow(X1_aug) * n_starts), nrow(X1_aug), n_starts)
W <- sweep(W, 2, sqrt(colSums(W^2)), "/")
for (i in seq_len(n_iter)) W <- fastica_update(X1_aug, W)
Lhat_ica <- t(X1_aug) %*% W   # n x n_starts

# alt_proj: store all binary vectors
set.seed(1)
V <- sign(matrix(rnorm(n * n_starts), n, n_starts))
V[V == 0] <- 1L
for (i in seq_len(n_iter)) V <- alt_proj_update(X1_aug, V)
Lhat_ap <- V                   # n x n_starts

# Absolute correlation of each result with each true group: n_starts x 9
cor_ica <- abs(cor(Lhat_ica, L9))
cor_ap  <- abs(cor(Lhat_ap,  L9))
```

For each start, we look at how many groups it correlates with above 0.5 — a single-group solution scores 1, a combination scores 2 or more.

```{r diag-ngroups}
n_groups_hit = function(cor_mat, thresh = 0.5) rowSums(cor_mat > thresh)

tab_ica <- table(n_groups_hit(cor_ica))
tab_ap  <- table(n_groups_hit(cor_ap))

cat("fastICA — number of groups with |cor| > 0.5:\n"); print(tab_ica)
cat("AltProj — number of groups with |cor| > 0.5:\n"); print(tab_ap)
```

We can also check whether the fastICA objective is higher for single-group solutions than for combos, using the weight vectors W already computed.

```{r diag-objective}
# fastICA objective for each start
P <- t(X1_aug) %*% W                         # n x n_starts
obj_ica <- colMeans(log(cosh(P)))

n_hit <- n_groups_hit(cor_ica)
boxplot(obj_ica ~ n_hit,
        xlab = "Number of groups with |cor| > 0.5",
        ylab = "fastICA objective  mean(log cosh(Xw))",
        main = "fastICA objective by solution type (9 non-overlapping groups)")
```

Heatmap of absolute correlations for each start (rows = starts sorted by best single-group correlation, columns = true groups):

```{r diag-heatmap, fig.width=8, fig.height=5}
par(mfrow = c(1, 2))

order_ica <- order(apply(cor_ica, 1, max), decreasing = TRUE)
image(t(cor_ica[order_ica, ]), zlim = c(0, 1), axes = FALSE,
      main = "fastICA", xlab = "True group", ylab = "Start (sorted)")
axis(1, at = seq(0, 1, length.out = 9), labels = 1:9)

order_ap <- order(apply(cor_ap, 1, max), decreasing = TRUE)
image(t(cor_ap[order_ap, ]), zlim = c(0, 1), axes = FALSE,
      main = "AltProj", xlab = "True group", ylab = "Start (sorted)")
axis(1, at = seq(0, 1, length.out = 9), labels = 1:9)

par(mfrow = c(1, 1))
```

## Diagnosing the 3 overlapping groups case

AltProj achieves a higher mean v'Pv than fastICA in this setting despite a much lower success rate, suggesting it is finding high-energy binary vectors that do not isolate single groups.

```{r diag-3groups}
K <- 3; p <- 1000; n <- 100; n_starts <- 100; n_iter <- 50
set.seed(1)
L3 <- matrix(-1, nrow = n, ncol = K)
for (i in 1:K) L3[sample(1:n, 20), i] <- 1
FF <- matrix(rnorm(p * K), nrow = p, ncol = K)
X3 <- L3 %*% t(FF) + rnorm(n * p, 0, 0.01)

X1     <- preprocess(X3, n.comp = 10)
X1_aug <- rbind(rep(1, ncol(X1)), X1)

# fastICA
set.seed(1)
W <- matrix(rnorm(nrow(X1_aug) * n_starts), nrow(X1_aug), n_starts)
W <- sweep(W, 2, sqrt(colSums(W^2)), "/")
for (i in seq_len(n_iter)) W <- fastica_update(X1_aug, W)
Lhat_ica <- t(X1_aug) %*% W

# alt_proj
set.seed(1)
V <- sign(matrix(rnorm(n * n_starts), n, n_starts))
V[V == 0] <- 1L
for (i in seq_len(n_iter)) V <- alt_proj_update(X1_aug, V)
Lhat_ap <- V

cor_ica <- abs(cor(Lhat_ica, L3))   # n_starts x 3
cor_ap  <- abs(cor(Lhat_ap,  L3))
```

```{r diag-3groups-heatmap, fig.width=8, fig.height=5}
par(mfrow = c(1, 2))

order_ica <- order(apply(cor_ica, 1, max), decreasing = TRUE)
image(t(cor_ica[order_ica, ]), zlim = c(0, 1), axes = FALSE,
      main = "fastICA", xlab = "True group", ylab = "Start (sorted)")
axis(1, at = seq(0, 1, length.out = 3), labels = 1:3)

order_ap <- order(apply(cor_ap, 1, max), decreasing = TRUE)
image(t(cor_ap[order_ap, ]), zlim = c(0, 1), axes = FALSE,
      main = "AltProj", xlab = "True group", ylab = "Start (sorted)")
axis(1, at = seq(0, 1, length.out = 3), labels = 1:3)

par(mfrow = c(1, 1))
```

## Simulation 6: 25 non-overlapping groups

Twenty-five equal non-overlapping groups of 20 (n=500). Using n.comp=25.

```{r sim6-25nonoverlap}
n <- 500; p <- 1000
set.seed(1)
L <- matrix(0, nrow = n, ncol = 25)
for (i in 1:25) { L[((i - 1) * 20 + 1):(i * 20), i] <- 1 }
FF <- matrix(rnorm(p * 25), nrow = p)
X  <- L %*% t(FF) + rnorm(n * p, 0, 0.01)

res6 <- run_comparison(X, L, n.comp = 25, label = "25 non-overlapping groups (n=500)")
```

## Summary

```{r summary-table}
results <- data.frame(
  Simulation = c(
    "3 overlapping groups (n=100)",
    "9 overlapping groups (n=100)",
    "4 non-overlapping groups (n=100)",
    "9 non-overlapping groups (n=225)",
    "Bifurcating tree (n=100, 6 branches)",
    "25 non-overlapping groups (n=500)"
  ),
  n.comp = c(10, 10, 10, 20, 10, 25),
  p = c("0.20", "0.20", "0.25", "0.11", "0.25-0.50", "0.04"),
  fastICA_exact  = round(100 * c(res1["fastICA_exact"],  res2["fastICA_exact"],  res3["fastICA_exact"],
                                  res4["fastICA_exact"],  res5["fastICA_exact"],  res6["fastICA_exact"])),
  Hybrid_exact   = round(100 * c(res1["Hybrid_exact"],   res2["Hybrid_exact"],   res3["Hybrid_exact"],
                                  res4["Hybrid_exact"],   res5["Hybrid_exact"],   res6["Hybrid_exact"])),
  AltProj_exact  = round(100 * c(res1["AltProj_exact"],  res2["AltProj_exact"],  res3["AltProj_exact"],
                                  res4["AltProj_exact"],  res5["AltProj_exact"],  res6["AltProj_exact"])),
  fastICA_degen  = round(100 * c(res1["fastICA_degen"],  res2["fastICA_degen"],  res3["fastICA_degen"],
                                  res4["fastICA_degen"],  res5["fastICA_degen"],  res6["fastICA_degen"])),
  Hybrid_degen   = round(100 * c(res1["Hybrid_degen"],   res2["Hybrid_degen"],   res3["Hybrid_degen"],
                                  res4["Hybrid_degen"],   res5["Hybrid_degen"],   res6["Hybrid_degen"])),
  AltProj_degen  = round(100 * c(res1["AltProj_degen"],  res2["AltProj_degen"],  res3["AltProj_degen"],
                                  res4["AltProj_degen"],  res5["AltProj_degen"],  res6["AltProj_degen"])),
  fastICA_ms     = round(c(res1["fastICA_ms"], res2["fastICA_ms"],
                            res3["fastICA_ms"], res4["fastICA_ms"], res5["fastICA_ms"], res6["fastICA_ms"])),
  Hybrid_ms      = round(c(res1["Hybrid_ms"],  res2["Hybrid_ms"],
                            res3["Hybrid_ms"],  res4["Hybrid_ms"],  res5["Hybrid_ms"],  res6["Hybrid_ms"])),
  AltProj_ms     = round(c(res1["AltProj_ms"], res2["AltProj_ms"],
                            res3["AltProj_ms"], res4["AltProj_ms"], res5["AltProj_ms"], res6["AltProj_ms"])),
  fastICA_beat   = round(100 * c(res1["fastICA_beat"], res2["fastICA_beat"], res3["fastICA_beat"],
                                  res4["fastICA_beat"], res5["fastICA_beat"], res6["fastICA_beat"])),
  Hybrid_beat    = round(100 * c(res1["Hybrid_beat"],  res2["Hybrid_beat"],  res3["Hybrid_beat"],
                                  res4["Hybrid_beat"],  res5["Hybrid_beat"],  res6["Hybrid_beat"])),
  AltProj_beat   = round(100 * c(res1["AltProj_beat"], res2["AltProj_beat"], res3["AltProj_beat"],
                                  res4["AltProj_beat"], res5["AltProj_beat"], res6["AltProj_beat"])),
  true_vPv       = round(c(res1["true_vPv"], res2["true_vPv"], res3["true_vPv"],
                            res4["true_vPv"], res5["true_vPv"], res6["true_vPv"]), 4)
)

knitr::kable(results,
  col.names = c("Simulation", "n.comp", "p",
                "ICA >0.99", "Hyb >0.99", "AP >0.99",
                "ICA degen%", "Hyb degen%", "AP degen%",
                "ICA (ms)", "Hyb (ms)", "AP (ms)",
                "ICA % ≥ v'Pv", "Hyb % ≥ v'Pv", "AP % ≥ v'Pv",
                "True min v'Pv"),
  caption   = "Per-start success rates for fastICA (Newton), Hybrid (5 gradient + 45 Newton steps), and AltProj, at two correlation thresholds; percentage of degenerate solutions; wall-clock time; and fraction of starts whose binarized v'Pv meets or exceeds the minimum true group v'Pv."
)
```

The Hybrid method performs noticeably worse than standard fastICA on non-overlapping group simulations (sims 3, 4, 6).
The likely reason is that a binary vector indicating a single non-overlapping group is a *minimum* of the log-cosh contrast function: because the group members are all identical in whitened space, the projection has a highly sub-Gaussian (near-Bernoulli) distribution, which log-cosh penalises rather than rewards.
The initial gradient-ascent steps therefore push the weight vector *away* from these solutions before Newton steps take over, leaving fewer starting points in the basin of attraction of the correct solution.
The standard fastICA Newton steps, by contrast, can converge to both maxima and minima of the contrast, and apparently reach the non-overlapping group solutions more reliably from random initialisation.

# High-noise simulations

We repeat all five simulations with `noise_sd = 4` instead of 0.01, chosen so that the true v'Pv is around 90% of n.

## Simulation 1: 3 overlapping groups (high noise)

```{r noise-sim1}
K <- 3; p <- 1000; n <- 100
set.seed(1)
L <- matrix(-1, nrow = n, ncol = K)
for (i in 1:K) { L[sample(1:n, 20), i] <- 1 }
FF <- matrix(rnorm(p * K), nrow = p, ncol = K)
X  <- L %*% t(FF) + rnorm(n * p, 0, 4)

nres1 <- run_comparison(X, L, n.comp = 10, label = "3 overlapping groups, high noise")
```

## Simulation 2: 9 overlapping groups (high noise)

```{r noise-sim2}
K <- 9; p <- 1000; n <- 100
set.seed(1)
L <- matrix(0, nrow = n, ncol = K)
for (i in 1:K) { L[sample(1:n, 20), i] <- 1 }
FF <- matrix(rnorm(p * K), nrow = p, ncol = K)
X  <- L %*% t(FF) + rnorm(n * p, 0, 4)

nres2 <- run_comparison(X, L, n.comp = 10, label = "9 overlapping groups, high noise")
```

## Simulation 3: 4 non-overlapping groups (high noise)

```{r noise-sim3}
n <- 100; p <- 1000
set.seed(1)
L <- matrix(0, nrow = n, ncol = 4)
L[1:25,   1] <- 1; L[26:50,  2] <- 1; L[51:75,  3] <- 1; L[76:100, 4] <- 1
FF <- matrix(rnorm(p * 4), nrow = p)
X  <- L %*% t(FF) + rnorm(n * p, 0, 4)

nres3 <- run_comparison(X, L, n.comp = 10, label = "4 non-overlapping groups, high noise")
```

## Simulation 4: 9 non-overlapping groups (high noise)

```{r noise-sim4}
n <- 225; p <- 1000
set.seed(1)
L <- matrix(0, nrow = n, ncol = 9)
for (i in 1:9) { L[((i - 1) * 25 + 1):(i * 25), i] <- 1 }
FF <- matrix(rnorm(p * 9), nrow = p)
X  <- L %*% t(FF) + rnorm(n * p, 0, 4)

nres4 <- run_comparison(X, L, n.comp = 20, label = "9 non-overlapping groups, high noise")
```

## Simulation 5: Bifurcating tree (high noise)

```{r noise-sim5}
n <- 100; p <- 1000
set.seed(1)
L <- matrix(0, nrow = n, ncol = 6)
L[1:50, 1] <- 1; L[51:100, 2] <- 1
L[1:25, 3] <- 1; L[26:50,  4] <- 1; L[51:75, 5] <- 1; L[76:100, 6] <- 1
FF <- matrix(rnorm(p * 6), nrow = p)
X  <- L %*% t(FF) + rnorm(n * p, 0, 4)

nres5 <- run_comparison(X, L, n.comp = 10, label = "Bifurcating tree, high noise")
```

## Simulation 6: 25 non-overlapping groups (high noise)

```{r noise-sim6}
n <- 500; p <- 1000
set.seed(1)
L <- matrix(0, nrow = n, ncol = 25)
for (i in 1:25) { L[((i - 1) * 20 + 1):(i * 20), i] <- 1 }
FF <- matrix(rnorm(p * 25), nrow = p)
X  <- L %*% t(FF) + rnorm(n * p, 0, 4)

nres6 <- run_comparison(X, L, n.comp = 25, label = "25 non-overlapping groups, high noise")
```

## Summary (high noise)

```{r noise-summary-table}
nresults <- data.frame(
  Simulation = c(
    "3 overlapping groups (n=100)",
    "9 overlapping groups (n=100)",
    "4 non-overlapping groups (n=100)",
    "9 non-overlapping groups (n=225)",
    "Bifurcating tree (n=100, 6 branches)",
    "25 non-overlapping groups (n=500)"
  ),
  n.comp = c(10, 10, 10, 20, 10, 25),
  p = c("0.20", "0.20", "0.25", "0.11", "0.25-0.50", "0.04"),
  fastICA_exact  = round(100 * c(nres1["fastICA_exact"],  nres2["fastICA_exact"],  nres3["fastICA_exact"],
                                  nres4["fastICA_exact"],  nres5["fastICA_exact"],  nres6["fastICA_exact"])),
  Hybrid_exact   = round(100 * c(nres1["Hybrid_exact"],   nres2["Hybrid_exact"],   nres3["Hybrid_exact"],
                                  nres4["Hybrid_exact"],   nres5["Hybrid_exact"],   nres6["Hybrid_exact"])),
  AltProj_exact  = round(100 * c(nres1["AltProj_exact"],  nres2["AltProj_exact"],  nres3["AltProj_exact"],
                                  nres4["AltProj_exact"],  nres5["AltProj_exact"],  nres6["AltProj_exact"])),
  fastICA_degen  = round(100 * c(nres1["fastICA_degen"],  nres2["fastICA_degen"],  nres3["fastICA_degen"],
                                  nres4["fastICA_degen"],  nres5["fastICA_degen"],  nres6["fastICA_degen"])),
  Hybrid_degen   = round(100 * c(nres1["Hybrid_degen"],   nres2["Hybrid_degen"],   nres3["Hybrid_degen"],
                                  nres4["Hybrid_degen"],   nres5["Hybrid_degen"],   nres6["Hybrid_degen"])),
  AltProj_degen  = round(100 * c(nres1["AltProj_degen"],  nres2["AltProj_degen"],  nres3["AltProj_degen"],
                                  nres4["AltProj_degen"],  nres5["AltProj_degen"],  nres6["AltProj_degen"])),
  fastICA_ms     = round(c(nres1["fastICA_ms"], nres2["fastICA_ms"],
                            nres3["fastICA_ms"], nres4["fastICA_ms"], nres5["fastICA_ms"], nres6["fastICA_ms"])),
  Hybrid_ms      = round(c(nres1["Hybrid_ms"],  nres2["Hybrid_ms"],
                            nres3["Hybrid_ms"],  nres4["Hybrid_ms"],  nres5["Hybrid_ms"],  nres6["Hybrid_ms"])),
  AltProj_ms     = round(c(nres1["AltProj_ms"], nres2["AltProj_ms"],
                            nres3["AltProj_ms"], nres4["AltProj_ms"], nres5["AltProj_ms"], nres6["AltProj_ms"])),
  fastICA_beat   = round(100 * c(nres1["fastICA_beat"], nres2["fastICA_beat"], nres3["fastICA_beat"],
                                  nres4["fastICA_beat"], nres5["fastICA_beat"], nres6["fastICA_beat"])),
  Hybrid_beat    = round(100 * c(nres1["Hybrid_beat"],  nres2["Hybrid_beat"],  nres3["Hybrid_beat"],
                                  nres4["Hybrid_beat"],  nres5["Hybrid_beat"],  nres6["Hybrid_beat"])),
  AltProj_beat   = round(100 * c(nres1["AltProj_beat"], nres2["AltProj_beat"], nres3["AltProj_beat"],
                                  nres4["AltProj_beat"], nres5["AltProj_beat"], nres6["AltProj_beat"])),
  true_vPv       = round(c(nres1["true_vPv"], nres2["true_vPv"], nres3["true_vPv"],
                            nres4["true_vPv"], nres5["true_vPv"], nres6["true_vPv"]), 2)
)

knitr::kable(nresults,
  col.names = c("Simulation", "n.comp", "p",
                "ICA >0.99", "Hyb >0.99", "AP >0.99",
                "ICA degen%", "Hyb degen%", "AP degen%",
                "ICA (ms)", "Hyb (ms)", "AP (ms)",
                "ICA % ≥ v'Pv", "Hyb % ≥ v'Pv", "AP % ≥ v'Pv",
                "True min v'Pv"),
  caption   = "High-noise results (noise_sd = 4). Hybrid = 5 gradient + 45 Newton steps."
)
```
