---
title: "Explorations of implementing the UDI framework in mash"
author: "Matthew Stephens "
date: 2017-06-25
output: html_document
---

<!-- The file analysis/chunks.R contains chunks that define default settings
shared across the workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```


# Preliminary

First, here is the code from Stephens (2013) PloS ONE

```{r}
#' @param VYX is (1/n) Y'X is d by p
#' @param VYY is (1/n) Y'Y is d by d
#' @param VXX a p-vector of the estimated variances of the SNP
#' @param U vector of length d of true/false
#' @param D vector of length d of true/false
#' @param n sample size the summaries were computed from
#' @param m a parameter in the prior (I believe)
#' @param d the number of phenotypes
#' @param sigmaa the parameter of the prior determining effect size
logBF.fromVSummaries = function(VYX,VYY,VXX,U,D,n,m,d,sigmaa){

  dd = sum(D)
  du= sum(U)
  p = dim(VYX)[2]

  if(du>0){
    LUU = chol(VYY[U,U,drop=FALSE]) # a du by du matrix
    VUD = VYY[U,D,drop=FALSE]      #a du by dd matrix of the correlations of Yu with Yd

    c = cbind(forwardsolve(t(LUU),VYX[U,,drop=FALSE]))#c solves LUU'c = phiU, c is a du by p matrix
    b = cbind(forwardsolve(t(LUU), VUD))  # b is du by dd, and solves LUU' b = VUD, so b'b = VUD' LUU^-1 LUU'^-1 VUD = VUD' (LUU'LUU)^-1 VUD = VUD'VYYU^-1 VUD
  } else{c=matrix(0,nrow=1,ncol=p); b=matrix(0,nrow=1,ncol=dd);}

  C = VXX - colSums(c*c)

  u = VYX[D,,drop=FALSE] - crossprod(b,c)

  V0 = VYY[D,D,drop=FALSE] - crossprod(b)
  L0 = chol(V0)
  a = forwardsolve(t(L0),u)
  lambda = sigmaa^(-2)  / (n*C)
  k = as.numeric(1/(1+lambda))
  return((dd/2) * log(1-k) - 0.5*(n+m-(d-sum(D)-sum(U)))*log(1-(k/C) *colSums(a*a)))
}
```

And now how we apply this to summary data based on the
methods in Stephens (2016)
```{r define}
#' @param Z a p by d matrix of Z scores
#' @param VYY a d by d matrix 
#' @param f a p vector containing the frequencies of the p SNPs
logBF.fromZ = function(Z,VYY,f,U,D,n,m,sigmaa){
  VXX = 2*f*(1-f)
  VYX = t(sqrt(VXX/n) * Z)
  logBF.fromVSummaries(VYX,VYY,VXX,U,D,n,m,d=ncol(Z),sigmaa)
}


```

Read in some example data
```{r read}
z = read.table("../data/bmass.HaemgenRBC2016.Vs2.NewSNPs.ZScores.hclust.vs1.txt",header=TRUE)
z = as.matrix(z[,2:7])

V = diag(6)
V[1,]=c(0.5,-0.47,0.80,-0.47,0.73,-0.13)
V[2,2:6] = c(0.5,0.12,0.87,0.12,0.03)
V[3,3:6] = c(0.5,0.04,0.93,-0.1)
V[4,4:6] = c(0.5,0.20,0.46)
V[5,5:6] =c(0.5,0.22)
V[6,6] = 0.5
V  = V+t(V)
```

And apply the method. Here we show how changes in n and f 
give the same BF if we scale the prior parameter sigma_a
like $1/\sqrt{2f(1-f)n}$.

First scaling in n:
```{r scaling_n}
UU = c(F,F,F,F,F,F)
DD = c(T,T,T,T,T,T)
p = nrow(z)
f = 0.5
lbf1=logBF.fromZ(z,V,rep(f,p), UU,DD,n=1000,m=0,sigmaa=sqrt(1/(2*f*(1-f)*1000)))
lbf2=logBF.fromZ(z,V,rep(f,p), UU,DD,n=10000,m=0,sigmaa=sqrt(1/(2*f*(1-f)*10000)))
plot(lbf1,lbf2)
abline(a=0,b=1,col=2)
```

Now try changing f:
```{r scaling_f}
f = 0.2
lbf3=logBF.fromZ(z,V,rep(f,p), UU,DD,n=10000,m=0,sigmaa=sqrt(1/(2*f*(1-f)*10000)))
plot(lbf2,lbf3)
abline(a=0,b=1,col=2)
```

Now check the mashr version:
```{r mashr_version}
library("mashr")
data = set_mash_data(z,1,V)
Ulist = list(null= mashr:::cov_all_zeros(data), all = cov_udi(data,c("D","D","D","D","D","D")))
temp = calc_lik_matrix(data,Ulist,log=TRUE)
plot(temp[,2]-temp[,1],lbf1,xlab="mash lBF",ylab="original lBF (n=1000)")
abline(a=0,b=1,col=2)
```
Pretty good agreement. But even better if we use the version
with n=10k:
```{r}
plot(temp[,2]-temp[,1],lbf2,xlab="mash lBF",ylab="original lBF (n=10k)")
abline(a=0,b=1,col=2)
```

Now try making the first two components unassociated
```{r}
UU = c(T,T,F,F,F,F)
DD = c(F,F,T,T,T,T)
lbf=logBF.fromZ(z,V,rep(f,p),UU,DD,n=10000,m=0,sigmaa=sqrt(1/(2*f*(1-f)*10000)))

data = set_mash_data(z,1,V)
Ulist = list(null= mashr:::cov_all_zeros(data), all = cov_udi(data,c("U","U","D","D","D","D")))
temp = calc_lik_matrix(data,Ulist,log=TRUE)
plot(temp[,2]-temp[,1],lbf,xlab="mash lBF",ylab="original lBF (n=10k)")
abline(a=0,b=1,col=2)
```

And making components 3-4 unassociated
```{r}
UU = c(F,F,T,T,F,F)
DD = c(T,T,F,F,T,T)
lbf=logBF.fromZ(z,V,rep(f,p),UU,DD,n=10000,m=0,sigmaa=sqrt(1/(2*f*(1-f)*10000)))

data = set_mash_data(z,1,V)
Ulist = list(null= mashr:::cov_all_zeros(data), all = cov_udi(data,c("D","D","U","U","D","D")))
temp = calc_lik_matrix(data,Ulist,log=TRUE)
plot(temp[,2]-temp[,1],lbf,xlab="mash lBF",ylab="original lBF (n=10k)")
abline(a=0,b=1,col=2)
```


And also making components 1 and 6 indirectly associated
```{r}
UU = c(F,F,T,T,F,F)
DD = c(F,T,F,F,T,F)
lbf=logBF.fromZ(z,V,rep(f,p),UU,DD,n=10000,m=0,sigmaa=sqrt(1/(2*f*(1-f)*10000)))

data = set_mash_data(z,1,V)
Ulist = list(null= mashr:::cov_all_zeros(data), all = cov_udi(data,c("I","D","U","U","D","I")))
temp = calc_lik_matrix(data,Ulist,log=TRUE)
plot(temp[,2]-temp[,1],lbf,xlab="mash lBF",ylab="original lBF (n=10k)")
abline(a=0,b=1,col=2)
```


Without any unaffected:
```{r}
UU = c(F,F,F,F,F,F)
DD = c(T,T,T,F,F,F)
lbf=logBF.fromZ(z,V,rep(f,p),UU,DD,n=10000,m=0,sigmaa=sqrt(1/(2*f*(1-f)*10000)))

data = set_mash_data(z,1,V)
Ulist = list(null= mashr:::cov_all_zeros(data), all = cov_udi(data,c("D","D","D","I","I","I")))
temp = calc_lik_matrix(data,Ulist,log=TRUE)
plot(temp[,2]-temp[,1],lbf,xlab="mash lBF",ylab="original lBF (n=10k)")
abline(a=0,b=1,col=2)
```

And another, just to check:
```{r}
UU = c(F,F,F,F,F,F)
DD = c(F,T,F,T,F,T)
lbf=logBF.fromZ(z,V,rep(f,p),UU,DD,n=10000,m=0,sigmaa=sqrt(1/(2*f*(1-f)*10000)))

data = set_mash_data(z,1,V)
Ulist = list(null= mashr:::cov_all_zeros(data), all = cov_udi(data,c("I","D","I","D","I","D")))
temp = calc_lik_matrix(data,Ulist,log=TRUE)
plot(temp[,2]-temp[,1],lbf,xlab="mash lBF",ylab="original lBF (n=10k)")
abline(a=0,b=1,col=2)
```



## Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
