---
title: "EB_subspace"
author: "Matthew Stephens"
date: "2026-06-19"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

I want to look at simple EB/variational/Bayesian approach to estimating a binary vector that lives approximately in a given subspace. You are given an $nxn$, rank $k$, projection matrix P and you want to find a vector $v$ such that $\rho(v) :=v'Pv/v'v$ is large. If $v$ lies entirely in the subspace defined by $P$ then $\rho(v) = v'v/v'v = 1$.

My derivations [here](https://www.overleaf.com/project/67c8cfa8666bbc398ebb08fa) suggest using a log-likelihood $l(v) \approx -(n-k-2b/2)\log(1-\rho(v))$, where $b$ is a hyperparameter. For now we make the approximation that $k,b$ are small so $l(v) \approx -n/2 \log(1-\rho(v))$. We also make the approximation $E_q(l(v)) = -(n/2) \log(1-\rho(q))$ where $\rho(q) = \bar{v}'P\bar{v}/<v'v>$. Here $\bar{v},<v'v>$ denote the expectations of $v,v'v$ under $q$. 

The variational EB approach then iterates: 
$$q \leftarrow EBNM\left( \frac{\mathbf{P}\bar{v}}{\rho(q)}, \frac{(1-\rho(q)) <v^Tv>}{n\rho(q)} \right)$$

For the Rademacher prior we have $<v'v>=n$ and the update can be written entirely in terms of the posterior mean:
$$\bar{v} = tanh(P\bar{v}/(1-\rho))$$
While the idea is to use $\rho=v'Pv/n$, I will also try fixing $\rho$ in these updates to see what happens. Note that this approach is not really EB because the prior is fixed Rademacher. TODO: we should try estimating a generalized (asymmetric) Rademacher prior, so this becomes a genuinely EB approach.


## Three group simulation (low noise)

Here I simulate 3 groups, with 20 members each.
```{r}
K=3
p = 1000
n = 100
set.seed(1)
L = matrix(-1,nrow=n,ncol=K)
for(i in 1:K){L[sample(1:n,20),i]=1}
FF = matrix(rnorm(p*K), nrow = p, ncol=K)

X = L %*% t(FF) + rnorm(n*p,0,1)
```

The column space of $X$ approximately contains three binary vectors (columns of $L$). Here I find the projection matrix corresponding to the column space of $X$ (approximated using a rank 3 matrix, which helps remove the noise.)
We can see the values of rho for the true L are very high.

```{r}
U = svd(X)$u[,1:3] 
P = U %*% t(U) 
true_rho = diag(t(L) %*% P %*% L/n)
true_rho
```

Now, initializing from a random vector. If we set rho=0 then the iterates converge to something close to 0 (as expected, because tanh is a contraction).
However, interestingly it does find a binary source!
```{r}
set.seed(1)
v.init = sample(c(-1,1),n,replace=TRUE)
maxiter = 10000

rho = 0
v = v.init
for(i in 1:maxiter){
  v = tanh(P %*% v/(1-rho))
}
t(v) %*% P %*% v/n # print out rho
plot(v)
cor(v,L) 
```

Try instead setting rho=0.5. It converges to a binary source again, but this time close to v in (-1,1).
```{r}
rho = 0.5
v = v.init
for(i in 1:maxiter){
  v = tanh(P %*% v/(1-rho))
}
t(v) %*% P %*% v/n # print out rho
plot(v)
cor(v,L)
```

This time we estimate rho. It converges to a binary source again, but this time close to v in (-1,1).
```{r}
rho = 0
v = v.init
for(i in 1:maxiter){
  rho = as.numeric(t(v) %*% P %*% v/n) # update rho
  v = tanh(P %*% v/(1-rho))
}
t(v) %*% P %*% v/n # print out rho
plot(v)
cor(v,L)
```


## Three group simulation (higher noise)

Here I repeat the above, with more noise:
```{r}
K=3
p = 1000
n = 100
set.seed(1)
L = matrix(-1,nrow=n,ncol=K)
for(i in 1:K){L[sample(1:n,20),i]=1}
FF = matrix(rnorm(p*K), nrow = p, ncol=K)

X = L %*% t(FF) + rnorm(n*p,0,10)
```

The column space of $X$ approximately contains three binary vectors (columns of $L$). The true value of rho is near 0.8.

```{r}
U = svd(X)$u[,1:3] 
P = U %*% t(U) 
true_rho = diag(t(L) %*% P %*% L/n)
true_rho
```

Now, initializing from a random vector. If we set rho=0 then the iterates converge to something close to 0 (as expected, because tanh is a contraction).
This time it does not really find a binary source!
```{r}
set.seed(1)
v.init.1 = sample(c(-1,1),n,replace=TRUE)
maxiter = 10000

rho = 0
v = v.init.1
for(i in 1:maxiter){
  v = tanh(P %*% v/(1-rho))
}
t(v) %*% P %*% v/n # print out rho
plot(v)
cor(v,L) 
```

Try instead setting rho=0.5. It is getting closer to a binary source.
```{r}
rho = 0.5
v = v.init.1
for(i in 1:maxiter){
  v = tanh(P %*% v/(1-rho))
}
t(v) %*% P %*% v/n # print out rho
plot(v)
cor(v,L)
```

This time we estimate rho. It converges to something closer to a binary source but does not get it exactly right.
```{r}
rho = 0
v = v.init.1
for(i in 1:maxiter){
  rho = as.numeric(t(v) %*% P %*% v/n) # update rho
  v = tanh(P %*% v/(1-rho))
}
t(v) %*% P %*% v/n # print out rho
plot(v)
cor(v,L)
```

Try a different random start: it gets closer (and also a larger rho).
```{r}
set.seed(3)
v.init.3 = sample(c(-1,1),n,replace=TRUE)
rho = 0
v = v.init.3
for(i in 1:maxiter){
  rho = as.numeric(t(v) %*% P %*% v/n) # update rho
  v = tanh(P %*% v/(1-rho))
}
t(v) %*% P %*% v/n # print out rho
plot(v)
cor(v,L)
```

Try EBICA
```{r}
set.seed(3)
v.init.3 = sample(c(-1,1),n,replace=TRUE)
rho = 0
v = v.init.3

for(i in 1:maxiter){
  v = tanh(n*P %*% v/as.numeric(sqrt(n * t(v) %*% P %*% v)))
}
t(v) %*% P %*% v/n # print out rho
plot(v)
cor(v,L)
```

## Try centering X

In the above I did not center X. So here I do so. This reduces the "true rho" substantially (below 0.5). The issue is that the original binary vectors are no longer in the subspace spanned by columns of Xcenter - the centering shifts the +-1 vectors so they are no longer +-1.
```{r}
Xcenter = scale(X,scale=FALSE)
Ucenter = svd(Xcenter)$u[,1:3] 
Pcenter = Ucenter %*% t(Ucenter) 
diag(t(L) %*% Pcenter %*% L/n)
```

The method now fails to find them (not suprisingly):
```{r}
set.seed(3)
v.init.3 = sample(c(-1,1),n,replace=TRUE)
rho = 0
v = v.init.3
for(i in 1:maxiter){
  rho = as.numeric(t(v) %*% Pcenter %*% v/n) # update rho
  v = tanh(Pcenter %*% v/(1-rho))
}
t(v) %*% Pcenter %*% v/n # print out rho
plot(v)
cor(v,L)
```

If we add an intercept we have a problem: the method finds the trivial solution of all 1s. (When I ran this about five times with different seeds it did eventually find one of the sources. But the trivial solution has the largest rho, so it tends to win.)
```{r}
Ucenter = cbind(rep(1/sqrt(n),n),Ucenter) # add an intercept
Pcenter = Ucenter %*% t(Ucenter) 
diag(t(L) %*% Pcenter %*% L/n) #rho is back to 0.8 or so

set.seed(3)
v.init.3 = sample(c(-1,1),n,replace=TRUE)
rho = 0
v = v.init.3
v = sample(c(-1,1),n,replace=TRUE)
for(i in 1:maxiter){
  rho = as.numeric(t(v) %*% Pcenter %*% v/n) # update rho
  v = tanh(Pcenter %*% v/(1-rho))
}
t(v) %*% Pcenter %*% v/n # print out rho
plot(v)
cor(v,L)
```
