---
title: "Trying method of moments for estimating prior in EB"
author: "Matthew Stephens"
date: 2017-12-09
output: html_document
---

<!-- The file analysis/chunks.R contains chunks that define default settings
shared across the workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

# Idea

The idea is to investigate method of moments estimation
for very fast EB procedures like EBayesThresh.

Suppose $x_j = b_j + N(0,s_j^2)$
and $b_j \sim \pi_0 \delta_0 + (1-\pi_0) DExp(a)$
where $Dexp$ means the double exponential (Laplace)
with mean $a$ (so rate parameter $1/a$).

Then the even moments of $b_j$ are
$E(b_j^n) = (1-\pi_0) n! a^n$.

Also, 2nd and 4th moments of $N(0,s_j^2)$ are
$s_j^2$ and $3s_j^4$.

Thus, second and 4th moments of $x_j$ are:
$$E(x_j^2) = s_j^2 + 2(1-\pi_0)a^2$$

$$E(x_j^4) = 3s_j^4 + 24(1-\pi_0)a^4 + 6 s_j^2 [2(1-\pi_0)a^2] $$.


Let $m_2 := E(x_j^2 - s_j^2)$ and 
$m_4:= E(x_j^4 - 3s_j^4)$. Then

$$E(m_2) = 2(1-\pi_0)a^2$$.
$$E(m_4) = 24[(1-\pi_0)a^2] + 12s_j^2 (1-\pi_0)a^2$$
$$= m_2 (12a^2 + 6s_j^2)$$
So we can solve to give
$$a^2 = (1/12) (m_4/m_2 - 6s_j^2)$$

OK, so I've been sloppy with subscripts on the $m_4$ and $m_2$... need to sort that out.


We we will try it with constant $s_j=1$.

# Simulate some data

```{r}
  eb_mm = function(x,s){
    m4 = mean(x^4-3*s^4)
    m2 = mean(x^2-s^2)
    a2 = ((m4/m2) - 6*mean(s^2))/12
    pi0 = 1- m2/(2*a2)
    if(a2<0){a2=0; pi0=1}
    if(pi0<0){pi0=0; a2 =m2/2}
    if(pi0>1){pi0=1; a2 = NA}
    
    return(list(pi0=pi0,a=sqrt(a2)))
  }
  set.seed(1)
  n=100000
  e = rnorm(n)
  b = rexp(n)
  x = b+e   
  eb_mm(x,1)
  EbayesThresh::wandafromx(x,1,FALSE)
  
  b2 =b
  b2[1:(n/2)] = 0
  x2 = b2 + e
  eb_mm(x2,1)
  EbayesThresh::wandafromx(x2,1,FALSE)
  
```


Try a case that is "nearly null"
```{r}
  b3 =b
  b3[1:(0.99*n)] = 0
  x3 = b3 + e
  eb_mm(x3,1)
  EbayesThresh::wandafromx(x3,1,FALSE)
```






# Thoughts

- hard part is when pi0 is very close to 1 but not 1.
That might be worth thinking about. Maybe some upper quantiles would be better than 4th moment?

- the case where s is not constant - probably want to work with E(x/s) rather than E(x) ?



## Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
