---
title: "Trying method of moments for estimating prior in EB"
author: "Matthew Stephens"
date: 2017-12-09
output: workflowr::wflow_html
---




# Idea

The idea is to investigate method of moments estimation
for very fast EB procedures like EBayesThresh.

Suppose $x_j = b_j + N(0,s_j^2)$
and $b_j \sim \pi_0 \delta_0 + (1-\pi_0) DExp(a)$
where $Dexp$ means the double exponential (Laplace)
with rate $a$.

Then the even moments of $b_j$ are
$E(b_j^n) = (1-\pi_0) n! (1/a)^n$.

Also, 2nd and 4th moments of $N(0,s_j^2)$ are
$s_j^2$ and $3s_j^4$.

Let $z_j := x_j/s_j$. The, second and 4th moments of $z_j$ are:
$$E(z_j^2) = 1 + 2(1-\pi_0)/(s_j^2 a^2)$$

$$E(z_j^4) = 3 + 24(1-\pi_0)/(a^4 s_j^4) + 6[2(1-\pi_0)]/[a^2 s_j^2] $$.


Let $m_{2j} := E(z_j^2 - 1)$ and 
$m_{4j}:= E(z_j^4 - 3)$. Then

$$E(m_{2j}) = 2(1-\pi_0)/(s_j^2 a^2)$$.

$$E(m_{4j}) = 24[(1-\pi_0)/s_j^4a^4] + 12(1-\pi_0)/(s_j^2 a^2)$$

Now we need to do somethign like sum the $m_{4j}$ and sum the $m_{2j}$ over
$j$ to get two equations, and solve.... so the rest of this needs correcting.


$$= E(m_{2j}) (12/s_j^2a^2 + 6)$$
So we can solve to give
$$a^2 = \sum(1/s_j^2) 12/ (m_4/m_2 - 6)$$

OK, so I've been sloppy with subscripts on the $m_4$ and $m_2$... need to sort that out.


We we will try it with constant $s_j=1$.

# Simulate some data

```{r}
  eb_mm = function(x,s){
    z = x/s
    S1 = mean(1/s^2)
    S2 = mean(1/s^4)
    
    m4 = mean(z^4) - 3
    m2 = mean(z^2) - 1

    a2 = 12*(S2/S1)*((m4/m2)- 6)
    
    pi0 = 1- a2*m2/(2*S1)
    if(a2<0){a2=0; pi0=1}
    if(pi0<0){pi0=0; a2 =2/m2}
    if(pi0>1){pi0=1; a2 = 1}
    
    return(list(pi0=pi0,a=sqrt(a2)))
  }
  set.seed(1)
  n=100000
  e = rnorm(n)
  b = rexp(n,rate=2)
  x = b+e   
  eb_mm(x,1)
  EbayesThresh::wandafromx(x,1,FALSE)
  
  b2 =b
  b2[1:(n/2)] = 0
  x2 = b2 + e
  eb_mm(x2,1)
  EbayesThresh::wandafromx(x2,1,FALSE)
  EbayesThresh:::wandafromx.mle(x2,1)
```


Try a case that is "nearly null". Note that here the original 
Ebayesthresh approach based on the beta function is less accurate, presumably
due to numeric issues.
```{r}
  b3 =b
  b3[1:(0.99*n)] = 0
  x3 = b3 + e
  eb_mm(x3,1)
  EbayesThresh::wandafromx(x3,1,FALSE)
  EbayesThresh:::wandafromx.mle(x3,1)
```



```{r}
  s = rgamma(n,1,1)
  x4 = b3 + s*e
  eb_mm(x4,s)
  EbayesThresh:::wandafromx.mle(x4,s)
```



# Thoughts

- hard part is when pi0 is very close to 1 but not 1.
That might be worth thinking about. Maybe some upper quantiles would be better than 4th moment?

- the case where s is not constant - probably want to work with E(x/s) rather than E(x) ?




