---
title: "EB Flash, n=p=1"
author: "Matthew Stephens"
date: 2017-06-14
output: html_document
---

<!-- The file analysis/chunks.R contains chunks that define default settings
shared across the workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

# Background

I was interested in whether we can try to differentiate
the flash likelihood at the "null" ($g_l=g_f=0$).

Let n=p=1. And Y (1 by 1 matrix) be given by factor model
$$Y = lf + e$$
where $l,f \sim N(0,s^2)$ and $e \sim N(0,1)$.

Integrating out $l$ and $f$ yields a likelihood for $s$:
$$L(s) = \int p(Y| l, f) p(l) p(f) dl df $$

My question is what is the derivative of the likelihood
(or log likelihood) with respect to $s$. Particularly what is it at $s=0$
(which is the 0-factor model)?

If we define $h$ to be the log-likelihood for $l,f$ 
$$h(l,f) = \log p(Y|l,f)$$
then is easy to show that

$$d^2h/dl^2 = f^2/s^2$$
and 
$$d^2h/dldf = -(Y-2lf)$$

so at $l=f=0$ the hessian is the matrix with 0 on diagonal and $-Y$ on off-diagonal. This matrix has eigenvalues $\pm Y$.

This suggests that the posterior distribution of $(l,f)$ will be approximated
by a multivariate normal with inverse covariance matrix with $1/s^2$ on diagonal and $Y$ on off-diagonal. Let's try with s=0.1.

```{r}
l = rnorm(100000,0,0.1)
f = rnorm(100000,0,0.1)
y = 2
Lik = dnorm((y-l*f),0,1)
ss = sample(1:100000,prob=Lik,replace=TRUE)
plot(l[ss],f[ss])
```

So we see the problem - indeed the posterior is approximated by this, but it looks just like the prior... So the ratio of the posterior to the prior will be 1.

## Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
