---
title: "fastica_asymmetric_03"
author: "Matthew Stephens"
date: "2026-08-03"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

After two "quick and dirty" attempts, I think I'm now ready to implement a proper version of  an asymmetric version of fastica. We can also think of this as a Newton-based method for optimizing the Empirical Bayes Projection model (EBProj).

### The Newton version of EBCD

Given a psd matrix $S=UDU'$ and a residual variance $\sigma^2$, define the objective function $J$ as:
$$J(w, g; \sigma^2, S) = \sum_{i=1}^n \log Z((UDw)_i, g, \sigma^2)$$
where
$$Z(x, g, \sigma^2) = \int g(\theta) \exp\left( \frac{2x \theta - \theta^2}{2\sigma^2} \right) d\theta. $$
The EBCD method (for fixed $\sigma^2$) can be implemented by maximizing $J$ subject to $w'Dw=1$ using a Newton method on $w$ (and, for example, EM method for $g$). Given the optimal $w,g$, the optimal $q$ is the EBNM posterior with prior $g^*$ and data $x=UDw^*,s=\sigma$.

For $g$ a binary prior with probabilities $1-p,p$ on $y_0,y_1$ we have
$$Z(x, g, \sigma^2) = (1-p)\exp\!\left(\frac{2xy_0 - y_0^2}{2\sigma^2}\right) + p\exp\!\left(\frac{2xy_1 - y_1^2}{2\sigma^2}\right).$$

Define the log unnormalized weights $\ell_j(x) = \log p_j + (2xy_j - y_j^2)/(2\sigma^2)$, with $p_0=1-p$ and $p_1=p$. Then the posterior probability of $y_1$ is
$$\pi_1(x) = \frac{e^{\ell_1(x)}}{e^{\ell_0(x)} + e^{\ell_1(x)}},$$
the posterior mean is
$$M(x;\sigma^2,g) = (1-\pi_1)y_0 + \pi_1 y_1,$$
and the derivative of $M$ with respect to $x$ (used in the Newton step) is
$$M'(x;\sigma^2,g) = \pi_1(1-\pi_1)(y_1-y_0)^2/\sigma^2.$$


### The EBProj approach

The EBProj method works on a projection matrix $P$ (if $S$ is a whitened then $P=S/n$ is a projection matrix). The objective is:
$$F_{\text{EBproj}}(w, g; \tau, P)
= \frac{n - k - 2b}{2} \left( \log \tau - \tau + 1 \right)  + J(w,g; \sigma^2_\text{eff}, S = nP)$$
where $\sigma^2_\text{eff} =\frac{n}{(n - k - 2b)\tau}$.

Optimizing the EBProj objective can iterate between optimizing $J$ over $w,g$ as above and then optimizing $\tau = 1/(1-\hat\rho)$ where $\hat\rho = \bar{\mathbf{v}}'P\bar{\mathbf{v}} / U_q$, $U_q = E_q[\mathbf{v}'\mathbf{v}]$ is the second moment, and $P=S/n$.


## Implementation

### Helper functions for the binary prior

```{r}
# log Z(x, g, sigma2) for binary prior g = (1-p) delta_{y0} + p delta_{y1}
log_Z_binary = function(x, y_0, y_1, p, sigma2) {
  lp0 = log(1 - p) + (2*x*y_0 - y_0^2) / (2*sigma2)
  lp1 = log(p)     + (2*x*y_1 - y_1^2) / (2*sigma2)
  m   = pmax(lp0, lp1)
  log(exp(lp0 - m) + exp(lp1 - m)) + m
}

# posterior probability of y_1
post_prob1 = function(x, y_0, y_1, p, sigma2) {
  lp0 = log(1 - p) + (2*x*y_0 - y_0^2) / (2*sigma2)
  lp1 = log(p)     + (2*x*y_1 - y_1^2) / (2*sigma2)
  m   = pmax(lp0, lp1)
  e0  = exp(lp0 - m);  e1 = exp(lp1 - m)
  e1 / (e0 + e1)
}

# posterior mean M(x; sigma2, g)
post_mean_binary = function(x, y_0, y_1, p, sigma2) {
  pi1 = post_prob1(x, y_0, y_1, p, sigma2)
  (1 - pi1)*y_0 + pi1*y_1
}

# dM/dx = pi1*(1-pi1)*(y1-y0)^2 / sigma2
post_mean_deriv_binary = function(x, y_0, y_1, p, sigma2) {
  pi1 = post_prob1(x, y_0, y_1, p, sigma2)
  pi1 * (1 - pi1) * (y_1 - y_0)^2 / sigma2
}
```

### Initialisation

Define a fit object (simply a list) that contains the data $U$, and the parameters to be updated (prior $g$, weights $w$). Subsequent update functions will take a fit as input and output a new fit.

```{r}
ebproj_init = function(U, b=0, c = 1, tau=NULL, sigma2=NULL, w=NULL) {
  n   = nrow(U)
  k   = ncol(U)
  nu  = n - k - 2*b
  if (!is.null(tau) && !is.null(sigma2))
    stop("specify at most one of tau and sigma2")
  if (is.null(tau) && is.null(sigma2)) tau = 1
  if (is.null(sigma2)) sigma2 = n / (nu * tau)
  if (is.null(tau))    tau    = n / (nu * sigma2)
  d   = rep(n, k)   # S = UDU' = nUU' = nP
  
  if(is.null(w)){
    w   = rnorm(k)
    w   = w / sqrt(sum(d * w^2))   # normalise to w'Dw = 1
  }
  
  fit = list(
    U         = U,
    d         = d,
    b         = b,
    n         = n,
    k         = k,
    nu        = nu,
    sigma2    = sigma2,
    w         = w,
    c         = c,     # scale factor; effective atoms are c*y_0, c*y_1
    y_0       = -1,
    y_1       = 1,
    p         = 0.5,
    objective = NULL,
    tau       = tau
  )
  return(fit)
}
```


This function initializes using a fastica strategy. It runs fastica to get $w$ (by setting $\sigma^2=c$) and then initializes $c$ to `mean(x tanh(x))` where $x=UDw$ are the pseudo-data (pre-shrunken source values).
```{r}
#initializes by first running fastICA (with c=sigma^2) to get w, and then
ebproj_init_fastica = function(U, b=0){
#  fit = ebproj_init(U, b)              # default tau=1, so sigma2 = n/nu
#  fit$c = fit$sigma2                   # set c = sigma2 = n/nu
  
  fit = ebproj_init(U, b, c=1, sigma2=1)              # default tau=1, so sigma2 = n/nu
  fit$c = fit$sigma2                   # set c = sigma2 = n/nu
  fit = ebproj_fit(fit, g_update = "none", fix_tau = TRUE)
  x = ebproj_calc_x(fit)
  
  fit = ebproj_init(U,b,w=fit$w,c=mean(x*tanh(x)), sigma2=mean(x*tanh(x)))
#  fit$c      = mean(x * tanh(x))
#  fit$sigma2 = fit$c                   # so that sigma2 = c
#  fit$tau    = fit$n / (fit$nu * fit$sigma2)
  return(fit)
}

ebproj_calc_x = function(fit){
  as.vector(fit$U %*% (fit$d * fit$w))
}
```

### Objective function

```{r}
ebproj_objective = function(fit) {
  ey0 = fit$c * fit$y_0;  ey1 = fit$c * fit$y_1   # effective atoms
  x   = ebproj_calc_x(fit)
  J   = sum(log_Z_binary(x, ey0, ey1, fit$p, fit$sigma2))
  obj = (fit$nu/2) * (log(fit$tau) - fit$tau + 1) + J
  return(obj)
}
```

The dual objective (eq 200 of the notes) is $F_{\text{EBproj}}(\mathbf{w}, g; \tau) = J(\mathbf{w}, g; \sigma^2_\text{eff}) + C(\tau)$.  It is equivalent to the primal at the optimal $q$ but is easier to compute because it avoids computing $\text{KL}(q\|g)$ explicitly.

The **primal** objective (eq 190/199) retains the full variational structure:
$$F_{\text{EBproj}}(q, g; \tau) = \frac{n - k - 2b}{2} \bigl(\log \tau - \tau(1 - \hat\rho) + 1\bigr) - \text{KL}(q \| g)$$
where $\hat\rho = \bar{\mathbf{v}}'\mathbf{P}\bar{\mathbf{v}} / U_q$, $\bar{\mathbf{v}} = E_q[\mathbf{v}]$ is the posterior mean, and $U_q = E_q[\mathbf{v}'\mathbf{v}]$ is the second moment.  The $\tau$ update targets this objective, so after a $\tau$ step the primal increases while the dual can decrease.

For the binary prior, $\text{KL}(q\|g) = \sum_i \bigl[(1-\pi_{1i})\log\tfrac{1-\pi_{1i}}{1-p} + \pi_{1i}\log\tfrac{\pi_{1i}}{p}\bigr]$ and $U_q = \sum_i \bigl[(1-\pi_{1i})(cy_0)^2 + \pi_{1i}(cy_1)^2\bigr]$.

```{r}
kl_binary = function(pi1, p) {
  pi0 = 1 - pi1
  # clamp to avoid log(0)
  pi0 = pmax(pi0, 1e-300);  pi1 = pmax(pi1, 1e-300)
  sum(pi0 * log(pi0 / (1 - p)) + pi1 * log(pi1 / p))
}

ebproj_primal_objective = function(fit) {
  ey0 = fit$c * fit$y_0;  ey1 = fit$c * fit$y_1
  x    = ebproj_calc_x(fit)
  pi1  = post_prob1(x, ey0, ey1, fit$p, fit$sigma2)
  vbar = post_mean_binary(x, ey0, ey1, fit$p, fit$sigma2)

  Uv   = as.vector(t(fit$U) %*% vbar)
  Uq   = sum((1 - pi1) * ey0^2 + pi1 * ey1^2)   # E_q[v'v] second moment
  # rho_hat = vbar' P vbar / U_q, with P = S/n and S = UDU'
  rho  = sum(fit$d * Uv^2) / (fit$n * Uq)

  kl   = kl_binary(pi1, fit$p)
  obj  = (fit$nu/2) * (log(fit$tau) - fit$tau * (1 - rho) + 1) - kl
  return(obj)
}
```

### Prior parameterisation and update modes

The binary prior is parameterised as $(c, g_0)$ where $g_0 = (1-p)\delta_{y_0} + p\delta_{y_1}$ is a **base prior** and $c > 0$ is a **scale factor**.  The effective prior is
$$g = (1-p)\delta_{cy_0} + p\delta_{cy_1},$$
so the effective atoms are $cy_0$ and $cy_1$.  This parameterisation is over-complete — multiplying $c$ by any positive constant and dividing $y_0, y_1$ by the same constant leaves $g$ unchanged — but the redundancy is useful because it separates "shape" from "scale".

A natural constraint is to require $g_0$ to have **mean 0 and unit variance**:
$$\mathbb{E}_{g_0}[v] = 0 \quad \text{and} \quad \text{Var}_{g_0}(v) = 1.$$
Under these constraints $g_0$ has a single free parameter $p$, with atoms determined by
$$y_1 = \sqrt{\frac{1-p}{p}}, \qquad y_0 = -\sqrt{\frac{p}{1-p}}.$$
The scale $c$ then controls the overall amplitude of $g$.

Three update functions are provided, selected via the `g_update` argument of `ebproj_fit`:

| `g_update` | function called | what is updated |
|------------|-----------------|-----------------|
| `"g"`      | `ebproj_update_g`  | full EM: $(p, y_0, y_1)$, rescale $g_0$ unit variance |
| `"c"`      | `ebproj_update_c`  | scale $c$ only; $g_0 = (p, y_0, y_1)$ fixed |
| `"g0"`     | `ebproj_update_g0` | constrained $g_0$: optimise $p$ with $y_0, y_1$ set by mean-0, var-1 constraint; $c$ fixed |
| `"none"`   | —               | freeze $g$ entirely |

**`ebproj_update_g`** (full EM): updates $p$, $y_0$, $y_1$ jointly via their EM closed forms, then rescales so that $\text{Var}(g_0) = 1$, storing the scale in $c$.

**`ebproj_update_c`**: fixes $g_0 = (p, y_0, y_1)$ and finds $c$ by exact 1-D maximisation of $J(w, cg_0)$ over $c > 0$ on the log scale.

**`ebproj_update_g0`**: fixes $c$ and optimises $p \in (0,1)$ subject to the mean-0, unit-variance constraint, via 1-D numerical maximisation.  After updating $p$ the atoms are recomputed as $y_1 = \sqrt{(1-p)/p}$, $y_0 = -\sqrt{p/(1-p)}$.

Note that only the last of these constrains $g_0$ to have zero mean, so that update is a bit inconsistent with the others. For example, if you estimate $g$ freely, and then perfom `ebproj_update_g0` the solution will likely change to satisfy the constraint.

### Update functions

Each function takes a fit object and returns a fit object.

The Newton update for $w$ (Section 2.3 of the notes) solves the constrained problem with the approximate Newton step
$$w_{\text{new}} = (\lambda I_k - \tilde{H})^{-1}(w_\text{target} - \tilde{H}w)$$
where $w_\text{target} = U^T M(x)$ and $\lambda = \max(w^T D w_\text{target},\, \max_j \tilde{H}_{jj} + \epsilon)$.  Five choices of diagonal preconditioner $\tilde{H}$ are supported (default `"ica"`):

1. **`"diag"`** (Full Diagonal): $\tilde{H}_{jj} = c_j d_j$ where $c_j = \sum_i U_{ij}^2 M'(x_i)$.
2. **`"trace"`** (Trace): $\tilde{H}_{jj} = c_\text{trace}\, d_j$ where $c_\text{trace} = \sum_i M'(x_i) S_{ii} / \operatorname{tr}(D)$ and $S_{ii} = \sum_j U_{ij}^2 d_j$.
3. **`"iso"`** (Isotropic): $\tilde{H}_{jj} = c_\text{iso}$ where $c_\text{iso} = (1/k)\sum_i M'(x_i) S_{ii}$.
4. **`"ica"`** (FastICA): $\tilde{H}_{jj} = c_\text{ica}$ where $c_\text{ica} = \sum_i M'(x_i)$ — the isotropic approximation under the uniform-feature-variance assumption $S_{ii} \approx k$ (Lemma 2.1).
5. **`"none"`** ($\tilde{H} = 0$): recovers the natural power / gradient step $w_\text{new} \propto w_\text{target}$.

```{r}
ebproj_update_w = function(fit, hessian = "ica", eps = 1e-6) {
  ey0 = fit$c * fit$y_0;  ey1 = fit$c * fit$y_1
  x        = ebproj_calc_x(fit)          # UDw, length n
  Mx       = post_mean_binary(x, ey0, ey1, fit$p, fit$sigma2)
  Mpx      = post_mean_deriv_binary(x, ey0, ey1, fit$p, fit$sigma2)

  w_target = as.vector(t(fit$U) %*% Mx)                    # U^T M(x), length k

  # diagonal of S: S_ii = sum_j U_ij^2 d_j
  S_diag = as.vector((fit$U^2) %*% fit$d)                  # length n

  H_diag = switch(hessian,
    diag  = { c_j = as.vector(t(fit$U^2) %*% Mpx); c_j * fit$d },
    trace = { c_t = sum(Mpx * S_diag) / sum(fit$d); c_t * fit$d },
    iso   = { c_i = sum(Mpx * S_diag) / fit$k; rep(c_i, fit$k) },
    ica   = { c_c = sum(Mpx); rep(c_c, fit$k) },
    none  = rep(0, fit$k),
    stop("hessian must be one of: 'diag', 'trace', 'iso', 'ica', 'none'")
  )

  # Lagrange multiplier; ensure lambda > max(H_diag) for positive definiteness
  lambda = max(sum(fit$d * fit$w * w_target), max(H_diag) + eps)

  w_new  = (w_target - H_diag * fit$w) / (lambda - H_diag)
  # normalise to w'Dw = 1
  scale  = sqrt(sum(fit$d * w_new^2))
  fit$w  = w_new / scale
  return(fit)
}

# Full EM update for binary prior: update (p, y_0, y_1), then rescale so that
# Var(g0) = 1, absorbing the scale into c.
ebproj_update_g = function(fit) {
  ey0 = fit$c * fit$y_0;  ey1 = fit$c * fit$y_1
  x   = ebproj_calc_x(fit)
  pi1 = post_prob1(x, ey0, ey1, fit$p, fit$sigma2)
  pi0 = 1 - pi1

  fit$p  = mean(pi1)
  y0_new = sum(pi0 * x) / sum(pi0)
  y1_new = sum(pi1 * x) / sum(pi1)

  # rescale so Var(g0) = 1; c carries the scale
  var_g  = (1 - fit$p) * y0_new^2 + fit$p * y1_new^2 -
           ((1 - fit$p) * y0_new + fit$p * y1_new)^2
  fit$c   = sqrt(var_g)
  fit$y_0 = y0_new / fit$c
  fit$y_1 = y1_new / fit$c
  return(fit)
}

# Scale-only update: fix the base prior g_0 = (p, y_0, y_1) and optimise c by
# maximising J(w, c*g_0) over c > 0 via 1-D exact optimisation.
ebproj_update_c = function(fit) {
  x = ebproj_calc_x(fit)
  obj_c = function(log_c) {
    cc = exp(log_c)
    sum(log_Z_binary(x, cc * fit$y_0, cc * fit$y_1, fit$p, fit$sigma2))
  }
  res   = optimize(obj_c, c(-5, 5), maximum = TRUE)
  fit$c = exp(res$maximum)
  return(fit)
}

# Constrained g_0 update: fix c, optimise p with y0, y1 determined by
# mean-0, unit-variance constraint: y1 = sqrt((1-p)/p), y0 = -sqrt(p/(1-p)).
ebproj_update_g0 = function(fit) {
  x = ebproj_calc_x(fit)
  obj_p = function(logit_p) {
    pp  = 1 / (1 + exp(-logit_p))
    y1  =  sqrt((1 - pp) / pp)
    y0  = -sqrt(pp / (1 - pp))
    sum(log_Z_binary(x, fit$c * y0, fit$c * y1, pp, fit$sigma2))
  }
  res      = optimize(obj_p, c(-5, 5), maximum = TRUE)
  pp       = 1 / (1 + exp(-res$maximum))
  fit$p    = pp
  fit$y_1  =  sqrt((1 - pp) / pp)
  fit$y_0  = -sqrt(pp / (1 - pp))
  return(fit)
}

# update tau using rho = vbar' P vbar / U_q, tau = 1/(1-rho); refresh sigma2
ebproj_update_tau = function(fit) {
  ey0  = fit$c * fit$y_0;  ey1 = fit$c * fit$y_1
  x    = ebproj_calc_x(fit)
  pi1  = post_prob1(x, ey0, ey1, fit$p, fit$sigma2)
  vbar = post_mean_binary(x, ey0, ey1, fit$p, fit$sigma2)

  # P = S/n = U D U'/n, so vbar' P vbar = (1/n) (U'vbar)' D (U'vbar)
  Uv         = as.vector(t(fit$U) %*% vbar)
  Uq         = sum((1 - pi1) * ey0^2 + pi1 * ey1^2)   # E_q[v'v] second moment
  rho        = sum(fit$d * Uv^2) / (fit$n * Uq)
  fit$tau    = 1 / (1 - rho)
  fit$sigma2 = fit$n / (fit$nu * fit$tau)
  return(fit)
}
```



### Main fitting loop


```{r}
# Updates tau, g, w in that order.
# g_update: "g"    = full EM update of (p, y_0, y_1), reset c = 1
#           "c"    = update scale c only, fix g_0
#           "g0"   = update p under mean-0, unit-var constraint, fix c
#           "none" = freeze g entirely
ebproj_fit = function(fit, hessian = "ica",
                      fix_w = FALSE, g_update = "g", fix_tau = FALSE,
                      max_iter = 200, tol = 1e-6, verbose = FALSE) {
  if (!fix_tau && fix_w)
    warning("updating tau without updating w: the objective may not increase")
  fit$objective = ebproj_objective(fit)

  for (iter in seq_len(max_iter)) {
    if (!fix_tau) fit = ebproj_update_tau(fit)

    fit = switch(g_update,
      g    = ebproj_update_g(fit),
      c    = ebproj_update_c(fit),
      g0   = ebproj_update_g0(fit),
      none = fit,
      stop("g_update must be one of: 'g', 'c', 'g0', 'none'")
    )

    if (!fix_w) fit = ebproj_update_w(fit, hessian = hessian)
    
    obj_new = ebproj_objective(fit)

    if (verbose)
      cat(sprintf("iter %3d  obj = %.6f  tau = %.2f  p = %.3f  c = %.4f y0 = %.2f  y1 = %.2f\n",
                  iter, obj_new, fit$tau, fit$p, fit$c, fit$y_0, fit$y_1))

    if (!is.null(fit$objective) && abs(obj_new - fit$objective) < tol) {
      fit$objective = obj_new
      break
    }
    fit$objective = obj_new
  }
  fit$iter = iter
  if(iter==max_iter){
    warning("Warning: iterations reached max_iter iterations")
  }
  return(fit)
}
```

## Simple test 1a (p=0.5)

Generate data from a single asymmetric binary source with $p = 0.5$.
```{r}
set.seed(1)
n = 200; p_dim = 1000; k = 10

# true source: binary with p = 0.5
p_true = 0.5
s_raw  = as.numeric(runif(n) < p_true)
s_true = (s_raw - p_true) / sqrt(p_true * (1 - p_true))   # zero-mean, unit-var

# mix into high-dimensional data
a = rnorm(p_dim)
a = a / sqrt(sum(a^2))
X = outer(a, s_true) + matrix(rnorm(p_dim * n, 0, 0.1), p_dim, n)
```

The data matrix $X$ is $p \times n$. 
To whiten $X$ we center it and take the first $k$ right singular vectors ($U$ is $n \times k$).
```{r}
Xcenter = X - rowMeans(X)
sv = svd(Xcenter, nu = 0, nv = k)
U  = sv$v              # n x k
```


### Fit fastica

Now fit the model. I start by using `ebproj_init_fastica` to run fastica:
it does this by using the `ebproj` updates, but setting `sigma2=c` and updating only `w`.
The init function then sets `sigma2=c=mean(x tanh(x))`, at which point the solution should be stationary at both $w$ and $c$. Here I check this by checking that the solution does not change when I update $c$ (fix tau and g0):
```{r}
fit = ebproj_init_fastica(U,b=k/2)
fit = ebproj_fit(fit, max_iter = 200, g_update = "none", fix_tau = TRUE, verbose = TRUE)
fit = ebproj_fit(fit, max_iter = 200, g_update = "c", fix_tau = TRUE, verbose = TRUE)
# check recovery: x = UDw gives the n-vector of projections
x_hat = ebproj_calc_x(fit)
cat(sprintf("Correlation with true source: %.4f\n", abs(cor(x_hat, s_true))))
```

### Run EBproj

Now we run the full EBproj updates from the fastica solution. 
```{r}
fit = ebproj_fit(fit, max_iter = 200, verbose = TRUE)
cat(sprintf("Converged in %d iterations\n", fit$iter))
cat(sprintf("Estimated p = %.3f (true = %.3f)\n", fit$p, p_true))
cat(sprintf("y_0 = %.3f, y_1 = %.3f\n", fit$y_0, fit$y_1))

# check recovery: x = UDw gives the n-vector of projections
x_hat = ebproj_calc_x(fit)
cat(sprintf("Correlation with true source: %.4f\n", abs(cor(x_hat, s_true))))
```

### Random initialization

Here we randomly initialize and plot objective over iterations to check convergence. (Note this example appears to be converging to the null solution, where the objective is 0).
```{r}
set.seed(2)
fit_verbose = ebproj_init(U, b = 1)
objs = numeric(50)
for (iter in 1:50) {
  fit_verbose = ebproj_update_tau(fit_verbose)
  fit_verbose = ebproj_update_g(fit_verbose)
  fit_verbose = ebproj_update_w(fit_verbose)
  objs[iter]  = ebproj_objective(fit_verbose)
}
plot(objs, type = "l", xlab = "iteration", ylab = "objective",
     main = "EBProj objective over iterations")
```

This is the same but using a gradient step instead of Newton:
```{r}
set.seed(2)
fit_verbose = ebproj_init(U, b = 1)
objs = numeric(50)
for (iter in 1:50) {
  fit_verbose = ebproj_update_tau(fit_verbose)
  fit_verbose = ebproj_update_g(fit_verbose)
  fit_verbose = ebproj_update_w(fit_verbose, hessian="none")
  objs[iter]  = ebproj_objective(fit_verbose)
}
plot(objs, type = "l", xlab = "iteration", ylab = "objective",
     main = "EBProj objective over iterations")
```

Here I note that the tau update is not actually guaranteed to increase the objective (nor the primal objective). This is because the tau update is designed to increase O(q,g,tau) for a fixed q,g, but in the way I am doing things (working with w,g,tau) the q defined by (w,tau) changes implicitly when tau changes. This is OK: the new tau is optimizing O(q,g,tau) for the current implicit q (=q(current_w,current_tau) assumed fixed), and as long as we go on to update w, it will find (implicitly) the new q that optimizes O(q,g,tau) this new tau. But the potential for non-monotonic behavior of the obective (after a tau update, before w update) is something to be aware of. And we should probably add a warning if the function is called to update tau with w fixed.

```{r}
set.seed(2)
fit_verbose = ebproj_init(U, b = 1)
objs = numeric(50)
objs_primal = numeric(50)
for (iter in 1:50) {
  fit_verbose = ebproj_update_tau(fit_verbose)
  objs[iter]  = ebproj_objective(fit_verbose)
  objs_primal[iter] = ebproj_primal_objective(fit_verbose)
}
plot(objs, type = "l", xlab = "iteration", ylab = "objective",
     main = "EBProj objective over iterations")
plot(objs_primal, type = "l", xlab = "iteration", ylab = "objective",
     main = "EBProj objective over iterations")
```

### fastica init

This is the trace of the objective initializing with fastica 

```{r}
# plot objective over iterations to check convergence
set.seed(2)
fit_verbose = ebproj_init_fastica(U, b = 1)
objs = numeric(50)
for (iter in 1:50) {
  fit_verbose = ebproj_update_tau(fit_verbose)
  fit_verbose = ebproj_update_g(fit_verbose)
  fit_verbose = ebproj_update_w(fit_verbose)
  objs[iter]  = ebproj_objective(fit_verbose)
}
plot(objs, type = "l", xlab = "iteration", ylab = "objective",
     main = "EBProj objective over iterations")
```

## Simple test 1b (p=0.1)

Generate data from a single asymmetric binary source with $p = 0.1$.
```{r}
set.seed(1)
n = 200; p_dim = 1000; k = 10

# true source: binary with p = 0.1
p_true = 0.1
s_raw  = as.numeric(runif(n) < p_true)
s_true = (s_raw - p_true) / sqrt(p_true * (1 - p_true))   # zero-mean, unit-var

# mix into high-dimensional data
a = rnorm(p_dim)
a = a / sqrt(sum(a^2))
X = outer(a, s_true) + matrix(rnorm(p_dim * n, 0, 0.1), p_dim, n)
```

The data matrix $X$ is $p \times n$.
To whiten $X$ we center it and take the first $k$ right singular vectors ($U$ is $n \times k$).
```{r}
Xcenter = X - rowMeans(X)
sv = svd(Xcenter, nu = 0, nv = k)
U  = sv$v              # n x k
```

Now fit the model. I initialize at fastica solution (run fastica by setting sigma2=c and
updating only w); then set $sigma2=c=mean(x tanh(x))$ and check that the solution does not change when I update c (fix tau and g0). fastica finds the true source in this instance, but it is actually a local minimum rather than local maximum of the fastica objective. 
```{r}
fit = ebproj_init_fastica(U, b = k/2)
fit = ebproj_fit(fit, max_iter = 200, g_update = "none", fix_tau = TRUE, verbose = TRUE)
fit = ebproj_fit(fit, max_iter = 200, g_update = "c", fix_tau = TRUE, verbose = TRUE)
# check recovery: x = UDw gives the n-vector of projections
x_hat = ebproj_calc_x(fit)
cat(sprintf("Correlation with true source: %.4f\n", abs(cor(x_hat, s_true))))
```


Now I run from the fastica solution. It kind of converges (slowly) to the null solution (c moving to 0). Even though the correlation with the true source remains high,
it is not really finding a good solution in my opinion; what I actually wanted it to do was to realize that the prior should be a 0.1,0.9 binary mixture and move there. I think the objective should be much higher at that point. 
```{r}
fit = ebproj_fit(fit, max_iter = 200, verbose = TRUE)
cat(sprintf("Converged in %d iterations\n", fit$iter))
cat(sprintf("Estimated p = %.3f (true = %.3f)\n", fit$p, p_true))
cat(sprintf("y_0 = %.3f, y_1 = %.3f\n", fit$y_0, fit$y_1))

# check recovery: x = UDw gives the n-vector of projections
x_hat = ebproj_calc_x(fit)
cat(sprintf("Correlation with true source: %.4f\n", abs(cor(x_hat, s_true))))
hist(x_hat)
```

It turns out this is solved elegantly by using a different update of $g$ which constrains g0 to be mean 0 and unit variance, and so there is a single parameter $p$ that controls $y0,y1,p$. This helps because in order to move the prior support (y0,y1) toward the xhat values, it has no option but to change p, which is exactly the right thing to do. 
```{r}
fit$c = 1
fit = ebproj_fit(fit, max_iter = 200, verbose=TRUE, g_update="g0")
```

Then we can switch to the general update if we want to refine even further:
```{r}
fit = ebproj_fit(fit, max_iter = 200, verbose=TRUE)
```


### Random start

Try a random start, and plot objective over iterations. Interestingly we see the non-monotonic behavior here, and it does not converge to the true source. 
```{r}
set.seed(2)
fit_verbose = ebproj_init(U, b = 1)
objs = numeric(50)
for (iter in 1:50) {
  fit_verbose = ebproj_update_tau(fit_verbose)
  fit_verbose = ebproj_update_g(fit_verbose)
  fit_verbose = ebproj_update_w(fit_verbose)
  objs[iter]  = ebproj_objective(fit_verbose)
}
plot(objs, type = "l", xlab = "iteration", ylab = "objective",
     main = "EBProj objective over iterations")
x_hat = ebproj_calc_x(fit_verbose)
cat(sprintf("Correlation with true source: %.4f\n", abs(cor(x_hat, s_true))))
hist(x_hat)
```


### Initialization with tau fixed

Try with tau fixed (ie sigma2 fixed) with sigma2 set to $c_golden$ (the inverse of the golden ratio, which obeys $c_\text{golden}^2 + c_\text{golden} = 1$). It finds a true source but it does so in a weird way, with objective worse than null, almost like it is finding a local minimum again.
```{r}
set.seed(1)
c_golden = 2/(1+sqrt(5))
fit = ebproj_init(U, b = k/2, c = c_golden, sigma2 = c_golden)
fit = ebproj_fit(fit, max_iter = 200, fix_tau=TRUE, verbose = TRUE)
x_hat = ebproj_calc_x(fit)
cat(sprintf("Correlation with true source: %.4f\n", abs(cor(x_hat, s_true))))
hist(x_hat)
```

Running from that soluion with the g0 update behaves similarly to how it behaves when initializing from fastica local minimum.
```{r}
fit = ebproj_fit(fit, max_iter = 200, verbose = TRUE, g_update = "g0")
x_hat = ebproj_calc_x(fit)
cat(sprintf("Correlation with true source: %.4f\n", abs(cor(x_hat, s_true))))
hist(x_hat)
```




## Simple test 1c (p=0.2)

Generate data from a single asymmetric binary source with $p = 0.2$. Here fastICA should work less well because this source has intermediate value of log cosh.
```{r}
set.seed(1)
n = 200; p_dim = 1000; k = 10

# true source: binary with p = 0.2
p_true = 0.2
s_raw  = as.numeric(runif(n) < p_true)
s_true = (s_raw - p_true) / sqrt(p_true * (1 - p_true))   # zero-mean, unit-var

# mix into high-dimensional data
a = rnorm(p_dim)
a = a / sqrt(sum(a^2))
X = outer(a, s_true) + matrix(rnorm(p_dim * n, 0, 0.1), p_dim, n)
```

The data matrix $X$ is $p \times n$.
To whiten $X$ we center it and take the first $k$ right singular vectors ($U$ is $n \times k$).
```{r}
Xcenter = X - rowMeans(X)
sv = svd(Xcenter, nu = 0, nv = k)
U  = sv$v              # n x k
```

Now fit the model. I initialize using fastica. This time the initial fastica run does not actually converge in 200 interation so the update does change things slightly. Here fastica does not find a true source.
```{r}
set.seed(3)
fit.ica = ebproj_init_fastica(U, b = k/2)
fit.ica = ebproj_fit(fit.ica, max_iter = 200, g_update = "none", fix_tau = TRUE, verbose = TRUE)
fit.ica = ebproj_fit(fit.ica, max_iter = 200, g_update = "c", fix_tau = TRUE, verbose = TRUE)
# check recovery: x = UDw gives the n-vector of projections
x_hat = ebproj_calc_x(fit.ica)
cat(sprintf("Correlation with true source: %.4f\n", abs(cor(x_hat, s_true))))
hist(x_hat)
```


Now I run from the fastica solution. It converges to something worse than the null, and does not find a true source.
```{r}
fit = ebproj_fit(fit.ica, max_iter = 200, verbose = TRUE)
cat(sprintf("Converged in %d iterations\n", fit$iter))
cat(sprintf("Estimated p = %.3f (true = %.3f)\n", fit$p, p_true))
cat(sprintf("y_0 = %.3f, y_1 = %.3f\n", fit$y_0, fit$y_1))

# check recovery: x = UDw gives the n-vector of projections
x_hat = ebproj_calc_x(fit)
cat(sprintf("Correlation with true source: %.4f\n", abs(cor(x_hat, s_true))))
hist(x_hat)
```

Now I run from the fastica solution but updating only g0. It is better than the null but not a true source.
```{r}
fit = ebproj_fit(fit.ica, max_iter = 200, g_update="g0",verbose = TRUE)
cat(sprintf("Converged in %d iterations\n", fit$iter))
cat(sprintf("Estimated p = %.3f (true = %.3f)\n", fit$p, p_true))
cat(sprintf("y_0 = %.3f, y_1 = %.3f\n", fit$y_0, fit$y_1))

# check recovery: x = UDw gives the n-vector of projections
x_hat = ebproj_calc_x(fit)
cat(sprintf("Correlation with true source: %.4f\n", abs(cor(x_hat, s_true))))
hist(x_hat)
```


### Another start

Interestingly, with a different random start fastica converges to the right solution (as a local minima).
```{r}
set.seed(2)
fit.ica = ebproj_init_fastica(U, b = k/2)
fit.ica = ebproj_fit(fit.ica, max_iter = 200, g_update = "none", fix_tau = TRUE, verbose = TRUE)
fit.ica = ebproj_fit(fit.ica, max_iter = 200, g_update = "c", fix_tau = TRUE, verbose = TRUE)
# check recovery: x = UDw gives the n-vector of projections
x_hat = ebproj_calc_x(fit.ica)
cat(sprintf("Correlation with true source: %.4f\n", abs(cor(x_hat, s_true))))
hist(x_hat)
```

Now I run from the fastica solution. If we do an unconstrained update the objective is bad but it stays near the fastica solution (so still a true source!). Updating g0 only finds a better objective value.
```{r}
fit.g = ebproj_fit(fit.ica, verbose = TRUE)
fit.g0 = ebproj_fit(fit.ica, g_update = "g0", verbose=TRUE)

# check recovery: x = UDw gives the n-vector of projections
x_hat = ebproj_calc_x(fit.g)
cat(sprintf("Correlation with true source: %.4f\n", abs(cor(x_hat, s_true))))
hist(x_hat)

# check recovery: x = UDw gives the n-vector of projections
x_hat = ebproj_calc_x(fit.g0)
cat(sprintf("Correlation with true source: %.4f\n", abs(cor(x_hat, s_true))))
hist(x_hat)
```



### Initialization with tau fixed

Try with tau fixed (ie sigma2 fixed) with sigma2 set to $c_golden$ (the inverse of the golden ratio, which obeys $c_\text{golden}^2 + c_\text{golden} = 1$). (I tried seeds 1-5. 1, 4 and 5 find the true source but 2-3 do not.)
```{r}
set.seed(1)
c_golden = 2/(1+sqrt(5))
fit = ebproj_init(U, b = 1, c = c_golden, sigma2 = c_golden)
fit = ebproj_fit(fit, max_iter = 200, fix_tau=TRUE, verbose = TRUE)
x_hat = ebproj_calc_x(fit)
cat(sprintf("Correlation with true source: %.4f\n", abs(cor(x_hat, s_true))))
```


```{r}
set.seed(2)
c_golden = 2/(1+sqrt(5))
fit = ebproj_init(U, b = 1, c = c_golden, sigma2 = c_golden)
fit = ebproj_fit(fit, max_iter = 200, fix_tau=TRUE, verbose = TRUE)
x_hat = ebproj_calc_x(fit)
cat(sprintf("Correlation with true source: %.4f\n", abs(cor(x_hat, s_true))))
hist(x_hat)
```

Here we try updating only g0 and fixing c too.  
```{r}
set.seed(5)
fit = ebproj_init(U, b = 1, c = c_golden, sigma2 = c_golden)
fit = ebproj_fit(fit, max_iter = 200, fix_tau=TRUE, g_update="g0", verbose = TRUE)
x_hat = ebproj_calc_x(fit)
cat(sprintf("Correlation with true source: %.4f\n", abs(cor(x_hat, s_true))))
hist(x_hat)
```



# Test 2: Single source p=0.2

### Test 2a: repeat Test 1c 

I was struck by the fact that in the test above, fastICA did sometimes work when p=0.2.
This was different from my (slightly different) simulation with p=0.2 in
fast_ICA_asymmetric.Rmd where log-cosh failed from all 100 random seeds.

Here I repeat Test 1c exactly as before but with 10 different seeds.
For each seed I run fastICA 100 times from different random starting points and document what proportion of the time it finds a true source. I do the same with EBproj

```{r}
# Runs fastica from n_starts different starting points and assesses correlation with the true source s_true for each one
run_fastica_seeds = function(U, s_true, b, n_starts = 100) {
  cors = numeric(n_starts)
  for (i in seq_len(n_starts)) {
    fit = ebproj_init_fastica(U, b = b)
    x_hat = ebproj_calc_x(fit)
    cors[i] = abs(cor(x_hat, s_true))
  }
  cors
}

run_ebproj_golden_seeds = function(U, s_true, b, n_starts = 100) {
  c_golden = 2 / (1 + sqrt(5))
  cors = numeric(n_starts)
  for (i in seq_len(n_starts)) {
    fit.init = ebproj_init(U, b = b, c = c_golden, sigma2 = c_golden)
    fit = suppressWarnings(ebproj_fit(fit.init, fix_tau = TRUE, g_update = "g0"))
    x_hat = ebproj_calc_x(fit)
    cors[i] = abs(cor(x_hat, s_true))
  }
  cors
}


run_experiment = function(p_true, n_data_seeds = 10, n_starts = 100,
                          cor_thresh = 0.9, master_seed = 42,
                          n = 200, p_dim = 1000, k = 10) {
  set.seed(master_seed)
  data_seeds = sample.int(1e6, n_data_seeds)
  results = vector("list", n_data_seeds)
  for (di in seq_len(n_data_seeds)) {
    set.seed(data_seeds[di])
    s_raw  = as.numeric(runif(n) < p_true)
    s_true = (s_raw - p_true) / sqrt(p_true * (1 - p_true))
    a = rnorm(p_dim); a = a / sqrt(sum(a^2))
    X = outer(a, s_true) + matrix(rnorm(p_dim * n, 0, 0.1), p_dim, n)
    Xcenter = X - rowMeans(X)
    U = svd(Xcenter, nu = 0, nv = k)$v

    cors_ica    = suppressWarnings(run_fastica_seeds(U, s_true, b = k/2, n_starts = n_starts))
    cors_golden = run_ebproj_golden_seeds(U, s_true, b = k/2, n_starts = n_starts)
    results[[di]] = list(data_seed   = data_seeds[di],
                         frac_ica    = mean(cors_ica    > cor_thresh),
                         frac_golden = mean(cors_golden > cor_thresh))
    cat(sprintf("data seed %7d: fastica %.2f  golden %.2f\n",
                data_seeds[di], results[[di]]$frac_ica, results[[di]]$frac_golden))
  }
  results
}

make_table = function(results, n_starts, cor_thresh) {
  frac_ica    = sapply(results, `[[`, "frac_ica")
  frac_golden = sapply(results, `[[`, "frac_golden")
  tab = data.frame(
    seed         = sapply(results, `[[`, "data_seed"),
    frac_fastica = round(frac_ica, 2),
    frac_golden  = round(frac_golden, 2)
  )
  tab = rbind(tab, data.frame(seed = NA,
                               frac_fastica = round(mean(frac_ica), 2),
                               frac_golden  = round(mean(frac_golden), 2)))
  rownames(tab) = c(paste0("seed ", seq_len(nrow(tab) - 1)), "mean")
  knitr::kable(tab, col.names = c("data seed", "frac found (fastICA)", "frac found (golden)"),
               caption = sprintf("Fraction of starts with |cor| > %.1f (%d starts each)",
                                 cor_thresh, n_starts))
}
```

```{r}
cat("=== p = 0.5 ===\n")
res_05 = run_experiment(p_true = 0.5)
make_table(res_05, n_starts = 100, cor_thresh = 0.9)
```

```{r}
cat("=== p = 0.2 ===\n")
res_02 = run_experiment(p_true = 0.2)
make_table(res_02, n_starts = 100, cor_thresh = 0.9)
```

```{r}
cat("=== p = 0.1 ===\n")
res_01 = run_experiment(p_true = 0.1)
make_table(res_01, n_starts = 100, cor_thresh = 0.9)
```


### Set k=20

With a larger number of (noise) dimensions and p=0.5 the performance of both methods drops, and the advantage of fastICA maybe bigger. For p=0.2 the opposite holds. For p=0.1 both methods remain good.
```{r}
res_05_k20 = run_experiment(p_true=0.5, k=20)
res_02_k20 = run_experiment(p_true=0.2, k=20)
res_01_k20 = run_experiment(p_true=0.1, k=20)
```


