---
title: "FastICA for Non-Whitened Data"
author: "Matthew Stephens"
date: "2026-07-16"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

Standard FastICA pre-whitens data: given $X$ ($p \times n$) with thin SVD
$X = UDV'$, it sets $Z = \sqrt{n}\,V'$ ($k \times n$) and maximizes
$E[G(w'z)]$ subject to $\|w\|^2 = 1$.  The constraint $\|w\|^2 = 1$ is
equivalent to $E[(w'z)^2] = n$ because $ZZ' = nI_k$ for whitened data.

Here we skip the whitening step and instead work with
$$Z = DV' \qquad (k \times n),$$
so that $ZZ' = D^2$.  The natural constraint to keep evaluation scale-free is
$$w'D^2 w = n,$$
which again enforces $E[(w'z)^2] = n$.  The FastICA fixed-point update itself
is unchanged; only the normalization step changes from $w \leftarrow w/\|w\|$
to $w \leftarrow w\sqrt{n}/\sqrt{w'D^2 w}$.

## Motivation: robustness to the number of PCs

In practice we do not know how many independent components are present.
Standard whitening requires a hard choice of $k$: too small and we miss
components, too large and we include pure-noise PCs.  The problem with
whitening is that all $k$ retained PCs are placed on equal footing — the
noise PCs get the same unit variance as the signal PCs.  This can overwhelm
the signal when $k$ is much larger than the true number of components.

With the non-whitened approach the $k$ PCs enter with their original singular
values $d_1 \ge d_2 \ge \cdots \ge d_k$.  Signal PCs have large $d_i$; excess
noise PCs have small $d_i$ and therefore small weight in the objective.  The
constraint $w'D^2 w = n$ automatically concentrates $w$ on the high-variance
directions, so inflating $k$ adds far less noise than in the whitened case.

To test this we use $k = 20$ PCs for all experiments below, even though the
true number of sources is smaller (1, 4, or 9).  We expect whitened FastICA
to degrade noticeably with the extra noise PCs, while non-whitened FastICA
should remain robust.

## Implementation

```{r}
# ---- whitened helpers (for comparison) ----

fastica_r1update = function(X, w) {
  w = w / sqrt(sum(w^2))
  P  = t(X) %*% w
  G  = tanh(P)
  G2 = 1 - tanh(P)^2
  w  = X %*% G - mean(G2) * ncol(X) * w
  w / sqrt(sum(w^2))
}

fastica_r1update_tlc = function(X, w, lambda = 1) {
  w = w / sqrt(sum(w^2))
  P  = t(X) %*% w
  G  = tanh(P) + 2 * lambda * abs(P)
  G2 = 1 - tanh(P)^2 + 2 * lambda * sign(P)
  w  = X %*% G - mean(G2) * ncol(X) * w
  w / sqrt(sum(w^2))
}

compute_objective     = function(X, w) mean(log(cosh(t(X) %*% w)))
compute_objective_tlc = function(X, w, lambda = 1) {
  P = t(X) %*% w
  mean(log(cosh(P)) + lambda * abs(P) * P)
}

prewhiten = function(X, n.comp) {
  X = X - rowMeans(X)
  sqrt(ncol(X)) * t(svd(X)$v[, 1:n.comp])
}

# ---- non-whitened helpers ----

# Returns list(Z, D2): Z = DV' (k x n), D2 = diag(d^2) (k x k)
preprocess_nonwhitened = function(X, n.comp) {
  X  = X - rowMeans(X)
  sv = svd(X, nu = 0, nv = n.comp)
  d  = sv$d[1:n.comp]
  Z  = diag(d, nrow = length(d)) %*% t(sv$v)   # k x n
  D2 = diag(d^2, nrow = length(d))              # k x k
  list(Z = Z, D2 = D2)
}

# Normalize w so that w'D2 w = n
normalize_nw = function(w, D2, n) {
  w * sqrt(n) / sqrt(as.numeric(t(w) %*% D2 %*% w))
}

fastica_r1update_nw = function(Z, D2, w) {
  n  = ncol(Z)
  w  = normalize_nw(w, D2, n)
  P  = t(Z) %*% w
  G  = tanh(P)
  G2 = 1 - tanh(P)^2
  w  = Z %*% G - mean(G2) * n * w
  normalize_nw(w, D2, n)
}

fastica_r1update_nw_tlc = function(Z, D2, w, lambda = 1) {
  n  = ncol(Z)
  w  = normalize_nw(w, D2, n)
  P  = t(Z) %*% w
  G  = tanh(P) + 2 * lambda * abs(P)
  G2 = 1 - tanh(P)^2 + 2 * lambda * sign(P)
  w  = Z %*% G - mean(G2) * n * w
  normalize_nw(w, D2, n)
}

compute_objective_nw = function(Z, D2, w) {
  n = ncol(Z)
  w = normalize_nw(w, D2, n)
  mean(log(cosh(t(Z) %*% w)))
}

compute_objective_nw_tlc = function(Z, D2, w, lambda = 1) {
  n = ncol(Z)
  w = normalize_nw(w, D2, n)
  P = t(Z) %*% w
  mean(log(cosh(P)) + lambda * abs(P) * P)
}

# ---- generic runner: works for both whitened and non-whitened ----
# update_fn(data, w, ...) and obj_fn(data, w, ...) where data is a list
# (non-whitened) or a matrix (whitened).

run_seeds = function(data, S_true, update_fn, obj_fn, n_seeds = 50,
                     n_iter = 200, ...) {
  obj    = numeric(n_seeds)
  maxcor = numeric(n_seeds)
  for (seed in seq_len(n_seeds)) {
    set.seed(seed)
    w = rnorm(nrow(if (is.list(data)) data$Z else data))
    for (i in seq_len(n_iter))
      w = update_fn(data, w, ...)
    obj[seed]    = obj_fn(data, w, ...)
    proj         = if (is.list(data)) t(data$Z) %*% w else t(data) %*% w
    maxcor[seed] = max(abs(cor(t(S_true), proj)))
  }
  list(obj = obj, maxcor = maxcor)
}

# Wrappers so run_seeds gets a single 'data' argument
update_nw = function(data, w, lambda = NULL) {
  if (is.null(lambda)) fastica_r1update_nw(data$Z, data$D2, w)
  else fastica_r1update_nw_tlc(data$Z, data$D2, w, lambda = lambda)
}
obj_nw = function(data, w, lambda = NULL) {
  if (is.null(lambda)) compute_objective_nw(data$Z, data$D2, w)
  else compute_objective_nw_tlc(data$Z, data$D2, w, lambda = lambda)
}

k_pca = 20   # number of PCs used for all tests (larger than true k)
```

## Test 1: Symmetric binary source (1 true source, 20 PCs)

The data have a single Rademacher source, but we extract `k_pca = 20` PCs.
Whitening inflates all 20 to unit variance; non-whitening retains the true
signal-to-noise contrast in the singular values.

### Singular values

```{r}
set.seed(10)
n = 200; p = 1000; k_true = 1
A     = matrix(rnorm(p * k_true), nrow = p)
S_rad = matrix(sample(c(-1, 1), n, replace = TRUE), nrow = 1)
sigma = 0.1
X_rad = A %*% S_rad + matrix(rnorm(p * n, 0, sigma), nrow = p)

sv_rad = svd(X_rad - rowMeans(X_rad), nu = 0, nv = k_pca)
plot(sv_rad$d[1:k_pca], type = "b", pch = 19, cex = 0.7,
     xlab = "PC index", ylab = "singular value",
     main = "Top 20 singular values (1 true source)")
abline(v = k_true + 0.5, lty = 2, col = "red")
legend("topright", "true k", lty = 2, col = "red")
```

The first singular value is much larger than the rest, which are pure noise.
Whitening would equalise all 20; non-whitening keeps this gap intact.

### Single-seed detailed comparison

```{r}
Z_rad  = prewhiten(X_rad, k_pca)
nw_rad = preprocess_nonwhitened(X_rad, k_pca)

run_one = function(data, update_fn, obj_fn, seed = 1, n_iter = 200, ...) {
  set.seed(seed)
  w = rnorm(nrow(if (is.list(data)) data$Z else data))
  for (i in seq_len(n_iter))
    w = update_fn(data, w, ...)
  proj = if (is.list(data)) t(data$Z) %*% w else t(data) %*% w
  list(w = w, proj = as.vector(proj), obj = obj_fn(data, w, ...))
}

r_wh_lc  = run_one(Z_rad,  fastica_r1update,     compute_objective)
r_wh_tlc = run_one(Z_rad,  fastica_r1update_tlc, compute_objective_tlc, lambda = 1)
r_nw_lc  = run_one(nw_rad, update_nw,            obj_nw)
r_nw_tlc = run_one(nw_rad, update_nw,            obj_nw, lambda = 1)

cat("whitened  log-cosh: obj =", round(r_wh_lc$obj,  4),
    "  max|cor| =", round(max(abs(cor(as.vector(S_rad), r_wh_lc$proj))),  3), "\n")
cat("whitened  TLC:      obj =", round(r_wh_tlc$obj, 4),
    "  max|cor| =", round(max(abs(cor(as.vector(S_rad), r_wh_tlc$proj))), 3), "\n")
cat("nonwhiten log-cosh: obj =", round(r_nw_lc$obj,  4),
    "  max|cor| =", round(max(abs(cor(as.vector(S_rad), r_nw_lc$proj))),  3), "\n")
cat("nonwhiten TLC:      obj =", round(r_nw_tlc$obj, 4),
    "  max|cor| =", round(max(abs(cor(as.vector(S_rad), r_nw_tlc$proj))), 3), "\n")
```

### Projections w'z

```{r fig.height=6}
par(mfrow = c(2, 2))
ord = order(as.vector(S_rad))
col = ifelse(as.vector(S_rad)[ord] == 1, "tomato", "steelblue")

plot_proj = function(proj, title) {
  plot(proj[ord], col = col, pch = 19, cex = 0.5,
       main = title, xlab = "sample (sorted by true source)", ylab = "w'z",
       cex.main = 0.85)
  abline(h = 0, lty = 2)
}

plot_proj(r_wh_lc$proj,  "whitened log-cosh")
plot_proj(r_wh_tlc$proj, "whitened TLC")
plot_proj(r_nw_lc$proj,  "nonwhitened log-cosh")
plot_proj(r_nw_tlc$proj, "nonwhitened TLC")
par(mfrow = c(1, 1))
```

Samples are sorted by true source value (+1 = red, -1 = blue).
A good recovery shows two clearly separated bands.

### Many seeds

```{r}
res_wh_lc  = run_seeds(Z_rad,  S_rad, fastica_r1update,     compute_objective)
res_wh_tlc = run_seeds(Z_rad,  S_rad, fastica_r1update_tlc, compute_objective_tlc, lambda = 1)
res_nw_lc  = run_seeds(nw_rad, S_rad, update_nw,            obj_nw)
res_nw_tlc = run_seeds(nw_rad, S_rad, update_nw,            obj_nw, lambda = 1)

cat("whitened  log-cosh: mean max|cor| =", round(mean(res_wh_lc$maxcor),  3),
    " frac > 0.9:", mean(res_wh_lc$maxcor  > 0.9), "\n")
cat("whitened  TLC:      mean max|cor| =", round(mean(res_wh_tlc$maxcor), 3),
    " frac > 0.9:", mean(res_wh_tlc$maxcor > 0.9), "\n")
cat("nonwhiten log-cosh: mean max|cor| =", round(mean(res_nw_lc$maxcor),  3),
    " frac > 0.9:", mean(res_nw_lc$maxcor  > 0.9), "\n")
cat("nonwhiten TLC:      mean max|cor| =", round(mean(res_nw_tlc$maxcor), 3),
    " frac > 0.9:", mean(res_nw_tlc$maxcor > 0.9), "\n")
```

## Test 2: Non-overlapping groups (4 true sources, 20 PCs)

```{r}
set.seed(1)
n = 100; p = 1000; k_true = 4
A = matrix(rnorm(p * k_true), nrow = p)
S = matrix(0, nrow = k_true, ncol = n)
S[1, 1:25] = 1; S[2, 26:50] = 1; S[3, 51:75] = 1; S[4, 76:100] = 1
sigma = 0.1
X  = A %*% S + matrix(rnorm(p * n, 0, sigma), nrow = p)

Z  = prewhiten(X, k_pca)
nw = preprocess_nonwhitened(X, k_pca)

res_wh_lc  = run_seeds(Z,  S, fastica_r1update,     compute_objective)
res_wh_tlc = run_seeds(Z,  S, fastica_r1update_tlc, compute_objective_tlc, lambda = 1)
res_nw_lc  = run_seeds(nw, S, update_nw,            obj_nw)
res_nw_tlc = run_seeds(nw, S, update_nw,            obj_nw, lambda = 1)

par(mfrow = c(2, 2))
hist(res_wh_lc$maxcor,  breaks = seq(0, 1, by = 0.05), main = "whitened log-cosh",    xlab = "max |cor|")
hist(res_wh_tlc$maxcor, breaks = seq(0, 1, by = 0.05), main = "whitened TLC",         xlab = "max |cor|")
hist(res_nw_lc$maxcor,  breaks = seq(0, 1, by = 0.05), main = "nonwhitened log-cosh", xlab = "max |cor|")
hist(res_nw_tlc$maxcor, breaks = seq(0, 1, by = 0.05), main = "nonwhitened TLC",      xlab = "max |cor|")
par(mfrow = c(1, 1))
```

## Test 3: Sparse binary sources (4 true sources, p=0.2, 20 PCs)

```{r}
set.seed(2)
n = 500; p = 1000; k_true = 4
prob_active = 0.2
A = matrix(rnorm(p * k_true), nrow = p)
S_sparse = matrix(0, nrow = k_true, ncol = n)
for (j in 1:k_true)
  S_sparse[j, sample(n, round(prob_active * n))] = 1
sigma = 0.1
X_sp = A %*% S_sparse + matrix(rnorm(p * n, 0, sigma), nrow = p)

Z_sp  = prewhiten(X_sp, k_pca)
nw_sp = preprocess_nonwhitened(X_sp, k_pca)

res_wh_lc_sp  = run_seeds(Z_sp,  S_sparse, fastica_r1update,     compute_objective,     n_seeds = 100)
res_wh_tlc_sp = run_seeds(Z_sp,  S_sparse, fastica_r1update_tlc, compute_objective_tlc, n_seeds = 100, lambda = 1)
res_nw_lc_sp  = run_seeds(nw_sp, S_sparse, update_nw,            obj_nw,                n_seeds = 100)
res_nw_tlc_sp = run_seeds(nw_sp, S_sparse, update_nw,            obj_nw,                n_seeds = 100, lambda = 1)

cat("whitened  log-cosh: mean max|cor| =", round(mean(res_wh_lc_sp$maxcor),  3),
    " frac > 0.9:", mean(res_wh_lc_sp$maxcor  > 0.9), "\n")
cat("whitened  TLC:      mean max|cor| =", round(mean(res_wh_tlc_sp$maxcor), 3),
    " frac > 0.9:", mean(res_wh_tlc_sp$maxcor > 0.9), "\n")
cat("nonwhiten log-cosh: mean max|cor| =", round(mean(res_nw_lc_sp$maxcor),  3),
    " frac > 0.9:", mean(res_nw_lc_sp$maxcor  > 0.9), "\n")
cat("nonwhiten TLC:      mean max|cor| =", round(mean(res_nw_tlc_sp$maxcor), 3),
    " frac > 0.9:", mean(res_nw_tlc_sp$maxcor > 0.9), "\n")

par(mfrow = c(2, 2))
hist(res_wh_lc_sp$maxcor,  breaks = seq(0, 1, by = 0.05), main = "whitened log-cosh",    xlab = "max |cor|")
hist(res_wh_tlc_sp$maxcor, breaks = seq(0, 1, by = 0.05), main = "whitened TLC",         xlab = "max |cor|")
hist(res_nw_lc_sp$maxcor,  breaks = seq(0, 1, by = 0.05), main = "nonwhitened log-cosh", xlab = "max |cor|")
hist(res_nw_tlc_sp$maxcor, breaks = seq(0, 1, by = 0.05), main = "nonwhitened TLC",      xlab = "max |cor|")
par(mfrow = c(1, 1))
```

## Test 4: 9 overlapping groups, 20 members each, n=100 (20 PCs)

Multi-component extraction via deflation.  After finding weight $w_1$, the
next search is restricted to vectors satisfying $w \perp_{D^2} w_1$, i.e.
$w_1' D^2 w = 0$, so the extracted components are orthogonal in the
$D^2$-inner-product sense (equivalently, orthogonal projections in the
original data space).  We extract 9 components from a 20-PC space.

```{r}
fastica_deflation_nw = function(nw, k, update_fn, obj_fn, n_iter = 200,
                                n_starts = 5, ...) {
  Z  = nw$Z
  D2 = nw$D2
  n  = ncol(Z)
  W  = matrix(0, nrow(Z), k)
  for (comp in seq_len(k)) {
    best_obj = -Inf; best_w = NULL
    for (s in seq_len(n_starts)) {
      set.seed(s + comp * 1000)
      w = rnorm(nrow(Z))
      if (comp > 1) {
        for (j in 1:(comp - 1))
          w = w - as.numeric(t(W[, j]) %*% D2 %*% w) /
                  as.numeric(t(W[, j]) %*% D2 %*% W[, j]) * W[, j]
      }
      w = normalize_nw(w, D2, n)
      for (i in seq_len(n_iter)) {
        w = update_fn(nw, w, ...)
        if (comp > 1) {
          for (j in 1:(comp - 1))
            w = w - as.numeric(t(W[, j]) %*% D2 %*% w) /
                    as.numeric(t(W[, j]) %*% D2 %*% W[, j]) * W[, j]
          w = normalize_nw(w, D2, n)
        }
      }
      o = obj_fn(nw, w, ...)
      if (o > best_obj) { best_obj = o; best_w = w }
    }
    W[, comp] = best_w
  }
  W
}

fastica_deflation_wh = function(X, k, update_fn, obj_fn, n_iter = 200,
                                n_starts = 5, ...) {
  W = matrix(0, nrow(X), k)
  for (comp in seq_len(k)) {
    best_obj = -Inf; best_w = NULL
    for (s in seq_len(n_starts)) {
      set.seed(s + comp * 1000)
      w = rnorm(nrow(X))
      if (comp > 1)
        w = w - W[, 1:(comp-1), drop=FALSE] %*%
                (t(W[, 1:(comp-1), drop=FALSE]) %*% w)
      w = w / sqrt(sum(w^2))
      for (i in seq_len(n_iter)) {
        w = update_fn(X, w, ...)
        if (comp > 1)
          w = w - W[, 1:(comp-1), drop=FALSE] %*%
                  (t(W[, 1:(comp-1), drop=FALSE]) %*% w)
        w = w / sqrt(sum(w^2))
      }
      o = obj_fn(X, w, ...)
      if (o > best_obj) { best_obj = o; best_w = w }
    }
    W[, comp] = best_w
  }
  W
}

set.seed(1)
n = 100; p = 1000; K = 9
L  = matrix(0, nrow = n, ncol = K)
for (i in 1:K) L[sample(n, 20), i] = 1
FF = matrix(rnorm(p * K), nrow = p, ncol = K)
sigma = 0.1
X9  = t(L %*% t(FF) + matrix(rnorm(n * p, 0, sigma), nrow = n))
Z9  = prewhiten(X9, k_pca)
nw9 = preprocess_nonwhitened(X9, k_pca)

W9_wh_lc  = fastica_deflation_wh(Z9,  K, fastica_r1update,     compute_objective,     n_starts = 10)
W9_wh_tlc = fastica_deflation_wh(Z9,  K, fastica_r1update_tlc, compute_objective_tlc, n_starts = 10, lambda = 1)
W9_nw_lc  = fastica_deflation_nw(nw9, K, update_nw,            obj_nw,                n_starts = 10)
W9_nw_tlc = fastica_deflation_nw(nw9, K, update_nw,            obj_nw,                n_starts = 10, lambda = 1)

cor9_wh_lc  = cor(L, t(Z9)    %*% W9_wh_lc)
cor9_wh_tlc = cor(L, t(Z9)    %*% W9_wh_tlc)
cor9_nw_lc  = cor(L, t(nw9$Z) %*% W9_nw_lc)
cor9_nw_tlc = cor(L, t(nw9$Z) %*% W9_nw_tlc)

cat("whitened log-cosh    — max |cor| per true source:\n")
print(round(apply(abs(cor9_wh_lc),  2, max), 3))
cat("whitened TLC         — max |cor| per true source:\n")
print(round(apply(abs(cor9_wh_tlc), 2, max), 3))
cat("nonwhitened log-cosh — max |cor| per true source:\n")
print(round(apply(abs(cor9_nw_lc),  2, max), 3))
cat("nonwhitened TLC      — max |cor| per true source:\n")
print(round(apply(abs(cor9_nw_tlc), 2, max), 3))
```

Extracted projections (whitened TLC vs nonwhitened TLC):

```{r fig.height=8}
proj9_wh_tlc = t(Z9)    %*% W9_wh_tlc
proj9_nw_tlc = t(nw9$Z) %*% W9_nw_tlc

par(mfrow = c(K, 2), mar = c(1, 2, 1.5, 0.5))
for (i in 1:K) {
  plot(proj9_wh_tlc[, i], main = paste0("whitened TLC comp ",    i),
       ylab = "", xlab = "", cex.main = 0.8)
  plot(proj9_nw_tlc[, i], main = paste0("nonwhitened TLC comp ", i),
       ylab = "", xlab = "", cex.main = 0.8)
}
par(mfrow = c(1, 1))
```

## Summary

The non-whitened FastICA replaces the whitened data $Z = \sqrt{n}V'$ with
$Z = DV'$ and changes only the normalization: $\|w\|^2 = 1$ becomes
$w'D^2 w = n$.  This preserves the fixed-point structure of the algorithm.
The deflation step uses the $D^2$-inner-product for orthogonalization instead
of the standard inner product.  When $k$ is chosen larger than the true number
of components, the non-whitened approach is more robust because excess noise
PCs — which have small singular values $d_i$ — contribute little to the
objective, whereas whitening equalises all $k$ PCs and amplifies noise.
