---
title: "Structure plots from fastICA sources on 1000 Genomes data"
author: "Matthew Stephens"
date: "2026-09-18"
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
    toc_depth: 2
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

## Introduction

This follows up [`fastica_1kg`](fastica_1kg.html), which fitted fastICA with
the $x|x|$ contrast to the 1000 Genomes genotype data and argued that the
objects worth taking seriously are the *unconstrained rank-1 maxima* rather
than the factors of any one rank-$r$ fit. Here we turn those sources into
structure plots.

A structure plot needs non-negative memberships, which signed ICA sources are
not. But the $x|x|$ sources are one-sided --- most loadings near zero, a
minority large and positive --- so their positive parts may serve directly.
Take
$$
A = \max(L, 0), \qquad X \approx AB + E \ \text{fitted by least squares},
$$
and *if* $B$ comes out non-negative, rescale each row of $B$ to sum to 1 and
scale the corresponding column of $A$ to compensate. The rows of $A$ are then
memberships on a common scale, ready for a stacked bar plot. Their sums need
not be 1, which `fastTopics::structure_plot` handles by letting bars differ in
height.

Two points about the centering, which is the thing that makes this awkward.
There is **no intercept column** in $A$, and none is needed: every individual
has at least one positive source, so the positive parts already span the mean.
Adding one is actively harmful --- with an intercept the remaining rows of $B$
become *increments* to allele frequency, which are necessarily signed, and
about half of $B$ goes negative. Fitting the uncentered $X$ with no intercept
is what keeps $B$ non-negative.

Whether $B \ge 0$ really holds is an empirical question, and it is the main
thing this analysis checks.

```{r load}
library(ggplot2)
library(cowplot)
library(fastTopics)

meta   <- readRDS("../data/tgp_meta.rds")
fit    <- readRDS("../output/tgp_fastica.rds")
struct <- readRDS("../output/tgp_structure.rds")

pop_order <- c("LWK","ESN","YRI","MSL","GWD","ACB","ASW",     # AFR
               "CLM","MXL","PUR","PEL",                       # AMR
               "TSI","IBS","GBR","CEU","FIN",                 # EUR
               "PJL","GIH","ITU","STU","BEB",                 # SAS
               "CDX","KHV","CHS","CHB","JPT")                 # EAS
meta$pop <- factor(as.character(meta$pop), levels = pop_order)

sp_palette <- c(AFR = "#E69F00", AMR = "#D55E00", EUR = "#7570B3",
                SAS = "#1B9E77", EAS = "#0072B2")
```

The sources come from
[`code/fit_tgp_fastica.R`](https://github.com/stephens999/misc/blob/master/code/fit_tgp_fastica.R),
which saves every fit made in the previous analysis: the rank-$r$ fits at
$K = 20$ and $K = 30$ on the full data, the $K = 30$ fit with the 40
sparse-factor individuals pruned, the two chromosome-half fits, and the
rank-1 maxima of each of those four whitened spaces. The least-squares $B$ and
its diagnostics are computed separately, in
[`code/fit_tgp_structure.R`](https://github.com/stephens999/misc/blob/master/code/fit_tgp_structure.R),
since that needs the genotype matrix; the chromosome halves are fit against
their own half of the SNPs.

## Does $B \ge 0$ hold?

```{r diag-table}
sets <- setdiff(names(struct), "half_match")
tab <- do.call(rbind, lapply(sets, function(nm) {
  s <- struct[[nm]]
  data.frame(sources = nm, K = s$K, `frac B < -0.01` = round(s$frac_lt, 4),
             min_B = round(s$min_B, 3), R2 = round(s$r2, 3),
             check.names = FALSE, row.names = NULL)
}))
tab[order(tab$`frac B < -0.01`), ]
```

For every set of **rank-1 maxima** --- full data, pruned, odd chromosomes, even
chromosomes --- $B$ is non-negative to within rounding: at most 0.06% of
entries below $-0.01$, against a mean $|B|$ of order $0.1$. Clipping those at
zero changes nothing.

For every **rank-$r$ fit** it fails: 10--15% of entries below $-0.01$, with
minima from $-0.26$ to $-0.81$. So the construction works for the maxima and
not for the rank-$r$ factors, in all four whitened spaces independently. That
is a fifth respect --- after the local-maximum test, basin sizes, replication
across SNP halves, and initialization experiments --- in which the maxima
behave like real structure and a rank-$r$ factor set does not.

The structure plots below therefore use maxima only.

## Building the memberships

```{r build-fn}
hm  <- struct$half_match
src <- list(
  `pruned maxima (K=14)` = list(L = fit$maxima$no_sparse$L, key = "pruned_maxima"),
  `full maxima (K=17)`   = list(L = fit$maxima$full$L,      key = "full_maxima"),
  `odd maxima (K=11)`    = list(L = fit$maxima$odd$L,       key = "odd_maxima"),
  `even maxima (K=10)`   = list(L = fit$maxima$even$L,      key = "even_maxima"),
  `odd K=30 stable`      = list(L = fit$split$odd$fit$L[, hm$stable, drop = FALSE],
                                key = "odd_k30_stable"),
  `even K=30 stable`     = list(L = fit$split$even$fit$L[, hm$perm[hm$stable], drop = FALSE],
                                key = "even_k30_stable"))

# A = pmax(L, 0) with columns rescaled so the rows of B sum to 1, factors
# ordered and labelled by the populations they carry.
memberships <- function(name) {
  L   <- src[[name]]$L
  r   <- struct[[src[[name]]$key]]$r
  pop <- meta$pop[match(rownames(L), meta$sample)]

  w <- pmax(L, 0)^2
  o <- order(colSums(w * as.integer(pop)) / colSums(w))
  A <- sweep(pmax(L[, o, drop = FALSE], 0), 2, r[o], "*")

  colnames(A) <- make.unique(apply(L[, o, drop = FALSE], 2, function(x) {
    v <- tapply(x, pop, mean); v <- sort(v[v > 0], decreasing = TRUE)
    paste(names(v)[v >= 0.3 * v[1]], collapse = "/")
  }))
  A <- A / mean(rowSums(A))     # bar heights on a scale of about 1

  # An individual with no positive loading on any source has no membership
  # representation at all; drop those and record how many.
  zero <- rowSums(A) == 0
  list(A = A[!zero, , drop = FALSE], pop = droplevels(pop[!zero]),
       n_dropped = sum(zero), o = o)
}

plot_structure <- function(name, normalize = FALSE, colors = NULL) {
  m <- memberships(name)
  A <- if (normalize) m$A / rowSums(m$A) else m$A
  if (is.null(colors)) colors <- Polychrome::kelly.colors()[2:(ncol(A) + 1)]
  structure_plot(A, grouping = m$pop, colors = colors,
                 gap = 16, verbose = FALSE) +
    labs(y = "membership",
         title = sprintf("%s%s%s", name,
                         if (normalize) " (normalized)" else " (unnormalized)",
                         if (m$n_dropped) sprintf(" - %d individual(s) dropped",
                                                  m$n_dropped) else ""))
}
```

The normalized plots show composition; the unnormalized ones keep the bar
heights, which carry information the normalized version discards --- a short bar
is an individual poorly explained by all the factors in that set.

```{r heights}
sapply(names(src), function(nm) {
  h <- rowSums(memberships(nm)$A)
  round(c(min = min(h), median = median(h), max = max(h)), 2)
})
```

## Pruned data, 14 maxima {.tabset}

The cleanest set: the 14 maxima of the pruned data, all of which replicate
across SNP halves and are reachable from random starts.

### Normalized

```{r pruned-norm, fig.width=9, fig.height=4}
plot_structure("pruned maxima (K=14)", normalize = TRUE)
```

### Unnormalized

```{r pruned-raw, fig.width=9, fig.height=4}
plot_structure("pruned maxima (K=14)")
```

## {-}

The normalized version is close to a hard population assignment --- LWK 97%,
ESN/YRI 98%, GWD 97%, TSI 93%, CDX 95% and JPT 92% on their own factors ---
with mixture where mixture is expected. ASW comes out 60% ESN/YRI with 11%
GBR/CEU and 9% LWK; CLM 37% PEL/MXL and 33% TSI/IBS; IBS 61% TSI/IBS and 25%
GBR/CEU; FIN 80% its own factor and 15% GBR/CEU; CHS 65% CHB/CHS and 30%
CDX/KHV; PJL splits 38/42 with STU/ITU/BEB. These are the expected admixture
patterns, read off ICA sources rather than fitted under a non-negativity
constraint.

## Full data, 17 maxima {.tabset}

The same for the unpruned data, where three of the 17 maxima are the sparse
relative and subgroup factors.

### Normalized

```{r full-norm, fig.width=9, fig.height=4}
plot_structure("full maxima (K=17)", normalize = TRUE)
```

### Unnormalized

```{r full-raw, fig.width=9, fig.height=4}
plot_structure("full maxima (K=17)")
```

## Chromosome halves, maxima {.tabset}

The stability check as structure plots: the halves are independent SNP sets, so
agreement between the two is evidence the memberships do not depend on which
markers were used. They find 11 and 10 maxima respectively.

### Odd, normalized

```{r odd-norm, fig.width=9, fig.height=4}
plot_structure("odd maxima (K=11)", normalize = TRUE)
```

### Odd, unnormalized

```{r odd-raw, fig.width=9, fig.height=4}
plot_structure("odd maxima (K=11)")
```

### Even, normalized

```{r even-norm, fig.width=9, fig.height=4}
plot_structure("even maxima (K=10)", normalize = TRUE)
```

### Even, unnormalized

```{r even-raw, fig.width=9, fig.height=4}
plot_structure("even maxima (K=10)")
```

## Stability-pruned K = 30 {.tabset}

The rank-$r$ fits fail the $B \ge 0$ check, but it is worth asking how much of
that is down to the factors that do not replicate. Matching the two half fits
factor to factor and keeping only the pairs that agree leaves
`r length(hm$stable)` of the 30, with matched correlations from
`r round(min(hm$match_cor[hm$stable]), 2)` to
`r round(max(hm$match_cor[hm$stable]), 2)`; the 15 discarded sit at
`r round(max(hm$match_cor[-hm$stable]), 2)` and below.

Pruning does help, but not enough: $B$ below $-0.01$ falls from 9.8% to 3.4% of
entries on the odd half and from 11.7% to 5.0% on the even, against 0.05% or
less for the maxima. So the non-replicating factors are part of the problem and
not the whole of it --- even the replicating rank-$r$ factors do not give
non-negative $B$ the way the maxima do.

Two further things to note when reading these plots. Restricting to 15 of the
30 factors leaves a fair number of individuals with no positive loading on any
retained factor --- 47 on the odd half and 17 on the even --- and those are
dropped, against one individual for the worst of the maxima sets. And the even
half has a single bar of height 9.6 against a 99th percentile of 2.2, so the
unnormalized plot there is dominated by one individual.

### Odd, normalized

```{r odd-stable-norm, fig.width=9, fig.height=4}
plot_structure("odd K=30 stable", normalize = TRUE)
```

### Odd, unnormalized

```{r odd-stable-raw, fig.width=9, fig.height=4}
plot_structure("odd K=30 stable")
```

### Even, normalized

```{r even-stable-norm, fig.width=9, fig.height=4}
plot_structure("even K=30 stable", normalize = TRUE)
```

### Even, unnormalized

```{r even-stable-raw, fig.width=9, fig.height=4}
plot_structure("even K=30 stable")
```

## Comparison with an ADMIXTURE run at K = 12

For an external point of reference, here is our 14-factor result recoloured to
match an ADMIXTURE-style run on the same data at $K = 12$, with that run's
$K = 12$ panel directly below it. The populations are in the same order and
the bar widths are proportional to sample size in both, so the two line up.

The colours were read off the $K = 12$ panel: for each population we took the
modal colour of its column, which is the colour of whichever component
dominates it. Our factors then take the colour of the component covering the
same populations. Where our $K = 14$ splits something the $K = 12$ does not,
the extra factor gets a darker shade of the same hue rather than a new colour,
so the correspondence stays readable:

```{r k12-palette}
k12_colors <- c(
  "LWK"             = "#0048f0",   # exact matches to the K = 12 components
  "ESN/YRI/ACB/ASW" = "#18f000",
  "GWD"             = "#f00000",
  "PUR"             = "#f000d8",
  "PEL/MXL"         = "#90f000",
  "TSI/IBS"         = "#a00090",   # darker: K = 12 has TSI with PUR/CLM
  "GBR/CEU"         = "#f06000",
  "FIN"             = "#f0d800",
  "PJL"             = "#00c0f0",
  "GIH"             = "#1800f0",
  "STU/ITU/BEB"     = "#0080a0",   # darker: K = 12 has these with PJL
  "CDX/KHV"         = "#00f0c0",
  "CHB/CHS"         = "#6000a0",   # darker: K = 12 has CHS with CDX/KHV
  "JPT"             = "#9000f0")

# The palette is keyed by factor label, so check it lines up with the plot.
m14 <- memberships("pruned maxima (K=14)")
stopifnot(identical(colnames(m14$A), names(k12_colors)))
```

```{r k12-ours, fig.width=9, fig.height=4}
plot_structure("pruned maxima (K=14)", normalize = TRUE,
               colors = unname(k12_colors)) +
  labs(title = "Ours: 14 maxima, pruned data, recoloured to match K = 12 below")
```

```{r k12-theirs, fig.width=9, fig.height=2.1, dpi=150}
img <- png::readPNG("../data/admixture_k12.png")
grid::grid.newpage()
grid::grid.raster(img)
```

Read together, the two agree on most of the structure and differ in resolution
in ways the $K$ difference predicts. Both give LWK its own component, both put
ESN/YRI/ACB/ASW together with ACB and ASW as gradients, both separate MSL/GWD
from the other African populations, both give FIN its own component, and both
group the four South Asian populations with GIH set apart.

The visible differences are the ones where our extra two factors buy
resolution: the $K = 12$ run puts TSI with CLM and PUR in a single component
(magenta) where we split PUR from TSI/IBS, and it groups CHS with CDX/KHV where
we split CDX/KHV from CHB/CHS. Our decomposition is also markedly less
gradual within populations --- their bars vary continuously across individuals
within a population, ours are closer to flat --- which follows from the $x|x|$
contrast rewarding one-sided factors.


## Correlations between the components' allele frequencies

Each row of $B$ is a component's estimated allele frequency profile across the
185,116 SNPs, so the $K \times K$ correlation matrix of $B$ says how similar
the components are as populations. Focusing on the $K = 14$ result:

```{r corB-setup}
m14 <- memberships("pruned maxima (K=14)")
C   <- struct$pruned_maxima$cor_B[m14$o, m14$o]
dimnames(C) <- list(colnames(m14$A), colnames(m14$A))
round(range(C[upper.tri(C)]), 3)
```

Every off-diagonal entry is large, between 0.60 and 0.99. That is expected
rather than informative on its own: because there is no intercept column, each
row of $B$ is an *absolute* frequency profile, and all fourteen share the
overall allele frequency spectrum of the data. What matters is the pattern in
those high values, so the colour scale below spans the observed range rather
than starting at zero.

```{r corB-heatmap, fig.width=6.5, fig.height=5.5}
heat <- function(M, title) {
  df <- data.frame(
    i = factor(rep(rownames(M), ncol(M)), levels = rev(rownames(M))),
    j = factor(rep(colnames(M), each = nrow(M)), levels = colnames(M)),
    r = as.vector(M))
  ggplot(df, aes(j, i, fill = r)) +
    geom_tile() +
    geom_text(aes(label = sprintf("%.2f", r)), size = 2.1,
              colour = ifelse(df$r > mean(range(df$r)), "white", "grey20")) +
    scale_fill_viridis_c(limits = range(M), option = "mako", direction = -1) +
    labs(x = NULL, y = NULL, fill = "cor", title = title) +
    theme_cowplot(font_size = 9) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
}
heat(C, "Correlation of estimated allele frequencies, K = 14")
```

The structure is the expected continental one, in four blocks along the
diagonal:

- **Africa** (LWK, ESN/YRI/ACB/ASW, GWD) mutually 0.96--0.98, and the most
  distinct from everything else at 0.60--0.74.
- **Europe and the Americas** (PUR, TSI/IBS, GBR/CEU, FIN) mutually 0.95--0.99.
- **South Asia** (PJL, GIH, STU/ITU/BEB) mutually 0.96--0.98.
- **East Asia** (CDX/KHV, CHB/CHS, JPT) mutually 0.97--0.98.

Two features are worth pointing out. Africa being the most differentiated block
is what the out-of-Africa history predicts, and it falls out of the
correlations without anything being fitted to produce it. And PEL/MXL does not
sit cleanly in any block: 0.88 with PUR, 0.83--0.84 with the European
components, but 0.85--0.86 with the East Asian ones, which is the signature of
Native American ancestry being closer to East Asia than to Europe. It is the
only component whose highest correlations are split across two continents.

Reordering by hierarchical clustering on $1 - r$ rather than by population
makes the grouping explicit:

```{r corB-clustered, fig.width=6.5, fig.height=5.5}
ord <- hclust(as.dist(1 - C), method = "average")$order
heat(C[ord, ord], "Same, ordered by hierarchical clustering")
```

Because all of them share the frequency spectrum, the raw correlations are a
weak summary: the interesting variation sits in the third decimal place. It is
worth removing the shared component, which is what the next subsection does.


### Pairwise scatter plots and outliers

The correlations above are aggregate; a scatter plot of each high-correlation
pair shows whether the agreement is uniform across SNPs or whether particular
SNPs break it. We take the `r sum(upper.tri(C) & C > 0.95)` pairs with
$r > 0.95$, fit a least-squares line to each, and flag SNPs whose standardized
residual exceeds 6 in absolute value.

```{r corB-pairs-setup}
B <- readRDS("../output/tgp_B_pruned_maxima.rds")[m14$o, ]
rownames(B) <- colnames(m14$A)

hi <- which(upper.tri(C) & C > 0.95, arr.ind = TRUE)
pairs_df <- do.call(rbind, lapply(seq_len(nrow(hi)), function(k) {
  i <- hi[k, 1]; j <- hi[k, 2]
  z <- local({ r <- residuals(lm(B[j, ] ~ B[i, ])); r / sd(r) })
  data.frame(pair = sprintf("%s vs %s\n(r = %.3f)", rownames(B)[i],
                            rownames(B)[j], C[i, j]),
             snp = colnames(B), x = B[i, ], y = B[j, ], z = z,
             row.names = NULL)
}))
pairs_df$outlier <- abs(pairs_df$z) > 6

# Nearest gene for every SNP, from the UCSC hg19 refGene table via
# code/annotate_snps.R. GRCh37 is the build these data use; the installed
# org.Hs.eg.db is GRCh38 and would be offset by hundreds of kb here.
genes <- readRDS("../output/tgp_snp_genes.rds")
pairs_df$gene <- genes$gene[match(pairs_df$snp, genes$snp)]
pairs_df$dist <- genes$dist[match(pairs_df$snp, genes$snp)]
# Label with the gene, marking those the SNP is not actually inside.
pairs_df$label <- ifelse(pairs_df$dist == 0, pairs_df$gene,
                         sprintf("%s*", pairs_df$gene))
# Within a panel, several outlier SNPs often sit in the same gene (the LCT
# region contributes 22 points to one panel). Label only the most extreme SNP
# per gene per panel; the region table below keeps the full detail.
pairs_df$label[pairs_df$outlier] <- local({
  d <- pairs_df[pairs_df$outlier, ]
  keep <- ave(abs(d$z), d$pair, d$gene, FUN = function(z) z == max(z)) == 1
  ifelse(keep, d$label, NA_character_)
})

# All flagged points are kept; the bulk is thinned for rendering only.
set.seed(1)
bulk <- pairs_df[!pairs_df$outlier, ]
bulk <- bulk[sample(nrow(bulk), 3e4), ]
plot_df <- rbind(bulk, pairs_df[pairs_df$outlier, ])
c(pairs = nrow(hi), flagged = sum(pairs_df$outlier),
  distinct_snps = length(unique(pairs_df$snp[pairs_df$outlier])))
```

```{r corB-pairs, fig.width=15, fig.height=14}
ggplot(plot_df, aes(x, y)) +
  geom_point(data = subset(plot_df, !outlier), size = 0.15, alpha = 0.25,
             colour = "grey55") +
  geom_point(data = subset(plot_df, outlier), size = 0.9, colour = "#D55E00") +
  ggrepel::geom_text_repel(data = subset(plot_df, outlier & !is.na(label)),
                           aes(label = label), size = 3.7, colour = "#8a3800",
                           fontface = "italic",
                           segment.size = 0.2, segment.colour = "grey55",
                           min.segment.length = 0, max.overlaps = Inf,
                           box.padding = 0.3, point.padding = 0.2) +
  facet_wrap(~ pair, ncol = 4, scales = "free") +
  labs(x = "allele frequency, first component",
       y = "allele frequency, second component",
       caption = "* = nearest gene, SNP not inside it") +
  theme_cowplot(font_size = 12) +
  theme(strip.text = element_text(size = 11, lineheight = 0.9),
        strip.background = element_rect(fill = "grey92", colour = NA))
```

The clouds are tight and essentially linear, so the high correlations are not
being produced by a few influential points --- but every pair has a small
number of clear outliers, and they are not scattered at random. Grouping the
flagged SNPs into genomic regions (same chromosome, gaps of at most 1 Mb)
shows how concentrated they are:

```{r corB-regions}
out <- subset(pairs_df, outlier)
out$chr <- sub(":.*", "", out$snp)
out$pos <- as.numeric(sub(".*:", "", out$snp))

# Assign each distinct SNP to a region.
u <- unique(out[, c("snp", "chr", "pos")])
u <- u[order(u$chr, u$pos), ]
u$region <- cumsum(c(1, (u$chr[-1] != u$chr[-nrow(u)]) |
                        (diff(u$pos) > 1e6)))
out$region <- u$region[match(out$snp, u$snp)]

region_tab <- do.call(rbind, lapply(split(out, out$region), function(d) {
  us <- unique(d[, c("snp", "pos", "gene", "dist")])
  data.frame(
    `genomic region` = sprintf("chr%s:%.2f-%.2f Mb", d$chr[1],
                               min(us$pos) / 1e6, max(us$pos) / 1e6),
    SNPs      = nrow(us),
    `max |z|` = round(max(abs(d$z)), 1),
    `nearby genes` = paste(head(unique(us$gene[order(us$dist)]), 5),
                           collapse = ", "),
    `pair comparisons` = paste(unique(sub("\\s*\\(r *=.*", "", d$pair)),
                               collapse = "; "),
    check.names = FALSE, stringsAsFactors = FALSE)
}))
region_tab <- region_tab[order(-region_tab$SNPs, -region_tab$`max |z|`), ]
knitr::kable(region_tab, row.names = FALSE)
```

`r nrow(region_tab)` regions hold the
`r length(unique(out$snp))` outlier SNPs, and two of them dominate.

**chr2:135.29--136.97 Mb** contains 21 of the 48, all from the two European
comparisons (PUR vs GBR/CEU and TSI/IBS vs GBR/CEU). The gene labels spread
across the region rather than pointing at one gene, but *LCT* and its
immediate neighbours *UBXN4*, *DARS1* and *DARS-AS1* are among them. This is
the lactase persistence locus, one of the strongest selective sweeps known in
humans, and exactly the sort of site whose allele frequency differs sharply
between European and non-European populations while the rest of the genome
does not. The sweep haplotype is long, which fits outliers spreading over more
than a megabyte into the flanking genes. Only two of the 21 SNPs are inside
*LCT* itself, so the locus is identified here by position rather than by the
labels.

**chr14:105.98--106.26 Mb** holds five, from the two East Asian comparisons.
The nearest annotated features are microRNAs at 18--62 kb; this is the
immunoglobulin heavy chain region, structurally complex and a common source of
mapping artifacts, so that cluster is more likely technical than selective.

The remaining 22 regions are single SNPs, but three of them recur in five
different pairs each --- *NAP1L1* (chr12), *FOXP1* (chr3) and *MIR4454*
(chr5). Each appears in all three African comparisons and in two South Asian
ones, so these are SNPs that break the agreement between closely related
components in more than one part of the world. *HLA-DRB5* appears as a second
classic high-differentiation locus, in the ESN/YRI vs GWD comparison.

#### Single-SNP outliers are probably technical

Selection acts on haplotypes, so a sweep should perturb a run of nearby
markers, not one in isolation. That makes the isolated single-SNP outliers
suspicious. The obvious objection is that these data are LD-thinned, so a real
signal might survive at only one marker --- but the thinning is not nearly
that severe:

```{r corB-density}
snp_all <- colnames(B); chr_all <- sub(":.*", "", snp_all)
pos_all <- as.numeric(sub(".*:", "", snp_all))
sp <- unlist(lapply(split(pos_all, chr_all), function(p) diff(sort(p))))
c(median_spacing_kb = round(median(sp) / 1e3, 1),
  mean_spacing_kb   = round(mean(sp) / 1e3, 1))
```

At a median spacing of 8.6 kb a 100 kb window holds about 15 markers, so a
sweep has plenty of opportunity to show up more than once. We can therefore
ask directly whether each outlier's neighbours carry any signal in the *same*
pair comparison:

```{r corB-neighbours}
Zmat <- sapply(seq_len(nrow(hi)), function(k)
  local({ r <- residuals(lm(B[hi[k, 2], ] ~ B[hi[k, 1], ])); r / sd(r) }))

nb_tab <- do.call(rbind, lapply(seq_len(ncol(Zmat)), function(k) {
  w6 <- which(abs(Zmat[, k]) > 6)
  if (!length(w6)) return(NULL)
  do.call(rbind, lapply(w6, function(i) {
    nb <- which(chr_all == chr_all[i] & abs(pos_all - pos_all[i]) <= 1e5 &
                seq_along(pos_all) != i)
    data.frame(snp = snp_all[i], z = Zmat[i, k], n_nb = length(nb),
               max_nb_z = if (length(nb)) max(abs(Zmat[nb, k])) else NA_real_)
  }))
}))
c(outliers = nrow(nb_tab),
  `with neighbours` = sum(nb_tab$n_nb > 0),
  `neighbour |z|>3` = sum(nb_tab$max_nb_z > 3, na.rm = TRUE),
  `neighbour |z|>6` = sum(nb_tab$max_nb_z > 6, na.rm = TRUE))

nb_tab$gene <- genes$gene[match(nb_tab$snp, genes$snp)]
iso <- subset(nb_tab, max_nb_z < 3)
head(iso[order(-abs(iso$z)), c("snp", "gene", "z", "n_nb", "max_nb_z")], 8)
```

Every outlier has neighbours, but only half of them have a neighbour reaching
even $|z| > 3$. And the pattern is the wrong way round for selection: the
*most* extreme outliers are the *most* isolated. `NAP1L1` reaches $z = 15.8$
with eleven neighbours whose largest $|z|$ is 2.1; `ARHGAP20` reaches $-11.7$
with twenty-three neighbours all below 2.3. No haplotype-level process
produces a fifteen-sigma deviation at one marker while two dozen markers
around it show nothing. These are the same SNPs that recur across five pair
comparisons each, which fits the same reading --- a bad marker is bad in every
comparison it enters.

So the useful division is not "clustered versus single" but that the two
clusters (LCT, IGH) behave like genomic signals and the isolated singletons
behave like marker-level failures.

##### What technical effects could do this?

Several mechanisms produce a frequency error at one SNP that differs between
populations, which is what it takes to move a single point off a pairwise
line:

- **Mis-mapping and paralogy.** Reads from a duplicated or repetitive region
  mapping to a single locus create a spurious variant whose apparent frequency
  tracks copy number, and copy number varies between populations. This is the
  most likely explanation for the chr14 IGH cluster, and it is suggestive that
  `MIR4454` appears twice in the outlier list at two different positions
  (chr5:34.7 Mb and chr5:40.8 Mb), which is what a multiply-mapping annotation
  looks like. `NAP1L1` is worth checking on the same grounds, since genes with
  processed pseudogenes elsewhere in the genome are classic mis-mapping
  targets.
- **Reference bias.** Reads carrying the non-reference allele align less
  readily, and the penalty is larger where a sample's haplotype is more
  diverged from the reference. Since divergence from the reference differs
  systematically by population, this biases allele frequencies differentially.
- **Batch and centre effects.** 1000 Genomes phase 3 samples were sequenced at
  several centres over several years, and populations were not evenly spread
  across batches. Differences in depth, library preparation and caller
  behaviour then align with population labels.
- **Imputation and phasing.** Phase 3 genotypes come from low-coverage
  sequencing with imputation, whose accuracy depends on how well each
  population is represented in the reference panel. A SNP that imputes badly
  in one population only will show exactly this signature.
- **Copy number variation overlapping the SNP**, where genotype calls reflect
  copy number rather than allele dosage.

One mechanism that is *not* a likely explanation here is ancestral allele
misassignment, even though the matrix was re-coded to derived allele counts. A
wrong ancestral call flips $p \to 1-p$ for every component at that SNP, which
moves the point to a line parallel to the fitted one --- so mis-polarized SNPs
would deviate in a consistent direction by a similar amount, whereas the
observed outliers scatter on both sides with a wide range of residuals.

Distinguishing among the remaining mechanisms is not possible from the
genotype matrix alone. It would need the per-SNP quality information that was
discarded in building this dataset: read depth, mapping quality, VQSR score,
Hardy-Weinberg departure computed within population, and overlap with
segmental duplication and CNV tracks. A cheap partial check would be to
intersect the isolated outliers with a segmental duplication track, which
would test the mis-mapping hypothesis directly.

#### Which signals look like real selection?

The neighbour test gives a way to rank the loci rather than just split them in
two. Widening the window to 500 kb and counting how many nearby markers are
also perturbed in the same comparison:

```{r corB-rank}
nb_wide <- do.call(rbind, lapply(seq_len(ncol(Zmat)), function(k) {
  w6 <- which(abs(Zmat[, k]) > 6)
  if (!length(w6)) return(NULL)
  do.call(rbind, lapply(w6, function(i) {
    n1 <- which(chr_all == chr_all[i] & abs(pos_all - pos_all[i]) <= 1e5 &
                seq_along(pos_all) != i)
    n5 <- which(chr_all == chr_all[i] & abs(pos_all - pos_all[i]) <= 5e5 &
                seq_along(pos_all) != i)
    data.frame(pair = sub("\\s*\\(r *=.*", "",
                          sprintf("%s vs %s", rownames(B)[hi[k, 1]],
                                  rownames(B)[hi[k, 2]])),
               snp = snp_all[i], gene = genes$gene[match(snp_all[i], genes$snp)],
               z = round(Zmat[i, k], 1), n_100kb = length(n1),
               max_nb_z = round(max(abs(Zmat[n1, k])), 1),
               nb_gt3_500kb = sum(abs(Zmat[n5, k]) > 3),
               nb_gt6_500kb = sum(abs(Zmat[n5, k]) > 6))
  }))
}))

in_cluster <- (nb_wide$snp %in% snp_all[chr_all == "2" &
                 pos_all > 1.35e8 & pos_all < 1.37e8]) |
              (nb_wide$snp %in% snp_all[chr_all == "14" &
                 pos_all > 1.05e8 & pos_all < 1.07e8])
out <- nb_wide[!in_cluster, ]
head(out[order(-out$nb_gt3_500kb, -out$max_nb_z), ], 8)
```

**HLA is the strongest candidate after *LCT*, and the region table above
understates it.** That table grouped SNPs passing $|z| > 6$, and the MHC
contributes only one such SNP --- but it has 56 markers within 100 kb, of
which the most extreme reaches $|z| = 5.9$, and **44 markers within 500 kb
exceed $|z| > 3$**. That is more perturbed neighbours than the *LCT* region
averages (22.5). A broad shoulder of moderately deviating markers with few
extreme ones is what long-standing balancing selection should look like,
as opposed to the tall narrow peak of a recent sweep, so the MHC being
invisible to a hard $|z| > 6$ cut is a property of the statistic rather than
of the locus.

***FADS1* also has genuine neighbour support** --- 13 markers within 100 kb
with a maximum neighbour of $|z| \approx 5$, and five or six above $|z| > 3$
in each of the two East Asian comparisons. It is therefore a useful
counterexample to a blanket "isolated SNP means artifact" rule: it is a
documented selection locus that contributes only one marker above the
threshold while still showing a real regional shoulder.

*KCNMB4* is marginal --- 24 neighbours, the best at $|z| = 5.4$, but only
three above 3, and no published selection result I can point to. I would not
push it.

Everything below that is isolated, and *FOXP1* is the clearest case for
treating these as technical: $z = -14.6$ with **zero** neighbours above
$|z| > 3$ anywhere within 500 kb.

```{r corB-rank-summary}
c(`non-cluster outliers` = sum(!in_cluster),
  `with any neighbour |z|>6 within 500kb` = sum(out$nb_gt6_500kb > 0),
  `LCT region, mean such neighbours` =
    round(mean(nb_wide$nb_gt6_500kb[in_cluster & sub(":.*", "", nb_wide$snp) == "2"]), 1),
  `IGH region, mean such neighbours` =
    round(mean(nb_wide$nb_gt6_500kb[in_cluster & sub(":.*", "", nb_wide$snp) == "14"]), 1))
```

Not one of the 41 non-cluster outliers has a neighbour reaching $|z| > 6$
within 500 kb, against an average of 9.1 in the *LCT* region. On this evidence
the ordering is *LCT* first by a wide margin, then HLA, then *FADS1*, with
everything else either technical or unresolved.

#### Cross-reference with published selection findings

Three of these regions match well-established selection signals, and in each
case the *direction* in our data is consistent with the published account. To
check that, we put the rows of $B$ on a comparable scale (dividing each by its
mean, since the row scale is not identified) and look at the relative
frequency by component:

```{r corB-selection}
Bn <- B / rowMeans(B)
freq_at <- function(snp) round(sort(Bn[, snp], decreasing = TRUE), 2)
freq_at("2:136570613")     # LCT
freq_at("11:61569830")     # FADS1 / rs174546
```

**LCT / MCM6 (chr2:135.3--137.0 Mb).** Lactase persistence is the textbook
example of a recent strong sweep in Europeans, with an unusually long
haplotype around the *MCM6* regulatory variant (Bersaglieri et al. 2004;
independent persistence alleles arose in African pastoralists, Tishkoff et al.
2007), and *LCT* is consistently among the top hits of genome-wide scans such
as Sabeti et al. 2007. Our data reproduce three features of that account. The
outliers are confined to the two comparisons involving GBR/CEU, which is the
extreme component at these SNPs; there is a north-south gradient across the
European components (GBR/CEU 0.13, FIN 0.50, TSI/IBS 0.68 at 2:136570613),
matching the well-documented cline in lactase persistence frequency; and the
signal spreads over more than a megabyte into the flanking genes, as a long
sweep haplotype should in an LD-thinned marker set.

**FADS1 (chr11:61.57 Mb).** The *FADS1/FADS2* fatty acid desaturase cluster
carries selection signals reported in several populations --- among the
strongest signals in ancient European DNA (Mathieson et al. 2015), in
Greenlandic Inuit (Fumagalli et al. 2015), and in South Asian populations with
long vegetarian dietary histories. `11:61569830` is rs174546, a known *FADS1*
regulatory variant. Here it appears as an outlier specifically in the two
comparisons involving CDX/KHV, which sits at 0.44 against 1.80 for CHB/CHS and
1.97 for JPT --- a four-fold difference *within* East Asia --- while the
African components are at 3.3. A locus with this much frequency structure
between closely related populations is exactly what breaks a pairwise
correlation.

**HLA (chr6:32.46 Mb, *HLA-DRB5*).** The MHC is the classic target of
balancing selection and one of the most polymorphic and differentiated regions
in the genome, so its appearance here is unsurprising. Our outlier comes from
the ESN/YRI vs GWD comparison, i.e. it distinguishes West African components
from each other.

The chr14:106 Mb cluster is a different matter: that is the immunoglobulin
heavy chain locus, which is structurally complex and segmentally duplicated.
Signals there are more plausibly mapping artifacts than selection, and I would
not treat those five SNPs as biology without independent checking.

For the remaining twenty-odd single-SNP regions I could not match anything to
a published selection result with confidence, and I have not invented
attributions for them. One point worth flagging to avoid a natural confusion:
the *FOXP1* hit on chromosome 3 is **not** *FOXP2*, the chromosome 7 gene
associated with the well-known selection and language literature. Note also
that this cross-reference is from background knowledge rather than a
systematic lookup; doing it properly would mean intersecting these regions
with a catalogue of published selection-scan results, which would also give a
null expectation for how many matches to expect by chance.

One more is worth naming on its own. `11:61569830` is an outlier in two pairs
and sits inside *FADS1*; it is rs174546, the same SNP Annie Xie's GBCD analysis
of these data picked out as a large-effect marker, reached here by a completely
different route.

So the answer to the question is: no, the high correlations are not driven by
outliers, and the outliers that do exist are mostly real biology at two or
three well-known loci rather than a problem with the decomposition.


### Removing the shared spectrum

Every row of $B$ is an absolute frequency profile, so all of them contain the
overall spectrum of the data and the raw correlations sit between 0.6 and 0.99.
Removing that shared component should expose the residual structure. The
obvious way to try it is to centre each SNP across the $K$ components,
subtracting $\bar m_j = \frac{1}{K}\sum_k B_{kj}$, and that does not work
well. It is worth understanding why, because the reason dictates what does.

Two facts about $B$:

```{r shared-facts}
mbar <- colMeans(B)
c(min_row_mean = min(rowMeans(B)), max_row_mean = max(rowMeans(B)),
  ratio = max(rowMeans(B)) / min(rowMeans(B))) |> round(3)
# how tightly the components' overall level and their spread move together
round(cor(rowMeans(B), apply(B, 1, sd)), 3)
# share of B's total sum of squares in its leading singular value
local({ d <- svd(B)$d; round(d[1]^2 / sum(d^2), 3) })
```

The components differ in overall level by a factor of 2.7, their level and
their spread move together almost perfectly ($r = 0.99$), and a single rank-1
term accounts for 92% of $B$. In other words the shared structure is
**multiplicative**, $B_{kj} \approx s_k m_j$, not additive. Subtracting an
unweighted mean removes $\bar s\, m_j$ and leaves $(s_k - \bar s)\,m_j$
behind --- still proportional to the shared spectrum, with a coefficient that
depends only on the component's scale. The spectrum is not removed, it is
retained with a scale-dependent loading:

```{r shared-why}
D  <- B - rep(1, nrow(B)) %o% mbar        # naive per-SNP centring
sv <- svd(D)
c(`rank-1 share of the deviations` = sv$d[1]^2 / sum(sv$d^2),
  `|cor| of its loading with row scale` = abs(cor(sv$u[, 1], rowMeans(B))),
  `|cor| of its profile with mbar`      = abs(cor(sv$v[, 1], mbar))) |> round(3)
```

Forty-three percent of the "deviations" are a rank-1 term whose loading
correlates 0.98 with the components' overall scale and whose profile correlates
0.97 with the spectrum we were trying to remove.

This also means the row scale of $B$ is the wrong thing to lean on: it is not
identified. $B = (A'A)^{-1}A'X$ with $A = \max(L, 0)$, and the columns of $A$
have whatever scale the sources happen to have, so the 2.7-fold range in row
level is a property of that arbitrary normalization rather than of the
populations. Only scale-invariant summaries of $B$ are meaningful.

So the shared component has to be removed in a scale-free way. Three natural
ways to do it:

```{r shared-variants}
K <- nrow(B)
variants <- list(
  `raw`                          = cor(t(B)),
  `naive per-SNP centring`       = cor(t(D)),
  `row-standardise, then centre` = cor(t(local({
      Z <- t(scale(t(B))); Z - rep(1, K) %o% colMeans(Z) }))),
  `remove rank 1 of B`           = cor(t(local({
      sv <- svd(B); B - sv$d[1] * sv$u[, 1] %o% sv$v[, 1] }))),
  `regress each row on mbar`     = cor(t(local({
      t(apply(B, 1, function(b) residuals(lm(b ~ mbar)))) }))))

# Score each by how well it separates continents: mean within-block
# correlation minus mean between-block.
blocks <- c("AFR","AFR","AFR","AMR","AMR","EUR","EUR","EUR",
            "SAS","SAS","SAS","EAS","EAS","EAS")
t(sapply(variants, function(M) {
  u <- upper.tri(M); same <- outer(blocks, blocks, "==")[u]
  round(c(within = mean(M[u][same]), between = mean(M[u][!same]),
          gap = mean(M[u][same]) - mean(M[u][!same])), 3)
}))
```

All three scale-free variants give a within-minus-between gap of 0.96, against
0.20 for the raw correlations and 0.65 for naive centring, and they agree with
each other closely:

```{r shared-agreement}
v <- variants[3:5]
round(c(`std-centre vs rank-1`  = max(abs(v[[1]] - v[[2]])),
        `std-centre vs regress` = max(abs(v[[1]] - v[[3]])),
        `rank-1 vs regress`     = max(abs(v[[2]] - v[[3]]))), 3)
```

```{r shared-heatmap, fig.width=6.5, fig.height=5.5}
R <- variants[["regress each row on mbar"]]
heat(R, "Correlation after removing the shared spectrum, K = 14")
```

This is a far more informative picture than the raw matrix. The four
continental blocks are now strongly positive internally --- East Asia
0.82--0.89, Europe 0.73--0.78, South Asia 0.60--0.79 --- and clearly negative
between, and the two features noted earlier become much more legible. PEL/MXL
sits apart from every block, with its only positive correlations being
0.20--0.25 with the three East Asian components, and PUR shows its European
affinity directly (0.53 with FIN, 0.19--0.26 with the other European
components) while being negative with East Asia.

A correction is in order here. An earlier version of this analysis reported
that this centring produced implausible values --- CHB/CHS with GIH at 0.94 ---
and concluded that removing the shared spectrum was a bad idea. That was a
mistake of mine: the matrix I computed applied the population reordering twice,
once through the row and column names and once through the indexing, so the
labels did not match the values. Under proper centring that pair is
`r round(variants[["naive per-SNP centring"]]["CHB/CHS", "GIH"], 2)`, and
scale-free it is
`r round(R["CHB/CHS", "GIH"], 2)` --- essentially uncorrelated, as it should be.
Removing the shared component is a good idea; it just has to be done in a way
that respects the unidentified row scale.

## Structure in the residuals

If the 14 factors capture the population structure, the residual
$E = X - AB$ should be close to noise. It is not.

The residual is $2464 \times 185116$ and never needs to be formed:
$EE' = G_X - MA' - AM' + A(BB')A'$ with $M = XB'$, all of which are available
or cheap. [`code/fit_tgp_residual.R`](https://github.com/stephens999/misc/blob/master/code/fit_tgp_residual.R)
builds that, centres it, whitens to 30 dimensions, and runs the same rank-1
search as before --- 1000 independent random starts, 500 iterations, clustered
at the same threshold.

```{r resid-load}
res <- readRDS("../output/tgp_residual.rds")
pop_r <- meta$pop[match(res$ids, meta$sample)]
c(residual_ss_fraction = round(res$ss_ratio, 3))
round(res$d[1:12])
```

The residual holds 32% of the sum of squares, and its spectrum is not flat:
two very large singular values, then 420, 406, 321, and only then the noise
plateau near 300. So several strong directions survive the fit.

```{r resid-maxima}
Lr <- res$L
w  <- pmax(Lr, 0)^2
o  <- order(colSums(w * as.integer(pop_r)) / colSums(w))
Lr <- Lr[, o, drop = FALSE]
sz <- res$size[o]
colnames(Lr) <- make.unique(apply(Lr, 2, function(x) {
  v <- tapply(x, pop_r, mean); v <- sort(v[v > 0], decreasing = TRUE)
  paste(names(v)[v >= 0.3 * v[1]], collapse = "/") }))

data.frame(factor = colnames(Lr), basin = sz,
           skewness = round(apply(Lr, 2, function(x)
             mean(x^3) / mean(x^2)^1.5), 1),
           `n > 2` = colSums(Lr > 2),
           max = round(apply(Lr, 2, max), 1),
           check.names = FALSE, row.names = NULL)
```

Six maxima, and they are of the two kinds seen throughout this work. Three are
population-level (skewness around 3, with 84--112 individuals loading above 2):
one on **MSL**, one on **CLM**, and a second **PJL** direction. Three are the
familiar individual-level kind (skewness 13--15, fewer than 10 individuals),
picking out the LWK pair NA19331/NA19334 and the STU pair HG03750/HG03754 ---
relatives that survived the earlier pruning, because that pruning was decided
on the full-data pool before these fits existed.

```{r resid-plot, fig.width=9, fig.height=5}
plot_resid <- function(L) {
  ord <- order(pop_r); nf <- ncol(L)
  df <- data.frame(idx = rep(seq_len(nrow(L)), nf),
                   loading = as.vector(L[ord, ]),
                   sp = rep(meta$super_pop[match(res$ids, meta$sample)][ord], nf),
                   factor = factor(rep(colnames(L), each = nrow(L)),
                                   levels = colnames(L)))
  brk <- tapply(seq_along(ord), pop_r[ord], mean)
  ggplot(df, aes(idx, loading, colour = sp)) +
    geom_hline(yintercept = 0, linewidth = 0.2, colour = "grey50") +
    geom_point(size = 0.3, alpha = 0.7) +
    facet_wrap(~ factor, ncol = 3, scales = "free_y") +
    scale_x_continuous(breaks = brk, labels = names(brk)) +
    scale_colour_manual(values = sp_palette) +
    labs(x = NULL, y = "loading", colour = "super-population") +
    theme_cowplot(font_size = 9) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 5),
          legend.position = "bottom") +
    guides(colour = guide_legend(override.aes = list(size = 3, alpha = 1),
                                 nrow = 1))
}
plot_resid(Lr)
```

### Is this new structure, or just what $\max(L,0)$ threw away?

Fitting with $A = \max(L,0)$ discards the negative part of every source, so
some residual structure is guaranteed. The question is whether these six
directions are inside the span of the 14 sources --- in which case they are an
artifact of the non-negative approximation --- or outside it, in which case the
source set itself is incomplete.

```{r resid-span}
L14 <- fit$maxima$no_sparse$L
data.frame(factor = colnames(Lr),
           R2_on_span = round(sapply(seq_len(ncol(Lr)), function(j)
             summary(lm(Lr[, j] ~ L14))$r.squared), 3),
           row.names = NULL)
```

They are mostly outside it: 0.001--0.05 for four of the six, and only 0.27 and
0.37 for the CLM and MSL directions. A cleaner version of the same test asks
how well the 14 sources can represent a population at all, by regressing a
pure population indicator on their span:

```{r resid-indicator}
sapply(c("LWK", "GWD", "MSL", "CLM", "PJL"), function(pp)
  round(summary(lm(as.numeric(pop_r == pp) ~ L14))$r.squared, 3))
```

That settles it. LWK and GWD, which have their own factors among the 14, are
represented almost perfectly (0.96 and 0.90). MSL and CLM are not (0.35 and
0.24) --- and MSL and CLM are precisely the two populations the residual search
recovers. They are real populations with no factor of their own in the
$K = 14$ solution, so the structure in the residual is genuinely missing from
the sources rather than lost in the non-negative step.

The practical reading is that $K = 14$ is too small for these data even though
it is what the rank-1 search converges on: the maxima it finds are real, but
they do not exhaust the population structure. MSL and CLM are visible in the
$K = 20$ and $K = 30$ rank-$r$ fits of the earlier analysis, where they appear
as ordinary population factors --- and MSL was one of the factors there that
dissolved when the orthogonality constraint was lifted. Taken together, these
are populations the $x|x|$ objective does not make a local maximum of, but
which are nonetheless present in the data.

## A binary-subset decomposition of the correlation matrix

The correlation matrix of the $K = 14$ components can be read as a sum of
contributions from shared ancestry. Model it as
$$
C \approx \sum_k d_k\, t_k t_k' + E, \qquad t_k \in \{0,1\}^{14},\ d_k \ge 0,
$$
so each binary vector picks a subset $S_k$ of components and adds $d_k$ to the
correlation of every pair inside it. With $K = 14$ we can simply enumerate all
$2^{14}$ subsets and let an $L_1$-penalized non-negative regression choose a
sparse set of them. We fit the **off-diagonal** entries only, and use the raw
correlation matrix with the shared component still in it --- the all-ones
subset is in the dictionary, so that component is something the model can
represent rather than something we have to remove first.

```{r subset-design}
Cm <- struct$pruned_maxima$cor_B[m14$o, m14$o]
dimnames(Cm) <- list(colnames(m14$A), colnames(m14$A))
Kc <- nrow(Cm)
short <- sub("/.*", "", colnames(m14$A))      # short labels for printing

# All subsets with at least two members (a singleton contributes no pairs).
Tm <- sapply(0:(2^Kc - 1), function(x) as.integer(intToBits(x))[1:Kc])
Tm <- Tm[, colSums(Tm) >= 2]

ij <- which(upper.tri(Cm), arr.ind = TRUE)
Xd <- t(sapply(seq_len(nrow(ij)), function(r) Tm[ij[r, 1], ] * Tm[ij[r, 2], ]))
yd <- Cm[upper.tri(Cm)]
c(pairs = length(yd), subsets = ncol(Tm))
```

### A caution about the solver

`glmnet` gives the wrong answer on this design. The columns are near-duplicates
--- the all-ones subset and the fourteen "all but one" subsets agree on at
least 78 of the 91 pairs --- and its screening rules discard the all-ones
column even though that column has the largest $x'y$ and should enter first.
The result is a solution that is not merely a different sparse representation
but has a strictly worse objective, so we use a plain active-set coordinate
descent instead and check the KKT conditions.

```{r subset-solver}
nnlasso <- function(X, y, lam, pf = rep(1, ncol(X)), tol = 1e-10) {
  n <- length(y); xx <- colSums(X^2)
  b <- numeric(ncol(X)); act <- integer(0); r <- y
  for (out in 1:100) {
    g    <- as.vector(crossprod(X, r)) / n - lam * pf
    viol <- setdiff(which(g > tol), act)
    if (!length(viol)) break
    act <- union(act, viol[order(-g[viol])][1:min(20, length(viol))])
    Xa <- X[, act, drop = FALSE]; ba <- b[act]; xa <- xx[act]; pa <- pf[act]
    r  <- y - Xa %*% ba
    for (it in 1:5000) {                       # CD on the active set
      md <- 0
      for (jj in seq_along(act)) {
        r  <- r + Xa[, jj] * ba[jj]
        nb <- max(0, sum(Xa[, jj] * r) / n - lam * pa[jj]) / (xa[jj] / n)
        md <- max(md, abs(nb - ba[jj])); ba[jj] <- nb
        r  <- r - Xa[, jj] * nb
      }
      if (md < 1e-13) break
    }
    b[act] <- ba; act <- act[b[act] > 0]; r <- as.vector(y - X %*% b)
  }
  b
}
```

### The path

```{r subset-path}
setname <- function(t) paste(short[t == 1], collapse = "+")
lams <- c(0.30, 0.15, 0.08, 0.04, 0.02, 0.01)
fits <- lapply(lams, function(l) nnlasso(Xd, yd, l))

do.call(rbind, lapply(seq_along(lams), function(i) {
  b <- fits[[i]]
  data.frame(lambda = lams[i], terms = sum(b > 1e-8),
             R2 = round(1 - sum((yd - Xd %*% b)^2) / sum(yd^2), 4),
             max_KKT = signif(max(as.vector(crossprod(Xd, yd - Xd %*% b)) /
                                  length(yd) - lams[i]), 2))
}))
```

The KKT residuals are at machine precision throughout, so these are the exact
solutions. The sequence in which subsets enter is the interesting part:

```{r subset-terms}
for (i in seq_along(lams)) {
  b <- fits[[i]]; nz <- which(b > 1e-8)
  if (length(nz) > 6) next
  cat(sprintf("lambda = %.2f\n", lams[i]))
  for (j in nz[order(-b[nz])])
    cat(sprintf("   d = %.3f  |S| = %2d  %s\n", b[j], sum(Tm[, j]),
                setname(Tm[, j])))
}
```

The decomposition is nested and reads as a population history:

1. **All fourteen**, $d \approx 0.65$, entering first and alone explaining 94%
   of the off-diagonal sum of squares. This is the shared allele frequency
   spectrum --- the thing that made the raw correlations all exceed 0.6.
2. **Everything except the three African components**, $d \approx 0.14$--$0.23$.
   The out-of-Africa split, recovered as the single most useful subset after
   the global term, taking the fit to 0.99.

The third term the lasso adds is *not* the obvious one, and that turns out to
be a defect of the penalty rather than a feature of the data.

### The $L_1$ penalty has a size bias

Looking at the heatmap, a three-component model ought to be: a shared
component, an African component, and an out-of-Africa component. The lasso
instead picks African $\cup$ European $\cup$ South Asian as its third subset,
which gives African pairs a fitted correlation of only $0.646 + 0.040 = 0.686$
where the observed values are 0.96--0.98. It is underfitting the African block
badly.

The reason is that an $L_1$ penalty on $d$ alone does not account for how many
pairs a subset touches. The African subset covers 3 of the 91 pairs and needs
$d \approx 0.33$ to fit them, costing 0.33 in penalty; a subset of size 10
covers 45 pairs and buys a comparable reduction in squared error for a cost of
0.09. Small-but-strong blocks are therefore systematically passed over. Fitting
the two candidate three-term models by unpenalized non-negative least squares
shows the cost:

```{r subset-three}
nnls_cols <- function(cols, iters = 3000) {
  Xa <- X_sub[, cols, drop = FALSE]; b <- numeric(length(cols))
  r <- yd; xx <- colSums(Xa^2)
  for (it in 1:iters) {
    md <- 0
    for (j in seq_along(cols)) {
      r  <- r + Xa[, j] * b[j]
      nb <- max(0, sum(Xa[, j] * r) / xx[j])
      md <- max(md, abs(nb - b[j])); b[j] <- nb
      r  <- r - Xa[, j] * nb
    }
    if (md < 1e-13) break
  }
  list(b = b, sse = sum((yd - Xa %*% b)^2))
}
X_sub  <- Xd
idx_of <- function(v) which(apply(Tm, 2, function(t) all(t == v)))
i_all  <- idx_of(rep(1, Kc))
i_afr  <- idx_of(c(1, 1, 1, rep(0, Kc - 3)))
i_ooa  <- idx_of(c(0, 0, 0, rep(1, Kc - 3)))
i_lasso3 <- idx_of(c(1,1,1,1,0,1,1,1,1,1,1,0,0,0))

sapply(list(`lasso: all + nonAFR + (AFR,EUR,SAS)` = c(i_all, i_ooa, i_lasso3),
            `all + AFR + out-of-Africa`           = c(i_all, i_afr, i_ooa)),
       function(cl) round(1 - nnls_cols(cl)$sse / sum(yd^2), 4))
```

The proposed model fits better with the same number of terms. And unpenalized
greedy forward selection --- which has no size bias, since it just asks which
single subset most reduces the squared error --- picks it out directly:

```{r subset-greedy}
sel <- integer(0)
for (step in 1:3) {
  sse <- sapply(seq_len(ncol(Xd)),
                function(j) if (j %in% sel) Inf else nnls_cols(c(sel, j))$sse)
  sel <- c(sel, which.min(sse))
  cat(sprintf("step %d: add %-30s R2 = %.4f\n", step, setname(Tm[, sel[step]]),
              1 - nnls_cols(sel)$sse / sum(yd^2)))
}
```

So the interpretable three-component model is the one the eye picks off the
heatmap: **a shared component** ($d = 0.645$), **an African component**
($d = 0.328$ on LWK, ESN/YRI/ACB/ASW, GWD) and **an out-of-Africa component**
($d = 0.232$ on the other eleven), reaching $R^2 = 0.996$ on the off-diagonals.
That a greedy search over all 16,369 subsets recovers exactly this, in this
order and with no tree structure imposed, is good evidence that the $x|x|$
components carry real ancestry information.

### Does normalizing the columns fix it?

The natural remedy is to put the columns on an equal footing before
penalizing. Since the columns are 0/1, scaling each to unit $L_2$ norm is the
same as a penalty factor of $\sqrt{\#\text{pairs}}$, half way between the
uniform penalty above and a per-pair one. All three were tried, with the same
solver:

| column scaling | penalty $\propto$ | behaviour |
|---|---|---|
| none | $1$ | African subset never enters in the sparse regime |
| unit $L_2$ norm | $\sqrt{\#\text{pairs}}$ | African enters 4th, at $\lambda = 0.005$ |
| per pair | $\#\text{pairs}$ | floods: 105 terms at the largest $\lambda$ tried |

```{r subset-l2}
pf   <- sqrt(colSums(Xd))            # = ||x_j||_2, i.e. unit-L2 columns
fit2 <- lapply(c(0.010, 0.008, 0.005), function(l) nnlasso(Xd, yd, l, pf = pf))
for (i in seq_along(fit2)) {
  b <- fit2[[i]]; nz <- which(b > 1e-8)
  cat(sprintf("lambda = %.3f  terms = %d  R2 = %.4f : %s\n",
              c(0.010, 0.008, 0.005)[i], length(nz),
              1 - sum((yd - Xd %*% b)^2) / sum(yd^2),
              paste(sapply(nz[order(-b[nz])], function(j) setname(Tm[, j])),
                    collapse = " | ")))
}
```

Normalizing helps --- the African block enters the solution at all, which it
never does under a uniform penalty --- but it still arrives fourth, behind the
same size-10 subset, and the four-term fit ($R^2 = 0.990$) is worse than the
greedy three-term one. The size bias is reduced, not removed: even with equal
column norms, a subset covering 45 pairs still has 15 times as many chances to
reduce the squared error as one covering 3.

### susieR

The real problem is that $L_1$ conflates selection with shrinkage: the penalty
that decides *which* subset enters is the same one that sets *how large* its
coefficient is, and the two want different scalings. `susieR` separates them.
Each single effect picks its column by Bayes factor and estimates the effect
size on its own, so there is no size bias, and the near-duplicate columns that
defeated `glmnet` are exactly what its credible sets are designed for.

```{r subset-susie}
nrm  <- sqrt(colSums(Xd^2))
Xn   <- sweep(Xd, 2, nrm, "/")          # unit L2 columns
sfit <- susieR::susie(Xn, yd, L = 10, intercept = FALSE, standardize = FALSE)
dhat <- coef(sfit)[-1] / nrm            # back to the d scale

c(converged = sfit$converged,
  R2 = round(1 - sum((yd - Xd %*% dhat)^2) / sum(yd^2), 4),
  all_positive = all(dhat[dhat != 0] > 0))

cs <- susieR::susie_get_cs(sfit, X = Xn)
do.call(rbind, lapply(seq_along(cs$cs), function(i) {
  j <- cs$cs[[i]][which.max(sfit$pip[cs$cs[[i]]])]
  data.frame(CS = i, size = length(cs$cs[[i]]),
             purity = round(cs$purity[i, "min.abs.corr"], 2),
             pip = round(sfit$pip[j], 3), d = round(dhat[j], 3),
             `|S|` = sum(Tm[, j]), subset = setname(Tm[, j]),
             check.names = FALSE, row.names = NULL)
}))
```

This is the decomposition we were after, and it is cleaner than anything the
lasso produced. Five credible sets, **every one a singleton with purity 1 and
PIP 1**, reaching $R^2 = 0.999$:

- **all fourteen**, $d = 0.59$ --- the shared spectrum
- **out-of-Africa** (the eleven non-African), $d = 0.25$
- **African** (LWK, ESN/YRI/ACB/ASW, GWD), $d = 0.30$ --- the block the lasso
  would not select at any penalty
- **East Asian** (CDX/KHV, CHB/CHS, JPT), $d = 0.14$
- **African $\cup$ European $\cup$ South Asian**, $d = 0.09$

The implied fitted values check out against the heatmap: African pairs get
$0.59 + 0.30 + 0.09 = 0.97$ against 0.96--0.98 observed, East Asian pairs
$0.59 + 0.25 + 0.14 = 0.97$ against 0.97--0.98, and African-to-East-Asian
pairs get the shared term alone, 0.59, against 0.60--0.62 observed.

`L` is not binding: `L = 6`, `L = 10` and `L = 15` all return the same five
credible sets with the same subsets and the same $d$ to three decimals, at
identical $R^2$ and an ELBO agreeing to 128.376 vs 128.377.

```{r subset-susie-L}
sapply(c(6, 10, 15), function(LL) {
  fl <- susieR::susie(Xn, yd, L = LL, intercept = FALSE, standardize = FALSE)
  dl <- coef(fl)[-1] / nrm
  c(L = LL, nCS = length(susieR::susie_get_cs(fl, X = Xn)$cs),
    R2 = round(1 - sum((yd - Xd %*% dl)^2) / sum(yd^2), 5),
    ELBO = round(susieR::susie_get_objective(fl), 3))
})
```

Two other things worth noting. Every coefficient came out positive without
non-negativity being imposed, so the constraint is not binding --- the data
prefer a non-negative decomposition on their own. And the credible sets are
all singletons despite a dictionary in which the top columns correlate above
0.99, which is the case `susieR` is built for and where both `glmnet` and the
greedy search were on shakier ground.

For a decomposition at this scale `susieR` is clearly the right tool. Greedy
selection gets the first three terms right; the $L_1$ path is reliable only
for the first two.

### Residuals after removing the binary components

```{r subset-resid, fig.width=6.5, fig.height=5.5}
resid_vec <- as.vector(yd - Xd %*% dhat)
Rm <- matrix(0, Kc, Kc, dimnames = dimnames(Cm))
Rm[upper.tri(Rm)] <- resid_vec
Rm <- Rm + t(Rm)

c(sd = round(sd(resid_vec), 4), min = round(min(resid_vec), 3),
  max = round(max(resid_vec), 3))

lim <- max(abs(Rm))
dfR <- data.frame(
  i = factor(rep(rownames(Rm), ncol(Rm)), levels = rev(rownames(Rm))),
  j = factor(rep(colnames(Rm), each = nrow(Rm)), levels = colnames(Rm)),
  r = as.vector(Rm))
ggplot(dfR, aes(j, i, fill = r)) +
  geom_tile() +
  geom_text(aes(label = ifelse(r == 0, "", sprintf("%+.02f", r))), size = 2.1,
            colour = "grey15") +
  scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B",
                       limits = c(-lim, lim)) +
  labs(x = NULL, y = NULL, fill = "residual",
       title = "Correlation residuals after the five binary components") +
  theme_cowplot(font_size = 9) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```

The residuals are an order of magnitude smaller than the correlations
themselves --- standard deviation about 0.026 against correlations of 0.6 to
0.99 --- but they are plainly not noise, and what is left tells us what the
five subsets miss.

```{r subset-resid-top}
ord <- order(-abs(resid_vec))[1:8]
data.frame(pair = paste(rownames(Cm)[ij[ord, 1]], "vs", colnames(Cm)[ij[ord, 2]]),
           observed = round(yd[ord], 3),
           residual = round(resid_vec[ord], 3), row.names = NULL)
```

Three things stand out.

**Europe and South Asia have no block of their own.** The model gives Africa
and East Asia their own subsets but not Europe or South Asia, and the residuals
show exactly that gap: TSI/IBS--GBR/CEU is the largest positive residual at
+0.063, with GBR/CEU--FIN at +0.048, and on the South Asian side
GIH--STU/ITU/BEB at +0.056 and PJL--STU/ITU/BEB at +0.044. Those blocks are
under-fitted, and the corresponding European-to-East-Asian entries are
over-fitted at about $-0.05$.

**PUR does not fit the subset structure at all.** Its entire row is positive
against Africa, the Americas and Europe (+0.04 to +0.06) and negative against
East Asia. A model built from binary memberships cannot represent a population
that is a genuine three-way admixture, so PUR shows up as a uniformly
under-predicted row rather than as a missing block.

**The structure is nested, not partitioned.** A tree would leave residuals
without this pattern of matched positive and negative blocks. What the
residuals describe is the difference between a strict hierarchy and real
population history with gene flow across the branches.

It is worth noting that `susieR` stopped at five components with `L = 15`
available, so it judged the European and South Asian blocks not worth adding
on 91 data points. The residual plot suggests they are real; with only 91
off-diagonal entries there is simply not much power to establish them.


## Caveats

- Individuals with no positive loading on any source in a set have no
  membership representation and are dropped from that plot: one for the odd
  maxima, none for the other three maxima sets, but 47 and 17 for the
  stability-pruned K = 30 sets. Each plot title reports its own count.

- $B \ge 0$ is empirical, not guaranteed. It holds for every maxima set here
  and fails for every rank-$r$ fit, so it should be rechecked for any new set
  of sources rather than assumed.

- The $R^2$ values are against the raw sum of squares of $X$, which is
  dominated by the mean. They are not evidence that individual genotypes are
  well predicted and should not be read as such.

- This is **not** the $X \approx 2QP$ parameterization with $Q$ non-negative,
  rows summing to 1, and $P$ a matrix of allele frequencies. That version is
  infeasible on these data. Making $Q$ non-negative with rows summing to 1
  requires a simplex whose vertices sit about 2.9 times further from the
  baseline than any observed individual, and the implied frequencies at those
  vertices then run from $-2.6$ to $3.9$, with 46% outside $[0,1]$. An
  enclosing simplex has to extrapolate past the data, and the frequency map is
  already marginal at the data itself (about 4% of fitted individual
  frequencies fall outside $[0,1]$). What is plotted here is the weaker,
  achievable thing: non-negative memberships on a common scale, with no claim
  that the rows of $B$ are population allele frequencies.
