---
title: "fastICA with x|x| contrast on HGDP haplotypes"
author: "Matthew Stephens"
date: "2026-09-18"
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
    toc_depth: 2
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

## Introduction

This repeats the analysis of [`fastica_1kg`](fastica_1kg.html) --- fastICA with
the asymmetric $G(x) = x|x|$ contrast --- on the Human Genome Diversity Panel
whole-genome data, and uses the recipe that analysis arrived at rather than
rediscovering it:

* whiten to $K$ components, here $K = 40$ and $K = 60$;
* find the local maxima of the objective with 1000 independent **rank-1** runs
  from random starts, and cluster the results, rather than reading factors off
  a single rank-$r$ fit.

The 1000 Genomes analysis reached that recipe the long way round. Its lesson
was that a single rank-$r$ fit is the unreliable step --- it can lose a real
factor to a bad local optimum --- while the rank-1 searches say directly which
directions are genuine maxima.

It also pruned the data: a small number of related individuals and sample
outliers dominated very sparse factors there, and dropping 40 of 2504
reorganized the whole landscape. **That step turns out not to be needed
here**, so the analysis below uses every individual. The evidence is kept in
one short section rather than run through the whole document. Part of the
reason is that these 929 samples contain no duplicates and no first-degree
relative pairs at all (checked below), so the relative pairs that dominated
the sparse 1000 Genomes factors are simply not present; and part is that
HGDP populations are small and drifted enough that a genuine population
factor is already narrow and sharp.

Two things differ from the 1000 Genomes setting. HGDP is smaller and much
more finely sampled --- 929 individuals from 54 populations in 7 regions,
against 2504 from 26 populations --- so there are far more populations than
components. And the data are statistically phased, so we work with the
$2 \times 929 = 1858$ **haplotypes** as rows rather than with individuals.

```{r load}
library(ggplot2)
library(cowplot)
library(fastTopics)

meta   <- readRDS("../data/hgdp_meta.rds")
fit    <- readRDS("../output/hgdp_fastica.rds")
struct <- readRDS("../output/hgdp_structure.rds")
kin    <- readRDS("../output/hgdp_kinship.rds")

region_palette <- c(AFRICA             = "#E69F00",
                    MIDDLE_EAST        = "#D55E00",
                    EUROPE             = "#7570B3",
                    CENTRAL_SOUTH_ASIA = "#1B9E77",
                    EAST_ASIA          = "#0072B2",
                    OCEANIA            = "#CC79A7",
                    AMERICA            = "#332288")

# One row of metadata per haplotype, in the row order of the fits.
hap_meta <- function(ids) {
  m <- meta[match(sub("_[12]$", "", ids), meta$sample), ]
  m$hap <- ids
  m
}
hm <- hap_meta(fit$hap_ids)

pop_order <- levels(meta$pop)
skew_obj  <- function(L) colMeans(L * abs(L))
```

## Data

The genotypes are the statistically phased autosomal VCF from the
[HGDP WGS release](https://ngs.sanger.ac.uk/production/hgdp/hgdp_wgs.20190516/statphase/)
(GRCh38, 929 samples), and the sample table is the
[metadata file](https://ngs.sanger.ac.uk/production/hgdp/hgdp_wgs.20190516/metadata/)
from the same release. That VCF is 5.9 Gb compressed and has tens of millions
of variants, so [`code/vcf_thin.c`](https://github.com/stephenslab/misc/blob/master/code/vcf_thin.c)
streams it once and keeps a thinned subset: biallelic `PASS` SNPs with an
`AA_ensembl` ancestral allele matching one of the two alleles, no missing call
in any sample, derived-allele frequency in $[0.05, 0.95]$, and at least 6 kb
from the previous SNP kept on that chromosome. Each kept SNP is written as
1858 derived-allele indicators, one per haplotype.

Distance thinning is a cruder stand-in for LD pruning than the $r^2$ pruning
behind the 1000 Genomes matrix, but it costs one pass over the file instead of
several. The 6 kb spacing is not a tuned choice, and the last section checks
what it buys: halving it from 12 kb, which doubles the SNP count, leaves every
maximum in place at $r \ge 0.98$.

[`code/prep_hgdp_data.R`](https://github.com/stephenslab/misc/blob/master/code/prep_hgdp_data.R)
turns that into `data/hgdp_hap_matrix.rds`, and
[`code/fit_hgdp_fastica.R`](https://github.com/stephenslab/misc/blob/master/code/fit_hgdp_fastica.R)
does everything below. As in the TGP script the columns are centred and
nothing else is done to them --- no division by $\sqrt{2p(1-p)}$ --- and the
left singular vectors come from the $1858 \times 1858$ centred Gram matrix
rather than from an SVD of the centred data.

```{r data-size}
c(haplotypes = length(fit$hap_ids), individuals = nrow(meta),
  SNPs = fit$nsnp, populations = nlevels(meta$pop), regions = nlevels(meta$region))
table(meta$region)
```

## Eigenvalue scree plot

```{r scree, fig.width=6, fig.height=3.2}
ggplot(data.frame(k = 1:60, d = fit$d[1:60]), aes(k, d)) +
  geom_point(size = 1.2) +
  geom_vline(xintercept = fit$K, linetype = "dashed", colour = "dodgerblue") +
  scale_x_continuous(breaks = c(1, 10, 20, 30, 40, 50, 60)) +
  labs(x = "component", y = "singular value") +
  theme_cowplot(font_size = 10)
round(fit$d[1:20])
```

The first four are well separated --- Africa, then Oceania and the Americas
--- and from about the tenth the spectrum is a slow decline with no elbow, so
$K$ is a choice about how many sources we are willing to look for rather
than a dimension the data hand us --- which is why both 40 and 60 are tried.
With 54 populations there are fewer components than populations at $K = 40$,
the opposite of the 1000 Genomes setting (30 components, 26 populations), and
that is worth keeping in mind below.

## Plotting functions

```{r plot-fns}
# Order factors so they roughly follow the (geographic) population order: each
# factor takes the mean population rank of the haplotypes it loads on, weighted
# by pmax(L,0)^2 so only the large positive loadings count.
pop_order_of <- function(L, m) {
  w <- pmax(L, 0)^2
  order(colSums(w * as.integer(m$pop)) / colSums(w))
}

# Populations carrying a factor: those whose mean loading is within `frac` of
# the largest. Capped at `top` names so that panel strips stay readable --- a
# factor spread over more than a few populations gets a trailing "+".
pop_label <- function(x, m, frac = 0.3, top = 3) {
  v <- tapply(x, m$pop, mean)
  v <- sort(v[v > 0], decreasing = TRUE)
  v <- names(v)[v >= frac * v[1]]
  paste0(paste(head(v, top), collapse = "/"), if (length(v) > top) "+" else "")
}

# Factors are referred to by short names (f01, f02, ...) everywhere, with the
# populations, objective, basin size and sparsity kept in a companion table.
name_maxima <- function(L, m, n_rand) {
  o <- pop_order_of(L, m)
  L <- L[, o, drop = FALSE]
  colnames(L) <- sprintf("f%02d", seq_len(ncol(L)))
  list(L   = L,
       tab = data.frame(factor      = colnames(L),
                        pops        = apply(L, 2, pop_label, m = m),
                        obj         = round(skew_obj(L), 2),
                        n_above2    = colSums(L > 2),
                        max_loading = round(apply(L, 2, max), 1),
                        rand_starts = n_rand[o],
                        row.names   = NULL))
}

# Loadings of every haplotype, x-axis grouped by sampling population.
plot_loadings <- function(L, m, tab, ncol = 2) {
  ord <- order(m$pop)
  nf  <- ncol(L)
  strip <- sprintf("%s  %s  n=%d", colnames(L), tab$pops, tab$n_above2)
  df  <- data.frame(
    idx     = rep(seq_len(nrow(L)), nf),
    loading = as.vector(L[ord, ]),
    region  = rep(m$region[ord], nf),
    factor  = factor(rep(strip, each = nrow(L)), levels = strip))
  brk <- tapply(seq_along(ord), m$pop[ord], mean)
  ggplot(df, aes(idx, loading, colour = region)) +
    geom_hline(yintercept = 0, linewidth = 0.2, colour = "grey50") +
    geom_point(size = 0.35, alpha = 0.7) +
    facet_wrap(~ factor, ncol = ncol, scales = "free_y") +
    # Population positions are haplotype indices, so the small populations sit
    # very close together; dodging the labels over two rows stops them
    # colliding without moving any point.
    scale_x_continuous(breaks = brk, labels = names(brk), expand = c(0.01, 0),
                       guide = guide_axis(n.dodge = 2)) +
    scale_colour_manual(values = region_palette) +
    labs(x = NULL, y = "loading", colour = "region") +
    theme_cowplot(font_size = 9) +
    theme(axis.text.x     = element_text(angle = 90, vjust = 0.5, hjust = 1,
                                         size = 5.5),
          strip.text      = element_text(size = 7),
          legend.position = "bottom") +
    guides(colour = guide_legend(override.aes = list(size = 3, alpha = 1),
                                 nrow = 1))
}

# Population-mean loadings as a heatmap, populations in geographic order.
plot_pop_heatmap <- function(L, m) {
  M  <- apply(L, 2, function(x) tapply(x, m$pop, mean))
  df <- data.frame(
    pop    = factor(rep(rownames(M), ncol(M)), levels = rev(pop_order)),
    factor = factor(rep(colnames(M), each = nrow(M)), levels = colnames(M)),
    value  = as.vector(M))
  # A few isolated populations (San, Surui) load an order of magnitude higher
  # than anything else, so a full-range scale renders every broad factor as
  # white. Clip the scale at the 98th percentile and squish the rest.
  lim <- as.numeric(quantile(abs(df$value), 0.98))
  ggplot(df, aes(factor, pop, fill = value)) +
    geom_tile() +
    scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B",
                         limits = c(-lim, lim), oob = scales::squish) +
    labs(x = NULL, y = NULL, fill = "mean\nloading") +
    theme_cowplot(font_size = 9) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 6),
          axis.text.y = element_text(size = 5.5))
}
```

## The 1000 rank-1 runs at $K = 40$

As in the 1000 Genomes analysis we run 1000 independent rank-1 fastICA runs in
parallel --- one column of `W` per run, each renormalized to unit length after
every update, with no orthogonalization between them --- from random starts in
the $K = 40$ whitened space. What the runs converge to are the local maxima of
the objective, and the number of runs landing in each is a rough measure of its
basin of attraction. Runs that found the same maximum are merged by clustering
on signed correlation with complete linkage; the correlation is signed, not
absolute, because $x|x|$ is not sign-symmetric and $w$ and $-w$ are genuinely
different solutions.

```{r r1-basics}
mx <- fit$fits$k40$maxima
c(largest_change_last_iter = signif(mx$converged, 3))
# Number of distinct maxima if we cut the tree at correlation 0.90 / 0.95 / 0.99.
mx$n_cut
c(n_maxima = ncol(mx$L),
  found_by_random_starts = sum(mx$n_random > 0),
  found_only_from_rank_r = sum(mx$n_random == 0))
```

```{r r1-order}
r1   <- name_maxima(mx$L, hm, mx$n_random)
L_r1 <- r1$L
r1$tab
```

The clustering is not sensitive to where we cut --- the same number of maxima
at a correlation of 0.90, 0.95 and 0.99 --- so these are well-separated
solutions, not a continuum.

The striking thing, against the 1000 Genomes result, is that **almost every
maximum is a population factor**. There is no analogue of the GIH trio that
took 307 of 1000 starts there, and no relative-pair factors at all in this
list. The only exceptions are the second Bedouin factor and the Hazara factor,
each loading on five or six individuals at up to 13, and both are taken up in
the next section. Everything else
is a population, a pair of neighbouring populations, or a regional cluster,
and the ordering runs cleanly from Africa through the Middle East, Europe,
Central/South Asia and East Asia to Oceania and the Americas.

```{r r1-plot, fig.width=10, fig.height=2 + 1.7 * ceiling(ncol(L_r1) / 2)}
plot_loadings(L_r1, hm, r1$tab)
```

```{r r1-heat, fig.width=9, fig.height=7}
plot_pop_heatmap(L_r1, hm)
```

### Basin size against objective

```{r r1-basins, fig.width=4.5, fig.height=3.5}
df <- data.frame(size = r1$tab$rand_starts, obj = skew_obj(L_r1))
ggplot(df, aes(size, obj)) + geom_point(size = 1.4) +
  labs(x = "random starts reaching this maximum", y = "objective") +
  theme_cowplot(font_size = 10)
round(cor(df$size, df$obj), 2)
```

Basin size and objective are more strongly related here than in the 1000
Genomes data (0.39 there), but the relationship is still not one a search
could rely on. The pattern behind it is geographic rather than numerical: the
maxima with large basins are the isolated, strongly drifted populations --- San,
Surui, Karitiana, Mbuti, the Papuan groups, Bougainville, Pima --- and the ones
random starts almost never reach are the broad, shallow factors spanning many
related populations, such as Russian/Orcadian/French, Brahui/Makrani/Balochi,
Burusho/Pathan/Sindhi and Japanese/Hezhen/Daur. Those have basins of zero: they
are found only by starting rank-1 at the corresponding rank-$r$ factor.

```{r r1-unreachable}
data.frame(factor = colnames(L_r1), pops = r1$tab$pops,
           obj = r1$tab$obj)[r1$tab$rand_starts == 0, ]
```

This is the same lesson as in the 1000 Genomes analysis, and it is the reason
for running the rank-$r$ fit at all: 1000 random rank-1 starts, on their own,
would miss `r sum(r1$tab$rand_starts == 0)` of the `r ncol(L_r1)` maxima,
including most of the broad continental ones.

## Is pruning needed? No

In the 1000 Genomes data a handful of related individuals and sample outliers
dominated very sparse factors, and removing them reorganized the landscape:
17 maxima became 14, three maxima that random starts could never reach became
reachable, and a new composite factor appeared. Here almost none of that
happens, and this section is the check.

The 1000 Genomes rule --- "fewer than 20 individuals load above 2" --- does not
transfer, and this is worth recording because applying it unchanged does real
damage. Every 1000 Genomes population has 60--100 individuals, so nothing that
narrow could be a population; HGDP populations run from 6 to 46, and a bare
count deletes whole populations --- San (6), Colombian (7), Surui (8) and Mbuti
(13) each have their own maximum and each looks "very sparse" by that rule.
Run as written it drops three entire populations and loses their factors,
while the fit converges just as cleanly and still reports a tidy list.

So the test is made relative to population size. A maximum is *screened in* if
it is narrow --- fewer than `r fit$prune_check$sparse_nmax` haplotypes load
above 2 --- and judged a nuisance only if the individuals it singles out
(loading above `r fit$prune_check$sparse_lmin` on either haplotype) are less
than `r fit$prune_check$sparse_fracmax` of the population most of them belong
to. `frac` is the column that does the work: the whole-population maxima sit
at 0.92--1.00 and the within-population ones at 0.13--0.36, with nothing in
between.

```{r prune-screen}
pc <- fit$prune_check
st <- pc$screen_tab
st$verdict <- ifelse(st$frac < pc$sparse_fracmax, "within-population",
                     "whole population")
st[order(st$frac), ]
```

Note the objective alone would not have separated these: the
within-population maxima score 0.52--0.75 and the whole-population ones
0.61--0.96, which overlap. It is the fraction, not the sharpness.

```{r prune-set}
c(individuals_dropped = length(pc$dropped),
  individuals_left    = nrow(meta) - length(pc$dropped),
  maxima_before       = ncol(fit$fits$k40$maxima$L),
  maxima_after        = ncol(pc$maxima$L))
table(droplevels(meta$pop[match(pc$dropped, meta$sample)]))
```

Twenty-seven individuals in five clusters of 4--8, from BantuKenya, Bedouin,
Druze, Makrani and Hazara.

### Are those clusters relatives?

The obvious reading, and the one the 1000 Genomes analysis would suggest, is
that clusters that size and that sharp are close relatives. On these data
that is mostly **wrong**, which is worth establishing rather than assuming.
[`code/fit_hgdp_kinship.R`](https://github.com/stephenslab/misc/blob/master/code/fit_hgdp_kinship.R)
computes KING-robust kinship for all 431,056 pairs from the same SNPs.

```{r kinship-counts}
kin$counts
```

There are **no duplicates and no first-degree pairs** among the 929. That is
itself informative: the full HGDP-CEPH panel of 1064 needs 77 individuals
removed to reach that state (Rosenberg 2006 defines the nested clean subsets
H1048, H971 and H952 for exactly this purpose), so these samples must come
from a relatedness-aware starting set, even though the sequencing paper's
supplement documents its QC in terms of coverage, error rates, array
concordance and cell-line artefacts rather than relatedness. Twenty pairs sit
in the second-degree band and 69 in the third, and they are concentrated in
Surui, Karitiana, Pima and Bougainville --- the smallest and most drifted
populations, where KING-robust is inflated by background relatedness and a
"second-degree" call need not be a pedigree relationship.

Against that background, here is the maximum kinship of each dropped
individual to anyone in the panel:

```{r kinship-dropped}
dm <- data.frame(sample = pc$dropped,
                 pop = as.character(meta$pop[match(pc$dropped, meta$sample)]),
                 max_kinship = round(kin$max_kinship[pc$dropped], 3))
aggregate(max_kinship ~ pop, dm, function(x)
  c(n = length(x), max = max(x), median = median(x)))
```

Only the BantuKenya cluster is clearly relatedness (0.121, second degree).
Hazara reaches 0.084 and Druze 0.077, both third-degree at best. **Bedouin
(0.037) and Makrani (0.034) show no detectable relatedness at all** --- those
are genuine fine-scale substructure within a population, not families.

The contrast is not blind to relatedness, though. Dropped individuals are
enriched for it by about fivefold:

```{r kinship-enrichment}
rel3 <- kin$ids[kin$max_kinship > 0.0442]
c(`dropped, in a 3rd-degree-or-closer pair` =
    mean(pc$dropped %in% rel3),
  `not dropped, same`                       =
    mean(setdiff(kin$ids, pc$dropped) %in% rel3))
```

So the narrow factors pick up relatedness where it exists, but on this panel
most of what they pick up is real population substructure. That makes the
decision not to prune easier rather than harder: pruning would have thrown
away four Bedouin-internal and four Makrani-internal signals that are not
artifacts of anything.

And the result of removing them:

```{r prune-compare}
hm_pc  <- hap_meta(pc$hap_ids)
L_pc   <- pc$maxima$L[, pop_order_of(pc$maxima$L, hm_pc), drop = FALSE]
Cpc    <- cor(L_r1[match(pc$hap_ids, fit$hap_ids), , drop = FALSE], L_pc)
data.frame(before = r1$tab$pops,
           after  = apply(L_pc, 2, pop_label, m = hm_pc)[apply(Cpc, 1, which.max)],
           r      = round(apply(Cpc, 1, max), 2), row.names = NULL)
c(unchanged = sum(apply(Cpc, 1, max) > 0.9),
  lost      = sum(apply(Cpc, 1, max) < 0.9),
  new       = sum(apply(Cpc, 2, max) < 0.9))
```

The only two maxima that go are the Bedouin and Hazara within-population
clusters --- the things we deliberately removed. Every other maximum survives
at $r \ge 0.95$ and nothing new appears. Pruning here buys the removal of two
factors we can simply recognize and set aside, at the cost of 27 individuals,
so the rest of this analysis does not prune.

Two plausible reasons it matters so much less than in the 1000 Genomes data.
The HGDP WGS panel may already have been screened for close relatives. And
HGDP populations are small and many are strongly drifted, so a genuine
population factor here is already narrow and sharp --- San loads 6
individuals at 12 --- and occupies the part of the $x|x|$ landscape that
relative pairs occupied in the 1000 Genomes data.

## $K = 60$

The obvious question is whether 40 components is enough, given 54 populations.
Repeating everything at $K = 60$ --- same data, same 1000 rank-1 runs, no
pruning:

```{r k60-basics}
mx60 <- fit$fits$k60$maxima
r60  <- name_maxima(mx60$L, hm, mx60$n_random)
L_60 <- r60$L
rbind(`K=40` = c(maxima = ncol(L_r1), rank_r_only = sum(r1$tab$rand_starts == 0)),
      `K=60` = c(maxima = ncol(L_60), rank_r_only = sum(r60$tab$rand_starts == 0)))
mx60$n_cut   # distinct maxima cutting at correlation 0.90 / 0.95 / 0.99
r60$tab
```

**The same 25 maxima.** Matching them to the $K = 40$ set:

```{r k60-match}
C60 <- cor(L_r1, L_60)
data.frame(`K=40` = r1$tab$pops, `K=60` = r60$tab$pops[apply(C60, 1, which.max)],
           r = round(apply(C60, 1, max), 3),
           starts_40 = r1$tab$rand_starts,
           starts_60 = r60$tab$rand_starts[apply(C60, 1, which.max)],
           check.names = FALSE, row.names = NULL)
c(matched_r_gt_0.9 = sum(apply(C60, 2, max) > 0.9),
  worst_r          = round(min(apply(C60, 2, max)), 3),
  k60_only         = sum(apply(C60, 2, max) < 0.9),
  k40_only         = sum(apply(C60, 1, max) < 0.9))
```

Every maximum matches one-to-one, the worst at
`r round(min(apply(cor(L_r1, L_60), 2, max)), 3)` and most above 0.99, and the
basin sizes barely move. So between 40 and 60 the whitening dimension makes no
difference to the unconstrained maxima: PCs 41--60 carry too little population
signal to create new ones. The user-facing answer is that 40 was already
enough, and that the maxima are a property of the data rather than of $K$.

### Factor plots

```{r k60-plot, fig.width=10, fig.height=2 + 1.7 * ceiling(ncol(L_60) / 2)}
plot_loadings(L_60, hm, r60$tab)
```

```{r k60-heat, fig.width=9, fig.height=7}
plot_pop_heatmap(L_60, hm)
```

### Where the extra 20 components do go

The rank-$r$ fit is a different matter: it must use all 60 directions, and its
total objective rises from `r round(sum(skew_obj(fit$fits$k40$fit$L)), 1)` at
$K = 40$ to `r round(sum(skew_obj(fit$fits$k60$fit$L)), 1)` at $K = 60$. So
the extra components are doing something --- they are just not creating new
unconstrained maxima. Here are all 60:

```{r k60-rankr-order}
rr60 <- fit$fits$k60$fit$L
rr60 <- rr60[, pop_order_of(rr60, hm), drop = FALSE]
colnames(rr60) <- sprintf("f%02d", seq_len(ncol(rr60)))
rr_tab <- data.frame(factor = colnames(rr60),
                     pops = apply(rr60, 2, pop_label, m = hm),
                     obj = round(skew_obj(rr60), 2),
                     n_above2 = colSums(rr60 > 2),
                     max_loading = round(apply(rr60, 2, max), 1),
                     row.names = NULL)
rr_tab
```

```{r k60-rankr-plot, fig.width=12, fig.height=2 + 1.6 * ceiling(ncol(rr60) / 3)}
# Three columns rather than two: 60 panels at two columns is an unreadably
# long figure.
plot_loadings(rr60, hm, rr_tab, ncol = 3)
```

Splitting them by whether they correspond to an unconstrained maximum tells
the story:

```{r k60-rankr-split}
Cx <- cor(rr60, L_60)
is_max <- apply(Cx, 1, max) > 0.9
c(matches_a_maximum = sum(is_max), does_not = sum(!is_max),
  maxima_represented = sum(apply(Cx, 2, max) > 0.9), of_maxima = ncol(L_60))
round(rbind(`matches a maximum` = summary(rr_tab$obj[is_max]),
            `does not`          = summary(rr_tab$obj[!is_max])), 2)
summary(rr_tab$n_above2[!is_max])
```

Only 19 of the 60 correspond to a maximum, and those are the good ones ---
objectives 0.60 to 0.94, median 0.82, in the same range as the maxima
themselves (0.58 to 0.96). The other 41 have objectives from 0.15 to 0.74
with a median of 0.27, and they are **not** narrow: a median of 64 haplotypes
load above 2 on them. They are diffuse, low-objective directions that exist
only because orthogonality requires 60 components to be produced, and they
dissolve if the constraint is lifted --- the same thing the 1000 Genomes
analysis saw with the factors that "slid away" under rank-1 refinement.

Note also that the rank-$r$ fit does not even represent all the maxima: 6 of
the 25 have no $K = 60$ factor correlating above 0.9 with them. Raising $K$
buys objective and diffuse extra directions, not resolution. The next section
adds that these factors also fail the non-negativity test the maxima pass.

## Stability: odd against even chromosomes

A different way to ask which maxima are real: fit the two halves of the SNPs
independently and keep only what both find. SNP names are `chrN:pos`, so we
split on chromosome parity, which keeps whole chromosomes together and so
cannot split residual LD across the two halves. Each half gets its own
centred SVD, its own $K = 40$ rank-$r$ fit and its own 1000 rank-1 runs ---
nothing is shared but the individuals.

```{r split-basics}
sp <- fit$split
odd  <- name_maxima(sp$odd$maxima$L,  hm, sp$odd$maxima$n_random)
even <- name_maxima(sp$even$maxima$L, hm, sp$even$maxima$n_random)
rbind(odd  = c(SNPs = sp$odd$n_snps,  maxima = ncol(odd$L)),
      even = c(SNPs = sp$even$n_snps, maxima = ncol(even$L)),
      all  = c(SNPs = fit$nsnp,       maxima = ncol(L_r1)))
```

Each half finds 20 maxima against 25 on the full data. Matching them to each
other:

```{r split-match}
Cs <- cor(odd$L, even$L)
data.frame(odd        = odd$tab$pops,
           odd_starts = odd$tab$rand_starts,
           even       = even$tab$pops[apply(Cs, 1, which.max)],
           even_starts = even$tab$rand_starts[apply(Cs, 1, which.max)],
           r          = round(apply(Cs, 1, max), 3), row.names = NULL)
c(odd_matched_0.9  = sum(apply(Cs, 1, max) > 0.9),
  even_matched_0.9 = sum(apply(Cs, 2, max) > 0.9),
  odd_matched_0.8  = sum(apply(Cs, 1, max) > 0.8),
  of               = ncol(odd$L))
```

Nineteen of the 20 pair up, and the agreement is close: 16 pairs above
$r = 0.9$ and three more between 0.83 and 0.89 --- Druze, Palestinian and
Burusho/Pathan, the broad shallow factors, which are the ones a halved SNP
set should estimate least precisely. Basin sizes across the matched pairs
correlate at 0.99.

```{r split-basins}
ok <- apply(Cs, 1, max) > 0.8
round(cor(odd$tab$rand_starts[ok],
          even$tab$rand_starts[apply(Cs, 1, which.max)][ok]), 2)
```

The single genuine disagreement is the odd half's eighth maximum, matching
its best even partner at $r =$ `r round(min(apply(Cs, 1, max)), 2)`:

```{r split-odd-only}
odd$tab$pops[which.min(apply(Cs, 1, max))]
```

### What the halves lose

More interesting is what both halves miss relative to the full data. They
miss the *same* seven maxima:

```{r split-vs-full}
Do <- cor(L_r1, odd$L); De <- cor(L_r1, even$L)
data.frame(full = r1$tab$pops,
           best_odd  = round(apply(Do, 1, max), 2),
           best_even = round(apply(De, 1, max), 2),
           row.names = NULL)
miss_o <- r1$tab$pops[apply(Do, 1, max) < 0.9]
miss_e <- r1$tab$pops[apply(De, 1, max) < 0.9]
c(missing_from_odd = length(miss_o), missing_from_even = length(miss_e),
  same_set = identical(sort(miss_o), sort(miss_e)))
miss_o
```

That the two independent halves lose exactly the same seven is the point ---
these are not random failures. Where each one goes:

```{r split-where}
mo <- apply(Do, 1, max) < 0.9
data.frame(full      = r1$tab$pops[mo],
           best_odd  = odd$tab$pops[apply(Do, 1, which.max)][mo],
           r_odd     = round(apply(Do, 1, max)[mo], 2),
           best_even = even$tab$pops[apply(De, 1, which.max)][mo],
           r_even    = round(apply(De, 1, max)[mo], 2), row.names = NULL)
```

They fall into three groups.

**Merged.** Basque and Sardinian are absorbed into a broad European factor
in both halves ($r$ = 0.64--0.79), and Russian/Orcadian/French partly with
them. Europe goes from three factors on the full data to one on the odd half
and two on the even --- the split is real but needs the full marker set.

```{r split-europe}
list(full = grep("Basque|Sardinian|Russian|Orcadian|French", r1$tab$pops, value = TRUE),
     odd  = grep("Basque|Sardinian|Russian|Orcadian|French", odd$tab$pops,  value = TRUE),
     even = grep("Basque|Sardinian|Russian|Orcadian|French", even$tab$pops, value = TRUE))
```

**Unresolved.** Japanese/Hezhen/Daur and Colombian/Maya have no counterpart
at all ($r \le 0.27$): East Asia drops from three factors to two and the
Americas from four to three. These are splits the halves cannot see rather
than splits they blur.

**The nuisance factors.** Hazara is the least reproducible thing in the whole
analysis --- $r$ of 0.04 and 0.09, effectively absent from both halves --- and
the Bedouin cluster is found by the odd half ($r = 0.86$) and not the even
($r = 0.13$). That is a point against those two factors rather than against
the data, and it agrees with the pruning check, which identified exactly
these two as within-population clusters. So two independent lines of evidence
say the same thing about them.

This is the same behaviour as in the 1000 Genomes analysis, where the halves
also merged the finer distinctions, and it is worth reading both ways: the
coarse structure is what replicates on half the markers, and the finer splits
the full data supports are real but need the markers to see.

### Factor plots for each half {.tabset}

#### Odd chromosomes

```{r split-odd-plot, fig.width=10, fig.height=2 + 1.7 * ceiling(ncol(odd$L) / 2)}
plot_loadings(odd$L, hm, odd$tab)
```

#### Even chromosomes

```{r split-even-plot, fig.width=10, fig.height=2 + 1.7 * ceiling(ncol(even$L) / 2)}
plot_loadings(even$L, hm, even$tab)
```

### {-}

```{r split-heat, fig.width=9, fig.height=7}
plot_pop_heatmap(odd$L, hm) + labs(title = "odd chromosomes")
plot_pop_heatmap(even$L, hm) + labs(title = "even chromosomes")
```

## Structure plots

Following [`fastica_1kg_structure`](fastica_1kg_structure.html), we turn the
maxima into structure plots. A structure plot needs non-negative memberships,
which signed ICA sources are not; but the $x|x|$ sources are one-sided, so
their positive parts may serve directly. Take
$$
A = \max(L, 0), \qquad X \approx AB + E \ \text{fitted by least squares},
$$
and *if* $B$ comes out non-negative, rescale each row of $B$ to sum to 1 and
scale the corresponding column of $A$ to compensate. The rows of $A$ are then
memberships on a common scale. There is **no intercept column** and $X$ is
left uncentered: with an intercept the rows of $B$ become increments to allele
frequency, which are necessarily signed. This is done in
[`code/fit_hgdp_structure.R`](https://github.com/stephenslab/misc/blob/master/code/fit_hgdp_structure.R).

One thing differs from the 1000 Genomes version: the rows here are
haplotypes, so a membership is a membership of a single chromosome, and an
individual appears as two adjacent bars that need not agree.

### Does $B \ge 0$ hold?

```{r struct-diag}
do.call(rbind, lapply(names(struct), function(nm) {
  z <- struct[[nm]]
  data.frame(sources = nm, K = z$K,
             `n B < -0.01` = z$n_lt, of = z$K * z$n_snps,
             frac = signif(z$frac_lt, 2),
             min_B = round(z$min_B, 3), `mean |B|` = round(z$mean_abs, 3),
             R2 = round(z$r2, 3), check.names = FALSE, row.names = NULL)
}))
```

The same split as in the 1000 Genomes data, and sharper. For the **rank-1
maxima** at both $K$, $B$ is non-negative to within rounding: 46 and 47
entries out of 5.4 million below $-0.01$, with minima of about $-0.014$
against a mean $|B|$ of 0.05. Clipping those at zero changes nothing.

For both **rank-$r$ fits** it fails, and it gets worse as $K$ grows: 5.6% of
entries below $-0.01$ at $K = 40$ and 8.3% at $K = 60$, with minima of
$-0.18$ and $-0.25$.

The $R^2$ column is the other half of the same point. It is the fit of
$X \approx \max(L,0) B$, so it measures how much of the data survives taking
positive parts. The 25 maxima reach 0.585, *better than the 60 rank-$r$
factors at 0.569* despite using less than half as many components, because
the maxima are one-sided and lose almost nothing to $\max(L, 0)$ while the
rank-$r$ factors are two-sided and lose a great deal.

The structure plots below therefore use maxima only.

### Building the memberships

```{r struct-build}
src <- list(
  `K = 40 maxima` = list(L = fit$fits$k40$maxima$L, key = "maxima_k40"),
  `K = 60 maxima` = list(L = fit$fits$k60$maxima$L, key = "maxima_k60"))

# A = pmax(L, 0) with columns rescaled so the rows of B sum to 1, factors in
# the same geographic order as everywhere else and labelled by the populations
# they carry.
memberships <- function(name) {
  L <- src[[name]]$L
  r <- struct[[src[[name]]$key]]$r
  o <- pop_order_of(L, hm)
  A <- sweep(pmax(L[, o, drop = FALSE], 0), 2, r[o], "*")
  colnames(A) <- make.unique(apply(L[, o, drop = FALSE], 2, pop_label,
                                   m = hm, top = 2))
  A <- A / mean(rowSums(A))     # bar heights on a scale of about 1

  # A haplotype with no positive loading on any source has no membership
  # representation at all; drop those and record how many.
  zero <- rowSums(A) == 0
  list(A = A[!zero, , drop = FALSE], pop = droplevels(hm$pop[!zero]),
       n_dropped = sum(zero))
}

plot_structure <- function(name, normalize = FALSE) {
  m <- memberships(name)
  A <- if (normalize) m$A / rowSums(m$A) else m$A
  set.seed(1)   # structure_plot orders within each group by t-SNE
  structure_plot(A, grouping = m$pop, gap = 8, verbose = FALSE,
                 colors = Polychrome::glasbey.colors(ncol(A) + 1)[-1]) +
    labs(y = "membership", fill = NULL,
         title = sprintf("%s (%s)%s", name,
                         if (normalize) "normalized" else "unnormalized",
                         if (m$n_dropped)
                           sprintf(" - %d haplotype(s) dropped", m$n_dropped)
                         else "")) +
    theme(axis.text.x     = element_text(angle = 90, vjust = 0.5, hjust = 1,
                                         size = 6),
          legend.text     = element_text(size = 6),
          legend.key.size = unit(0.32, "cm"))
}
```

The normalized plots show composition; the unnormalized ones keep the bar
heights, which carry information the normalized version discards --- a short
bar is a haplotype poorly explained by all the factors in the set.

### $K = 40$ maxima {.tabset}

#### Normalized

```{r struct-40-norm, fig.width=11, fig.height=5}
plot_structure("K = 40 maxima", normalize = TRUE)
```

#### Unnormalized

```{r struct-40-raw, fig.width=11, fig.height=5}
plot_structure("K = 40 maxima")
```

### {-}

### $K = 60$ maxima {.tabset}

The maxima are the same, so these should be --- and are --- the same plots.
They are shown for completeness.

#### Normalized

```{r struct-60-norm, fig.width=11, fig.height=5}
plot_structure("K = 60 maxima", normalize = TRUE)
```

#### Unnormalized

```{r struct-60-raw, fig.width=11, fig.height=5}
plot_structure("K = 60 maxima")
```

### {-}

### What the memberships say

```{r struct-pop-means}
mp <- memberships("K = 40 maxima")
P  <- mp$A / rowSums(mp$A)
M  <- apply(P, 2, function(x) tapply(x, mp$pop, mean))
# For each population, the factors making up at least 10% of it on average.
data.frame(
  pop = rownames(M),
  composition = apply(M, 1, function(x) {
    x <- sort(x[x >= 0.1], decreasing = TRUE)
    paste(sprintf("%s %d%%", names(x), round(100 * x)), collapse = ", ")
  }), row.names = NULL)
```

Most populations come out close to a hard assignment --- San 98%, PapuanHighlands
97%, Mbuti and Japanese and Colombian and Lahu 96%, Biaka and Bougainville and
PapuanSepik 95% --- and where there is mixture it is where geography says it
should be:

* **BantuSouthAfrica is 57% the West/Central African factor and 24% San**,
  the Khoe-San ancestry in southern African Bantu speakers.
* **Europe shows a south-north gradient** rather than discrete clusters:
  Sardinian 87% and Basque 83% on their own factors, Tuscan 35%
  Russian/Orcadian with 26% Sardinian and 17% Basque, BergamoItalian 39/31/15,
  French 58/18/14, Russian 86%.
* **A Central/South Asian cline** runs Brahui 86% and Makrani 81% on
  Brahui/Makrani, through Sindhi at 45/40 and Pathan at 49/29, to Burusho 89%
  on Burusho/Pathan.
* **East Asia is a Lahu/Dai--Japanese/Hezhen gradient**, with Han split
  47/47, Yi 47/43, Naxi 48/41, Tujia 51/44, and the Siberian populations
  pulled onto Yakut/Oroqen.
* **Uygur is the most mixed population in the panel**, with no factor of its
  own: 29% Japanese/Hezhen, 15% Yakut/Oroqen, 14% Burusho/Pathan and 13%
  Russian/Orcadian --- the Silk Road signal.
* **The Americas are nearly pure**, 86--96% on their own factors.

These are the expected admixture patterns, read off ICA sources rather than
fitted under a non-negativity constraint.

Not pruning has a visible cost here, and it is worth naming. Two of the 25
factors are the within-population clusters from the pruning check, and they
absorb membership that would otherwise go to a population factor: the Bedouin
are split 46% Bedouin and 22% `Bedouin.1`, and BantuKenya picks up 15% on the
Hazara cluster factor. Hazara itself now has 45% on a factor of its own,
where in the pruned fit it was a genuinely admixed population with none. So
the unpruned memberships are readable, but the two nuisance factors have to
be recognized as such rather than read as ancestry.

The unnormalized plots add something the normalized ones cannot show. Bar
heights run from about 0.4 to 1.6 around a median near 1, and the low end is
systematic rather than noise: the populations with the shortest bars are the
ones that sit *between* factors rather than owning one.

```{r struct-heights-bypop}
mk <- memberships("K = 40 maxima")
mh <- sort(tapply(rowSums(mk$A), mk$pop, mean))
round(head(mh, 6), 2)          # shortest bars
round(rev(tail(mh, 6)), 2)     # tallest
round(quantile(rowSums(mk$A), c(0, 0.01, 0.5, 0.99, 1)), 2)
```

The shortest bars are BantuKenya and BantuSouthAfrica, the two admixed
African populations, then Uygur and Adygei, the two admixed populations
between West Eurasia and East Asia, then Yoruba and Tu. The tallest are Lahu,
Japanese, Mandenka, Russian, Colombian and Burusho --- populations that
essentially own a factor each. So the bar height is reading as "how well any
one source explains this haplotype", which is the diagnostic it is supposed
to provide and which the normalized plot discards. Unlike the 1000 Genomes
case no single bar dominates: the 99th percentile is 1.57 against a maximum
of 1.84.

## Structure in the residuals

The structure fit gives a residual, $E = X - AB$, and the obvious question is
whether the $x|x|$ search finds anything in it. So we run exactly the same
procedure on $E$ that the main analysis ran on $X$: centre, whiten to 40
dimensions, 1000 random rank-1 starts, cluster.
[`code/fit_hgdp_residual.R`](https://github.com/stephenslab/misc/blob/master/code/fit_hgdp_residual.R)
does this, and never forms $E$ --- with $M = XB'$ its Gram matrix is
$$
EE' = XX' - MA' - AM' + A(BB')A',
$$
and the centred version follows from that and $\mu_E$ as usual.

```{r res-load}
res <- readRDS("../output/hgdp_residual.rds")
c(sources_removed = res$n_sources,
  residual_ss_fraction = round(res$ss_ratio, 3),
  centred_var_fraction = round(sum(res$d^2) / sum(fit$d^2), 3))
```

The 25 non-negative sources take out 58% of the raw sum of squares and 89% of
the *centred* variance. What is left is not flat, though:

```{r res-scree, fig.width=6, fig.height=3.2}
ggplot(data.frame(k = 1:40, d = res$d[1:40]), aes(k, d)) +
  geom_point(size = 1.2) +
  labs(x = "component", y = "singular value of the centred residual") +
  theme_cowplot(font_size = 10)
round(res$d[1:12])
round(res$d[1]^2 / sum(res$d^2), 3)   # share of residual variance in PC1
```

One direction dominates completely --- 57% of the remaining centred variance
--- then two more stand a little above a flat plateau from component 4 on,
which is what noise looks like.

### What the 1000 rank-1 runs find

```{r res-maxima}
hm_res <- hap_meta(res$ids)
c(n_maxima = ncol(res$L), largest_basin = max(res$size))
res$n_cut   # cutting at correlation 0.90 / 0.95 / 0.99
o_r <- pop_order_of(res$L, hm_res)
L_res <- res$L[, o_r, drop = FALSE]
colnames(L_res) <- sprintf("r%02d", seq_len(ncol(L_res)))
res_tab <- data.frame(factor = colnames(L_res),
                      pops = apply(L_res, 2, pop_label, m = hm_res),
                      obj = round(skew_obj(L_res), 2),
                      n_above2 = colSums(L_res > 2),
                      max_loading = round(apply(L_res, 2, max), 1),
                      basin = res$size[o_r], row.names = NULL)
res_tab
```

Four maxima against 25 on the data itself, and 928 of the 1000 starts go to
one of them. They are also genuinely new directions rather than leftovers of
the sources we removed:

```{r res-vs-original}
round(apply(cor(res$L, fit$fits$k40$maxima$L), 1, function(z) max(abs(z))), 2)
```

```{r res-plot, fig.width=10, fig.height=2 + 1.7 * ceiling(ncol(L_res) / 2)}
plot_loadings(L_res, hm_res, res_tab)
```

### The dominant residual direction is model misfit, not new structure

The leading residual maximum loads on BantuKenya at a mean of 6.8 and
BantuSouthAfrica at 2.3, with essentially nothing anywhere else --- the two
admixed African populations. That is a strong hint, and the residual PC1
makes it exact. Compare it with the structure-plot bar heights, which measure
how well the non-negative model explains each haplotype:

```{r res-pc1-vs-height, fig.width=4.6, fig.height=3.6}
h   <- rowSums(mk$A)                 # bar height per haplotype, from above
pc1 <- res$U_c[, 1]
if (cor(pc1, h) > 0) pc1 <- -pc1     # orient so positive = poorly explained
round(cor(pc1, h), 3)
ggplot(data.frame(pc1 = pc1, h = h, region = hm_res$region),
       aes(pc1, h, colour = region)) +
  geom_point(size = 0.7, alpha = 0.8) +
  scale_colour_manual(values = region_palette) +
  labs(x = "residual PC1 (oriented)", y = "structure bar height") +
  theme_cowplot(font_size = 9) + theme(legend.position = "none")
```

The correlation is $-0.983$ at the haplotype level and $-0.967$ at the
population level. **The dominant direction in the residual is not population
structure the sources missed; it is the sources failing on the admixed
individuals.** A model of 25 one-sided, non-negative sources represents a
haplotype well when one source owns it and badly when it is a mixture, and
the residual's first component is precisely that failure, sorted.

That also explains why the rank-1 search finds so little. The residual is
89% smaller than the data in centred variance, most of what remains is the
misfit axis rather than a source, and the four maxima it does find are all
low-objective --- 0.40 to 0.75 against 0.58 to 0.96 for the real ones. None
of them is a new population: `r01` is the admixed African pair, `r02` a
cluster inside Druze, `r04` a handful of Makrani, and `r03` a broad, shallow
West Eurasian gradient over Bedouin, Adygei and Tuscan with a maximum
loading under 8. There is no hidden 26th population factor here.

It is worth being clear about what this does and does not test. It says the
*non-negative* representation leaves nothing source-like behind, which is a
statement about $A = \max(L, 0)$ and $B$, not about the ICA fit itself: the
residual is defined by discarding the negative parts of the sources, so some
of what it contains is the information those negative parts carried.

## A worked example: Mandenka and Yoruba

One factor covers both Mandenka and Yoruba, but it does not cover them
equally, and following that through explains both the bar-height diagnostic
and the residual result better than either does on its own.

```{r wk-setup}
nc  <- c("Mandenka", "Yoruba", "BantuKenya", "BantuSouthAfrica")
j1  <- which.max(colMeans(pmax(L_r1, 0)[hm$pop == "Mandenka", , drop = FALSE]))
round(rbind(`loading on the shared factor` = tapply(L_r1[, j1], hm$pop, mean)[nc],
            `structure bar height`         = tapply(rowSums(mk$A), mk$pop, mean)[nc]), 2)
```

Mandenka load at 5.5 and Yoruba at 2.6 --- a factor of 2.1 --- and the bar
heights follow. Note the ordering continues into the two Bantu populations.

### It is not the obvious things

```{r wk-nulls}
rbind(`drift ||f_pop - mu||^2`    = round(res$pop_dev[nc]),
      `heterozygosity x 1000`     = round(1000 * res$pop_het[nc], 1),
      `centred haplotype norm^2`  = round(tapply(res$cx2, hm$pop, mean)[nc]),
      `sample size (individuals)` = table(meta$pop)[nc])
```

Mandenka and Yoruba are the same size (22 individuals each), equally
heterozygous, and equally far from the global mean --- so this is not "Yoruba
are less differentiated". Nor is the missing loading picked up elsewhere:

```{r wk-elsewhere}
round(colMeans(mk$A[mk$pop == "Yoruba", -j1, drop = FALSE]), 3) -> other
round(c(`Yoruba on the shared factor` = mean(mk$A[mk$pop == "Yoruba", j1]),
        `Yoruba on all other factors` = sum(other)), 3)
```

### The whole gap is one component

Write the factor as a unit direction $w$ in the whitened space and split the
population-mean loading over the 40 whitened PCs. The loading of a population
is $\sum_k \bar z_k w_k$, so each PC makes a separate contribution.

```{r wk-decompose}
Z  <- sqrt(length(fit$hap_ids)) * fit$U_c[, 1:fit$K[1], drop = FALSE]
U_w <- t(Z)
w  <- as.vector(U_w %*% L_r1[, j1]) / length(fit$hap_ids)
zm <- colMeans(Z[hm$pop == "Mandenka", , drop = FALSE])
zy <- colMeans(Z[hm$pop == "Yoruba",   , drop = FALSE])
dec <- data.frame(PC = seq_along(w), w = round(w, 3),
                  Mandenka = round(zm * w, 3), Yoruba = round(zy * w, 3),
                  diff = round((zm - zy) * w, 3))
head(dec[order(-abs(dec$diff)), ], 5)
c(total_gap = round(sum((zm - zy) * w), 2),
  from_PC1_3 = round(sum(((zm - zy) * w)[1:3]), 2),
  from_PC22  = round(((zm - zy) * w)[22], 2))
```

**PC22 accounts for 2.7 of the 2.9 gap; PC1--3 account for none of it.** On
the axes that carry most of the variance the two populations are
indistinguishable. And PC22 is not a general axis --- it is a within-Africa
one, with only the four Niger-Congo populations away from zero:

```{r wk-pc22}
m22 <- tapply(Z[, 22], hm$pop, mean)
round(m22[abs(m22) > 1], 2)                    # every other population
round(range(m22[abs(m22) <= 1]), 2)            # ... lies in this range
c(d1 = round(fit$d[1]), d6 = round(fit$d[6]), d22 = round(fit$d[22]))
```

It runs Mandenka at one end and BantuKenya at the other with Yoruba in
between, which is a Mande-versus-Bantu contrast. It is also uniform within
Mandenka rather than driven by a few individuals:

```{r wk-pc22-within}
z <- Z[hm$pop == "Mandenka", 22]
round(c(min = min(z), mean = mean(z), max = max(z), sd = sd(z)), 2)
```

### Why the contrast uses it

Two things combine. Whitening puts PC22 ($d = 213$) on the same footing as
PC1 ($d = 1384$), so a modest real axis becomes cheap to exploit. And $x|x|$
rewards concentration rather than evenness, so the optimizer is happy to
spend weight there: $w$ puts 0.57 on PC6, the shared West African axis, and
0.43 on PC22.

Comparing the fitted direction with two alternatives makes the trade explicit
--- one with PC22 removed, and one pointing at the four Niger-Congo
populations equally:

```{r wk-alternatives}
obj_of <- function(v) { v <- v / sqrt(sum(v^2)); q <- as.vector(Z %*% v)
                        mean(q * abs(q)) }
load_of <- function(v) { v <- v / sqrt(sum(v^2)); q <- as.vector(Z %*% v)
                         round(tapply(q, hm$pop, mean)[nc], 2) }
nsharp <- function(v) { v <- v / sqrt(sum(v^2)); sum(as.vector(Z %*% v) > 2) }

cand <- list(`fitted factor` = w,
             `PC22 removed`  = replace(w, 22, 0),
             `even West African` = colMeans(Z[hm$pop %in% nc, , drop = FALSE]))
data.frame(objective   = round(sapply(cand, obj_of), 3),
           n_above2    = sapply(cand, nsharp),
           t(sapply(cand, load_of)))
```

An even West African direction exists and is a perfectly reasonable ancestry
axis, but it is not a local maximum: tilting onto PC22 turns a broad plateau
over 130 haplotypes into a spike over 89 and raises the objective from 0.835
to 0.849. **The contrast prefers an anchored source to a fair one.**

### A prediction, and a check

If that is the mechanism then PC22 has to be available, so below $K = 22$ the
factor should even out. Re-running the rank-1 search across a range of $K$
(300 starts each, which finds fewer maxima than the 1000 used elsewhere but
is ample for this):

```{r wk-ksweep}
r1_update <- function(U, W) {
  P <- t(U) %*% W
  W <- U %*% (2 * abs(P)) - sweep(W, 2, colSums(2 * sign(P)), "*")
  sweep(W, 2, sqrt(colSums(W^2)) + 1e-15, "/")
}
do.call(rbind, lapply(c(10, 15, 20, 25, 30, 40), function(k) {
  U <- sqrt(nrow(fit$U_c)) * t(fit$U_c[, 1:k, drop = FALSE])
  set.seed(1)
  Wk <- matrix(rnorm(k * 300), k, 300)
  Wk <- sweep(Wk, 2, sqrt(colSums(Wk^2)), "/")
  for (i in 1:500) Wk <- r1_update(U, Wk)
  L  <- t(U) %*% Wk
  M  <- L[, match(unique(cutree(hclust(as.dist(1 - cor(L)), method = "complete"),
                                h = 0.05)),
                  cutree(hclust(as.dist(1 - cor(L)), method = "complete"),
                         h = 0.05)), drop = FALSE]
  v  <- tapply(M[, which.max(colMeans(pmax(M, 0)[hm$pop == "Mandenka", ,
                                                 drop = FALSE]))], hm$pop, mean)
  data.frame(K = k, Mandenka = round(v["Mandenka"], 2),
             Yoruba = round(v["Yoruba"], 2),
             ratio = round(v["Mandenka"] / v["Yoruba"], 2), row.names = NULL)
}))
```

The ratio is 1.0 up to $K = 20$ and 2.1 from $K = 25$, switching exactly where
PC22 enters the whitened space.

### Why not a Yoruba factor as well?

If PC6 separates the Niger-Congo populations from everyone else and PC22
separates Mandenka from Yoruba, the plane they span ought to contain a
Mandenka-anchored direction *and* a Yoruba-anchored one, and the natural
expectation is two factors rather than Yoruba sitting part-way along the
Mandenka axis.

The first thing to check is whether PC22 is the only axis separating them.
It is not:

```{r wk-otherpcs}
Zall <- sqrt(nrow(fit$U_c)) * fit$U_c
iM <- hm$pop == "Mandenka"; iY <- hm$pop == "Yoruba"
sep <- data.frame(PC = seq_len(ncol(Zall)), d = round(fit$d[seq_len(ncol(Zall))]),
                  Mandenka = round(colMeans(Zall[iM, ]), 2),
                  Yoruba   = round(colMeans(Zall[iY, ]), 2))
sep$diff <- round(sep$Mandenka - sep$Yoruba, 2)
head(sep[order(-abs(sep$diff)), ], 4)
```

**PC31 separates them too**, in the opposite sense: Yoruba $+2.47$ against
Mandenka $-1.02$. And it is a population-level axis, not a few outliers ---
within Yoruba it runs from 0.7 to 4.3 with sd 0.70. So Yoruba is genuinely
distinguishable from Mandenka in the whitened space, and a Yoruba-anchored
direction exists and looks reasonable:

```{r wk-yoruba-dir}
nrm  <- function(v) v / sqrt(sum(v^2))
objv <- function(v) { q <- as.vector(Z %*% nrm(v)); mean(q * abs(q)) }
Zpop <- t(sapply(levels(hm$pop), function(pp) colMeans(Z[hm$pop == pp, , drop = FALSE])))
v_Y  <- Zpop["Yoruba", ] - colMeans(Zpop[rownames(Zpop) != "Yoruba", , drop = FALSE])
round(sort(tapply(as.vector(Z %*% nrm(v_Y)), hm$pop, mean), decreasing = TRUE)[1:4], 2)
round(objv(v_Y), 3)
```

Plotting the three PCs pairwise makes the situation concrete. Points are
haplotypes; filled circles are population centroids; the arrows are the
fitted factor and the best Yoruba direction, projected onto each plane.

```{r wk-pcpairs, fig.width=12, fig.height=4.8}
w_fit <- nrm(as.vector(t(Z) %*% L_r1[, j1]) / nrow(Z))   # the fitted direction
v_Y   <- nrm(v_Y)

pc_pair_plot <- function(a, b) {
  nc4  <- c("Mandenka", "Yoruba", "BantuSouthAfrica", "BantuKenya")
  oaf  <- c("Biaka", "Mbuti", "San")
  grp  <- factor(ifelse(as.character(hm$pop) %in% nc4, as.character(hm$pop),
                 ifelse(as.character(hm$pop) %in% oaf, "other African",
                        "rest of world")),
                 levels = c(nc4, "other African", "rest of world"))
  pal <- c(Mandenka = "#D55E00", Yoruba = "#0072B2",
           BantuSouthAfrica = "#009E73", BantuKenya = "#CC79A7",
           `other African` = "#E69F00", `rest of world` = "grey80")
  df  <- data.frame(x = Z[, a], y = Z[, b], grp = grp)
  df  <- df[order(df$grp == "rest of world", decreasing = TRUE), ]
  cen <- data.frame(x = Zpop[c(nc4, oaf), a], y = Zpop[c(nc4, oaf), b],
                    grp = factor(c(nc4, rep("other African", 3)),
                                 levels = levels(grp)))
  sc  <- 0.8 * max(abs(c(Z[, a], Z[, b])))
  arr <- data.frame(xe = c(w_fit[a], v_Y[a]) * sc,
                    ye = c(w_fit[b], v_Y[b]) * sc,
                    lab = c("fitted factor", "best Yoruba direction"))
  ggplot(df, aes(x, y)) +
    geom_hline(yintercept = 0, colour = "grey70", linewidth = 0.2) +
    geom_vline(xintercept = 0, colour = "grey70", linewidth = 0.2) +
    geom_point(aes(colour = grp), size = 0.7, alpha = 0.75) +
    geom_point(data = cen, aes(fill = grp), shape = 21, size = 3,
               colour = "black", stroke = 0.4) +
    geom_segment(data = arr, aes(x = 0, y = 0, xend = xe, yend = ye,
                                 linetype = lab),
                 arrow = grid::arrow(length = grid::unit(0.16, "cm")),
                 linewidth = 0.5) +
    scale_colour_manual(values = pal, name = NULL) +
    scale_fill_manual(values = pal, guide = "none") +
    scale_linetype_manual(values = c(`fitted factor` = "solid",
                                     `best Yoruba direction` = "22"),
                          name = NULL) +
    labs(x = paste0("PC", a), y = paste0("PC", b)) +
    theme_cowplot(font_size = 9)
}
ps   <- lapply(list(c(6, 22), c(6, 31), c(22, 31)),
               function(ab) pc_pair_plot(ab[1], ab[2]))
leg  <- get_legend(ps[[1]] + theme(legend.position = "bottom") +
                   guides(colour = guide_legend(override.aes = list(size = 2.5),
                                                nrow = 1)))
plot_grid(plot_grid(plotlist = lapply(ps, function(q)
                      q + theme(legend.position = "none")), nrow = 1),
          leg, ncol = 1, rel_heights = c(1, 0.22))
```

Three things are worth reading off this. PC6 is not "West African versus the
world" --- it separates the four Niger-Congo populations (positive) from
Biaka, Mbuti and San (strongly negative), with the rest of the world at
zero. In the PC6 planes the four Niger-Congo groups are almost on top of one
another, which is the sense in which they share an ancestry component. And
in the **PC22--PC31 plane on the right they separate cleanly into three
directions**: Mandenka to the right, Yoruba up and to the left, BantuKenya
down and to the left, with the rest of the world as a tight cloud at the
origin. Yoruba is a corner of that cloud in its own right, not a point on a
line between the other two --- which is exactly why one might expect it to
get its own factor.

Yoruba is cleanly the top population on the best Yoruba direction. But its
objective is 0.698 against 0.849 for the fitted factor, and it is **not a
local maximum**.
Running rank-1 from five Yoruba-anchored starts, including ones on which
Mandenka loads strongly negative, every one climbs to the same
Mandenka-anchored optimum:

```{r wk-yoruba-starts}
zy <- Zpop["Yoruba", ]; zm <- Zpop["Mandenka", ]; zbk <- Zpop["BantuKenya", ]
starts <- list(`Yoruba centroid`          = zy,
               `orthogonal to the factor` = zy - sum(zy * w_fit) * w_fit,
               `Yoruba minus Mandenka`    = zy - zm,
               `Yoruba minus mean(M, BK)` = zy - 0.5 * (zm + zbk),
               `max Yoruba margin`        = v_Y)
do.call(rbind, lapply(names(starts), function(nm) {
  v <- nrm(starts[[nm]]); W1 <- matrix(v, ncol = 1)
  l0 <- tapply(as.vector(Z %*% v), hm$pop, mean)
  for (i in 1:500) W1 <- r1_update(t(Z), W1)
  v1 <- as.vector(W1); l1 <- tapply(as.vector(Z %*% v1), hm$pop, mean)
  data.frame(start = nm, obj_start = round(objv(v), 3),
             Mandenka_start = round(l0["Mandenka"], 2),
             Yoruba_start   = round(l0["Yoruba"], 2),
             obj_end = round(objv(v1), 3),
             Mandenka_end = round(l1["Mandenka"], 2),
             Yoruba_end   = round(l1["Yoruba"], 2), row.names = NULL)
}))
```

Nor is it hiding behind the Mandenka factor. Deflating that direction out and
running the full 1000-start search in the 39-dimensional complement produces
no Yoruba maximum either --- the only Niger-Congo maximum that appears is
BantuKenya/BantuSouthAfrica, and with a basin of one:

```{r wk-deflate}
Q  <- qr.Q(qr(cbind(w_fit, diag(ncol(Z)))))[, 2:ncol(Z)]
Zd <- Z %*% Q
set.seed(1)
Wd <- matrix(rnorm((ncol(Z) - 1) * 1000), ncol(Z) - 1, 1000)
Wd <- sweep(Wd, 2, sqrt(colSums(Wd^2)), "/")
for (i in 1:500) Wd <- r1_update(t(Zd), Wd)
Ld <- Zd %*% Wd
cld <- cutree(hclust(as.dist(1 - cor(Ld)), method = "complete"), h = 0.05)
jd  <- match(unique(cld), cld); Md <- Ld[, jd, drop = FALSE]
labd <- apply(Md, 2, pop_label, m = hm)
afr <- grep("Mandenka|Yoruba|Bantu", labd)
data.frame(factor = labd[afr],
           basin  = as.integer(table(cld)[as.character(cld[jd])])[afr],
           t(round(apply(Md[, afr, drop = FALSE], 2,
                         function(x) tapply(x, hm$pop, mean)[nc]), 2)),
           row.names = NULL)
```

So the absence of a Yoruba factor is not a search failure and not a
consequence of Yoruba being undifferentiated. It is that on *both* of the
minor African axes, the largest excursion belongs to someone else:

```{r wk-excursions}
round(Zpop[nc, c(6, 22, 31)], 2)
```

Mandenka owns the $+$ end of PC22 at $4.39$; BantuKenya owns the $-$ end of
PC31 at $-3.21$. Yoruba's own excursion, $+2.47$ on PC31, is the smaller
one, and because all four populations share the same large PC1/PC2/PC6
African component, a direction built on it cannot avoid bringing
BantuSouthAfrica (3.03) up with it. The result is a shorter, broader tail ---
0.698 against 0.849 --- and gradient ascent walks away from it. Under $x|x|$
the biggest excursion in a neighbourhood takes the factor, and the
second-biggest gets represented as a fraction of it.

### Every population as a starting point

That suggests a general check: start rank-1 at each of the 54 population
centroids in turn and see where it goes. This is the most targeted
initialization available --- if a population has a factor of its own, this
should find it.

```{r wk-all-centroids}
W0 <- sweep(t(Zpop), 2, sqrt(colSums(t(Zpop)^2)), "/")
obj0 <- apply(W0, 2, objv)
Wc <- W0
for (i in 1:500) Wc <- r1_update(t(Z), Wc)
Cc  <- cor(Z %*% Wc, L_r1)
land <- colnames(L_r1)[apply(Cc, 1, which.max)]
lab  <- r1$tab$pops[apply(Cc, 1, which.max)]
own  <- mapply(function(p, l) grepl(p, l, fixed = TRUE), levels(hm$pop), lab)
c(`lands on a maximum containing itself` = sum(own),
  `slides to a neighbour`                = sum(!own),
  `distinct endpoints`                   = length(unique(land)),
  `endpoints that are new maxima`        = sum(apply(Cc, 1, max) < 0.9))
```

Thirty-two of the 54 populations reach a maximum that includes them; the
other 22 slide onto a neighbour. Crucially **none of the 54 finds anything
new** --- every endpoint is one of the 25 maxima already known, at $r = 1.00$.
The populations that slide:

```{r wk-sliders}
data.frame(population = levels(hm$pop)[!own],
           obj_at_its_own_centroid = round(obj0[!own], 3),
           lands_on = lab[!own],
           obj_there = round(apply(Wc, 2, objv)[!own], 3),
           row.names = NULL)
```

They are exactly the populations that sit inside the cloud rather than at a
corner of it: the Han-adjacent East Asian groups collapsing onto
Lahu/Dai or Japanese/Hezhen, the Italian and French groups onto Sardinian or
Basque, Sindhi and Balochi onto their South Asian neighbours, Uygur onto
Yakut/Oroqen, and BantuKenya onto Mandenka/Yoruba. In every case the
objective rises along the way, so these are not near-misses --- the centroid
of such a population is not close to being a maximum at all.

### What it explains

Because the sources are one-sided, only the positive end of PC22 gets a
factor. The negative end --- Yoruba, then BantuSouthAfrica, then BantuKenya
--- gets nothing, so that part of their variation is simply not modelled.
Both of the earlier diagnostics are reading this:

* the **bar heights** order exactly along PC22 (Mandenka 1.23, Yoruba 0.64,
  BantuSouthAfrica 0.54, BantuKenya 0.40), because a short bar is a
  haplotype no source covers;
* the **largest residual maximum** is BantuKenya at 6.8 and
  BantuSouthAfrica at 2.3 --- the unmodelled negative end of PC22 coming back
  as the single biggest thing left in the data.

So neither is a data problem. They are the same fact seen twice: the $x|x|$
contrast traded an interpretable ancestry axis for a sharper one-sided
source, on a minor axis that whitening made cheap. It is worth knowing that
this is a choice the method makes rather than a statement about African
population structure, and that lowering $K$ --- or presumably using a
symmetric contrast --- reverses it.

The wider lesson is about what "25 maxima" means. The maxima are the largest
excursions of the whitened cloud, not a partition of the populations. A
population can be perfectly well separated from its neighbours --- Yoruba is,
on PC31 --- and still get no factor, because a nearby population has a bigger
excursion and the search converges there from every direction. So the count
should not be read as "25 ancestries": at $K = 40$ the data supports 25
one-sided directions, and a population is represented in its own right only
if it wins its local neighbourhood, not merely if it is distinguishable.

## Does the SNP density matter?

The 6 kb spacing was chosen to put the SNP count in the same range as the
185k of the 1000 Genomes matrix, not from any argument about LD, so it is
worth knowing whether it matters. The whole pipeline was also run at 12 kb,
which gives 140,628 SNPs against 215,468 --- a factor of 1.53 --- and
`output/hgdp_fastica_12kb.rds` holds the maxima from that run.

```{r density-load}
alt <- readRDS("../output/hgdp_fastica_12kb.rds")
hm_alt <- hap_meta(alt$hap_ids)
c(`6kb` = fit$nsnp, `12kb` = alt$nsnp)
```

The eigenvalues agree to three decimals once scaled by $\sqrt{p}$, so the
extra SNPs are not changing the subspace fastICA works in:

```{r density-scree}
round(rbind(`6kb`  = fit$d[1:12] / sqrt(fit$nsnp),
            `12kb` = alt$d[1:12] / sqrt(alt$nsnp)), 3)
```

Matching the maxima across the two densities:

```{r density-match}
alt_o <- pop_order_of(alt$maxima$L, hm_alt)
L_alt <- alt$maxima$L[, alt_o, drop = FALSE]
Cd    <- cor(L_r1, L_alt)
data.frame(`6kb`      = r1$tab$pops,
           starts_6   = r1$tab$rand_starts,
           `12kb`     = apply(L_alt, 2, pop_label, m = hm_alt)[apply(Cd, 1, which.max)],
           starts_12  = alt$maxima$n_random[alt_o][apply(Cd, 1, which.max)],
           r          = round(apply(Cd, 1, max), 3),
           check.names = FALSE, row.names = NULL)
c(n_6kb = ncol(L_r1), n_12kb = ncol(L_alt),
  matched_r_gt_0.9 = sum(apply(Cd, 2, max) > 0.9),
  worst_matched_r  = round(min(apply(Cd, 2, max)), 3))
```

Every one of the `r ncol(L_alt)` maxima found at 12 kb reappears at 6 kb, the
worst match being $r =$ `r round(min(apply(cor(L_r1, alt$maxima$L), 2, max)), 3)`,
and the basin sizes track closely:

```{r density-basins, fig.width=4.2, fig.height=3.6}
j  <- apply(Cd, 1, which.max)
ok <- apply(Cd, 1, max) > 0.9
db <- data.frame(six = r1$tab$rand_starts[ok],
                 twelve = alt$maxima$n_random[alt_o][j][ok])
ggplot(db, aes(twelve, six)) +
  geom_abline(slope = 1, linetype = "dashed", colour = "grey60") +
  geom_point(size = 1.4) +
  labs(x = "random starts, 12 kb", y = "random starts, 6 kb") +
  theme_cowplot(font_size = 10)
round(cor(db$twelve, db$six), 3)
```

The one difference is that 6 kb finds `r ncol(L_r1) - ncol(L_alt)` maximum
that 12 kb does not, and it is not new population structure: it is the Hazara
cluster, one of the within-population groups the pruning check identifies. At
12 kb that cluster is present too, but only as a rank-$r$ factor --- rank-1
started there converges to it, while no random start does. With more SNPs it
becomes a genuine rank-1 maximum with a basin of its own. The same marginal
wobble shows up in the pruning check: the two densities agree on 26 of the
individuals to drop, and disagree on a Palestinian cluster of four that is a
narrow rank-$r$ factor at 12 kb and not at 6 kb.

```{r density-pruning}
c(dropped_6kb  = length(pc$dropped), dropped_12kb = length(alt$pruned$dropped),
  in_both      = length(intersect(pc$dropped, alt$pruned$dropped)),
  only_6kb     = length(setdiff(pc$dropped, alt$pruned$dropped)),
  only_12kb    = length(setdiff(alt$pruned$dropped, pc$dropped)))
meta[match(setdiff(alt$pruned$dropped, pc$dropped), meta$sample), c("sample", "pop")]
```

So the SNP count is not what is limiting this analysis. Between 140k and 215k
SNPs the subspace, the maxima, their objectives and their basins are all
stable to two or three decimals, and the only things that move are the
marginal nuisance factors, which are recognizable in either case. That is
reassuring about the 6 kb choice but it is a narrow check --- it says the
answer is stable to *how many* SNPs are taken, not that distance thinning is
picking a good set.

## Comment

The recipe from the 1000 Genomes analysis transfers, but the thing it was
built to fix is largely absent here. There, dropping 40 of 2504 individuals
turned a landscape with 17 maxima, wildly uneven basins and three maxima that
random starts could never reach into one with 14 maxima, all population
factors. Here the landscape is *already* almost all population factors:
`r ncol(L_r1) - 2` of the `r ncol(L_r1)` maxima are populations or groups of
neighbouring populations with no pruning at all, and removing the 27
individuals behind the other two changes nothing else. So the analysis does
not prune.

Four checks come out the same way, which is the main thing to take from this.
The maxima do not depend on the whitening dimension (identical at $K = 40$
and $K = 60$, worst match $r = 0.94$), they do not depend on the SNP density
(every 12 kb maximum reappears at 6 kb, worst $r = 0.98$, basin sizes
correlated at 0.99), they do not depend on the 27 individuals the pruning
check removes, and 18 of the 25 are found independently by each half of the
chromosomes, matching each other at $r$ up to 0.99 with basin sizes again
correlated at 0.99. Whatever else they are, they are not artifacts of those
choices.

The stability split also sharpens what the seven non-replicating maxima are.
Both halves lose exactly the same seven, which is itself informative: Basque,
Sardinian and Russian/Orcadian/French merge into a broad European factor,
Japanese/Hezhen and Colombian/Maya are not resolved at all, and the Bedouin
and Hazara clusters --- the two the pruning check flags as within-population
--- are the least reproducible things in the analysis. The finer population
splits are real but need the full marker set; the two nuisance factors are
called out by the split and by the pruning check independently.

The rank-$r$ fits behave quite differently, and consistently worse. They are
the only route to `r sum(r1$tab$rand_starts == 0)` of the maxima --- so they
are still needed --- but as a factorization in their own right they fail on
both counts we can check: $B$ goes 5.6% negative at $K = 40$ and 8.3% at
$K = 60$, against essentially zero for the maxima, and the 25 maxima capture
more of the data after taking positive parts ($R^2 = 0.585$) than 60 rank-$r$
factors do ($0.569$). At $K = 60$ only 19 of the 60 factors correspond to a
maximum; the other 41 are diffuse, low-objective directions (median objective
0.27 against 0.82 for the 19) that exist because orthogonality requires 60
components to be produced. Raising $K$ from 40 to 60 buys a higher objective
and 20 more of those, and no new structure.

Running the same search on the residual of the non-negative fit closes this
off. The 25 sources take out 89% of the centred variance; of what is left,
57% sits in a single direction that correlates at $-0.98$ with the
structure-plot bar heights --- it is the model failing on admixed
individuals, not structure it missed --- and the 1000 rank-1 starts find only
four low-objective maxima, none of them a new population. That is a
statement about the non-negative representation rather than about the ICA
fit, since the residual is defined by throwing away the negative parts of
the sources, but within that scope nothing source-like is left behind.

One methodological point worth keeping even though the pruning was dropped.
The count-based sparsity rule from the TGP script measures the wrong thing
once populations can be smaller than the threshold: applied unchanged to HGDP
it deletes San, Colombian and Surui, three real populations with three of the
highest objectives in the fit. Nothing in the output would flag it --- the fit
converges just as cleanly and still reports a tidy list of population factors,
with three populations quietly missing. Testing sparsity relative to the size
of the population involved is a one-line change and fixes it.
