---
title: "FastICA for Asymmetric Binary Sources, part 2"
author: "Matthew Stephens (using Claude)"
date: "2026-07-30"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---


## Introduction

This is another try that I think can ultimately be ignored, as it was done before I fully understood the problem. Keeping it for completeness. Here is the prompt I started Claude with (generated in conversation with Gemini):

I want to follow up on fastica_asymmetric with a second analysis (fastica_asymmetric_02.Rmd) which takes a slightly different approach, based on what we learned. I have a writeup for you. can you implement the objective function and optimization. Please make the implementation modular, so separate out functions to update w by a newton step and optimize c by some appropriate methods. here is the writeup:

Generalizing FastICA for Asymmetric Binary Sources: A Generative Maximum Likelihood Approach

1. Introduction and Motivation
The standard FastICA algorithm, operating on whitened data $Y$ with the objective to maximize the non-Gaussianity of projections $x = Yw$, often relies on the $\log(\cosh(x))$ contrast function. This specific objective implicitly assumes a generative model where the source signal is a symmetric binary variable (Rademacher) corrupted by Gaussian noise, with the noise variance related to the signal standard deviation $c$ by $\sigma^2 = c$.
This report outlines an extension of this framework to recover asymmetric binary sources.
2. Generative Model and Constraints
Let the observed one-dimensional projection be $x = u + e$ where $u$ denotes the signal (source) and $e$ denotes the noise. Since the whitened data $x$ has variance 1, we assume that $u, e$ have variance $c^2, (1-c^2)$.
The Source ($u$): We assume $u$ is an asymmetric binary source taking values $y_1$ with probability $p$ and $y_0$ with probability $1-p$ and satisfying:

1. Zero Mean: $\mathbb{E}[u] = (1-p)y_0 + p y_1 = 0$
2. Variance $c^2$: $\text{Var}(u) = c^2$, where $0 < c < 1$.

Solving this system yields: 
$$y_1 = c \sqrt{\frac{1-p}{p}} \quad \text{and} \quad y_0 = -c \sqrt{\frac{p}{1-p}}$$
The Noise ($e$): 
$$e \sim \mathcal{N}(0, 1-c^2)$$
3. Deriving the Log-Likelihood Objective
The marginal probability of observing a projection $x_i$ is a Gaussian mixture: 
$$p(x_i) = (1-p) \mathcal{N}(x_i \mid y_0, 1-c^2) + p \mathcal{N}(x_i \mid y_1, 1-c^2)$$
Taking the log-likelihood over $n$ samples, expanding the squared terms, and applying the strict constraints of whitened data—specifically that $\sum x_i^2 = n$ (since $w^T w = 1$) and $\sum x_i = 0$ (since the data is centered)—allows us to drop several terms that become constants with respect to $w$.
The resulting average log-likelihood objective takes the form of a heavily shifted and scaled softplus function: 
$$\mathcal{J}(w, p, c) = K(p, c) + \frac{1}{n} \sum_{i=1}^n \log(1 + \exp(s_i))$$
Where the internal state is $s_i = \alpha x_i - \beta + \gamma$, parameterized by:

* Scale: $\alpha = \frac{c}{(1-c^2)\sqrt{p(1-p)}}$
* Shift: $\beta = \frac{c^2(1-2p)}{2(1-c^2)p(1-p)}$
* Log-odds: $\gamma = \log\left(\frac{p}{1-p}\right)$
* Variance Penalty: $K(p, c) = -\frac{1}{2}\log(2\pi(1-c^2)) - \frac{1-p+c^2 p}{2(1-c^2)(1-p)} + \log(1-p)$

4. Proposed Optimization Strategy
Directly optimizing $\mathcal{J}(w, p, c)$ jointly over all three parameters may not be necessary (or perhaps helpful). The crucial (high-dimensional) part is optimizing $w$, which can be done efficiently using a fastICA-like algorithm for fixed $p,c$. The optimization landscape over $w$ will be relatively gentle if $c$ is intermediate, but become less so as $c$ gets close to 1 (ideally, as we hone in on a source). For now we propose a simple nested, alternating optimization strategy:
Phase 1: Coarse Grid over $p$
We know that fastICA ($p=0.5$) is already quite good at finding binary sources. (It is ideally honed to find balanced sources, but somewhat fortuitously unbalanced sources are sometimes local minima of the log-likelihood, and so fastICA can also find those). Based on this it seems getting $p$ exactly right may not not strictly necessary and so we will start by using a coarse grid (e.g., $P = \{0.01, 0.05, 0.2, 0.5\}$). The following phases are run independently for each $p \in P$.
Phase 2: Fixed-Variance Initialization (The "Golden" Start)
To prevent early collapse toward $c \to 1$, we temporarily fix the variance tradeoff. By enforcing the standard FastICA assumption where the noise variance equals the signal standard deviation ($\sigma^2 = c$), and combining this with the total variance constraint ($\sigma^2 = 1 - c^2$), we get: 
$$1 - c^2 = c \implies c^2 + c - 1 = 0$$
The positive root of this equation yields a value that is the reciprocal of the golden ratio: 
$$c_{golden} = \frac{\sqrt{5} - 1}{2} \approx 0.618$$
Fixing $c = c_{golden}$, we optimize $w$ using a FastICA-style approximate Newton update. Let $g(x_i) = \alpha \sigma(s_i)$ and $g'(x_i) = \alpha^2 \sigma(s_i)(1 - \sigma(s_i))$, where $\sigma(\cdot)$ is the standard logistic sigmoid. The update step is: 
$$w^+ = \frac{1}{n} Y^T g(x) - \overline{g'} \cdot w$$$$w \leftarrow \frac{w^+}{\Vert{}w^+\Vert{}}$$
 We iterate this until $w$ converges within a standard tolerance.
Phase 3: Alternating Refinement of $c$ and $w$
Once the direction $w$ is roughly aligned with the source, we relax the variance constraint and alternate:

1. Update $c$: Fix $w$ and $p$. Maximize the 1D objective $\mathcal{J}(c)$ bounded on $c \in (0, 1)$ using a standard bounded scalar solver (e.g., Brent's method). The objective is $\mathcal{J}(c) = -\frac{1}{2}\log(1-c^2) - \frac{1}{2(1-p)(1-c^2)} + \frac{1}{n} \sum_{i=1}^n \text{softplus}\left( \alpha(c) x_i - \beta(c) + \gamma \right)$.
2. Update $w$: Recalculate $\alpha$ and $\beta$ with the new $c$, and run a few iterations of the Newton update from Phase 2. Repeat these two steps until joint convergence.

Phase 4: Final Selection
Evaluate the full log-likelihood $\mathcal{J}(w_p, p, c_p)$ for the converged candidates from each probability in the grid. Ultimately we might select the tuple that maximizes the global objective but for now we will just look at how this algorithm behaves.


## Claude's Introduction

The previous analysis ([fastica_asymmetric](fastica_asymmetric.html)) showed that
standard log-cosh fastICA fails for asymmetric binary sources, and that an
alternating algorithm that also optimises $p$ substantially helps.  However, that
approach used an ad-hoc M-estimator or golden-ratio anchor for the variance
parameter $c$ and optimised $p$ for fixed $c$.

Here we take a cleaner, fully generative approach.  For a projection
$x = Yw$ of whitened data, we posit:

$$x_i = u_i + e_i, \quad u_i \sim \text{Bernoulli}(p)\text{ binary (zero-mean, var }c^2\text{)},
  \quad e_i \sim \mathcal{N}(0, 1-c^2)$$

and maximise the resulting marginal log-likelihood jointly over the direction
$w$, the mixing weight $p$, and the variance allocation $c \in (0,1)$.

The key structural insight is that the full whitened data constraint ($\sum x_i^2 = n$)
allows several terms to be absorbed into constants, leaving a tractable
`softplus`-based objective.  We then optimise it with a nested strategy:

1. **Coarse grid** over $p \in P = \{0.01, 0.05, 0.2, 0.5\}$.
2. **Phase 2** (per $p$): fix $c = c_{\text{golden}} = (\sqrt{5}-1)/2 \approx 0.618$,
   run Newton iterations to convergence.
3. **Phase 3** (per $p$): alternate — optimise $c$ by Brent's method; run a few
   Newton steps to re-align $w$.
4. **Phase 4**: compare converged objectives across the $p$-grid.

## Helpers

```{r}
prewhiten = function(X, n.comp) {
  X = X - rowMeans(X)
  sqrt(ncol(X)) * t(svd(X)$v[, 1:n.comp])
}
```

## Objective function

The prior support points satisfying zero-mean and variance $c^2$ are

$$y_1 = c\sqrt{\frac{1-p}{p}}, \qquad y_0 = -c\sqrt{\frac{p}{1-p}}$$

and the marginal likelihood is a Gaussian mixture with shared variance $1-c^2$.
Exploiting the whitened-data constraints $\sum x_i = 0$ and $\sum x_i^2 = n$,
the average log-likelihood simplifies to

$$\mathcal{J}(w, p, c) = K(p, c) + \frac{1}{n}\sum_{i=1}^n \text{softplus}(s_i)$$

where $s_i = \alpha x_i - \beta + \gamma$ and

$$\alpha = \frac{c}{(1-c^2)\sqrt{p(1-p)}}, \quad
  \beta = \frac{c^2(1-2p)}{2(1-c^2)p(1-p)}, \quad
  \gamma = \log\!\frac{p}{1-p}$$

$$K(p,c) = -\tfrac{1}{2}\log(2\pi(1-c^2)) - \frac{1-p+c^2 p}{2(1-c^2)(1-p)} + \log(1-p)$$

```{r}
# Alpha, beta, gamma parameters from (p, c)
asym2_params = function(p, c) {
  list(
    alpha = c / ((1 - c^2) * sqrt(p * (1-p))),
    beta  = c^2 * (1 - 2*p) / (2 * (1 - c^2) * p * (1-p)),
    gamma = log(p / (1-p))
  )
}

softplus = function(z) {
  # log(1 + exp(z)), numerically stable for large and small z
  pmax(z, 0) + log1p(exp(-abs(z)))
}

# K(p, c): additive terms in J that depend only on (p, c), not w
asym2_K = function(p, c) {
  -0.5 * log(2 * pi * (1 - c^2)) -
    (1 - p + c^2 * p) / (2 * (1 - c^2) * (1 - p)) +
    log(1 - p)
}

# Full average log-likelihood given projections x = Y^T w
asym2_obj = function(x, p, c) {
  par = asym2_params(p, c)
  s   = par$alpha * x - par$beta + par$gamma
  asym2_K(p, c) + mean(softplus(s))
}
```

## Newton update for $w$

The score $g(x_i)$ and its derivative entering the Newton step are

$$g(x_i) = \alpha\,\sigma(s_i), \qquad
  g'(x_i) = \alpha^2\,\sigma(s_i)(1 - \sigma(s_i))$$

where $\sigma(\cdot)$ is the logistic sigmoid.  The fastICA-style Newton step is

$$w^+ = \frac{1}{n} Y^\top g(x) - \overline{g'(x)}\, w, \qquad
  w \leftarrow \frac{w^+}{\|w^+\|}$$

```{r}
# One Newton step for w (fixed p, c); returns normalised w
asym2_w_step = function(Y, w, p, c) {
  x   = as.vector(t(Y) %*% w)
  par = asym2_params(p, c)
  s   = par$alpha * x - par$beta + par$gamma
  sig = plogis(s)
  g   = par$alpha * sig
  gp  = par$alpha^2 * sig * (1 - sig)
  n   = ncol(Y)
  w_new = as.vector(Y %*% g) / n - mean(gp) * w
  w_new / sqrt(sum(w_new^2))
}

# Iterate Newton steps until convergence (fixed p, c)
asym2_w_optimize = function(Y, w, p, c, tol = 1e-6, max_iter = 500) {
  w = w / sqrt(sum(w^2))
  for (iter in seq_len(max_iter)) {
    w_old = w
    w = asym2_w_step(Y, w, p, c)
    if (1 - abs(sum(w * w_old)) < tol) break
  }
  w
}
```

## Optimize $c$ by Brent's method

For fixed $w$ and $p$, the objective as a function of $c$ alone is

$$\mathcal{J}(c) = -\tfrac{1}{2}\log(1-c^2) - \frac{1-p+c^2 p}{2(1-c^2)(1-p)}
  + \frac{1}{n}\sum_{i=1}^n \text{softplus}\!\left(\alpha(c)\,x_i - \beta(c) + \gamma\right)$$

We maximise this over $c \in (\epsilon, 1-\epsilon)$ using R's `optimize`.

```{r}
# Optimize c by Brent's method (fixed projections x, and p)
asym2_c_optimize = function(x, p, eps = 1e-4) {
  obj_c = function(c) {
    par = asym2_params(p, c)
    s   = par$alpha * x - par$beta + par$gamma
    -0.5 * log(1 - c^2) -
      (1 - p + c^2 * p) / (2 * (1 - c^2) * (1 - p)) +
      mean(softplus(s))
  }
  optimize(obj_c, c(eps, 1 - eps), maximum = TRUE)$maximum
}
```

## Full algorithm for one $p$

Given a fixed $p$, run the three-phase strategy and return the converged $(w, c)$
and the final objective value.

```{r}
C_GOLDEN = (sqrt(5) - 1) / 2   # ≈ 0.618

asym2_one_p = function(Y, p, w_init = NULL,
                        tol = 1e-6, max_iter = 50,
                        n_newton_refine = 10) {
  m = nrow(Y)
  w = if (is.null(w_init)) rnorm(m) else w_init
  w = w / sqrt(sum(w^2))

  # Phase 2: fix c = c_golden, optimize w to convergence
  c = C_GOLDEN
  w = asym2_w_optimize(Y, w, p, c)

  # Phase 3: alternate c (Brent) and w (Newton)
  for (iter in seq_len(max_iter)) {
    w_old = w;  c_old = c
    x = as.vector(t(Y) %*% w)
    c = asym2_c_optimize(x, p)
    w = asym2_w_optimize(Y, w, p, c, max_iter = n_newton_refine)
    if (1 - abs(sum(w * w_old)) < tol && abs(c - c_old) < tol) break
  }
  x = as.vector(t(Y) %*% w)
  list(w = w, p = p, c = c, obj = asym2_obj(x, p, c), iter = iter)
}
```

## Full algorithm: grid over $p$

```{r}
P_GRID = c(0.01, 0.05, 0.2, 0.5)

# Returns a list of results, one per p in p_grid.
# w_init is shared across all p (same starting direction).
asym2_fastica = function(Y, p_grid = P_GRID, w_init = NULL, ...) {
  m = nrow(Y)
  w0 = if (is.null(w_init)) { v = rnorm(m); v / sqrt(sum(v^2)) } else
         w_init / sqrt(sum(w_init^2))
  lapply(p_grid, \(p) asym2_one_p(Y, p, w_init = w0, ...))
}

# Convenience: run asym2_fastica and pick the result with highest obj
asym2_fastica_best = function(Y, p_grid = P_GRID, w_init = NULL, ...) {
  res = asym2_fastica(Y, p_grid, w_init, ...)
  best = which.max(sapply(res, `[[`, "obj"))
  res[[best]]
}
```

## Sanity checks

### $p=0.5$, $c = c_\text{golden}$ recovers log-cosh score

The generative model with $\sigma^2 = c$ (noise variance equals signal SD) combined
with the total variance constraint $\sigma^2 = 1 - c^2$ gives $c = c_\text{golden}$.
At $p = 0.5$, $c = c_\text{golden}$: $1 - c^2 = c$, so
$\alpha = c\,/\,(c\cdot\tfrac{1}{2}) = 2$, $\beta = 0$, $\gamma = 0$, giving
$s_i = 2x_i$ and

$$g(x_i) = 2\sigma(2x_i) = 1 + \tanh(x_i)$$

Since the data is centred, $\tfrac{1}{n}Y^\top\mathbf{1} = 0$, so the Newton step
becomes $w^+ \propto Y^\top\tanh(x) - \overline{\mathrm{sech}^2(x)}\,w$, exactly
log-cosh fastICA.

```{r}
x_grid = seq(-3, 3, length.out = 300)
par05  = asym2_params(0.5, C_GOLDEN)
g05    = par05$alpha * plogis(par05$alpha * x_grid - par05$beta + par05$gamma)

plot(x_grid, g05 - 1, type = "l", col = "steelblue", lwd = 2,
     xlab = "x", ylab = "g(x) - 1",
     main = sprintf("Score at p=0.5, c=c_golden (%.4f): g(x)-1 vs tanh(x)", C_GOLDEN))
lines(x_grid, tanh(x_grid), col = "tomato", lty = 2, lwd = 2)
legend("topleft", c("g(x) - 1,  p=0.5, c=c_golden", "tanh(x)"),
       col = c("steelblue","tomato"), lty = 1:2, lwd = 2, bty = "n")
```

The curves are numerically identical.

### Objective at golden-ratio start

The golden-ratio anchor satisfies $c^2 + c = 1$, i.e. $1 - c^2 = c$, so
$\alpha = c/(c \cdot \sqrt{p(1-p)}) = 1/\sqrt{p(1-p)}$ and the softplus
argument at $p = 0.5$ becomes $2x_i$, consistent with the log-cosh link.

```{r}
par05_gr = asym2_params(0.5, C_GOLDEN)
cat(sprintf("alpha at p=0.5, c_golden: %.6f  (1/sqrt(p(1-p)) = %.6f)\n",
    par05_gr$alpha, 1/sqrt(0.5*0.5)))
cat(sprintf("K(0.5, c_golden) = %.4f\n", asym2_K(0.5, C_GOLDEN)))
```



## Experiment helpers

Run the method 100 times for a single fixed $p$, returning the converged
objective, $c$, and correlation with the true sources for each seed.

```{r}
# 100 seeds for one fixed p; returns maxcor, c, obj per seed
run_seeds_p = function(Y, S_true, p, n_seeds = 100) {
  maxcor = numeric(n_seeds)
  cs     = numeric(n_seeds)
  objs   = numeric(n_seeds)
  for (seed in seq_len(n_seeds)) {
    set.seed(seed)
    res          = asym2_one_p(Y, p)
    maxcor[seed] = max(abs(cor(t(S_true), t(Y) %*% res$w)))
    cs[seed]     = res$c
    objs[seed]   = res$obj
  }
  list(maxcor = maxcor, c = cs, obj = objs)
}

# One panel: objectives ranked descending, coloured by success
plot_obj_p = function(res, p, ...) {
  succ = res$maxcor > 0.9
  ord  = order(res$obj, decreasing = TRUE)
  cols = ifelse(succ[ord], "steelblue", "tomato")
  plot(seq_along(res$obj), res$obj[ord],
       pch = 19, cex = 0.6, col = cols,
       xlab = "rank", ylab = "J(w, p, c)",
       main = sprintf("p = %.2f  (%.0f%% found)", p, 100*mean(succ)),
       ...)
}

# Four-panel figure: one panel per p in P_GRID
plot_by_p = function(results_list, p_grid, title) {
  ylims = range(sapply(results_list, \(r) range(r$obj)))
  par(mfrow = c(1, length(p_grid)))
  for (i in seq_along(p_grid))
    plot_obj_p(results_list[[i]], p_grid[i], ylim = ylims)
  mtext(title, side = 3, line = -1.5, outer = TRUE, font = 2)
  par(mfrow = c(1, 1))
}
```

## Data

Reproduce the datasets from the first analysis.

```{r}
set.seed(1)
n = 100; p_dim = 1000; K = 9
L    = matrix(0, nrow = n, ncol = K)
for (i in 1:K) L[sample(n, 20), i] = 1
FF   = matrix(rnorm(p_dim * K), nrow = p_dim)
X9   = t(L %*% t(FF) + matrix(rnorm(n * p_dim, 0, 0.1), nrow = n))
Z9   = prewhiten(X9, K)
S9   = t(L)

set.seed(2)
L_01  = matrix(0, nrow = n, ncol = K)
for (i in 1:K) L_01[sample(n, 10), i] = 1
FF_01 = matrix(rnorm(p_dim * K), nrow = p_dim)
X9_01 = t(L_01 %*% t(FF_01) + matrix(rnorm(n * p_dim, 0, 0.1), nrow = n))
Z9_01 = prewhiten(X9_01, K)
S9_01 = t(L_01)

set.seed(10)
n_k20 = 200; p_k20 = 1000; k20 = 20
A_k20    = matrix(rnorm(p_k20), nrow = p_k20)
S_k20_05 = matrix(sample(c(-1, 1), n_k20, replace = TRUE), nrow = 1)
X_k20_05 = A_k20 %*% S_k20_05 +
           matrix(rnorm(p_k20 * n_k20, 0, 0.1), nrow = p_k20)
Z_k20_05 = prewhiten(X_k20_05, k20)

set.seed(11)
S_k20_02_raw = matrix(as.numeric(runif(n_k20) < 0.2), nrow = 1)
S_k20_02     = (S_k20_02_raw - 0.2) / sqrt(0.2 * 0.8)
X_k20_02 = A_k20 %*% S_k20_02 +
           matrix(rnorm(p_k20 * n_k20, 0, 0.1), nrow = p_k20)
Z_k20_02 = prewhiten(X_k20_02, k20)

set.seed(12)
S_k20_01_raw = matrix(as.numeric(runif(n_k20) < 0.1), nrow = 1)
S_k20_01     = (S_k20_01_raw - 0.1) / sqrt(0.1 * 0.9)
X_k20_01 = A_k20 %*% S_k20_01 +
           matrix(rnorm(p_k20 * n_k20, 0, 0.1), nrow = p_k20)
Z_k20_01 = prewhiten(X_k20_01, k20)
```

## Simple test: 9 overlapping groups, $p \approx 0.1$, $k = 9$

Before the multi-seed experiments we run `asym2_one_p` once (seed 1) for each
assumed $p$ in the grid on the 9-groups $p \approx 0.1$ dataset, and compare
the recovered source and final objective side by side.

```{r}
set.seed(1)
w_init_ex = rnorm(K)
res_ex = lapply(P_GRID, \(p) asym2_one_p(Z9_01, p, w_init = w_init_ex))
```

```{r fig.width=10, fig.height=8}
par(mfrow = c(2, length(P_GRID)))

# Top row: scatter of best recovered source vs its true source
for (i in seq_along(P_GRID)) {
  x_hat = as.vector(t(Z9_01) %*% res_ex[[i]]$w)
  cors  = cor(t(S9_01), x_hat)
  best  = which.max(abs(cors))
  plot(S9_01[best, ], x_hat,
       pch = 19, cex = 0.5, col = "steelblue",
       xlab = "true source", ylab = "projection",
       main = sprintf("p=%.2f  cor=%.3f\nc=%.3f  obj=%.4f",
                      P_GRID[i], cors[best],
                      res_ex[[i]]$c, res_ex[[i]]$obj))
}

# Bottom row: histogram of projection values
for (i in seq_along(P_GRID)) {
  x_hat = as.vector(t(Z9_01) %*% res_ex[[i]]$w)
  hist(x_hat, breaks = 30, col = "steelblue", border = "white",
       main = sprintf("p=%.2f histogram", P_GRID[i]),
       xlab = "projection value")
}

par(mfrow = c(1, 1))
```

```{r}
cat("9-groups p≈0.1, k=9, seed 1:\n")
for (i in seq_along(P_GRID)) {
  x_hat = as.vector(t(Z9_01) %*% res_ex[[i]]$w)
  cors  = cor(t(S9_01), x_hat)
  cat(sprintf("  assumed p=%.2f  maxcor=%.4f  c=%.4f  obj=%.4f\n",
      P_GRID[i], max(abs(cors)), res_ex[[i]]$c, res_ex[[i]]$obj))
}
```

To check whether `c` is being maximised (not minimised), we plot
$\mathcal{J}(c)$ against $c$ for each assumed $p$, using the final converged
projection $x = Yw$ for that run.  The vertical dashed line marks the returned
optimum.

```{r fig.width=10, fig.height=4}
c_scan = seq(0.01, 0.99, by = 0.005)

# objective as function of c alone, same formula as asym2_c_optimize
obj_c_vec = function(x, p, c_vec) {
  sapply(c_vec, \(c) {
    par = asym2_params(p, c)
    s   = par$alpha * x - par$beta + par$gamma
    -0.5 * log(1 - c^2) -
      (1 - p + c^2 * p) / (2 * (1 - c^2) * (1 - p)) +
      mean(softplus(s))
  })
}

par(mfrow = c(1, length(P_GRID)))
for (i in seq_along(P_GRID)) {
  x_hat  = as.vector(t(Z9_01) %*% res_ex[[i]]$w)
  obj_cv = obj_c_vec(x_hat, P_GRID[i], c_scan)
  plot(c_scan, obj_cv, type = "l", col = "steelblue", lwd = 2,
       xlab = "c", ylab = "J(c)",
       main = sprintf("p = %.2f  (c* = %.3f)", P_GRID[i], res_ex[[i]]$c))
  abline(v = res_ex[[i]]$c, lty = 2, col = "tomato", lwd = 1.5)
}
par(mfrow = c(1, 1))
```

## 9 overlapping groups ($k = 9$), $p \approx 0.2$

```{r}
res9_by_p = lapply(P_GRID, \(p) run_seeds_p(Z9, S9, p))
```

```{r fig.width=10, fig.height=4}
plot_by_p(res9_by_p, P_GRID, "9-groups p≈0.2, k=9")
```

```{r}
cat("9-groups p≈0.2: frac > 0.9 per assumed p\n")
for (i in seq_along(P_GRID))
  cat(sprintf("  p = %.2f  frac > 0.9 = %.2f  mean_c = %.3f\n",
      P_GRID[i],
      mean(res9_by_p[[i]]$maxcor > 0.9),
      mean(res9_by_p[[i]]$c)))
```

## 9 overlapping groups ($k = 9$), $p \approx 0.1$

```{r}
res9_01_by_p = lapply(P_GRID, \(p) run_seeds_p(Z9_01, S9_01, p))
```

```{r fig.width=10, fig.height=4}
plot_by_p(res9_01_by_p, P_GRID, "9-groups p≈0.1, k=9")
```

```{r}
cat("9-groups p≈0.1: frac > 0.9 per assumed p\n")
for (i in seq_along(P_GRID))
  cat(sprintf("  p = %.2f  frac > 0.9 = %.2f  mean_c = %.3f\n",
      P_GRID[i],
      mean(res9_01_by_p[[i]]$maxcor > 0.9),
      mean(res9_01_by_p[[i]]$c)))
```

## Single source, over-complete whitening ($k = 20$)

### $p_\text{true} = 0.5$

```{r}
res_k20_05_by_p = lapply(P_GRID, \(p) run_seeds_p(Z_k20_05, S_k20_05, p))
```

```{r fig.width=10, fig.height=4}
plot_by_p(res_k20_05_by_p, P_GRID, "1-source k=20 p_true=0.5")
```

```{r}
cat("1-source k=20 p_true=0.5: frac > 0.9 per assumed p\n")
for (i in seq_along(P_GRID))
  cat(sprintf("  p = %.2f  frac > 0.9 = %.2f  mean_c = %.3f\n",
      P_GRID[i],
      mean(res_k20_05_by_p[[i]]$maxcor > 0.9),
      mean(res_k20_05_by_p[[i]]$c)))
```

### $p_\text{true} = 0.2$

```{r}
res_k20_02_by_p = lapply(P_GRID, \(p) run_seeds_p(Z_k20_02, S_k20_02, p))
```

```{r fig.width=10, fig.height=4}
plot_by_p(res_k20_02_by_p, P_GRID, "1-source k=20 p_true=0.2")
```

```{r}
cat("1-source k=20 p_true=0.2: frac > 0.9 per assumed p\n")
for (i in seq_along(P_GRID))
  cat(sprintf("  p = %.2f  frac > 0.9 = %.2f  mean_c = %.3f\n",
      P_GRID[i],
      mean(res_k20_02_by_p[[i]]$maxcor > 0.9),
      mean(res_k20_02_by_p[[i]]$c)))
```

### $p_\text{true} = 0.1$

```{r}
res_k20_01_by_p = lapply(P_GRID, \(p) run_seeds_p(Z_k20_01, S_k20_01, p))
```

```{r fig.width=10, fig.height=4}
plot_by_p(res_k20_01_by_p, P_GRID, "1-source k=20 p_true=0.1")
```

```{r}
cat("1-source k=20 p_true=0.1: frac > 0.9 per assumed p\n")
for (i in seq_along(P_GRID))
  cat(sprintf("  p = %.2f  frac > 0.9 = %.2f  mean_c = %.3f\n",
      P_GRID[i],
      mean(res_k20_01_by_p[[i]]$maxcor > 0.9),
      mean(res_k20_01_by_p[[i]]$c)))
```
