---
title: "fastica_asymmetric_05"
author: "Matthew Stephens"
date: "2026-09-22"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

This file extends `fastica_asymmetric_03.Rmd` and `fastica_asymmetric_04.Rmd` from the
centred **binary** prior to a centred, scaled **point-exponential** prior
$$g_0 = \pi_0 \delta_0 + (1-\pi_0)\,\text{Exp}(1).$$
As in `04`, the emphasis is on running many Newton updates for $w$ **in parallel**
(one column of $W$ per random start), with the option to update the
hyperparameters $\pi_0$, $c$ and $\sigma^2$ at every iteration.

The motivation for replacing binary by point-exponential is that the binary prior
commits to a single non-zero effect size, whereas the point-exponential has a
continuous slab and so can represent sources whose non-zero values vary in
magnitude, while still being strongly skewed. The skewness is the property that
drives the asymmetric objective, so we should retain the behaviour that made the
binary version work, while being less rigid about effect sizes.

### Parameterisation

Following the `g_0`/`c` split used in `03`/`04`, we constrain the **base prior**
$g_0$ to have mean 0 and variance 1, and carry the scale in $c$. Let
$$\theta_0 \sim \pi_0 \delta_0 + (1-\pi_0)\,\text{Exp}(1), \qquad
  m_0 = \mathbb{E}\theta_0 = 1-\pi_0, \qquad
  v_0 = \text{Var}(\theta_0) = 2(1-\pi_0) - (1-\pi_0)^2,$$
and define
$$g_0 = \text{law of } \frac{\theta_0 - m_0}{\sqrt{v_0}}, \qquad
  g = \text{law of } c\,g_0 .$$
So $g$ has mean 0 and **variance $c^2$**, and $\pi_0$ is the single shape
parameter, exactly parallel to $p$ in the binary case. It is convenient to write
$g$ directly in terms of an exponential scale $s$ and a shift $\mu$:
$$s = \frac{c}{\sqrt{v_0}}, \qquad \mu = s\,(1-\pi_0),$$
$$g(\theta) = \pi_0\,\delta_{-\mu}(\theta) \;+\; (1-\pi_0)\,\tfrac{1}{s}e^{-(\theta+\mu)/s}\,\mathbb{1}(\theta > -\mu).$$
That is: a point mass at $-\mu$, plus an exponential slab starting at $-\mu$.

The default settings suggested below are $\pi_0 = 0.5$ and $c^2 = \sigma^2 = 0.5$,
so that prior variance plus residual variance equals 1. This plays the role that
$c = c_\text{golden}$, $\sigma^2 = c$ played for the binary prior (where
$c_\text{golden}^2 + c_\text{golden} = 1$).

## The objective and its derivatives

As before the objective is $J(w) = \sum_i \log Z(x_i)$ with $x = U^Tw$ and
$$Z(x) = \int g(\theta)\exp\!\left(\frac{2x\theta - \theta^2}{2\sigma^2}\right)d\theta.$$

Substituting $u = \theta + \mu$ (so $u \ge 0$ and $u$ has the *uncentred* prior
$g_u = \pi_0\delta_0 + (1-\pi_0)\text{Exp}(1/s)$) and writing $\tilde x = x + \mu$,
$$2x\theta - \theta^2 = 2\tilde x u - u^2 - 2x\mu - \mu^2,$$
so
$$\log Z(x) = -\frac{2x\mu + \mu^2}{2\sigma^2} + \log A(\tilde x), \qquad
  A(\tilde x) = \int g_u(u)\exp\!\left(\frac{2\tilde x u - u^2}{2\sigma^2}\right)du .$$

For the slab, completing the square with $m = \tilde x - \sigma^2/s$ gives
$$-\frac{(u-\tilde x)^2}{2\sigma^2} - \frac{u}{s} = \frac{m^2}{2\sigma^2} - \frac{(u-m)^2}{2\sigma^2} - \frac{\tilde x^2}{2\sigma^2} + \frac{\tilde x^2}{2\sigma^2},$$
and hence
$$\int_0^\infty \tfrac{1}{s}e^{-u/s}\exp\!\left(\frac{2\tilde xu-u^2}{2\sigma^2}\right)du
 = \tfrac{1}{s}\sqrt{2\pi}\,\sigma\, e^{m^2/(2\sigma^2)}\,\Phi(m/\sigma).$$
So, with log-scale unnormalised component weights
$$\ell_0 = \log \pi_0, \qquad
  \ell_1 = \log(1-\pi_0) - \log s + \tfrac12\log(2\pi\sigma^2) + \frac{m^2}{2\sigma^2} + \log\Phi(m/\sigma),$$
we have $\log A(\tilde x) = \text{logsumexp}(\ell_0, \ell_1)$.
Working on the log scale matters here: $m^2/(2\sigma^2)$ overflows easily, and
$\log\Phi$ must be evaluated with `pnorm(..., log.p = TRUE)` to stay accurate in
the left tail.

The quantities needed for the Newton step follow from
$\partial_x \log Z = \mathbb{E}[\theta\mid x]/\sigma^2$ and
$\partial_x^2 \log Z = \text{Var}[\theta\mid x]/\sigma^2$:
$$M(x) = \mathbb{E}[u\mid \tilde x] - \mu, \qquad M'(x) = \text{Var}[u\mid \tilde x]/\sigma^2 .$$
The posterior on $u$ is $\bar\pi_0 \delta_0 + (1-\bar\pi_0)\,TN(m,\sigma^2;\,u>0)$
with $\bar\pi_1 = e^{\ell_1 - \log A}$. Writing $z = m/\sigma$ and
$\lambda = \phi(z)/\Phi(z)$ for the inverse Mills ratio,
$$\mathbb{E}[u \mid u>0] = m + \sigma\lambda, \qquad
  \text{Var}[u \mid u>0] = \sigma^2\left(1 - z\lambda - \lambda^2\right),$$
and then $\mathbb{E}[u] = \bar\pi_1(m+\sigma\lambda)$,
$\mathbb{E}[u^2] = \bar\pi_1(\text{Var}[u\mid u>0] + (m+\sigma\lambda)^2)$.
(The variance formula suffers cancellation for very negative $z$; it is clamped at 0
below. In practice $z$ stays in a benign range for the settings used here.)

Note that $\log Z$ contains a term linear in $x$ (through $-2x\mu/(2\sigma^2)$),
just as the binary $\log Z$ did for $p \ne 0.5$. It is handled automatically by
$M(x)$, so no special treatment is needed in the update.

## Implementation

Everything is vectorised over columns: `P` is an $n \times m$ matrix of $x$ values
(one column per start) and the hyperparameters may be scalars or length-$m$
vectors, so that each start can carry its own $(\pi_0, c, \sigma^2)$.

```{r}
pe_v0 = function(pi0) 2*(1-pi0) - (1-pi0)^2   # Var of the Exp(1) spike-and-slab

# broadcast length-1 or length-m vector to an n x m matrix (constant down columns)
pe_bc = function(v, n, m)
  if (length(v) == 1) matrix(v, n, m) else matrix(v, n, m, byrow = TRUE)

# Per-observation posterior quantities for the centred, unit-variance
# point-exponential base prior scaled by c (so Var(g) = c^2).
pe_stats = function(P, pi0, cc, s2) {
  n = nrow(P); m = ncol(P)
  pi0 = pmin(pmax(pi0, 1e-6), 1 - 1e-6)
  PI0 = pe_bc(pi0, n, m);  CC = pe_bc(cc, n, m);  S2 = pe_bc(s2, n, m)
  SS  = CC / sqrt(pe_v0(PI0))          # exponential scale s
  SIG = sqrt(S2)
  MU  = SS * (1 - PI0)                 # centring shift: the atom sits at -MU
  MM  = P + MU - S2 / SS               # mean of the untruncated normal in u
  Z   = MM / SIG
  logPhi = pnorm(Z, log.p = TRUE)

  l0 = log(PI0)
  l1 = log1p(-PI0) - log(SS) + 0.5*log(2*pi*S2) + MM^2/(2*S2) + logPhi
  mx = pmax(l0, l1)
  logA = mx + log(exp(l0 - mx) + exp(l1 - mx))

  pi1 = exp(l1 - logA)                        # posterior P(slab)
  lam = exp(dnorm(Z, log = TRUE) - logPhi)    # inverse Mills ratio phi/Phi
  m1  = MM + SIG*lam                          # E[u | slab]
  v1  = pmax(S2*(1 - Z*lam - lam^2), 0)       # Var[u | slab]
  Eu  = pi1*m1
  Eu2 = pi1*(v1 + m1^2)

  list(logZ = -(2*P*MU + MU^2)/(2*S2) + logA,
       pi1  = pi1,
       M    = Eu - MU,                        # posterior mean of theta
       Mp   = pmax(Eu2 - Eu^2, 0) / S2,       # dM/dx = Var(theta)/sigma^2
       M2   = Eu2 - 2*MU*Eu + MU^2,           # E[theta^2]
       m1   = m1, s = SS)
}

pe_logZ = function(x, pi0, cc, s2)
  as.vector(pe_stats(matrix(x, ncol = 1), pi0, cc, s2)$logZ)
```

### Objective and Newton update

`U` is $k \times n$ (rows scaled so $\sum_i U_{ji}^2 = n$) and `W` is
$k \times m$, as in `04`. Since we only care about centred sources there is no
intercept row.

```{r}
# J = sum_i log Z(x_i), returned as a length-m vector
pe_objective = function(U, W, pi0, cc, s2)
  colSums(pe_stats(crossprod(U, W), pi0, cc, s2)$logZ)

# EBproj objective F = nu/2 (log tau - tau + 1) + J,  with tau = n/(nu sigma^2)
pe_ebproj_objective = function(U, W, pi0, cc, s2, nu, n) {
  tau = n / (nu * s2)
  (nu/2)*(log(tau) - tau + 1) + pe_objective(U, W, pi0, cc, s2)
}

# Newton update with the fastICA ("ica") diagonal Hessian approximation
pe_newton_update = function(U, W, pi0, cc, s2, hessian = "ica") {
  st = pe_stats(crossprod(U, W), pi0, cc, s2)
  Wn = U %*% st$M
  if (hessian == "ica") Wn = Wn - sweep(W, 2, colSums(st$Mp), "*")
  sweep(Wn, 2, sqrt(colSums(Wn^2)) + 1e-15, "/")
}
```

### Hyperparameter updates

Two families are provided.

**Grid updates** do exact coordinate ascent in $J$ for $\pi_0$ and $c$. Since
$x = U^Tw$ does not change during a hyperparameter update, a whole grid can be
swept for all starts at once, so the cost is (grid size) matrix passes per
iteration. A parabolic refinement around the best grid point gives sub-grid
resolution.

**EM updates** are the exact EM updates for the point-exponential in $u$-space
holding the centring shift $\mu$ fixed at its current value, after which $\mu$ is
recomputed. Because the constraint $\mu = s(1-\pi_0)$ couples $\mu$ to
$(\pi_0, c)$, this is not an exact EM for the constrained model and is not
guaranteed to increase $J$ — but it is $O(n)$ rather than $O(n\times\text{grid})$,
and (see below) it turns out to behave considerably *better* than exact
coordinate ascent.

```{r}
# vectorised 1-D maximiser: grid sweep + parabolic refinement.
# f(t) returns a length-m vector of objective values at transformed value t.
pe_opt_1d = function(f, grid, m) {
  G = length(grid); V = matrix(0, G, m)
  for (g in seq_len(G)) V[g, ] = f(grid[g])
  i  = max.col(t(V), ties.method = "first")
  ii = seq_len(m); step = grid[2] - grid[1]
  v0 = V[cbind(pmax(i-1,1), ii)]; v1 = V[cbind(i, ii)]; v2 = V[cbind(pmin(i+1,G), ii)]
  den = v0 - 2*v1 + v2
  off = ifelse(abs(den) > 1e-12 & i > 1 & i < G, 0.5*(v0 - v2)/den, 0)
  list(par = grid[i] + pmin(pmax(off, -1), 1)*step, at_edge = (i == 1 | i == G))
}

logit = function(p) log(p/(1-p));  expit = function(z) 1/(1+exp(-z))

pe_update_pi0_grid = function(P, pi0, cc, s2,
                              grid = seq(logit(0.01), logit(0.99), length.out = 31))
  expit(pe_opt_1d(function(t) colSums(pe_stats(P, expit(t), cc, s2)$logZ), grid, ncol(P))$par)

pe_update_c_grid = function(P, pi0, cc, s2,
                            grid = seq(log(0.05), log(5), length.out = 31))
  exp(pe_opt_1d(function(t) colSums(pe_stats(P, pi0, exp(t), s2)$logZ), grid, ncol(P))$par)

# EM-style update of (pi0, c), holding the centring shift fixed within the step
pe_update_g_em = function(P, pi0, cc, s2, update_pi0 = TRUE, update_c = TRUE) {
  st = pe_stats(P, pi0, cc, s2)
  ss = cc / sqrt(pe_v0(pi0))
  if (update_c)   ss  = colSums(st$pi1 * st$m1) / colSums(st$pi1)
  if (update_pi0) pi0 = pmin(pmax(colMeans(1 - st$pi1), 1e-6), 1 - 1e-6)
  list(pi0 = pi0, c = ss * sqrt(pe_v0(pi0)))
}

# sigma^2 via the EBproj fixed point tau = 1/(1-rho), rho = vbar' P vbar / U_q.
# In this parameterisation P_proj = U'U/n, so rho = ||U M||^2 / (n * sum(M2)).
pe_update_s2 = function(U, P, pi0, cc, s2, nu, n, rho_max = 0.999) {
  st  = pe_stats(P, pi0, cc, s2)
  rho = colSums((U %*% st$M)^2) / (n * colSums(st$M2))
  n / (nu * (1/(1 - pmin(rho, rho_max))))
}
```

### Main parallel loop

```{r}
pe_fit = function(U, n_starts = 100, n_iter = 50, pi0 = 0.5, cc = sqrt(0.5), s2 = 0.5,
                  b = 0, W = NULL, update_pi0 = FALSE, update_c = FALSE,
                  update_s2 = FALSE, g_method = c("grid", "em"),
                  hessian = "ica", trace = FALSE) {
  g_method = match.arg(g_method)
  k = nrow(U); n = ncol(U); nu = n - k - 2*b
  if (is.null(W)) {
    W = matrix(rnorm(k*n_starts), k, n_starts)
    W = sweep(W, 2, sqrt(colSums(W^2)), "/")
  }
  m = ncol(W)
  pi0 = rep(pi0, length.out = m); cc = rep(cc, length.out = m); s2 = rep(s2, length.out = m)
  Jtr = if (trace) matrix(NA_real_, n_iter, m) else NULL

  for (it in seq_len(n_iter)) {
    P = crossprod(U, W)
    if (update_s2) s2 = pe_update_s2(U, P, pi0, cc, s2, nu, n)
    if (update_pi0 || update_c) {
      if (g_method == "em") {
        g = pe_update_g_em(P, pi0, cc, s2, update_pi0, update_c); pi0 = g$pi0; cc = g$c
      } else {
        if (update_pi0) pi0 = pe_update_pi0_grid(P, pi0, cc, s2)
        if (update_c)   cc  = pe_update_c_grid(P, pi0, cc, s2)
      }
    }
    W = pe_newton_update(U, W, pi0, cc, s2, hessian = hessian)
    if (trace) Jtr[it, ] = pe_objective(U, W, pi0, cc, s2)
  }
  list(W = W, pi0 = pi0, c = cc, s2 = s2, nu = nu, n = n, k = k,
       J = pe_objective(U, W, pi0, cc, s2),
       F = pe_ebproj_objective(U, W, pi0, cc, s2, nu, n), Jtrace = Jtr)
}
```

## Numerical checks

Check $\log Z$ against brute-force numerical integration, $M$ against
$\sigma^2\,\partial_x \log Z$, $M'$ against $\partial_x M$, and confirm that the
prior really does have mean 0 and variance $c^2$.

```{r}
logZ_num = function(x, pi0, cc, s2) {
  ss = cc/sqrt(pe_v0(pi0)); mu = ss*(1-pi0)
  f = function(th) exp((2*x*th - th^2)/(2*s2)) * (1/ss)*exp(-(th+mu)/ss)
  log(pi0*exp((2*x*(-mu) - mu^2)/(2*s2)) +
      (1-pi0)*integrate(f, -mu, Inf, rel.tol = 1e-12)$value)
}

xs   = c(-3, -1, -0.2, 0, 0.5, 1, 3, 6)
pars = list(c(0.5, sqrt(0.5), 0.5), c(0.2, 0.3, 0.7), c(0.9, 1.5, 0.2), c(0.05, 0.8, 1.3))
for (p in pars) {
  pi0 = p[1]; cc = p[2]; s2 = p[3]; h = 1e-5
  a  = pe_logZ(xs, pi0, cc, s2)
  b  = sapply(xs, logZ_num, pi0 = pi0, cc = cc, s2 = s2)
  st = pe_stats(matrix(xs, ncol = 1), pi0, cc, s2)
  Mnum  = s2*(pe_logZ(xs+h, pi0, cc, s2) - pe_logZ(xs-h, pi0, cc, s2))/(2*h)
  Mpnum = (pe_stats(matrix(xs+h, ncol=1), pi0, cc, s2)$M -
           pe_stats(matrix(xs-h, ncol=1), pi0, cc, s2)$M)/(2*h)
  set.seed(1); N = 2e6
  th0 = ifelse(runif(N) < pi0, 0, rexp(N))
  g   = cc*(th0 - (1-pi0))/sqrt(pe_v0(pi0))
  cat(sprintf("pi0=%.2f c=%.2f s2=%.2f | logZ err %.1e | M err %.1e | Mp err %.1e | prior mean %.4f var %.4f (target 0, %.4f)\n",
      pi0, cc, s2, max(abs(a-b)), max(abs(st$M-Mnum)), max(abs(st$Mp-Mpnum)),
      mean(g), var(g), cc^2))
}
```

All agree to essentially machine precision.

## Shape of the objective

`log Z(x)` for a range of $\pi_0$, at the default $c^2 = \sigma^2 = 0.5$. Unlike
the binary case these curves are strongly **asymmetric**: they are close to linear
on the left (where the atom dominates) and curve upward on the right (where the
slab dominates). Larger $\pi_0$ gives a sharper knee.

```{r}
c_val  = sqrt(0.5); s2_val = 0.5
pi0s   = c(0.1, 0.3, 0.5, 0.7, 0.8, 0.9)
cols   = palette.colors(length(pi0s), palette = "Okabe-Ito")
x_grid = seq(-5, 5, length.out = 500)
logZ_mat = sapply(pi0s, function(q) pe_logZ(x_grid, q, c_val, s2_val))
matplot(x_grid, logZ_mat, type = "l", lty = 1, col = cols, lwd = 2,
        xlab = "x", ylab = "log Z(x)", main = "log Z(x) by pi_0  (c^2 = sigma^2 = 0.5)")
legend("top", legend = paste0("pi_0 = ", pi0s), col = cols, lty = 1, lwd = 2, cex = 0.8)
```

### Expected objective by source type

We restrict attention throughout to **centred (zero-mean) sources**, so there is
no intercept to profile over. For a zero-mean binary source with weight
$p_\text{s}$ and unit variance the atoms are
$$a_0 = -p_s d, \quad a_1 = (1-p_s)d, \quad d = \frac{1}{\sqrt{p_s(1-p_s)}},$$
i.e. the usual $a_1 = \sqrt{(1-p_s)/p_s}$, $a_0 = -\sqrt{p_s/(1-p_s)}$. A source
is detectable as a maximum of this objective if
$\mathbb{E}[\log Z]$ sits above the standard normal baseline.

```{r}
E_logZ_binary = function(p_s, pi0, cc, s2) {
  a1 = sqrt((1-p_s)/p_s); a0 = -sqrt(p_s/(1-p_s))
  (1-p_s)*pe_logZ(a0, pi0, cc, s2) + p_s*pe_logZ(a1, pi0, cc, s2)
}
E_logZ_gauss = function(pi0, cc, s2)
  integrate(function(x) pe_logZ(x, pi0, cc, s2)*dnorm(x), -8, 8)$value

p_srcs    = c(0.02, 0.05, 0.1, 0.2, 0.3, 0.4, 0.5)
ps_grid   = seq(0.01, 0.99, by = 0.01)
E_mat     = sapply(pi0s, function(q) sapply(ps_grid, function(ps) E_logZ_binary(ps, q, c_val, s2_val)))
E_gauss   = sapply(pi0s, function(q) E_logZ_gauss(q, c_val, s2_val))

matplot(ps_grid, E_mat, type = "l", lty = 1, col = cols, lwd = 2,
        xlab = expression(p[source]), ylab = expression(E[source]*"[log Z(x)]"),
        main = "Expected objective, zero-mean binary sources")
for (i in seq_along(pi0s)) abline(h = E_gauss[i], col = cols[i], lty = 2, lwd = 1.2)
legend("topright", legend = c(paste0("pi_0 = ", pi0s), "Gaussian (dashed)"),
       col = c(cols, "grey50"), lty = c(rep(1, length(pi0s)), 2), lwd = 2, cex = 0.7)
```

This is qualitatively different from the binary objective, and is the main
structural finding of this file. For the binary prior, $p_\text{obj}$ selected
*which* sources were favoured — $p_\text{obj}=0.5$ favoured balanced sources,
small $p_\text{obj}$ favoured sparse ones. Here the expected objective is
**monotone decreasing in $p_\text{source}$ at every $\pi_0$**: sparse/skewed
sources always score above common ones, and $\pi_0$ only tunes how steeply.

The flip side is that a balanced ($p_s = 0.5$) binary source falls *below* the
Gaussian baseline for every $\pi_0$. So this objective is a detector for skewed
sources specifically, and should not be expected to find symmetric ones. That is
inherent: the prior is one-sided, so it cannot reward a symmetric source.

```{r}
for (q in c(0.3, 0.5, 0.7, 0.9)) {
  e0 = sapply(p_srcs, function(ps) E_logZ_binary(ps, q, c_val, s2_val))
  gb = E_logZ_gauss(q, c_val, s2_val)
  cat(sprintf("pi_0=%.1f  Gaussian=%.4f | E[logZ] by p_src (%s): %s\n", q, gb,
      paste(p_srcs, collapse = " "), paste(sprintf("%.4f", e0), collapse = " ")))
}
```

### The detectability boundary

Because $\mathbb{E}[\log Z]$ is monotone decreasing in $p_\text{source}$, for each
$\pi_0$ there is a single crossing $p^*(\pi_0)$ at which it meets the Gaussian
baseline. Sources with $p_\text{source} < p^*$ are detectable (they sit above the
baseline, so they are maxima of the objective) and those with
$p_\text{source} > p^*$ are not. Plotting $p^*$ against $\pi_0$ therefore
summarises, in one curve, which sources each $\pi_0$ can find.

$\pi_0 = 0$ is included: that is the pure centred exponential slab with no spike,
which is perfectly well defined here ($v_0 = 1$, so $s = c$).

```{r}
p_cross = function(pi0, cc, s2) {
  gb = E_logZ_gauss(pi0, cc, s2)
  h  = function(p) E_logZ_binary(p, pi0, cc, s2) - gb
  lo = 1e-4; hi = 1 - 1e-4
  if (h(lo) < 0) return(NA_real_)   # nothing detectable
  if (h(hi) > 0) return(1)          # everything detectable
  uniroot(h, c(lo, hi), tol = 1e-10)$root
}

pi0_grid = seq(0, 0.99, length.out = 100)
p_star   = sapply(pi0_grid, p_cross, cc = c_val, s2 = s2_val)

plot(pi0_grid, p_star, type = "n", ylim = c(0, 0.5),
     xlab = expression(pi[0]), ylab = expression(p^"*"),
     main = "Detectability boundary (c^2 = sigma^2 = 0.5)")
polygon(c(pi0_grid, rev(pi0_grid)), c(p_star, rep(0, length(p_star))),
        col = "grey90", border = NA)
lines(pi0_grid, p_star, lwd = 2)
abline(h = 0.2, lty = 3, col = "red")
text(0.5, 0.13, "detectable", cex = 0.9)
text(0.5, 0.47, "not detectable", cex = 0.9, col = "grey40")
legend("bottomleft", legend = c(expression(p^"*"), "p_source = 0.2 (simulation)"),
       col = c("black", "red"), lty = c(1, 3), lwd = c(2, 1), bty = "n", cex = 0.8)
```

```{r}
knitr::kable(
  data.frame(pi_0 = c(0, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 0.95, 0.99),
             p_star = round(sapply(c(0, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 0.95, 0.99),
                                   p_cross, cc = c_val, s2 = s2_val), 4)),
  col.names = c("pi_0", "p*"),
  caption = "Largest detectable p_source for each pi_0")
```

The shape of this curve is the opposite of what one might guess. Making the prior
sparser does **not** widen the range of sparse sources that can be found: $p^*$ is
essentially flat at about 0.45 for $\pi_0 \lesssim 0.2$ and then *falls*, down to
about 0.21 by $\pi_0 = 0.99$. Increasing $\pi_0$ narrows the detectable set rather
than shifting it towards sparser sources. The reason is that the Gaussian baseline
is not fixed — it moves with $\pi_0$ too — and raising $\pi_0$ pulls the source
curves down faster than it pulls the baseline down.

Two practical consequences. First, there may be reason to keep $\pi_0$ from being too high, certainly there seems no reason to initialize it high. Second, the
boundary never reaches 0.5, which is the analytic statement of the earlier remark
that a balanced binary source will not be detected by this one-sided objective.

This also explains the somewhat counter-intuitive behaviour in the $\pi_0$ sweep in the simulation below. The true sources
there have $p_\text{source} = 0.2$ (red line): it sits comfortably inside the
detectable region for $\pi_0 \le 0.7$, is getting close to the boundary by
$\pi_0 = 0.9$ ($p^* = 0.31$), and at $\pi_0 = 0.99$ would be right on it
($p^* = 0.21$) — which matches recovery being perfect up to 0.7 and collapsing at
0.9. So maybe matching $\pi_0$ to the "true" null value is not going to be advantageous in practice? This may require more investigation.

## Simulation: nine groups with $p \approx 0.2$

The same simulation used in `04`: $K=9$ binary loading vectors, each with 20
non-zero entries among $n=100$ observations. One deliberate difference from `04`:
because we are interested in **centred sources**, `U` is *not* augmented with an intercept row. (Essentially I have moved on from my original idea of including an intercept, which I originally introduced to deal with the fact that fastICA with log-cosh cannot find asymmetric sources; we are now fixing that by using asymmetric contrast functions, which in hindsight seems much the preferable route. Indeed, the ICA theory seems to depend on centered sources, so we keep the theory this way).

```{r}
set.seed(1)
n = 100; p_dim = 1000; K = 9
L = matrix(0, n, K); for (i in 1:K) L[sample(n, 20), i] = 1
FF = matrix(rnorm(p_dim*K), p_dim)
Y  = L %*% t(FF) + matrix(rnorm(n*p_dim, 0, 0.1), n)

n.comp = 9
Y  = scale(Y, scale = FALSE)
U  = sqrt(nrow(Y)) * t(svd(Y)$u[, 1:n.comp])
```

### Fixed hyperparameters

Run 1000 starts in parallel at the default $\pi_0 = 0.5$, $c^2 = \sigma^2 = 0.5$.

```{r}
set.seed(1)
system.time(fit0 <- pe_fit(U, n_starts = 1000, n_iter = 50, trace = TRUE))

Lhat = crossprod(U, fit0$W)
bestcor = apply(abs(cor(Lhat, L)), 1, max)
plot(fit0$J, xlab = "start", ylab = "J", main = "objective by start (pi_0 = 0.5)")
plot(bestcor, fit0$J, xlab = "best |cor| with a true source", ylab = "J")

cat(sprintf("frac of starts with |cor| > 0.9 : %.2f\n", mean(bestcor > 0.9)))
cat(sprintf("|cor| at the best-objective start: %.3f\n", bestcor[which.max(fit0$J)]))
cat("best |cor| achieved for each true source:\n")
print(round(apply(abs(cor(Lhat, L)), 2, max), 3))
cat(sprintf("frac of (iteration, start) pairs where J decreased: %.5f\n",
            mean(diff(fit0$Jtrace) < -1e-8)))
```

Every one of the 1000 starts converges to a true source, all nine sources are
recovered, the best solution by objective is a true source, and the Newton update
is empirically monotone for all but a handful of steps. Plotting the two extreme
solutions:

```{r}
plot(Lhat[, which.max(fit0$J)], ylab = "x", main = "best start (by objective)")
plot(Lhat[, which.min(fit0$J)], ylab = "x", main = "worst start (by objective)")
```

### Sensitivity to $\pi_0$

```{r}
for (q in c(0, 0.02, 0.05, 0.1, 0.3, 0.5, 0.7, 0.8, 0.9)) {
  set.seed(1)
  f  = pe_fit(U, n_starts = 1000, n_iter = 50, pi0 = q, cc = sqrt(0.5), s2 = 0.5)
  Lh = crossprod(U, f$W); bcr = apply(abs(cor(Lh, L)), 1, max)
  cat(sprintf("pi_0=%.2f : frac>0.9 = %.2f   best-obj |cor| = %.3f   #sources found = %d\n",
      q, mean(bcr > 0.9), bcr[which.max(f$J)], sum(apply(abs(cor(Lh, L)), 2, max) > 0.9)))
}
```

Recovery is perfect for every $\pi_0$ from 0 up to 0.7 — including $\pi_0 = 0$,
the pure exponential slab with no spike at all — degrades slightly at
0.8, and falls apart at 0.9 (where only 6 of the 9 sources are found). So the
method is insensitive to $\pi_0$ over a wide range, and $\pi_0 = 0.5$ remains a
reasonable default. 

### Comparison with the binary prior

The binary code from `04`, run on the same data for reference:

```{r}
binary_newton_update = function(U, W, p = 0.5, c = 2/(1+sqrt(5)), s2 = c) {
  y1 = sqrt((1-p)/p); y0 = -sqrt(p/(1-p)); ey0 = c*y0; ey1 = c*y1
  P   = t(U) %*% W
  lp0 = log(1-p) + (2*P*ey0 - ey0^2)/(2*s2)
  lp1 = log(p)   + (2*P*ey1 - ey1^2)/(2*s2)
  m = pmax(lp0, lp1); e0 = exp(lp0-m); e1 = exp(lp1-m); pi1 = e1/(e0+e1)
  Mx  = (1-pi1)*ey0 + pi1*ey1
  Mpx = pi1*(1-pi1)*(ey1-ey0)^2/s2
  W = U %*% Mx - sweep(W, 2, colSums(Mpx), "*")
  sweep(W, 2, sqrt(colSums(W^2)) + 1e-15, "/")
}
binary_objective = function(U, W, p = 0.5, c = 2/(1+sqrt(5)), s2 = c) {
  y1 = sqrt((1-p)/p); y0 = -sqrt(p/(1-p)); ey0 = c*y0; ey1 = c*y1
  P   = t(U) %*% W
  lp0 = log(1-p) + (2*P*ey0 - ey0^2)/(2*s2)
  lp1 = log(p)   + (2*P*ey1 - ey1^2)/(2*s2)
  m = pmax(lp0, lp1); colSums(log(exp(lp0-m) + exp(lp1-m)) + m)
}

for (p in c(0.5, 0.25, 0.1, 0.05)) {
  set.seed(1)
  W = matrix(rnorm(nrow(U)*1000), nrow(U), 1000)
  W = sweep(W, 2, sqrt(colSums(W^2)), "/")
  for (i in 1:50) W = binary_newton_update(U, W, p = p)
  Lh = crossprod(U, W); bcr = apply(abs(cor(Lh, L)), 1, max); o = binary_objective(U, W, p = p)
  cat(sprintf("binary p_obj=%.2f : frac>0.9 = %.2f   best-obj |cor| = %.3f   #sources = %d\n",
      p, mean(bcr > 0.9), bcr[which.max(o)], sum(apply(abs(cor(Lh, L)), 2, max) > 0.9)))
}
```

Without the intercept the binary prior also does perfectly — but only at
$p_\text{obj} = 0.25$ and $0.1$. At $p_\text{obj} = 0.5$ (i.e. plain fastICA) and
at $p_\text{obj} = 0.05$ it fails completely, recovering *none* of the nine
sources from any of the 1000 starts.

So this example illustrates a key advantage of the point-exponential: **robustness to the hyperparameters**. The
point-exponential is at 100% for every $\pi_0$ in $[0, 0.7]$ (no spike at all
up to a 70% spike), whereas the binary prior has a
narrow window around $p_\text{obj} \approx 0.1$–$0.25$ and is useless on either
side of it. 

## Updating the hyperparameters

Now we turn on the updates. `b = k/2` is used so that $\nu = n - k - 2b$ matches
the EBproj convention from `03`.

```{r}
run = function(lab, ...) {
  set.seed(1); t0 = Sys.time()
  f  = pe_fit(U, n_starts = 300, n_iter = 50, b = nrow(U)/2, ...)
  Lh = crossprod(U, f$W); bcr = apply(abs(cor(Lh, L)), 1, max); im = which.max(f$F)
  cat(sprintf("%-22s frac>0.9=%.2f  bestF|cor|=%.3f  pi0=%.3f c=%.3f s2=%.4f  #src=%d  (%.1fs)\n",
      lab, mean(bcr > 0.9), bcr[im], f$pi0[im], f$c[im], f$s2[im],
      sum(apply(abs(cor(Lh, L)), 2, max) > 0.9), as.numeric(Sys.time()-t0, units = "secs")))
  invisible(f)
}

run("fixed")
run("pi0 (grid)",   update_pi0 = TRUE)
run("c (grid)",     update_c   = TRUE)
run("pi0+c (grid)", update_pi0 = TRUE, update_c = TRUE)
run("pi0+c (em)",   update_pi0 = TRUE, update_c = TRUE, g_method = "em")
run("s2 only",      update_s2  = TRUE)
fit_all_grid = run("all (grid)", update_pi0 = TRUE, update_c = TRUE, update_s2 = TRUE)
fit_all_em   = run("all (em)",   update_pi0 = TRUE, update_c = TRUE, update_s2 = TRUE,
                   g_method = "em")
```

Almost everything works: every configuration except
`all (grid)` recovers a true source from all 300 starts, and all of them put a
true source at the global maximum. Three observations.

1. **Updating $\pi_0$ and $c$ is safe and well behaved.** $\pi_0$ drifts up from
   0.5 to about 0.59–0.65 and $c$ from 0.707 to about 0.95–1.05, both plausible
   for sources that are 20% dense. 

2. **`grid` and `em` reach nearly the same answer.** The table above only shows
   the single best-objective start, so it is not on its own evidence that the two
   methods agree; the chunk below compares all 300 starts, which are paired
   (same seed, hence the same starting `W`).

3. **$\sigma^2$ drops to about 0.01, but this is correct, not a collapse.** The
   simulation is near-noiseless, so a small $\sigma^2$ is the right answer;
   `s2 only` still recovers a true source from every start. See the section
   below, where varying the noise confirms the update is tracking it. The one
   configuration that does suffer is `all (grid)`, which drops to 0.61.

```{r}
set.seed(1); fg = pe_fit(U, n_starts = 300, n_iter = 50, b = nrow(U)/2,
                         update_pi0 = TRUE, update_c = TRUE, g_method = "grid")
set.seed(1); fe = pe_fit(U, n_starts = 300, n_iter = 50, b = nrow(U)/2,
                         update_pi0 = TRUE, update_c = TRUE, g_method = "em")

cat(sprintf("within grid: pi0 sd = %.2e, c sd = %.2e  (across 300 starts)\n", sd(fg$pi0), sd(fg$c)))
cat(sprintf("within em  : pi0 sd = %.2e, c sd = %.2e\n", sd(fe$pi0), sd(fe$c)))
cat(sprintf("paired em - grid: pi0 %+.4f (max|.| %.4f),  c %+.4f (max|.| %.4f),  J %+.4f\n",
    median(fe$pi0 - fg$pi0), max(abs(fe$pi0 - fg$pi0)),
    median(fe$c   - fg$c),   max(abs(fe$c   - fg$c)), median(fe$J - fg$J)))

xg = crossprod(U, fg$W); xe = crossprod(U, fe$W)
same = sapply(seq_len(300), function(i) abs(cor(xg[, i], xe[, i])))
cat(sprintf("paired |cor(x_grid, x_em)|: median %.4f, frac > 0.999 = %.3f, min %.4f\n",
    median(same), mean(same > 0.999), min(same)))
```

Two separate things are going on here, and they are worth keeping apart.

*Within* a method the 300 starts end up at essentially the same $(\pi_0, c)$ —
the standard deviation across starts is around $10^{-5}$. That is not a
degeneracy in the code (the updates are genuinely per-start: mid-run the spread
is sd $\approx 0.02$ in $\pi_0$, and all 300 final values are distinct). It
happens because every start converges to one of the nine true sources, and in
this simulation those nine are interchangeable by construction — each is exactly
20/100 dense — so they all imply the same prior. On data with heterogeneous
sources this spread would be real, which is the reason for carrying
per-start hyperparameters at all.

*Between* methods the difference is a near-constant offset: `em` ends about
$+0.005$ higher in $\pi_0$ and $+0.012$ higher in $c$ than `grid`, i.e. under 1%
and about 1% respectively, with `em` also very slightly higher in $J$. So
"agree" here means agreement to about 1% in the hyperparameters, and the same
recovered source for 97% of the paired starts — not on all of them: 3% of pairs
end on different sources. Given that, the choice between them is essentially
about cost, and `em` is roughly 30x faster.

### Is the $\sigma^2$ update really collapsing?

```{r}
nu = ncol(U) - nrow(U) - 2*(nrow(U)/2)
set.seed(1)
Wd = matrix(rnorm(nrow(U)*5), nrow(U), 5); Wd = sweep(Wd, 2, sqrt(colSums(Wd^2)), "/")
pi0d = rep(0.5, 5); ccd = rep(sqrt(0.5), 5); s2d = rep(0.5, 5)
cat("iter |   rho (5 starts)                     |  sigma^2\n")
for (it in 1:12) {
  P = crossprod(U, Wd); st = pe_stats(P, pi0d, ccd, s2d)
  rho = colSums((U %*% st$M)^2)/(ncol(U)*colSums(st$M2))
  s2d = pe_update_s2(U, P, pi0d, ccd, s2d, nu, ncol(U))
  Wd  = pe_newton_update(U, Wd, pi0d, ccd, s2d)
  cat(sprintf("%4d | %s | %s\n", it, paste(sprintf("%.4f", rho), collapse = " "),
              paste(sprintf("%.4f", s2d), collapse = " ")))
}
```

$\rho$ climbs to about 0.99 within a dozen iterations and $\sigma^2$ falls by
almost two orders of magnitude, which looks alarming. It is not: **$\rho \approx 1$
is the right answer for this simulation.**

First, note what $\rho$ actually measures. Since $UU^T = nI$, the matrix $U^TU/n$
is the orthogonal projection $P$ onto the $k$-dimensional row space of $U$, so
$$\rho = \frac{\|UM\|^2}{n\sum_i M_i^2} = \frac{M^TPM}{M^TM},$$
the fraction of $\|M\|^2$ lying in the fitted subspace. It is a signal-fraction,
and it equals 1 when $M \in \text{span}(U)$ — not, as one might guess, only when
$M = x$.

Second, the simulation is almost noiseless: the Gaussian noise has sd 0.1 against
a signal built from $\mathcal{N}(0,1)$ factors, so the signal is over 99% of the
total variance, and the recovered sources have $|\text{cor}| = 1.000$ to four
decimal places. A signal fraction near 1 is simply correct here, and the
corresponding $\sigma^2$ really should be near 0.

The test that settles it is to vary the noise and ask whether the fitted $\rho$
and $\sigma^2$ track it:

```{r}
run_noise = function(noise_sd, seed = 1, iters = 60, n_starts = 200) {
  set.seed(seed)
  nn = 100; pp = 1000; KK = 9
  LL = matrix(0, nn, KK); for (i in 1:KK) LL[sample(nn, 20), i] = 1
  Fm  = matrix(rnorm(pp*KK), pp)
  Sig = LL %*% t(Fm)
  Yn  = scale(Sig + matrix(rnorm(nn*pp, 0, noise_sd), nn), scale = FALSE)
  sig_frac = sum(scale(Sig, scale = FALSE)^2)/sum(Yn^2)
  Un = sqrt(nn) * t(svd(Yn)$u[, 1:9])
  kk = nrow(Un); nun = nn - kk - 2*(kk/2)
  set.seed(2)
  Wn = matrix(rnorm(kk*n_starts), kk, n_starts); Wn = sweep(Wn, 2, sqrt(colSums(Wn^2)), "/")
  q = rep(0.5, n_starts); cn = rep(sqrt(0.5), n_starts); sn = rep(0.5, n_starts)
  for (it in seq_len(iters)) {
    Pn = crossprod(Un, Wn)
    sn = pe_update_s2(Un, Pn, q, cn, sn, nun, nn)
    Wn = pe_newton_update(Un, Wn, q, cn, sn)
  }
  Pn = crossprod(Un, Wn); stn = pe_stats(Pn, q, cn, sn)
  rho = colSums((Un %*% stn$M)^2)/(nn*colSums(stn$M2))
  bcr = apply(abs(cor(Pn, LL)), 1, max)
  cat(sprintf("noise sd %4.1f | signal frac %.4f | rho %.4f | sigma^2 %.4f | frac>0.9 %.2f | med |cor| %.3f\n",
      noise_sd, sig_frac, median(rho), median(sn), mean(bcr > 0.9), median(bcr)))
}
for (sd_ in c(0.1, 0.3, 1, 2, 4, 8)) run_noise(sd_)
```

$\rho$ falls monotonically from 0.99 to 0.43 as the noise rises, and $\sigma^2$
rises correspondingly from 0.01 to 0.69. The update is estimating the noise
level, and doing it sensibly. Recovery holds up to noise sd 2 and degrades
beyond, which is a property of the problem rather than of the update.

Two corrections to what one might have guessed:

- The point-exponential prior is **bounded on the left** (the atom sits at
  $-\mu$, and the slab only extends to the right), so there is no "unbounded
  support lets $M$ track $x$" story. Empirically the left boundary is not even
  active at convergence.
- The binary prior behaves the same way, not differently: running the same
  $\tau$ update with a binary prior on this data also drives $\rho$ to about
  0.985 (no intercept). Bounded support does not protect it, because nothing
  needs protecting.

There is one genuine degeneracy, but it is about the intercept rather than the
prior. At convergence $x$ is essentially two-valued (it is a binary source
recovered almost exactly, within-cluster sd $\approx 0.008$ against an overall sd
of 1.005), and *any* function of a two-valued $x$ is exactly affine in $x$. So
$M \in \text{span}(x, \mathbf{1})$. Without an intercept row $\mathbf{1}$ is
orthogonal to $\text{span}(U)$, so $\rho$ saturates just below 1. With the
intercept row $\mathbf{1} \in \text{span}(U)$, so $M$ lies exactly in the fitted
subspace and $\rho = 1$ to machine precision, sending $\sigma^2$ to its floor
regardless of the actual noise. That is a reason to leave the intercept out, and
another instance of the intercept being the thing that breaks this method.

### Recommended configuration

Based on the above: fix $\sigma^2$, update $(\pi_0, c)$ by EM.

```{r}
set.seed(1)
fit_rec = pe_fit(U, n_starts = 1000, n_iter = 50, pi0 = 0.5, cc = sqrt(0.5), s2 = 0.5,
                 b = nrow(U)/2, update_pi0 = TRUE, update_c = TRUE, g_method = "em")
Lh  = crossprod(U, fit_rec$W); bcr = apply(abs(cor(Lh, L)), 1, max)
cat(sprintf("frac>0.9 = %.2f   best-obj |cor| = %.3f   #sources = %d\n",
    mean(bcr > 0.9), bcr[which.max(fit_rec$J)], sum(apply(abs(cor(Lh, L)), 2, max) > 0.9)))
plot(bcr, fit_rec$J, xlab = "best |cor| with a true source", ylab = "J",
     main = "recommended configuration")
hist(fit_rec$pi0, breaks = 30, main = "fitted pi_0 across starts", xlab = "pi_0")
hist(fit_rec$c,   breaks = 30, main = "fitted c across starts",    xlab = "c")
```

## Summary

- The centred, scaled point-exponential prior has closed-form $\log Z$, posterior
  mean and posterior variance in terms of $\Phi$ and the inverse Mills ratio, all
  of which vectorise cleanly, so thousands of Newton starts run in parallel in
  about a second.
- Unlike the binary objective, the expected objective is monotone decreasing in
  $p_\text{source}$ for every $\pi_0$: skewed sources always beat balanced ones.
  The objective cannot detect symmetric sources at all.
- The detectability boundary $p^*(\pi_0)$ — where the expected objective crosses
  the Gaussian baseline — is flat at about 0.45 for $\pi_0 \lesssim 0.2$ and then
  *falls* to about 0.21 by $\pi_0 = 0.99$. Raising $\pi_0$ narrows the set of
  detectable sources rather than shifting it towards sparser ones, so there is no
  reason to push $\pi_0$ high; $p^*$ never reaches 0.5.
- **Dropping the intercept row matters more than any other choice here.** With it,
  recovery on the nine-group simulation was 0.82 and several hyperparameter
  updates misbehaved; without it, recovery is 1.00 and almost every configuration
  works. The degenerate solutions that dominated the earlier diagnosis were the
  intercept.
- With fixed $\pi_0 = 0.5$, $c^2 = \sigma^2 = 0.5$ every start recovers a true
  source and all nine are found. The binary prior also reaches 1.00, but only in a
  narrow window $p_\text{obj} \approx 0.1$–$0.25$; at $0.5$ and $0.05$ it finds
  nothing. The point-exponential is at 1.00 across $\pi_0 \in [0, 0.7]$, so the
  gain is **robustness to the shape hyperparameter**, not a higher ceiling.
- Updating $(\pi_0, c)$ is well behaved with the intercept gone: `grid` and `em`
  agree on $\pi_0 \approx 0.6$, $c \approx 1.0$, and `em` is ~30x cheaper.
- The EBproj $\tau$ update is **not** broken for this prior. It drives
  $\rho \to 0.99$ and $\sigma^2 \to 0.01$ here because that is correct: the
  simulation is over 99% signal. Varying the noise sd from 0.1 to 8 moves the
  fitted $\rho$ monotonically from 0.99 to 0.43 and $\sigma^2$ from 0.01 to 0.69,
  so the update is estimating the noise. The binary prior behaves the same way.
  The one real degeneracy is again the intercept: at convergence $x$ is
  two-valued, so $M$ is affine in $x$, and if $\mathbf{1} \in \text{span}(U)$
  then $\rho = 1$ exactly whatever the noise.

Open questions: whether a prior on $c$ or a constraint such as
$c^2 + \sigma^2 = 1$ maintained *during* the updates restores well-behaved
$\sigma^2$ estimation; whether the robustness to $\pi_0$ seen here holds on
harder problems (more components, sparser or more variable sources) where the
binary window might be narrower still; and whether a two-sided version
(point-exponential plus its mirror image, or point-Laplace) recovers the ability
to detect symmetric sources without losing the advantage on skewed ones.
