---
title: "FastICA for Non-Whitened Data"
author: "Matthew Stephens"
date: "2026-07-16"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

### Standard whitened FastICA

Given $X$ ($p \times n$) with thin SVD $X = UDV'$, standard FastICA
pre-whitens by setting $Z_{\text{wh}} = \sqrt{n}\,V'$ ($k \times n$), so
that the sample covariance $E[zz'] = I_k$.  It maximises $E[G(w'z)]$
subject to $\|w\|^2 = 1$ (equivalently $E[(w'z)^2] = 1$).  Because the
covariance is the identity, the Newton fixed-point update requires no matrix
inversion:
$$w \leftarrow E[z\,g(w'z)] - E[g'(w'z)]\,w, \qquad
  w \leftarrow w/\|w\|.$$

### Non-whitened FastICA

We instead work with
$$Z = DV' \quad (k \times n), \qquad
  C = E[zz'] = \tfrac{1}{n}D^2.$$
The scale-free constraint is $E[(w'z)^2] = 1$, i.e.\ $w'Cw = 1$, i.e.\
$w'D^2w = n$.  The Newton update for a constrained problem with non-identity
covariance requires left-multiplying the gradient by $C^{-1} = nD^{-2}$:
$$w \leftarrow C^{-1}E[z\,g(w'z)] - E[g'(w'z)]\,w
  = nD^{-2}E[z\,g(w'z)] - E[g'(w'z)]\,w, \qquad
  w \leftarrow w\sqrt{n}/\sqrt{w'D^2 w}.$$
In code, since $Z\mathbf{g}/n = E[zg(w'z)]$, the term $nD^{-2}E[zg]$
is computed as `(Z %*% G) / diag(D2)` (elementwise division by $d_i^2$).

### Exact equivariance

The two algorithms are **exactly equivariant** under the linear map
$$w_{\text{wh}} = C^{1/2}w = \tfrac{1}{\sqrt{n}}Dw, \qquad
  w_{\text{nw}} = C^{-1/2}w_{\text{wh}} = \sqrt{n}D^{-1}w_{\text{wh}},$$
which preserves projections: $w_{\text{wh}}'z_{\text{wh}} = w_{\text{nw}}'z_{\text{nw}}$.

**Proof.**  Substitute $z_{\text{wh}} = C^{-1/2}z = \sqrt{n}D^{-1}z$ and
$w_{\text{wh}} = C^{1/2}w_{\text{nw}} = \tfrac{D}{\sqrt{n}}w_{\text{nw}}$
into the whitened update:
$$\tfrac{D}{\sqrt{n}}\,w^+ =
  E\!\left[\sqrt{n}D^{-1}z\cdot g(w'z)\right] - E[g']\,\tfrac{D}{\sqrt{n}}w
  = \sqrt{n}D^{-1}E[zg(w'z)] - E[g']\tfrac{D}{\sqrt{n}}w.$$
Left-multiplying by $\tfrac{\sqrt{n}}{D} = \sqrt{n}D^{-1}$ recovers exactly
the non-whitened update:
$$w^+ = nD^{-2}E[zg(w'z)] - E[g']\,w. \qquad \checkmark$$
The normalisation steps also correspond: $\|w_{\text{wh}}\|=1 \Leftrightarrow
w_{\text{nw}}'D^2w_{\text{nw}} = n$.  Therefore the two iterations trace
**identical trajectories** at every step, and have the same fixed points
with the same stability.

### Why non-whitened outperforms whitened in practice

The dynamics are identical given *corresponding* initialisations, but a
**random** $w_{\text{nw}}$ does not correspond to a random $w_{\text{wh}}$.
From a random $w_{\text{nw}}$, the equivalent whitened start is
$$w_{\text{wh}} \propto \tfrac{D}{\sqrt{n}}w_{\text{nw}},$$
which is biased toward large-$d_i$ (signal) directions.  Whitened FastICA
starts from a uniformly random direction on the $k$-sphere, giving only a
$\sim k_{\text{true}}/k$ chance of landing near the signal subspace.

In other words: the non-whitened parameterisation gives every random start an
implicit warm start toward the signal, without changing the algorithm's
fixed-point structure at all.

## Motivation: robustness to the number of PCs

In practice we do not know how many independent components are present.
Standard whitening requires a hard choice of $k$: too small and we miss
components, too large and we include pure-noise PCs.  Whitening places all
$k$ PCs on equal footing, so inflating $k$ adds pure-noise directions that
are equally likely starting points for the iteration.

The non-whitened parameterisation's random initialisation is naturally biased
toward high-variance (signal) PCs, so extra noise PCs cause much less harm.
We test this by using $k = 20$ PCs throughout, even when the true number of
sources is 1, 4, or 9.

## Implementation

```{r}
# ---- whitened helpers (for comparison) ----

fastica_r1update = function(X, w) {
  w = w / sqrt(sum(w^2))
  P  = t(X) %*% w
  G  = tanh(P)
  G2 = 1 - tanh(P)^2
  w  = X %*% G - mean(G2) * ncol(X) * w
  w / sqrt(sum(w^2))
}

fastica_r1update_tlc = function(X, w, lambda = 1) {
  w = w / sqrt(sum(w^2))
  P  = t(X) %*% w
  G  = tanh(P) + 2 * lambda * abs(P)
  G2 = 1 - tanh(P)^2 + 2 * lambda * sign(P)
  w  = X %*% G - mean(G2) * ncol(X) * w
  w / sqrt(sum(w^2))
}

compute_objective     = function(X, w) mean(log(cosh(t(X) %*% w)))
compute_objective_tlc = function(X, w, lambda = 1) {
  P = t(X) %*% w
  mean(log(cosh(P)) + lambda * abs(P) * P)
}

prewhiten = function(X, n.comp) {
  X = X - rowMeans(X)
  sqrt(ncol(X)) * t(svd(X)$v[, 1:n.comp])
}

# ---- non-whitened helpers ----

# Returns list(Z, D2): Z = DV' (k x n), D2 = diag(d^2) (k x k)
preprocess_nonwhitened = function(X, n.comp) {
  X  = X - rowMeans(X)
  sv = svd(X, nu = 0, nv = n.comp)
  d  = sv$d[1:n.comp]
  Z  = diag(d, nrow = length(d)) %*% t(sv$v)   # k x n
  D2 = diag(d^2, nrow = length(d))              # k x k
  list(Z = Z, D2 = D2)
}

# Normalize w so that w'D2 w = n
normalize_nw = function(w, D2, n) {
  w * sqrt(n) / sqrt(as.numeric(t(w) %*% D2 %*% w))
}

fastica_r1update_nw = function(Z, D2, w) {
  n  = ncol(Z)
  w  = normalize_nw(w, D2, n)
  P  = t(Z) %*% w
  G  = tanh(P)
  G2 = 1 - tanh(P)^2
  w  = (Z %*% G) / diag(D2) - mean(G2) * w   # n * D2^{-1} %*% E[zg] - E[g'] * w
  normalize_nw(w, D2, n)
}

fastica_r1update_nw_tlc = function(Z, D2, w, lambda = 1) {
  n  = ncol(Z)
  w  = normalize_nw(w, D2, n)
  P  = t(Z) %*% w
  G  = tanh(P) + 2 * lambda * abs(P)
  G2 = 1 - tanh(P)^2 + 2 * lambda * sign(P)
  w  = (Z %*% G) / diag(D2) - mean(G2) * w
  normalize_nw(w, D2, n)
}

compute_objective_nw = function(Z, D2, w) {
  n = ncol(Z)
  w = normalize_nw(w, D2, n)
  mean(log(cosh(t(Z) %*% w)))
}

compute_objective_nw_tlc = function(Z, D2, w, lambda = 1) {
  n = ncol(Z)
  w = normalize_nw(w, D2, n)
  P = t(Z) %*% w
  mean(log(cosh(P)) + lambda * abs(P) * P)
}

# ---- generic runner: works for both whitened and non-whitened ----
# update_fn(data, w, ...) and obj_fn(data, w, ...) where data is a list
# (non-whitened) or a matrix (whitened).

run_seeds = function(data, S_true, update_fn, obj_fn, n_seeds = 50,
                     n_iter = 200, ...) {
  obj    = numeric(n_seeds)
  maxcor = numeric(n_seeds)
  for (seed in seq_len(n_seeds)) {
    set.seed(seed)
    w = rnorm(nrow(if (is.list(data)) data$Z else data))
    for (i in seq_len(n_iter))
      w = update_fn(data, w, ...)
    obj[seed]    = obj_fn(data, w, ...)
    proj         = if (is.list(data)) t(data$Z) %*% w else t(data) %*% w
    maxcor[seed] = max(abs(cor(t(S_true), proj)))
  }
  list(obj = obj, maxcor = maxcor)
}

# Wrappers so run_seeds gets a single 'data' argument
update_nw = function(data, w, lambda = NULL) {
  if (is.null(lambda)) fastica_r1update_nw(data$Z, data$D2, w)
  else fastica_r1update_nw_tlc(data$Z, data$D2, w, lambda = lambda)
}
obj_nw = function(data, w, lambda = NULL) {
  if (is.null(lambda)) compute_objective_nw(data$Z, data$D2, w)
  else compute_objective_nw_tlc(data$Z, data$D2, w, lambda = lambda)
}

k_pca = 20   # number of PCs used for all tests (larger than true k)
```

## Test 1: Symmetric binary source (1 true source, 20 PCs)

The data have a single Rademacher source, but we extract `k_pca = 20` PCs.
Whitening inflates all 20 to unit variance; non-whitening retains the true
signal-to-noise contrast in the singular values.

### Singular values

```{r}
set.seed(10)
n = 200; p = 1000; k_true = 1
A     = matrix(rnorm(p * k_true), nrow = p)
S_rad = matrix(sample(c(-1, 1), n, replace = TRUE), nrow = 1)
sigma = 0.1
X_rad = A %*% S_rad + matrix(rnorm(p * n, 0, sigma), nrow = p)

sv_rad = svd(X_rad - rowMeans(X_rad), nu = 0, nv = k_pca)
plot(sv_rad$d[1:k_pca], type = "b", pch = 19, cex = 0.7,
     xlab = "PC index", ylab = "singular value",
     main = "Top 20 singular values (1 true source)")
abline(v = k_true + 0.5, lty = 2, col = "red")
legend("topright", "true k", lty = 2, col = "red")
```

The first singular value is much larger than the rest, which are pure noise.
Whitening would equalise all 20; non-whitening keeps this gap intact.

### Single-seed detailed comparison

```{r}
Z_rad  = prewhiten(X_rad, k_pca)
nw_rad = preprocess_nonwhitened(X_rad, k_pca)

run_one = function(data, update_fn, obj_fn, seed = 1, n_iter = 200, ...) {
  set.seed(seed)
  w = rnorm(nrow(if (is.list(data)) data$Z else data))
  for (i in seq_len(n_iter))
    w = update_fn(data, w, ...)
  proj = if (is.list(data)) t(data$Z) %*% w else t(data) %*% w
  list(w = w, proj = as.vector(proj), obj = obj_fn(data, w, ...))
}

r_wh_lc  = run_one(Z_rad,  fastica_r1update,     compute_objective)
r_wh_tlc = run_one(Z_rad,  fastica_r1update_tlc, compute_objective_tlc, lambda = 1)
r_nw_lc  = run_one(nw_rad, update_nw,            obj_nw)
r_nw_tlc = run_one(nw_rad, update_nw,            obj_nw, lambda = 1)

cat("whitened  log-cosh: obj =", round(r_wh_lc$obj,  4),
    "  max|cor| =", round(max(abs(cor(as.vector(S_rad), r_wh_lc$proj))),  3), "\n")
cat("whitened  TLC:      obj =", round(r_wh_tlc$obj, 4),
    "  max|cor| =", round(max(abs(cor(as.vector(S_rad), r_wh_tlc$proj))), 3), "\n")
cat("nonwhiten log-cosh: obj =", round(r_nw_lc$obj,  4),
    "  max|cor| =", round(max(abs(cor(as.vector(S_rad), r_nw_lc$proj))),  3), "\n")
cat("nonwhiten TLC:      obj =", round(r_nw_tlc$obj, 4),
    "  max|cor| =", round(max(abs(cor(as.vector(S_rad), r_nw_tlc$proj))), 3), "\n")
```

### Projections w'z

```{r fig.height=6}
par(mfrow = c(2, 2))
ord = order(as.vector(S_rad))
col = ifelse(as.vector(S_rad)[ord] == 1, "tomato", "steelblue")

plot_proj = function(proj, title) {
  plot(proj[ord], col = col, pch = 19, cex = 0.5,
       main = title, xlab = "sample (sorted by true source)", ylab = "w'z",
       cex.main = 0.85)
  abline(h = 0, lty = 2)
}

plot_proj(r_wh_lc$proj,  "whitened log-cosh")
plot_proj(r_wh_tlc$proj, "whitened TLC")
plot_proj(r_nw_lc$proj,  "nonwhitened log-cosh")
plot_proj(r_nw_tlc$proj, "nonwhitened TLC")
par(mfrow = c(1, 1))
```

Samples are sorted by true source value (+1 = red, -1 = blue).
A good recovery shows two clearly separated bands.

### Do the two methods share fixed points?

If the two iterations have the same fixed points, then initialising each
method from the other's converged solution should leave it unchanged.
We run 50 seeds, converge each method independently, then continue iterating
from the other's solution and check whether the projection changes.

```{r}
n_seeds_cross = 50
results_cross = data.frame(
  seed        = seq_len(n_seeds_cross),
  wh_maxcor   = NA_real_,   # whitened from random start
  nw_maxcor   = NA_real_,   # nonwhitened from random start
  wh_from_nw  = NA_real_,   # whitened continued from nonwhitened solution
  nw_from_wh  = NA_real_    # nonwhitened continued from whitened solution
)

for (seed in seq_len(n_seeds_cross)) {
  set.seed(seed)
  w0 = rnorm(k_pca)

  # converge whitened
  w_wh = w0
  for (i in seq_len(2000)) w_wh = fastica_r1update(Z_rad, w_wh)

  # converge nonwhitened
  w_nw = w0
  for (i in seq_len(2000)) w_nw = update_nw(nw_rad, w_nw)

  # continue whitened from nonwhitened solution
  w_wh_from_nw = w_nw
  for (i in seq_len(2000)) w_wh_from_nw = fastica_r1update(Z_rad, w_wh_from_nw)

  # continue nonwhitened from whitened solution
  w_nw_from_wh = w_wh
  for (i in seq_len(2000)) w_nw_from_wh = update_nw(nw_rad, w_nw_from_wh)

  # correct transformation: u = D*w_nw = sqrt(n)*w_wh
  # so w_nw = sqrt(n)*w_wh/d  and  w_wh = d*w_nw/sqrt(n)
  d_rad = sv_rad$d[1:k_pca]
  w_wh_from_nw = d_rad * w_nw / sqrt(n)
  for (i in seq_len(2000)) w_wh_from_nw = fastica_r1update(Z_rad, w_wh_from_nw)

  w_nw_from_wh = sqrt(n) * w_wh / d_rad
  for (i in seq_len(2000)) w_nw_from_wh = update_nw(nw_rad, w_nw_from_wh)

  mc    = function(w) max(abs(cor(as.vector(S_rad), as.vector(t(Z_rad)    %*% w))))
  mc_nw = function(w) max(abs(cor(as.vector(S_rad), as.vector(t(nw_rad$Z) %*% w))))

  results_cross[seed, "wh_maxcor"]  = mc(w_wh)
  results_cross[seed, "nw_maxcor"]  = mc_nw(w_nw)
  results_cross[seed, "wh_from_nw"] = mc(w_wh_from_nw)
  results_cross[seed, "nw_from_wh"] = mc_nw(w_nw_from_wh)
}

cat("Whitened from random start:          mean max|cor| =",
    round(mean(results_cross$wh_maxcor),  3), "\n")
cat("Nonwhitened from random start:       mean max|cor| =",
    round(mean(results_cross$nw_maxcor),  3), "\n")
cat("Whitened init'd from NW solution:    mean max|cor| =",
    round(mean(results_cross$wh_from_nw), 3), "\n")
cat("Nonwhitened init'd from WH solution: mean max|cor| =",
    round(mean(results_cross$nw_from_wh), 3), "\n")
```

If the methods share fixed points, "whitened init'd from NW solution" should
stay at the NW solution (high max|cor|), and "nonwhitened init'd from WH
solution" should stay at the WH solution (low max|cor| for failing seeds).
If they do *not* share fixed points, each will move away from the other's solution.


### Many seeds (200 iterations)

```{r}
res_wh_lc  = run_seeds(Z_rad,  S_rad, fastica_r1update,     compute_objective)
res_wh_tlc = run_seeds(Z_rad,  S_rad, fastica_r1update_tlc, compute_objective_tlc, lambda = 1)
res_nw_lc  = run_seeds(nw_rad, S_rad, update_nw,            obj_nw)
res_nw_tlc = run_seeds(nw_rad, S_rad, update_nw,            obj_nw, lambda = 1)

cat("whitened  log-cosh: mean max|cor| =", round(mean(res_wh_lc$maxcor),  3),
    " frac > 0.9:", mean(res_wh_lc$maxcor  > 0.9), "\n")
cat("whitened  TLC:      mean max|cor| =", round(mean(res_wh_tlc$maxcor), 3),
    " frac > 0.9:", mean(res_wh_tlc$maxcor > 0.9), "\n")
cat("nonwhiten log-cosh: mean max|cor| =", round(mean(res_nw_lc$maxcor),  3),
    " frac > 0.9:", mean(res_nw_lc$maxcor  > 0.9), "\n")
cat("nonwhiten TLC:      mean max|cor| =", round(mean(res_nw_tlc$maxcor), 3),
    " frac > 0.9:", mean(res_nw_tlc$maxcor > 0.9), "\n")
```

## Test 2: Non-overlapping groups (4 true sources, 20 PCs)

```{r}
set.seed(1)
n = 100; p = 1000; k_true = 4
A = matrix(rnorm(p * k_true), nrow = p)
S = matrix(0, nrow = k_true, ncol = n)
S[1, 1:25] = 1; S[2, 26:50] = 1; S[3, 51:75] = 1; S[4, 76:100] = 1
sigma = 0.1
X  = A %*% S + matrix(rnorm(p * n, 0, sigma), nrow = p)

Z  = prewhiten(X, k_pca)
nw = preprocess_nonwhitened(X, k_pca)

res_wh_lc  = run_seeds(Z,  S, fastica_r1update,     compute_objective)
res_wh_tlc = run_seeds(Z,  S, fastica_r1update_tlc, compute_objective_tlc, lambda = 1)
res_nw_lc  = run_seeds(nw, S, update_nw,            obj_nw)
res_nw_tlc = run_seeds(nw, S, update_nw,            obj_nw, lambda = 1)

par(mfrow = c(2, 2))
hist(res_wh_lc$maxcor,  breaks = seq(0, 1, by = 0.05), main = "whitened log-cosh",    xlab = "max |cor|")
hist(res_wh_tlc$maxcor, breaks = seq(0, 1, by = 0.05), main = "whitened TLC",         xlab = "max |cor|")
hist(res_nw_lc$maxcor,  breaks = seq(0, 1, by = 0.05), main = "nonwhitened log-cosh", xlab = "max |cor|")
hist(res_nw_tlc$maxcor, breaks = seq(0, 1, by = 0.05), main = "nonwhitened TLC",      xlab = "max |cor|")
par(mfrow = c(1, 1))
```

## Test 3: Sparse binary sources (4 true sources, p=0.2, 20 PCs)

```{r}
set.seed(2)
n = 100; p = 1000; k_true = 4
prob_active = 0.2
A = matrix(rnorm(p * k_true), nrow = p)
S_sparse = matrix(0, nrow = k_true, ncol = n)
for (j in 1:k_true)
  S_sparse[j, sample(n, round(prob_active * n))] = 1
sigma = 0.1
X_sp = A %*% S_sparse + matrix(rnorm(p * n, 0, sigma), nrow = p)

Z_sp  = prewhiten(X_sp, k_pca)
nw_sp = preprocess_nonwhitened(X_sp, k_pca)

res_wh_lc_sp  = run_seeds(Z_sp,  S_sparse, fastica_r1update,     compute_objective,     n_seeds = 100)
res_wh_tlc_sp = run_seeds(Z_sp,  S_sparse, fastica_r1update_tlc, compute_objective_tlc, n_seeds = 100, lambda = 1)
res_nw_lc_sp  = run_seeds(nw_sp, S_sparse, update_nw,            obj_nw,                n_seeds = 100)
res_nw_tlc_sp = run_seeds(nw_sp, S_sparse, update_nw,            obj_nw,                n_seeds = 100, lambda = 1)

cat("whitened  log-cosh: mean max|cor| =", round(mean(res_wh_lc_sp$maxcor),  3),
    " frac > 0.9:", mean(res_wh_lc_sp$maxcor  > 0.9), "\n")
cat("whitened  TLC:      mean max|cor| =", round(mean(res_wh_tlc_sp$maxcor), 3),
    " frac > 0.9:", mean(res_wh_tlc_sp$maxcor > 0.9), "\n")
cat("nonwhiten log-cosh: mean max|cor| =", round(mean(res_nw_lc_sp$maxcor),  3),
    " frac > 0.9:", mean(res_nw_lc_sp$maxcor  > 0.9), "\n")
cat("nonwhiten TLC:      mean max|cor| =", round(mean(res_nw_tlc_sp$maxcor), 3),
    " frac > 0.9:", mean(res_nw_tlc_sp$maxcor > 0.9), "\n")

par(mfrow = c(2, 2))
hist(res_wh_lc_sp$maxcor,  breaks = seq(0, 1, by = 0.05), main = "whitened log-cosh",    xlab = "max |cor|")
hist(res_wh_tlc_sp$maxcor, breaks = seq(0, 1, by = 0.05), main = "whitened TLC",         xlab = "max |cor|")
hist(res_nw_lc_sp$maxcor,  breaks = seq(0, 1, by = 0.05), main = "nonwhitened log-cosh", xlab = "max |cor|")
hist(res_nw_tlc_sp$maxcor, breaks = seq(0, 1, by = 0.05), main = "nonwhitened TLC",      xlab = "max |cor|")
par(mfrow = c(1, 1))
```

## Test 4: 9 overlapping groups, 20 members each, n=100 (20 PCs, no deflation)

Single-component extraction over many seeds, same as Tests 1–3.  Each seed
finds one direction; we report the max |cor| with any of the 9 true sources.

```{r}
set.seed(1)
n = 100; p = 1000; K = 9
L  = matrix(0, nrow = n, ncol = K)
for (i in 1:K) L[sample(n, 20), i] = 1
FF = matrix(rnorm(p * K), nrow = p, ncol = K)
sigma = 0.1
X9  = t(L %*% t(FF) + matrix(rnorm(n * p, 0, sigma), nrow = n))
sv9 = svd(X9 - rowMeans(X9), nu = 0, nv = k_pca)
Z9  = prewhiten(X9, k_pca)
nw9 = preprocess_nonwhitened(X9, k_pca)
```

### Singular values

```{r}
plot(sv9$d[1:k_pca], type = "b", pch = 19, cex = 0.7,
     xlab = "PC index", ylab = "singular value",
     main = "Top 20 singular values (9 true sources)")
abline(v = K + 0.5, lty = 2, col = "red")
legend("topright", "true K", lty = 2, col = "red")
```

### Single-seed detailed comparison

```{r fig.height=8}
r9_wh_lc  = run_one(Z9,  fastica_r1update,     compute_objective)
r9_wh_tlc = run_one(Z9,  fastica_r1update_tlc, compute_objective_tlc, lambda = 1)
r9_nw_lc  = run_one(nw9, update_nw,            obj_nw)
r9_nw_tlc = run_one(nw9, update_nw,            obj_nw, lambda = 1)

cor_fmt = function(proj) round(max(abs(cor(L, proj))), 3)
cat("whitened  log-cosh: obj =", round(r9_wh_lc$obj,  4), "  max|cor| =", cor_fmt(r9_wh_lc$proj),  "\n")
cat("whitened  TLC:      obj =", round(r9_wh_tlc$obj, 4), "  max|cor| =", cor_fmt(r9_wh_tlc$proj), "\n")
cat("nonwhiten log-cosh: obj =", round(r9_nw_lc$obj,  4), "  max|cor| =", cor_fmt(r9_nw_lc$proj),  "\n")
cat("nonwhiten TLC:      obj =", round(r9_nw_tlc$obj, 4), "  max|cor| =", cor_fmt(r9_nw_tlc$proj), "\n")

lbls = c("whitened log-cosh", "whitened TLC", "nonwhitened log-cosh", "nonwhitened TLC")
par(mfrow = c(2, 2))
for (i in seq_along(lbls)) {
  res = list(r9_wh_lc, r9_wh_tlc, r9_nw_lc, r9_nw_tlc)[[i]]
  best_src = which.max(abs(cor(L, res$proj)))
  ord = order(L[, best_src])
  col = ifelse(L[ord, best_src] == 1, "tomato", "steelblue")
  plot(res$proj[ord], col = col, pch = 19, cex = 0.5,
       main = paste0(lbls[i], "\nmax|cor|=", cor_fmt(res$proj), " (src ", best_src, ")"),
       xlab = "sample (sorted by best-matching source)", ylab = "w'z", cex.main = 0.8)
  abline(h = 0, lty = 2)
}
par(mfrow = c(1, 1))
```

### Many seeds

```{r}
res9_wh_lc  = run_seeds(Z9,  t(L), fastica_r1update,     compute_objective,     n_seeds = 100)
res9_wh_tlc = run_seeds(Z9,  t(L), fastica_r1update_tlc, compute_objective_tlc, n_seeds = 100, lambda = 1)
res9_nw_lc  = run_seeds(nw9, t(L), update_nw,            obj_nw,                n_seeds = 100)
res9_nw_tlc = run_seeds(nw9, t(L), update_nw,            obj_nw,                n_seeds = 100, lambda = 1)

cat("whitened  log-cosh: mean max|cor| =", round(mean(res9_wh_lc$maxcor),  3),
    " frac > 0.9:", mean(res9_wh_lc$maxcor  > 0.9), "\n")
cat("whitened  TLC:      mean max|cor| =", round(mean(res9_wh_tlc$maxcor), 3),
    " frac > 0.9:", mean(res9_wh_tlc$maxcor > 0.9), "\n")
cat("nonwhiten log-cosh: mean max|cor| =", round(mean(res9_nw_lc$maxcor),  3),
    " frac > 0.9:", mean(res9_nw_lc$maxcor  > 0.9), "\n")
cat("nonwhiten TLC:      mean max|cor| =", round(mean(res9_nw_tlc$maxcor), 3),
    " frac > 0.9:", mean(res9_nw_tlc$maxcor > 0.9), "\n")

par(mfrow = c(2, 2))
hist(res9_wh_lc$maxcor,  breaks = seq(0, 1, by = 0.05), main = "whitened log-cosh",    xlab = "max |cor|")
hist(res9_wh_tlc$maxcor, breaks = seq(0, 1, by = 0.05), main = "whitened TLC",         xlab = "max |cor|")
hist(res9_nw_lc$maxcor,  breaks = seq(0, 1, by = 0.05), main = "nonwhitened log-cosh", xlab = "max |cor|")
hist(res9_nw_tlc$maxcor, breaks = seq(0, 1, by = 0.05), main = "nonwhitened TLC",      xlab = "max |cor|")
par(mfrow = c(1, 1))
```

### Cross-initialization test

```{r}
results_cross9 = data.frame(
  seed        = seq_len(n_seeds_cross),
  wh_maxcor   = NA_real_,
  nw_maxcor   = NA_real_,
  wh_from_nw  = NA_real_,
  nw_from_wh  = NA_real_
)

for (seed in seq_len(n_seeds_cross)) {
  set.seed(seed)
  w0 = rnorm(k_pca)

  w_wh = w0
  for (i in seq_len(2000)) w_wh = fastica_r1update_tlc(Z9, w_wh)

  w_nw = w0
  for (i in seq_len(2000)) w_nw = update_nw(nw9, w_nw, lambda = 1)

  d9 = sv9$d[1:k_pca]
  w_wh_from_nw = d9 * w_nw / sqrt(n)
  for (i in seq_len(2000)) w_wh_from_nw = fastica_r1update_tlc(Z9, w_wh_from_nw)

  w_nw_from_wh = sqrt(n) * w_wh / d9
  for (i in seq_len(2000)) w_nw_from_wh = update_nw(nw9, w_nw_from_wh, lambda = 1)

  mc_wh = function(w) max(abs(cor(L, as.vector(t(Z9)    %*% w))))
  mc_nw = function(w) max(abs(cor(L, as.vector(t(nw9$Z) %*% w))))

  results_cross9[seed, "wh_maxcor"]  = mc_wh(w_wh)
  results_cross9[seed, "nw_maxcor"]  = mc_nw(w_nw)
  results_cross9[seed, "wh_from_nw"] = mc_wh(w_wh_from_nw)
  results_cross9[seed, "nw_from_wh"] = mc_nw(w_nw_from_wh)
}

cat("Whitened from random start:          mean max|cor| =",
    round(mean(results_cross9$wh_maxcor),  3), "\n")
cat("Nonwhitened from random start:       mean max|cor| =",
    round(mean(results_cross9$nw_maxcor),  3), "\n")
cat("Whitened init'd from NW solution:    mean max|cor| =",
    round(mean(results_cross9$wh_from_nw), 3), "\n")
cat("Nonwhitened init'd from WH solution: mean max|cor| =",
    round(mean(results_cross9$nw_from_wh), 3), "\n")
```

## Test 5: 9 overlapping groups, deflation (20 PCs)

Multi-component extraction via deflation.  After finding weight $w_1$, the
next search is restricted to vectors satisfying $w \perp_{D^2} w_1$, i.e.
$w_1' D^2 w = 0$, so the extracted components are orthogonal in the
$D^2$-inner-product sense (equivalently, orthogonal projections in the
original data space).  We extract 9 components from a 20-PC space.

```{r}
fastica_deflation_nw = function(nw, k, update_fn, obj_fn, n_iter = 200,
                                n_starts = 5, ...) {
  Z  = nw$Z
  D2 = nw$D2
  n  = ncol(Z)
  W  = matrix(0, nrow(Z), k)
  for (comp in seq_len(k)) {
    best_obj = -Inf; best_w = NULL
    for (s in seq_len(n_starts)) {
      set.seed(s + comp * 1000)
      w = rnorm(nrow(Z))
      if (comp > 1) {
        for (j in 1:(comp - 1))
          w = w - as.numeric(t(W[, j]) %*% D2 %*% w) /
                  as.numeric(t(W[, j]) %*% D2 %*% W[, j]) * W[, j]
      }
      w = normalize_nw(w, D2, n)
      for (i in seq_len(n_iter)) {
        w = update_fn(nw, w, ...)
        if (comp > 1) {
          for (j in 1:(comp - 1))
            w = w - as.numeric(t(W[, j]) %*% D2 %*% w) /
                    as.numeric(t(W[, j]) %*% D2 %*% W[, j]) * W[, j]
          w = normalize_nw(w, D2, n)
        }
      }
      o = obj_fn(nw, w, ...)
      if (o > best_obj) { best_obj = o; best_w = w }
    }
    W[, comp] = best_w
  }
  W
}

fastica_deflation_wh = function(X, k, update_fn, obj_fn, n_iter = 200,
                                n_starts = 5, ...) {
  W = matrix(0, nrow(X), k)
  for (comp in seq_len(k)) {
    best_obj = -Inf; best_w = NULL
    for (s in seq_len(n_starts)) {
      set.seed(s + comp * 1000)
      w = rnorm(nrow(X))
      if (comp > 1)
        w = w - W[, 1:(comp-1), drop=FALSE] %*%
                (t(W[, 1:(comp-1), drop=FALSE]) %*% w)
      w = w / sqrt(sum(w^2))
      for (i in seq_len(n_iter)) {
        w = update_fn(X, w, ...)
        if (comp > 1)
          w = w - W[, 1:(comp-1), drop=FALSE] %*%
                  (t(W[, 1:(comp-1), drop=FALSE]) %*% w)
        w = w / sqrt(sum(w^2))
      }
      o = obj_fn(X, w, ...)
      if (o > best_obj) { best_obj = o; best_w = w }
    }
    W[, comp] = best_w
  }
  W
}

W9_wh_lc  = fastica_deflation_wh(Z9,  K, fastica_r1update,     compute_objective,     n_starts = 10)
W9_wh_tlc = fastica_deflation_wh(Z9,  K, fastica_r1update_tlc, compute_objective_tlc, n_starts = 10, lambda = 1)
W9_nw_lc  = fastica_deflation_nw(nw9, K, update_nw,            obj_nw,                n_starts = 10)
W9_nw_tlc = fastica_deflation_nw(nw9, K, update_nw,            obj_nw,                n_starts = 10, lambda = 1)

cor9_wh_lc  = cor(L, t(Z9)    %*% W9_wh_lc)
cor9_wh_tlc = cor(L, t(Z9)    %*% W9_wh_tlc)
cor9_nw_lc  = cor(L, t(nw9$Z) %*% W9_nw_lc)
cor9_nw_tlc = cor(L, t(nw9$Z) %*% W9_nw_tlc)

cat("whitened log-cosh    — max |cor| per true source:\n")
print(round(apply(abs(cor9_wh_lc),  2, max), 3))
cat("whitened TLC         — max |cor| per true source:\n")
print(round(apply(abs(cor9_wh_tlc), 2, max), 3))
cat("nonwhitened log-cosh — max |cor| per true source:\n")
print(round(apply(abs(cor9_nw_lc),  2, max), 3))
cat("nonwhitened TLC      — max |cor| per true source:\n")
print(round(apply(abs(cor9_nw_tlc), 2, max), 3))
```

Extracted projections (whitened TLC vs nonwhitened TLC):

```{r fig.height=8}
proj9_wh_tlc = t(Z9)    %*% W9_wh_tlc
proj9_nw_tlc = t(nw9$Z) %*% W9_nw_tlc

par(mfrow = c(K, 2), mar = c(1, 2, 1.5, 0.5))
for (i in 1:K) {
  plot(proj9_wh_tlc[, i], main = paste0("whitened TLC comp ",    i),
       ylab = "", xlab = "", cex.main = 0.8)
  plot(proj9_nw_tlc[, i], main = paste0("nonwhitened TLC comp ", i),
       ylab = "", xlab = "", cex.main = 0.8)
}
par(mfrow = c(1, 1))
```

## Summary

The non-whitened FastICA replaces the whitened data $Z = \sqrt{n}V'$ with
$Z = DV'$ and changes only the normalization: $\|w\|^2 = 1$ becomes
$w'D^2 w = n$.  This preserves the fixed-point structure of the algorithm.
The deflation step uses the $D^2$-inner-product for orthogonalization instead
of the standard inner product.  When $k$ is chosen larger than the true number
of components, the non-whitened approach is more robust because excess noise
PCs — which have small singular values $d_i$ — contribute little to the
objective, whereas whitening equalises all $k$ PCs and amplifies noise.
