---
title: "Preconditioned FastICA"
author: "Matthew Stephens"
date: "2026-07-17"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

This analysis explores variance-prioritized initialization for non-whitened
FastICA, extending the approach in
[fastica_nonwhitened](fastica_nonwhitened.html) and using the Tilted
Log-Cosh (TLC) contrast from [fastica_skew](fastica_skew.html).

In non-whitened FastICA we work with $Z = DV'$ ($k \times n$) and the
constraint $w'D^2 w = n$.  The Newton fixed-point update is
$$w \leftarrow nD^{-2}E[z\,g(w'z)] - E[g'(w'z)]\,w, \qquad
  w \leftarrow w\sqrt{n}/\sqrt{w'D^2 w}.$$
A random $w$ in the non-whitened parameterisation already implicitly biases
convergence toward high-variance PCs.  The goal here is to make that bias
stronger, to further improve the success rate from random starts.

We try two approaches and explain why the first fails.

## Approach 1: Step-space preconditioning (fails)

### Idea

Scale the Newton *displacement* $\Delta w = w_{target} - w$ by a
time-decaying diagonal matrix $P_t$ before applying it:
$$P_{base} = \frac{D^2}{\max d_i^2}, \quad
  \alpha_t = e^{-t/\tau}, \quad
  P_t = \alpha_t P_{base} + (1-\alpha_t)I,$$
$$w_{new} = \text{normalize}(w + P_t\,\Delta w).$$

### Implementation

```{r}
# ---- shared helpers ----

preprocess_nonwhitened = function(X, n.comp) {
  X  = X - rowMeans(X)
  sv = svd(X, nu = 0, nv = n.comp)
  d  = sv$d[1:n.comp]
  Z  = diag(d, nrow = length(d)) %*% t(sv$v)
  D2 = diag(d^2, nrow = length(d))
  list(Z = Z, D2 = D2, d = d)
}

normalize_nw = function(w, D2, n) {
  w * sqrt(n) / sqrt(as.numeric(t(w) %*% D2 %*% w))
}

newton_target_nw = function(Z, D2, w, lambda = 0) {
  d2   = diag(D2)
  proj = as.vector(t(Z) %*% w)
  G    = tanh(proj) + 2 * lambda * abs(proj)
  G2   = 1 - tanh(proj)^2 + 2 * lambda * sign(proj)
  as.vector(Z %*% G) / d2 - mean(G2) * w
}

fastica_r1update_nw = function(Z, D2, w, lambda = 0) {
  n = ncol(Z)
  w = normalize_nw(w, D2, n)
  normalize_nw(newton_target_nw(Z, D2, w, lambda), D2, n)
}

fastica_step_precond = function(Z, D2, w_init = NULL, tau = 10, lambda = 0,
                                tol = 1e-6, t_max = 500) {
  n     = ncol(Z); d2 = diag(D2)
  Pbase = d2 / max(d2)
  w = if (is.null(w_init)) rnorm(nrow(Z)) else w_init
  w = normalize_nw(w, D2, n)
  for (t in seq_len(t_max)) {
    w_target = newton_target_nw(Z, D2, w, lambda)
    Pt       = exp(-t / tau) * Pbase + (1 - exp(-t / tau))
    w_new    = normalize_nw(w + Pt * (w_target - w), D2, n)
    sim = abs(as.numeric(t(w_new) %*% D2 %*% w)) / n
    w   = w_new
    if ((1 - sim) < tol) break
  }
  w
}

k_pca    = 20
tau_vals = c(1, 5, 10, 20, 50)
```

### Data: 1 true source, 20 PCs (Rademacher)

```{r}
set.seed(10)
n = 200; p = 1000
A     = matrix(rnorm(p), nrow = p)
S_rad = matrix(sample(c(-1, 1), n, replace = TRUE), nrow = 1)
X_rad = A %*% S_rad + matrix(rnorm(p * n, 0, 0.1), nrow = p)
nw_rad = preprocess_nonwhitened(X_rad, k_pca)

plot(nw_rad$d, type = "b", pch = 19, cex = 0.7,
     xlab = "PC index", ylab = "singular value",
     main = "Singular values (1 true source, k=20)")
abline(v = 1.5, lty = 2, col = "red")
legend("topright", "true k", lty = 2, col = "red", bty = "n")
```

### Why it fails: single-seed trajectory (τ=1 vs τ=4)

```{r}
trace_step_precond = function(Z, D2, s, w_init, tau, lambda = 0, n_iter = 100) {
  n = ncol(Z); d2 = diag(D2); k = nrow(Z)
  Pbase = d2 / max(d2)
  w = normalize_nw(w_init, D2, n)
  cors = numeric(n_iter); pc_wt = matrix(0, n_iter, k)
  for (t in seq_len(n_iter)) {
    w_target = newton_target_nw(Z, D2, w, lambda)
    Pt       = exp(-t / tau) * Pbase + (1 - exp(-t / tau))
    w        = normalize_nw(w + Pt * (w_target - w), D2, n)
    cors[t]  = abs(cor(as.vector(s), as.vector(t(Z) %*% w)))
    pc_wt[t,] = as.vector(w)^2 * d2 / sum(as.vector(w)^2 * d2)
  }
  list(cors = cors, pc_wt = pc_wt)
}

trace_nw = function(Z, D2, s, w_init, lambda = 0, n_iter = 100) {
  n = ncol(Z); d2 = diag(D2); k = nrow(Z)
  w = normalize_nw(w_init, D2, n)
  cors = numeric(n_iter); pc_wt = matrix(0, n_iter, k)
  for (t in seq_len(n_iter)) {
    w = fastica_r1update_nw(Z, D2, w, lambda)
    cors[t]   = abs(cor(as.vector(s), as.vector(t(Z) %*% w)))
    pc_wt[t,] = as.vector(w)^2 * d2 / sum(as.vector(w)^2 * d2)
  }
  list(cors = cors, pc_wt = pc_wt)
}

set.seed(1); w0 = rnorm(k_pca)
tr_nw = trace_nw(nw_rad$Z, nw_rad$D2, S_rad, w0)
tr_s1 = trace_step_precond(nw_rad$Z, nw_rad$D2, S_rad, w0, tau = 1)
tr_s4 = trace_step_precond(nw_rad$Z, nw_rad$D2, S_rad, w0, tau = 4)
```

```{r}
plot(tr_nw$cors, type = "l", col = "steelblue", lwd = 2, ylim = c(0, 1),
     xlab = "iteration", ylab = "|cor(w'z, s)|",
     main = "Step-space preconditioning: correlation trajectory (seed 1)")
lines(tr_s1$cors, col = "forestgreen", lwd = 2)
lines(tr_s4$cors, col = "tomato",      lwd = 2)
abline(h = 0.9, lty = 2, col = "grey60")
legend("bottomright",
       c("standard NW", expression(tau==1), expression(tau==4)),
       col = c("steelblue", "forestgreen", "tomato"), lwd = 2, bty = "n")
```

```{r fig.height=8}
plot_pc_traj = function(pc_wt, cors, title, top = 5) {
  cols = c("tomato", "steelblue", "forestgreen", "darkorange", "purple")
  n_iter = nrow(pc_wt)
  plot(NULL, xlim = c(1, n_iter), ylim = c(0, 1),
       xlab = "iteration", ylab = "D2-weighted loading fraction", main = title)
  for (j in 1:top) lines(pc_wt[, j], col = cols[j], lwd = 1.5)
  par(new = TRUE)
  plot(cors, type = "l", lty = 2, col = "black", lwd = 2,
       xlim = c(1, n_iter), ylim = c(0, 1), axes = FALSE, xlab = "", ylab = "")
  legend("topright", c(paste0("PC", 1:top), "|cor|"),
         col = c(cols[1:top], "black"), lty = c(rep(1, top), 2),
         lwd = c(rep(1.5, top), 2), bty = "n", cex = 0.8)
}

par(mfrow = c(3, 1))
plot_pc_traj(tr_nw$pc_wt, tr_nw$cors, "Standard NW")
plot_pc_traj(tr_s1$pc_wt, tr_s1$cors, expression("Step-precond " * tau == 1))
plot_pc_traj(tr_s4$pc_wt, tr_s4$cors, expression("Step-precond " * tau == 4))
par(mfrow = c(1, 1))
```

### Why this happens

The step-space preconditioner with $P_{t,i} \approx 0$ for noise PCs
effectively **freezes** those components: $w_{new,i} \approx w_i + 0 \cdot
\Delta w_i = w_i$.  Only the PC 1 component receives a full Newton step.

The trouble is that $w_{target}$ is computed from $t(Z) w$, which depends on
*all* components of $w$, including the frozen noise PCs.  At every
iteration the PC 1 Newton update is evaluated against stale, never-updated
noise-PC coordinates.  This creates a persistent mismatch between the assumed
and true projection, causing the iterate to orbit rather than converge.
With $\tau = 1$ the preconditioner decays before this mismatch builds up;
with $\tau = 4$ it persists long enough to derail the iteration.

## Approach 2: Homotopy (continuation) method

### Idea

Instead of distorting the step direction, temporarily distort *the target*
and let it morph back to the true Newton target.  Define
$$S_t = \alpha_t \frac{\text{diag}(D^2)}{\max d_i^2} + (1-\alpha_t)\mathbf{1},
  \quad \alpha_t = e^{-t/\tau},$$
and replace the Newton target with a biased version before normalizing:
$$w_{new} = \text{normalize}(S_t \circ w_{target}).$$

**Early iterations**: minor-PC coordinates of $w_{target}$ are shrunk, so
the iterate is pulled toward high-variance PCs.

**Late iterations** ($\alpha_t \to 0$): $S_t \to \mathbf{1}$, and we
normalize the unmodified Newton target — identical to standard FastICA.

Crucially, $w_{target}$ is always computed from the current $w$ with no
component frozen.  The full Newton coupling is preserved at every step.

### Implementation

```{r}
fastica_homotopy_nw = function(Z, D2, w_init = NULL, tau = 5, lambda = 0,
                               tol = 1e-6, t_max = 500) {
  n     = ncol(Z)
  d2    = diag(D2)
  Sbase = d2 / max(d2)
  w = if (is.null(w_init)) rnorm(nrow(Z)) else w_init
  w = normalize_nw(w, D2, n)
  for (t in seq_len(t_max)) {
    w_target = newton_target_nw(Z, D2, w, lambda)
    S_t      = exp(-t / tau) * Sbase + (1 - exp(-t / tau))
    w_new    = normalize_nw(w_target * S_t, D2, n)
    sim = abs(as.numeric(t(w_new) %*% D2 %*% w)) / n
    w   = w_new
    if ((1 - sim) < tol) break
  }
  w
}

run_seeds_homotopy = function(nw, S_true, tau = 5, lambda = 0, n_seeds = 100,
                              tol = 1e-6, t_max = 500) {
  maxcor = numeric(n_seeds)
  for (seed in seq_len(n_seeds)) {
    set.seed(seed)
    w = fastica_homotopy_nw(nw$Z, nw$D2,
                             w_init  = rnorm(nrow(nw$Z)),
                             tau     = tau,
                             lambda  = lambda,
                             tol     = tol,
                             t_max   = t_max)
    proj         = as.vector(t(nw$Z) %*% w)
    maxcor[seed] = max(abs(cor(t(S_true), proj)))
  }
  maxcor
}

run_seeds_nw = function(nw, S_true, lambda = 0, n_seeds = 100, n_iter = 500) {
  maxcor = numeric(n_seeds)
  for (seed in seq_len(n_seeds)) {
    set.seed(seed)
    w = rnorm(nrow(nw$Z))
    for (i in seq_len(n_iter)) w = fastica_r1update_nw(nw$Z, nw$D2, w, lambda)
    proj         = as.vector(t(nw$Z) %*% w)
    maxcor[seed] = max(abs(cor(t(S_true), proj)))
  }
  maxcor
}
```

### Single-seed trajectory

```{r}
trace_homotopy = function(Z, D2, s, w_init, tau, lambda = 0, n_iter = 100) {
  n = ncol(Z); d2 = diag(D2); k = nrow(Z)
  Sbase = d2 / max(d2)
  w = normalize_nw(w_init, D2, n)
  cors = numeric(n_iter); pc_wt = matrix(0, n_iter, k)
  for (t in seq_len(n_iter)) {
    w_target = newton_target_nw(Z, D2, w, lambda)
    S_t      = exp(-t / tau) * Sbase + (1 - exp(-t / tau))
    w        = normalize_nw(w_target * S_t, D2, n)
    cors[t]  = abs(cor(as.vector(s), as.vector(t(Z) %*% w)))
    pc_wt[t,] = as.vector(w)^2 * d2 / sum(as.vector(w)^2 * d2)
  }
  list(cors = cors, pc_wt = pc_wt)
}

set.seed(1); w0 = rnorm(k_pca)
tr_h5  = trace_homotopy(nw_rad$Z, nw_rad$D2, S_rad, w0, tau = 5)
tr_h10 = trace_homotopy(nw_rad$Z, nw_rad$D2, S_rad, w0, tau = 10)
```

```{r}
plot(tr_nw$cors,  type = "l", col = "steelblue", lwd = 2, ylim = c(0, 1),
     xlab = "iteration", ylab = "|cor(w'z, s)|",
     main = "Homotopy: correlation trajectory (seed 1)")
lines(tr_h5$cors,  col = "forestgreen", lwd = 2)
lines(tr_h10$cors, col = "tomato",      lwd = 2)
abline(h = 0.9, lty = 2, col = "grey60")
legend("bottomright",
       c("standard NW", expression(tau==5), expression(tau==10)),
       col = c("steelblue", "forestgreen", "tomato"), lwd = 2, bty = "n")
```

```{r fig.height=6}
par(mfrow = c(2, 1))
plot_pc_traj(tr_h5$pc_wt,  tr_h5$cors,  expression("Homotopy " * tau == 5))
plot_pc_traj(tr_h10$pc_wt, tr_h10$cors, expression("Homotopy " * tau == 10))
par(mfrow = c(1, 1))
```

## Test 1: 1 true source, 20 PCs (Rademacher)

Four methods: standard NW and homotopy, each with log-cosh (λ=0) and TLC (λ=1).
As shown above, TLC has the wrong objective for the symmetric Rademacher source
and is expected to fail here; it is included to demonstrate this.

```{r}
mc1 = list(
  nw_lc  = run_seeds_nw(nw_rad,       S_rad, lambda = 0),
  nw_tlc = run_seeds_nw(nw_rad,       S_rad, lambda = 1),
  ho_lc  = run_seeds_homotopy(nw_rad, S_rad, tau = 5, lambda = 0),
  ho_tlc = run_seeds_homotopy(nw_rad, S_rad, tau = 5, lambda = 1)
)

methods1 = c("NW log-cosh", "NW TLC", "Homotopy log-cosh", "Homotopy TLC")
for (i in seq_along(mc1))
  cat(sprintf("%-20s  mean=%.3f  frac>0.9=%.2f\n",
      methods1[i], mean(mc1[[i]]), mean(mc1[[i]] > 0.9)))

par(mfrow = c(2, 2))
for (i in seq_along(mc1))
  hist(mc1[[i]], breaks = seq(0, 1, by = 0.05),
       main = methods1[i], xlab = "max |cor|")
par(mfrow = c(1, 1))
```

### Why does TLC fail on the Rademacher source?

The Rademacher source is symmetric ($P(s=+1) = P(s=-1) = 0.5$).
The TLC objective is $G(z) = \log\cosh(z) + \lambda z|z|$, where the
extra term $z|z|$ rewards *skewness*.  For a perfectly symmetric source
$E[s|s|] = E[s] = 0$, so TLC adds zero contrast at the true source
direction.  In a finite sample, however, noise PCs can appear skewed by
chance.  TLC then treats those spuriously skewed noise directions as
better optima than the symmetric true source — this is an *objective
landscape* problem, not a convergence failure.

We verify this directly: compute both objectives at the LC solution,
TLC solution, and the pure PC1 direction, and check whether TLC's
converged solution genuinely has a higher TLC score than PC1.

```{r}
obj_lc  = function(Z, D2, w) {
  w = normalize_nw(w, D2, ncol(Z))
  mean(log(cosh(as.vector(t(Z) %*% w))))
}
obj_tlc = function(Z, D2, w, lambda = 1) {
  w = normalize_nw(w, D2, ncol(Z))
  p = as.vector(t(Z) %*% w)
  mean(log(cosh(p)) + lambda * abs(p) * p)
}

k   = nrow(nw_rad$Z)
n   = ncol(nw_rad$Z)
d2v = diag(nw_rad$D2)

set.seed(1); w0 = rnorm(k_pca)
w_lc  = w0; for (i in 1:500) w_lc  = fastica_r1update_nw(nw_rad$Z, nw_rad$D2, w_lc,  lambda = 0)
w_tlc = w0; for (i in 1:500) w_tlc = fastica_r1update_nw(nw_rad$Z, nw_rad$D2, w_tlc, lambda = 1)

e1      = rep(0, k_pca); e1[1] = 1
e1_norm = e1 * sqrt(n / d2v[1])   # PC1 direction, normalised to w'D2 w = n

cat("--- Log-cosh objective ---\n")
cat("  LC solution:       ", round(obj_lc(nw_rad$Z, nw_rad$D2, w_lc),    5), "\n")
cat("  TLC solution:      ", round(obj_lc(nw_rad$Z, nw_rad$D2, w_tlc),   5), "\n")
cat("  PC1 (true source): ", round(obj_lc(nw_rad$Z, nw_rad$D2, e1_norm), 5), "\n")
cat("\n--- TLC objective (lambda=1) ---\n")
cat("  LC solution:       ", round(obj_tlc(nw_rad$Z, nw_rad$D2, w_lc),    5), "\n")
cat("  TLC solution:      ", round(obj_tlc(nw_rad$Z, nw_rad$D2, w_tlc),   5), "\n")
cat("  PC1 (true source): ", round(obj_tlc(nw_rad$Z, nw_rad$D2, e1_norm), 5), "\n")
cat("\n--- Correlation with true source ---\n")
cat("  LC solution:  ", round(abs(cor(as.vector(S_rad), as.vector(t(nw_rad$Z) %*% w_lc))),    3), "\n")
cat("  TLC solution: ", round(abs(cor(as.vector(S_rad), as.vector(t(nw_rad$Z) %*% w_tlc))),   3), "\n")
cat("  PC1:          ", round(abs(cor(as.vector(S_rad), as.vector(t(nw_rad$Z) %*% e1_norm))), 3), "\n")
```

The TLC solution has a *higher* TLC objective than PC1, confirming this
is a landscape problem: TLC has found a genuine (but wrong) optimum at a
dense, apparently-skewed noise direction.  The log-cosh objective is
higher at PC1 than at the TLC solution, so log-cosh correctly identifies
the true source.  TLC should only be used for skewed/sparse sources.

### τ sensitivity for homotopy (log-cosh)

```{r}
mc1_tau = lapply(tau_vals, function(tau)
  run_seeds_homotopy(nw_rad, S_rad, tau = tau, lambda = 0))

cat("frac > 0.9 by tau (log-cosh):\n")
for (i in seq_along(tau_vals))
  cat(" tau =", tau_vals[i], ":", round(mean(mc1_tau[[i]] > 0.9), 3), "\n")
```

## Test 2: 9 overlapping groups, 20 PCs

Standard log-cosh fails here; TLC is needed (see [fastica_skew](fastica_skew.html)).
We compare all four methods.

```{r}
set.seed(1)
n = 100; p = 1000; K = 9
L  = matrix(0, nrow = n, ncol = K)
for (i in 1:K) L[sample(n, 20), i] = 1
FF = matrix(rnorm(p * K), nrow = p)
X9 = t(L %*% t(FF) + matrix(rnorm(n * p, 0, 0.1), nrow = n))
nw9 = preprocess_nonwhitened(X9, k_pca)

mc9 = list(
  nw_lc  = run_seeds_nw(nw9,       t(L), lambda = 0),
  nw_tlc = run_seeds_nw(nw9,       t(L), lambda = 1),
  ho_lc  = run_seeds_homotopy(nw9, t(L), tau = 5, lambda = 0),
  ho_tlc = run_seeds_homotopy(nw9, t(L), tau = 5, lambda = 1)
)

methods9 = methods1
for (i in seq_along(mc9))
  cat(sprintf("%-20s  mean=%.3f  frac>0.9=%.2f\n",
      methods9[i], mean(mc9[[i]]), mean(mc9[[i]] > 0.9)))

par(mfrow = c(2, 2))
for (i in seq_along(mc9))
  hist(mc9[[i]], breaks = seq(0, 1, by = 0.05),
       main = methods9[i], xlab = "max |cor|")
par(mfrow = c(1, 1))
```

### τ sensitivity for homotopy TLC on 9-source problem

```{r}
mc9_tau = lapply(tau_vals, function(tau)
  run_seeds_homotopy(nw9, t(L), tau = tau, lambda = 1))

cat("frac > 0.9 by tau (TLC):\n")
for (i in seq_along(tau_vals))
  cat(" tau =", tau_vals[i], ":", round(mean(mc9_tau[[i]] > 0.9), 3), "\n")
```

## Summary

**Step-space preconditioning** (Approach 1) fails for $\tau \geq 4$.
When $P_{t,i} \approx 0$ for noise PCs, those components are effectively
frozen while only PC 1 is updated.  But $w_{target}$ depends on
$Z'w$ — the full projection using all components — so the PC 1 Newton
step is computed against stale noise-PC values at every iteration.
This persistent mismatch causes oscillation rather than convergence.

**Homotopy method** (Approach 2) avoids this by computing $w_{target}$
from the full current $w$, then shrinking minor-PC coordinates of that
target before normalizing.  No component is ever frozen; the full Newton
coupling is preserved.  As $\alpha_t \to 0$ the bias vanishes and we
recover standard Newton with quadratic convergence.

**TLC contrast** ($\lambda = 1$) is needed for skewed/sparse sources
(e.g. the 9-source overlapping group problem with p=0.2), where standard
log-cosh falls below the Gaussian baseline and fails.  TLC should *not*
be used for symmetric sources (e.g. Rademacher p=0.5): the $z|z|$ term
adds zero contrast at the true source but can reward finite-sample
skewness in noise directions, producing a genuinely higher TLC objective
at a wrong (noise) direction than at the true source.
