---
title: "EB FastICA with Asymmetric Prior"
author: "Matthew Stephens (using Claude)" 
date: "2026-07-29"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

Standard fastICA maximizes the log-cosh contrast function, which is equivalent
to assuming a symmetric binary (Rademacher) prior on the independent components.
For sparse or skewed sources (e.g. indicator variables where the "on" fraction
$p \ll 0.5$), the expected log-cosh contrast falls **below** the Gaussian
baseline, causing fastICA to actively avoid the true source direction.

This analysis generalizes the framework to an **asymmetric** Rademacher prior
parameterized by its up-probability $p$. We alternate between:

1. Updating the projection direction $w$ via a Newton-like fixed-point step
   (for fixed $p$).
2. Updating $p$ by maximizing the marginal log-likelihood (for fixed $w$).

At $p = 0.5$ the algorithm exactly reduces to standard log-cosh fastICA.
Two scale anchors are compared:

- **M-estimator** ($c = s$): robust but slightly over-inflated variance.
- **Golden-ratio** ($c \approx 0.618\,s$): satisfies $c^2 + cs = s^2$, so the
  assumed generative variance equals the empirical data variance.

## Helpers

```{r}
prewhiten = function(X, n.comp) {
  X = X - rowMeans(X)
  sqrt(ncol(X)) * t(svd(X)$v[, 1:n.comp])
}
```

## Implementation

With $\sigma^2 = cs$, the score function and its derivative are:

$$M(x) = \frac{y_1}{cs}\,\pi(x) + \frac{y_0}{cs}\,(1-\pi(x)), \qquad
  M'(x) = \frac{\pi(x)(1-\pi(x))}{s^2\,p(1-p)}$$

where $\pi(x) = \sigma(\Delta(x))$ is the logistic sigmoid of

$$\Delta(x) = \frac{1}{\sqrt{p(1-p)}}\!\left(\frac{x}{s}
  - \frac{c\,(1-2p)}{2s\,\sqrt{p(1-p)}}\right) + \log\frac{p}{1-p}$$

and $y_1 = c\sqrt{(1-p)/p} > 0$, $y_0 = -c\sqrt{p/(1-p)} < 0$ are the prior
support points.

```{r}
asym_score = function(x, p, c, s) {
  y1     = c * sqrt((1-p)/p)
  y0     = -c * sqrt(p/(1-p))
  kappa3 = (1 - 2*p) / sqrt(p*(1-p))
  Delta  = (x/s - c*kappa3/(2*s)) / sqrt(p*(1-p)) + log(p/(1-p))
  pi_x   = plogis(Delta)
  list(
    M  = (y1/(c*s)) * pi_x + (y0/(c*s)) * (1 - pi_x),
    Mp = pi_x * (1 - pi_x) / (s^2 * p * (1-p))
  )
}
```

The marginal log-likelihood in $p$ for fixed projections $x = Yw$:

$$J(p) = \frac{1}{n}\sum_{i=1}^n \log\!\left(
  p\,e^{\,x_i y_1/(cs)\,-\,y_1^2/(2cs)}
  + (1-p)\,e^{\,x_i y_0/(cs)\,-\,y_0^2/(2cs)}\right)$$

```{r}
asym_obj_p = function(p, x, c, s) {
  y1  = c * sqrt((1-p)/p)
  y0  = -c * sqrt(p/(1-p))
  a1  = x * y1/(c*s) - y1^2/(2*c*s)
  a0  = x * y0/(c*s) - y0^2/(2*c*s)
  lp  = log(p); l1p = log(1-p)
  m   = pmax(lp + a1, l1p + a0)
  mean(m + log(exp(lp + a1 - m) + exp(l1p + a0 - m)))
}
```

The Newton-like fixed-point update for $w$ follows the standard fastICA pattern:
$$w \leftarrow \tfrac{1}{n}Y M(x) - \overline{M'(x)}\,w, \qquad
  w \leftarrow w / \|w\|$$

```{r}
fastica_asym_r1 = function(Y, s = 1, anchor = c("M", "golden"),
                            tol = 1e-6, max_iter = 500, eps = 0.01,
                            w_init = NULL) {
  anchor = match.arg(anchor)
  c      = if (anchor == "M") s else s * (sqrt(5)-1)/2
  m = nrow(Y); n = ncol(Y)
  w = if (is.null(w_init)) rnorm(m) else w_init
  w = w / sqrt(sum(w^2))
  p = 0.5
  for (iter in seq_len(max_iter)) {
    w_old = w; p_old = p
    x  = as.vector(t(Y) %*% w)
    sc = asym_score(x, p, c, s)
    w  = as.vector(Y %*% sc$M) / n - mean(sc$Mp) * w
    w  = w / sqrt(sum(w^2))
    x   = as.vector(t(Y) %*% w)
    opt = optimize(\(pp) asym_obj_p(pp, x, c, s), c(eps, 1-eps), maximum = TRUE)
    p   = opt$maximum
    if (1 - abs(sum(w * w_old)) < tol && abs(p - p_old) < tol) break
  }
  list(w = w, p = p, iter = iter, c = c)
}
```

Standard log-cosh fastICA for comparison:

```{r}
fastica_r1update = function(X, w) {
  w = w / sqrt(sum(w^2))
  P = t(X) %*% w
  G  = tanh(P); G2 = 1 - tanh(P)^2
  w  = X %*% G - mean(G2) * ncol(X) * w
  w / sqrt(sum(w^2))
}

run_seeds_lc = function(Y, S_true, n_seeds = 100, n_iter = 200) {
  maxcor = numeric(n_seeds)
  for (seed in seq_len(n_seeds)) {
    set.seed(seed)
    w = rnorm(nrow(Y))
    for (i in seq_len(n_iter)) w = fastica_r1update(Y, w)
    maxcor[seed] = max(abs(cor(t(S_true), t(Y) %*% w)))
  }
  maxcor
}

run_seeds_asym = function(Y, S_true, anchor, n_seeds = 100) {
  maxcor = numeric(n_seeds); ps = numeric(n_seeds)
  for (seed in seq_len(n_seeds)) {
    set.seed(seed)
    res = fastica_asym_r1(Y, anchor = anchor, w_init = rnorm(nrow(Y)))
    maxcor[seed] = max(abs(cor(t(S_true), t(Y) %*% res$w)))
    ps[seed]     = min(res$p, 1 - res$p)   # probability of the rare state
  }
  list(maxcor = maxcor, p = ps)
}

# Warm-start variant: run log-cosh to convergence, then hand off to asymmetric
run_seeds_asym_warm = function(Y, S_true, anchor, n_seeds = 100,
                                n_iter_lc = 200) {
  maxcor = numeric(n_seeds); ps = numeric(n_seeds)
  for (seed in seq_len(n_seeds)) {
    set.seed(seed)
    w = rnorm(nrow(Y))
    for (i in seq_len(n_iter_lc)) w = fastica_r1update(Y, w)
    res = fastica_asym_r1(Y, anchor = anchor, w_init = w)
    maxcor[seed] = max(abs(cor(t(S_true), t(Y) %*% res$w)))
    ps[seed]     = min(res$p, 1 - res$p)
  }
  list(maxcor = maxcor, p = ps)
}
```

## Sanity check: $p = 0.5$ recovers log-cosh

At $p = 0.5$, $s = 1$: $\Delta(x) = 2x$, $\pi(x) = (1 + \tanh x)/2$, and
$M(x) = \tanh(x)$ — the standard fastICA score.

```{r}
x_grid = seq(-3, 3, length.out = 300)
sc05   = asym_score(x_grid, p = 0.5, c = 1, s = 1)

plot(x_grid, sc05$M, type = "l", col = "steelblue", lwd = 2,
     xlab = "x", ylab = "M(x)",
     main = "Score function at p = 0.5 vs tanh(x)")
lines(x_grid, tanh(x_grid), col = "tomato", lty = 2, lwd = 2)
legend("topleft", c("asymmetric M(x), p = 0.5", "tanh(x)"),
       col = c("steelblue", "tomato"), lty = c(1, 2), lwd = 2, bty = "n")
```

The curves are numerically identical.

## Theoretical motivation

When does log-cosh fail? For a standardised binary source $(\text{Bernoulli}(p)$,
zero mean, unit variance), the expected log-cosh contrast is:

```{r}
expected_logcosh = function(p) {
  z1 =  sqrt((1-p)/p); z0 = -sqrt(p/(1-p))
  p * log(cosh(z1)) + (1-p) * log(cosh(z0))
}
lc_stable = function(z) abs(z) + log1p(exp(-2*abs(z))) - log(2)
lc_gauss  = integrate(\(z) lc_stable(z) * dnorm(z), -Inf, Inf)$value

pvec = seq(0.01, 0.99, by = 0.01)
lc   = sapply(pvec, expected_logcosh)

plot(pvec, lc, type = "l", col = "steelblue", lwd = 2,
     xlab = "p  (probability of positive state)",
     ylab = "E[log cosh(x)]",
     main = "Log-cosh contrast vs Gaussian baseline")
abline(h = lc_gauss, lty = 2, col = "grey40")
legend("top", c("E[log cosh], binary source", "Gaussian baseline"),
       col = c("steelblue", "grey40"), lty = c(1, 2), lwd = 2, bty = "n")
```

Standard fastICA with log-cosh **actively avoids** sources with $p < $ about
$0.1$ or $p > 0.9$: the contrast falls below the Gaussian baseline, so the
algorithm prefers noise directions over the true source.

## Score functions for varying $p$

As $p$ decreases below 0.5 the score shifts and steepens, penalising the
positive tail more heavily — appropriate for sources that are rarely "on".

```{r}
pvec2 = c(0.05, 0.1, 0.2, 0.3, 0.5)
cols  = c("purple", "tomato", "darkorange", "steelblue", "black")
plot(NULL, xlim = c(-3, 3), ylim = c(-2.5, 2.5),
     xlab = "x", ylab = "M(x)",
     main = "Asymmetric score functions  (c = s = 1)")
for (i in seq_along(pvec2))
  lines(x_grid, asym_score(x_grid, pvec2[i], 1, 1)$M, col = cols[i], lwd = 2)
legend("topleft", paste0("p = ", pvec2), col = cols, lwd = 2, bty = "n")
abline(h = 0, lty = 3, col = "grey60")
```

## Main test: 9 overlapping groups ($k = 9$, $p \approx 0.2$)

This is the canonical case where log-cosh fails: 9 sparse binary sources
(each active in 20 out of 100 samples), whitened to $k = 9$.

```{r}
set.seed(1)
n = 100; p_dim = 1000; K = 9
L  = matrix(0, nrow = n, ncol = K)
for (i in 1:K) L[sample(n, 20), i] = 1
FF = matrix(rnorm(p_dim * K), nrow = p_dim)
X9 = t(L %*% t(FF) + matrix(rnorm(n * p_dim, 0, 0.1), nrow = n))
Z9 = prewhiten(X9, K)
S9 = t(L)
```

```{r}
mc9_lc  = run_seeds_lc(Z9,       S9, n_seeds = 100)
mc9_M   = run_seeds_asym(Z9,      S9, "M",      n_seeds = 100)
mc9_gr  = run_seeds_asym(Z9,      S9, "golden", n_seeds = 100)
mc9_grw = run_seeds_asym_warm(Z9, S9, "golden", n_seeds = 100)

cat("9-groups (p_true ~ 0.2, k = 9, n_seeds = 100):\n")
cat(sprintf("  log-cosh               mean = %.3f   frac > 0.9 = %.2f\n",
    mean(mc9_lc), mean(mc9_lc > 0.9)))
cat(sprintf("  asym M-est (random)    mean = %.3f   frac > 0.9 = %.2f   mean_p = %.3f\n",
    mean(mc9_M$maxcor),   mean(mc9_M$maxcor   > 0.9), mean(mc9_M$p)))
cat(sprintf("  asym golden (random)   mean = %.3f   frac > 0.9 = %.2f   mean_p = %.3f\n",
    mean(mc9_gr$maxcor),  mean(mc9_gr$maxcor  > 0.9), mean(mc9_gr$p)))
cat(sprintf("  asym golden (warm)     mean = %.3f   frac > 0.9 = %.2f   mean_p = %.3f\n",
    mean(mc9_grw$maxcor), mean(mc9_grw$maxcor > 0.9), mean(mc9_grw$p)))
```

```{r}
par(mfrow = c(2, 2))
hist(mc9_lc,         breaks = seq(0,1,0.05), main = "log-cosh",              xlab = "max|cor|")
hist(mc9_M$maxcor,   breaks = seq(0,1,0.05), main = "asym M-est (random)",   xlab = "max|cor|")
hist(mc9_gr$maxcor,  breaks = seq(0,1,0.05), main = "asym golden (random)",  xlab = "max|cor|")
hist(mc9_grw$maxcor, breaks = seq(0,1,0.05), main = "asym golden (warm)",    xlab = "max|cor|")
par(mfrow = c(1, 1))
```

Log-cosh fails completely. The asymmetric golden-ratio anchor succeeds from
random starts (95%), but the warm-start variant drops to 65%: log-cosh
converges to wrong directions on this problem, and the asymmetric method
inherits those bad starting points.

### Why does the golden-ratio anchor do better?

The two anchors produce different score functions via the $c$-dependent bias
term in $\Delta(x)$. The golden-ratio anchor ($c = 0.618s$) satisfies
$c^2 + cs = s^2$, so the total assumed generative variance matches the empirical
variance $s^2$. The M-estimator ($c = s$) over-inflates the assumed variance
to $2s^2$, shifting the logistic midpoint and softening the asymmetry penalty.

We can visualise this: at $p = 0.2$, $s = 1$, the two anchors produce
noticeably different score functions:

```{r}
x_grid2 = seq(-4, 4, length.out = 400)
sc_M  = asym_score(x_grid2, p = 0.2, c = 1,              s = 1)
sc_gr = asym_score(x_grid2, p = 0.2, c = (sqrt(5)-1)/2,  s = 1)

plot(x_grid2, sc_M$M,  type = "l", col = "steelblue", lwd = 2,
     xlab = "x", ylab = "M(x)",
     main = "Score functions at p = 0.2: M-estimator vs golden-ratio")
lines(x_grid2, sc_gr$M, col = "tomato", lwd = 2)
lines(x_grid2, tanh(x_grid2), col = "grey50", lty = 2, lwd = 1.5)
legend("topleft",
       c("M-estimator (c = s)", "golden-ratio (c = 0.618s)", "tanh  (p = 0.5)"),
       col = c("steelblue", "tomato", "grey50"),
       lty = c(1, 1, 2), lwd = c(2, 2, 1.5), bty = "n")
```

## Sanity check: symmetric Rademacher ($p = 0.5$, $k = 9$)

Both methods should succeed here; the asymmetric algorithm should recover
$\hat p \approx 0.5$ automatically.  We compare three variants:

- **log-cosh**: standard fastICA.
- **asym golden (random start)**: alternating optimization from a random $w$.
- **asym golden (warm start)**: log-cosh run to convergence first, then hand
  the resulting $w$ to the asymmetric optimizer.

The warm start tests whether the degradation seen with random starts is
purely an initialization issue.

```{r}
set.seed(2)
S_sym = matrix(sample(c(-1, 1), K * n, replace = TRUE), nrow = K)
X_sym = t(S_sym) %*% t(FF) + matrix(rnorm(n * p_dim, 0, 0.1), nrow = n)
Z_sym = prewhiten(t(X_sym), K)

mc_s_lc   = run_seeds_lc(Z_sym, S_sym, n_seeds = 100)
mc_s_gr   = run_seeds_asym(Z_sym,      S_sym, "golden", n_seeds = 100)
mc_s_grw  = run_seeds_asym_warm(Z_sym, S_sym, "golden", n_seeds = 100)

cat("Symmetric Rademacher (p_true = 0.5, k = 9):\n")
cat(sprintf("  log-cosh               mean = %.3f   frac > 0.9 = %.2f\n",
    mean(mc_s_lc), mean(mc_s_lc > 0.9)))
cat(sprintf("  asym golden (random)   mean = %.3f   frac > 0.9 = %.2f   mean_p = %.3f\n",
    mean(mc_s_gr$maxcor),  mean(mc_s_gr$maxcor  > 0.9), mean(mc_s_gr$p)))
cat(sprintf("  asym golden (warm)     mean = %.3f   frac > 0.9 = %.2f   mean_p = %.3f\n",
    mean(mc_s_grw$maxcor), mean(mc_s_grw$maxcor > 0.9), mean(mc_s_grw$p)))
```

## Asymmetry parameter recovery ($k = 1$)

Using $k = 1$ whitening (which isolates each source exactly), we verify that
the estimated $\hat p$ tracks the true sparse fraction. Because of sign
ambiguity in the ICA direction, we report $\min(\hat p,\, 1-\hat p)$, i.e.
the probability of the rare state.

```{r}
set.seed(99)
n_rec = 500; p_dim_rec = 1000
A_rec = matrix(rnorm(p_dim_rec), nrow = p_dim_rec)

p_trues  = c(0.05, 0.10, 0.15, 0.20, 0.30, 0.40, 0.50)
p_hat_M  = numeric(length(p_trues))
p_hat_gr = numeric(length(p_trues))

for (j in seq_along(p_trues)) {
  pt = p_trues[j]
  S  = matrix(as.numeric(runif(n_rec) < pt), nrow = 1)
  S  = (S - pt) / sqrt(pt * (1-pt))
  X  = A_rec %*% S + matrix(rnorm(p_dim_rec * n_rec, 0, 0.1), nrow = p_dim_rec)
  Z  = prewhiten(X, 1)
  set.seed(1)
  r_M  = fastica_asym_r1(Z, anchor = "M",      w_init = rnorm(1))
  r_gr = fastica_asym_r1(Z, anchor = "golden",  w_init = rnorm(1))
  p_hat_M[j]  = min(r_M$p,  1 - r_M$p)
  p_hat_gr[j] = min(r_gr$p, 1 - r_gr$p)
}

plot(p_trues, p_hat_M, pch = 19, col = "steelblue",
     xlim = c(0, 0.52), ylim = c(0, 0.52),
     xlab = "true p  (sparse fraction)",
     ylab = "estimated p  (rare-state probability)",
     main = "Asymmetry parameter recovery  (k = 1 whitening)")
points(p_trues, p_hat_gr, pch = 17, col = "tomato")
abline(0, 1, lty = 2, col = "grey50")
legend("topleft", c("M-estimator", "golden-ratio"),
       col = c("steelblue","tomato"), pch = c(19,17), bty = "n")
```

Both anchors track the true sparse fraction closely across $p \in [0.05, 0.5]$.

## Summary

The asymmetric fastICA algorithm alternates between a Newton-like fixed-point
update for $w$ (identical to standard fastICA at $p = 0.5$) and 1D
optimization of $p$. Key findings from 100 random seeds each:

| Setting | log-cosh | asym golden (random) | asym golden (warm) |
|---------|----------|------------|-------------|
| Symmetric Rademacher ($p=0.5$, $k=9$) | **0.99** | 0.77 | **0.99** |
| 9 overlapping groups ($p \approx 0.2$, $k=9$) | 0.00 | **0.95** | 0.65 |

- **Log-cosh fails completely** for the 9-groups case because sparse sources
  have $E[\log\cosh] < $ Gaussian baseline, causing the algorithm to prefer
  noise directions.
- **The golden-ratio anchor** ($c \approx 0.618s$) dramatically outperforms
  the M-estimator anchor.  Its assumed generative variance matches the empirical
  variance ($c^2 + cs = s^2$), giving a better-calibrated asymmetry penalty.
- **Warm-starting has opposite effects depending on source type**:
  - *Symmetric sources*: a random start lets $p$ drift from 0.5 before $w$
    has converged, degrading performance (77%).  Warm-starting from log-cosh
    avoids this and fully recovers 99% success.
  - *Asymmetric sources*: log-cosh converges to a wrong direction (noise PC),
    and the asymmetric method then inherits that bad start.  A random start
    performs better (95% vs 65%) because it can stumble onto the sparse source
    from a more neutral initial position.
- **Asymmetry recovery** ($k=1$ whitening): $\hat p = \min(p, 1-p)$ correctly
  tracks the true sparse fraction over the range $[0.05, 0.5]$.
- **Practical guidance**: use random starts when sources are expected to be
  sparse or asymmetric; use warm starts from log-cosh when sources are expected
  to be symmetric.
