---
title: "fastica_asymmetric_03"
author: "Matthew Stephens"
date: "2026-08-03"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

After two "quick and dirty" attempts, I think I'm now ready to implement a proper version of  an asymmetric version of fastica. We can also think of this as a Newton-based method for optimizing the Empirical Bayes Projection model (EBProj).

### The Newton version of EBCD

Given a psd matrix $S=UDU'$ and a residual variance $\sigma^2$, define the objective function $J$ as:
$$J(w, g; \sigma^2, S) = \sum_{i=1}^n \log Z((UDw)_i, g, \sigma^2)$$
where
$$Z(x, g, \sigma^2) = \int g(\theta) \exp\left( \frac{2x \theta - \theta^2}{2\sigma^2} \right) d\theta. $$
The EBCD method (for fixed $\sigma^2$) can be implemented by maximizing $J$ subject to $w'Dw=1$ using a Newton method on $w$ (and, for example, EM method for $g$). Given the optimal $w,g$, the optimal $q$ is the EBNM posterior with prior $g^*$ and data $x=UDw^*,s=\sigma$.

For $g$ a binary prior with probabilities $1-p,p$ on $y_0,y_1$ we have
$$Z(x, g, \sigma^2) = (1-p)\exp\!\left(\frac{2xy_0 - y_0^2}{2\sigma^2}\right) + p\exp\!\left(\frac{2xy_1 - y_1^2}{2\sigma^2}\right).$$

Define the log unnormalized weights $\ell_j(x) = \log p_j + (2xy_j - y_j^2)/(2\sigma^2)$, with $p_0=1-p$ and $p_1=p$. Then the posterior probability of $y_1$ is
$$\pi_1(x) = \frac{e^{\ell_1(x)}}{e^{\ell_0(x)} + e^{\ell_1(x)}},$$
the posterior mean is
$$M(x;\sigma^2,g) = (1-\pi_1)y_0 + \pi_1 y_1,$$
and the derivative of $M$ with respect to $x$ (used in the Newton step) is
$$M'(x;\sigma^2,g) = \pi_1(1-\pi_1)(y_1-y_0)^2/\sigma^2.$$


### The EBProj approach

The EBProj method works on a projection matrix $P$ (if $S$ is a whitened then $P=S/n$ is a projection matrix). The objective is:
$$F_{\text{EBproj}}(w, g; \tau, P)
= \frac{n - k - 2b}{2} \left( \log \tau - \tau + 1 \right)  + J(w,g; \sigma^2_\text{eff}, S = nP)$$
where $\sigma^2_\text{eff} =\frac{n}{(n - k - 2b)\tau}$.

Optimizing the EBProj objective can iterate between optimizing $J$ over $w,g$ as above and then optimizing $\tau = 1/(1-\hat\rho)$ where $\hat\rho = \bar{\mathbf{v}}'P\bar{\mathbf{v}} / U_q$, $U_q = E_q[\mathbf{v}'\mathbf{v}]$ is the second moment, and $P=S/n$.


## Implementation

### Helper functions for the binary prior

```{r}
# log Z(x, g, sigma2) for binary prior g = (1-p) delta_{y0} + p delta_{y1}
log_Z_binary = function(x, y_0, y_1, p, sigma2) {
  lp0 = log(1 - p) + (2*x*y_0 - y_0^2) / (2*sigma2)
  lp1 = log(p)     + (2*x*y_1 - y_1^2) / (2*sigma2)
  m   = pmax(lp0, lp1)
  log(exp(lp0 - m) + exp(lp1 - m)) + m
}

# posterior probability of y_1
post_prob1 = function(x, y_0, y_1, p, sigma2) {
  lp0 = log(1 - p) + (2*x*y_0 - y_0^2) / (2*sigma2)
  lp1 = log(p)     + (2*x*y_1 - y_1^2) / (2*sigma2)
  m   = pmax(lp0, lp1)
  e0  = exp(lp0 - m);  e1 = exp(lp1 - m)
  e1 / (e0 + e1)
}

# posterior mean M(x; sigma2, g)
post_mean_binary = function(x, y_0, y_1, p, sigma2) {
  pi1 = post_prob1(x, y_0, y_1, p, sigma2)
  (1 - pi1)*y_0 + pi1*y_1
}

# dM/dx = pi1*(1-pi1)*(y1-y0)^2 / sigma2
post_mean_deriv_binary = function(x, y_0, y_1, p, sigma2) {
  pi1 = post_prob1(x, y_0, y_1, p, sigma2)
  pi1 * (1 - pi1) * (y_1 - y_0)^2 / sigma2
}
```

### Initialisation

Define a fit object (simply a list) that contains the data $U$, and the parameters to be updated (prior $g$, weights $w$). Subsequent update functions will take a fit as input and output a new fit.

```{r}
ebproj_init = function(U, b=0, tau=1) {
  k   = ncol(U)
  n = nrow(U)
  d = rep(n, k) # So that S=UDU'= nUU' = nP
  w   = rnorm(k)
  w   = w / sqrt(sum(d * w^2))   # normalise to w'Dw = 1
  fit = list(
    U         = U,
    d         = d,
    b         = b,
    w         = w,
    c         = 1,     # scale factor; effective atoms are c*y_0, c*y_1
    y_0       = -1,
    y_1       = 1,
    p         = 0.5,
    objective = NULL,
    tau       = tau
  )
  return(fit)
}
```


This function initializes using a fastica strategy. It runs fastica to get $w$ (by setting $\sigma^2=c$) and then initializes $c$ to `mean(x tanh(x))` where $x=UDw$ are the pseudo-data (pre-shrunken source values). 
```{r}
#initializes by first running fastICA (with c=sigma^2) to get w, and then 
ebproj_init_fastica = function(U,b=0){
  fit = ebproj_init(U,b)
  n      = nrow(fit$U)
  k      = ncol(fit$U)
  nu     = n - k - 2*fit$b
  sigma2 = n / (nu * fit$tau)
  fit$c = sigma2
  fit = ebproj_fit(fit,fix_g=TRUE, fix_tau= TRUE)
  x = ebproj_calc_x(fit)
  fit$c = mean(x * tanh(x))
  fit$tau = n/(fit$c * nu) # so that sigma2 = fit$c
  return(fit)
}

ebproj_calc_x = function(fit){
  as.vector(fit$U %*% (fit$d * fit$w))
}
```

### Objective function

```{r}
ebproj_objective = function(fit) {
  n      = nrow(fit$U)
  k      = ncol(fit$U)
  nu     = n - k - 2*fit$b
  sigma2 = n / (nu * fit$tau)

  ey0 = fit$c * fit$y_0;  ey1 = fit$c * fit$y_1   # effective atoms
  x = ebproj_calc_x(fit)
  J   = sum(log_Z_binary(x, ey0, ey1, fit$p, sigma2))
  obj = (nu/2) * (log(fit$tau) - fit$tau + 1) + J
  return(obj)
}
```

The dual objective (eq 200 of the notes) is $F_{\text{EBproj}}(\mathbf{w}, g; \tau) = J(\mathbf{w}, g; \sigma^2_\text{eff}) + C(\tau)$.  It is equivalent to the primal at the optimal $q$ but is easier to compute because it avoids computing $\text{KL}(q\|g)$ explicitly.

The **primal** objective (eq 190/199) retains the full variational structure:
$$F_{\text{EBproj}}(q, g; \tau) = \frac{n - k - 2b}{2} \bigl(\log \tau - \tau(1 - \hat\rho) + 1\bigr) - \text{KL}(q \| g)$$
where $\hat\rho = \bar{\mathbf{v}}'\mathbf{P}\bar{\mathbf{v}} / U_q$, $\bar{\mathbf{v}} = E_q[\mathbf{v}]$ is the posterior mean, and $U_q = E_q[\mathbf{v}'\mathbf{v}]$ is the second moment.  The $\tau$ update targets this objective, so after a $\tau$ step the primal increases while the dual can decrease.

For the binary prior, $\text{KL}(q\|g) = \sum_i \bigl[(1-\pi_{1i})\log\tfrac{1-\pi_{1i}}{1-p} + \pi_{1i}\log\tfrac{\pi_{1i}}{p}\bigr]$ and $U_q = \sum_i \bigl[(1-\pi_{1i})(cy_0)^2 + \pi_{1i}(cy_1)^2\bigr]$.

```{r}
kl_binary = function(pi1, p) {
  pi0 = 1 - pi1
  # clamp to avoid log(0)
  pi0 = pmax(pi0, 1e-300);  pi1 = pmax(pi1, 1e-300)
  sum(pi0 * log(pi0 / (1 - p)) + pi1 * log(pi1 / p))
}

ebproj_primal_objective = function(fit) {
  n      = nrow(fit$U)
  k      = ncol(fit$U)
  nu     = n - k - 2*fit$b
  sigma2 = n / (nu * fit$tau)

  ey0 = fit$c * fit$y_0;  ey1 = fit$c * fit$y_1
  x    = ebproj_calc_x(fit)
  pi1  = post_prob1(x, ey0, ey1, fit$p, sigma2)
  vbar = post_mean_binary(x, ey0, ey1, fit$p, sigma2)

  Uv   = as.vector(t(fit$U) %*% vbar)
  Uq   = sum((1 - pi1) * ey0^2 + pi1 * ey1^2)   # E_q[v'v] second moment
  # rho_hat = vbar' P vbar / U_q, with P = S/n and S = UDU'
  rho  = sum(fit$d * Uv^2) / (n * Uq)

  kl   = kl_binary(pi1, fit$p)
  obj  = (nu/2) * (log(fit$tau) - fit$tau * (1 - rho) + 1) - kl
  return(obj)
}
```

### Prior parameterisation and update modes

The binary prior is parameterised as $(c, g_0)$ where $g_0 = (1-p)\delta_{y_0} + p\delta_{y_1}$ is a **base prior** and $c > 0$ is a **scale factor**.  The effective prior is
$$g = (1-p)\delta_{cy_0} + p\delta_{cy_1},$$
so the effective atoms are $cy_0$ and $cy_1$.  This parameterisation is intentionally over-complete — multiplying $c$ by any positive constant and dividing $y_0, y_1$ by the same constant leaves $g$ unchanged — but the redundancy is useful because it separates "shape" from "scale".

Three update modes are provided, controlled by `fix_g` and `fix_g0` in `ebproj_fit`:

| `fix_g` | `fix_g0` | behaviour |
|---------|----------|-----------|
| `TRUE`  | —        | freeze $g$ entirely; no prior update |
| `FALSE` | `TRUE`   | fix base prior $g_0 = (p, y_0, y_1)$; optimise scale $c$ only |
| `FALSE` | `FALSE`  | full EM update of $(p, y_0, y_1)$; set $c = 1$ afterwards |

**Full update** (`fix_g0 = FALSE`): standard EM, updating $p$, $y_0$, $y_1$ using the effective atoms as observations.  Because the parameterisation is over-complete we reset $c = 1$ after each step (the scale is absorbed into $y_0, y_1$).

**Scale-only update** (`fix_g0 = TRUE`): $g_0$ is held fixed and $c$ is found by exact 1-D maximisation of $J(w, cg_0)$ over $c > 0$ (optimising on the log scale).  This is useful when $g_0$ has been pre-specified (e.g., from a previous run or a known prior shape) and only the amplitude needs to be estimated.

### Update functions

Each function takes a fit object and returns a fit object.

The Newton update for $w$ (Section 2.3 of the notes) solves the constrained problem with the approximate Newton step
$$w_{\text{new}} = (\lambda I_k - \tilde{H})^{-1}(w_\text{target} - \tilde{H}w)$$
where $w_\text{target} = U^T M(x)$ and $\lambda = \max(w^T D w_\text{target},\, \max_j \tilde{H}_{jj} + \epsilon)$.  Five choices of diagonal preconditioner $\tilde{H}$ are supported (default `"ica"`):

1. **`"diag"`** (Full Diagonal): $\tilde{H}_{jj} = c_j d_j$ where $c_j = \sum_i U_{ij}^2 M'(x_i)$.
2. **`"trace"`** (Trace): $\tilde{H}_{jj} = c_\text{trace}\, d_j$ where $c_\text{trace} = \sum_i M'(x_i) S_{ii} / \operatorname{tr}(D)$ and $S_{ii} = \sum_j U_{ij}^2 d_j$.
3. **`"iso"`** (Isotropic): $\tilde{H}_{jj} = c_\text{iso}$ where $c_\text{iso} = (1/k)\sum_i M'(x_i) S_{ii}$.
4. **`"ica"`** (FastICA): $\tilde{H}_{jj} = c_\text{ica}$ where $c_\text{ica} = \sum_i M'(x_i)$ — the isotropic approximation under the uniform-feature-variance assumption $S_{ii} \approx k$ (Lemma 2.1).
5. **`"none"`** ($\tilde{H} = 0$): recovers the natural power / gradient step $w_\text{new} \propto w_\text{target}$.

```{r}
ebproj_update_w = function(fit, hessian = "ica", eps = 1e-6) {
  n      = nrow(fit$U)
  k      = ncol(fit$U)
  nu     = n - k - 2*fit$b
  sigma2 = n / (nu * fit$tau)

  ey0 = fit$c * fit$y_0;  ey1 = fit$c * fit$y_1
  x = ebproj_calc_x(fit)          # UDw, length n
  Mx       = post_mean_binary(x, ey0, ey1, fit$p, sigma2)
  Mpx      = post_mean_deriv_binary(x, ey0, ey1, fit$p, sigma2)

  w_target = as.vector(t(fit$U) %*% Mx)                    # U^T M(x), length k

  # diagonal of S: S_ii = sum_j U_ij^2 d_j
  S_diag = as.vector((fit$U^2) %*% fit$d)                  # length n

  H_diag = switch(hessian,
    diag  = { c_j = as.vector(t(fit$U^2) %*% Mpx); c_j * fit$d },
    trace = { c_t = sum(Mpx * S_diag) / sum(fit$d); c_t * fit$d },
    iso   = { c_i = sum(Mpx * S_diag) / k; rep(c_i, k) },
    ica   = { c_c = sum(Mpx); rep(c_c, k) },
    none  = rep(0, k),
    stop("hessian must be one of: 'diag', 'trace', 'iso', 'ica', 'none'")
  )

  # Lagrange multiplier; ensure lambda > max(H_diag) for positive definiteness
  lambda = max(sum(fit$d * fit$w * w_target), max(H_diag) + eps)

  w_new  = (w_target - H_diag * fit$w) / (lambda - H_diag)
  # normalise to w'Dw = 1
  scale  = sqrt(sum(fit$d * w_new^2))
  fit$w  = w_new / scale
  return(fit)
}

# Full EM update for binary prior: update (p, y_0, y_1), reset c = 1.
# Since (c, y_0, y_1, p) ~ (1, c*y_0, c*y_1, p) the parameterisation is
# over-complete; we absorb c into y_0, y_1 after each full update.
ebproj_update_g = function(fit) {
  n      = nrow(fit$U)
  k      = ncol(fit$U)
  nu     = n - k - 2*fit$b
  sigma2 = n / (nu * fit$tau)

  ey0 = fit$c * fit$y_0;  ey1 = fit$c * fit$y_1
  x = ebproj_calc_x(fit)
  
  pi1 = post_prob1(x, ey0, ey1, fit$p, sigma2)
  pi0 = 1 - pi1

  fit$p   = mean(pi1)
  fit$y_0 = sum(pi0 * x) / sum(pi0)   # effective atom absorbed into y_0
  fit$y_1 = sum(pi1 * x) / sum(pi1)   # effective atom absorbed into y_1
  fit$c   = 1                          # reset after absorbing scale
  return(fit)
}

# Scale-only update: fix the base prior g_0 = (p, y_0, y_1) and optimise c by
# maximising J(w, c*g_0) over c > 0 via 1-D exact optimisation.
ebproj_update_c = function(fit) {
  n      = nrow(fit$U)
  k      = ncol(fit$U)
  nu     = n - k - 2*fit$b
  sigma2 = n / (nu * fit$tau)

  x = ebproj_calc_x(fit)
  obj_c = function(log_c) {
    cc = exp(log_c)
    sum(log_Z_binary(x, cc * fit$y_0, cc * fit$y_1, fit$p, sigma2))
  }
  res   = optimize(obj_c, c(-5, 5), maximum = TRUE)
  fit$c = exp(res$maximum)
  return(fit)
}

# update tau using rho = vbar' P vbar / U_q, tau = 1/(1-rho)
ebproj_update_tau = function(fit) {
  n      = nrow(fit$U)
  k      = ncol(fit$U)
  nu     = n - k - 2*fit$b
  sigma2 = n / (nu * fit$tau)

  ey0  = fit$c * fit$y_0;  ey1 = fit$c * fit$y_1
  x    = ebproj_calc_x(fit)
  pi1  = post_prob1(x, ey0, ey1, fit$p, sigma2)
  vbar = post_mean_binary(x, ey0, ey1, fit$p, sigma2)

  # P = S/n = U D U'/n, so vbar' P vbar = (1/n) (U'vbar)' D (U'vbar)
  Uv      = as.vector(t(fit$U) %*% vbar)
  Uq      = sum((1 - pi1) * ey0^2 + pi1 * ey1^2)   # E_q[v'v] second moment
  rho     = sum(fit$d * Uv^2) / (n * Uq)
  fit$tau = 1 / (1 - rho)
  return(fit)
}
```



### Main fitting loop


```{r}
# Updates tau, g, w in that order (does the Newton step last. This way if w has a good initial value then the updated tau,g will be good and so the next update of w will use those good tau, g)
# fix_g   = TRUE            : skip all prior updates (freeze g entirely)
# fix_g   = FALSE, fix_g0 = TRUE  : update only c (fix base prior g_0)
# fix_g   = FALSE, fix_g0 = FALSE : full update of (p, y_0, y_1), reset c = 1
ebproj_fit = function(fit, hessian = "ica",
                      fix_w = FALSE, fix_g = FALSE, fix_g0 = FALSE, fix_tau = FALSE,
                      max_iter = 100, tol = 1e-6, verbose = FALSE) {
  fit$objective = ebproj_objective(fit)

  for (iter in seq_len(max_iter)) {
    if (!fix_tau) fit = ebproj_update_tau(fit)
    
    if (!fix_g) {
      if (fix_g0) fit = ebproj_update_c(fit)   # only scale
      else        fit = ebproj_update_g(fit)    # full (p, y_0, y_1) update
    }
    
    if (!fix_w) fit = ebproj_update_w(fit, hessian = hessian)
    
    obj_new = ebproj_objective(fit)

    if (verbose)
      cat(sprintf("iter %3d  obj = %.6f  tau = %.4f  p = %.4f  y0 = %.3f  y1 = %.3f\n",
                  iter, obj_new, fit$tau, fit$p, fit$y_0, fit$y_1))

    if (!is.null(fit$objective) && abs(obj_new - fit$objective) < tol) {
      fit$objective = obj_new
      break
    }
    fit$objective = obj_new
  }
  fit$iter = iter
  return(fit)
}
```

## Simple test 1a (p=0.5)

Generate data from a single asymmetric binary source with $p = 0.5$.
```{r}
set.seed(1)
n = 200; p_dim = 1000; k = 10

# true source: binary with p = 0.5
p_true = 0.5
s_raw  = as.numeric(runif(n) < p_true)
s_true = (s_raw - p_true) / sqrt(p_true * (1 - p_true))   # zero-mean, unit-var

# mix into high-dimensional data
a = rnorm(p_dim)
a = a / sqrt(sum(a^2))
X = outer(a, s_true) + matrix(rnorm(p_dim * n, 0, 0.1), p_dim, n)
```

The data matrix $X$ is $p \times n$. 
To whiten $X$ we center it and take the first $k$ right singular vectors ($U$ is $n \times k$).
```{r}
Xcenter = X - rowMeans(X)
sv = svd(Xcenter, nu = 0, nv = k)
U  = sv$v              # n x k
```


### Fit fastica

Now fit the model. I start by using `ebproj_init_fastica` to run fastica:
it does this by using the `ebproj` updates, but setting `sigma2=c` and updating only `w`.
The init function then sets `sigma2=c=mean(x tanh(x))`, at which point the solution should be stationary at both $w$ and $c$. Here I check this by checking that the solution does not change when I update $c$ (fix tau and g0):
```{r}
fit = ebproj_init_fastica(U,b=k/2)
fit = ebproj_fit(fit, max_iter = 200, fix_g = TRUE, fix_tau = TRUE, verbose = TRUE)
fit = ebproj_fit(fit, max_iter = 200, fix_g0 = TRUE, fix_tau = TRUE, verbose = TRUE)
# check recovery: x = UDw gives the n-vector of projections
x_hat = ebproj_calc_x(fit)
cat(sprintf("Correlation with true source: %.4f\n", abs(cor(x_hat, s_true))))
```

### Run EBproj

Now we run the full EBproj updates from the fastica solution. 
```{r}
fit = ebproj_fit(fit, max_iter = 200, verbose = TRUE)
cat(sprintf("Converged in %d iterations\n", fit$iter))
cat(sprintf("Estimated p = %.3f (true = %.3f)\n", fit$p, p_true))
cat(sprintf("y_0 = %.3f, y_1 = %.3f\n", fit$y_0, fit$y_1))

# check recovery: x = UDw gives the n-vector of projections
x_hat = ebproj_calc_x(fit)
cat(sprintf("Correlation with true source: %.4f\n", abs(cor(x_hat, s_true))))
```

### Random initialization

Here we randomly initializz and plot objective over iterations to check convergence. (Note this example appears to be converging to the null solution, where the objective is 0).
```{r}
set.seed(2)
fit_verbose = ebproj_init(U, b = 1)
objs = numeric(50)
for (iter in 1:50) {
  fit_verbose = ebproj_update_tau(fit_verbose)
  fit_verbose = ebproj_update_g(fit_verbose)
  fit_verbose = ebproj_update_w(fit_verbose)
  objs[iter]  = ebproj_objective(fit_verbose)
}
plot(objs, type = "l", xlab = "iteration", ylab = "objective",
     main = "EBProj objective over iterations")
```

This is the same but using a gradient step instead of Newton:
```{r}
set.seed(2)
fit_verbose = ebproj_init(U, b = 1)
objs = numeric(50)
for (iter in 1:50) {
  fit_verbose = ebproj_update_tau(fit_verbose)
  fit_verbose = ebproj_update_g(fit_verbose)
  fit_verbose = ebproj_update_w(fit_verbose, hessian="none")
  objs[iter]  = ebproj_objective(fit_verbose)
}
plot(objs, type = "l", xlab = "iteration", ylab = "objective",
     main = "EBProj objective over iterations")
```

Here I note that the tau update is not actually guaranteed to increase the objective (nor the primal objective). This is because the tau update is designed to increase O(q,g,tau) for a fixed q,g, but in the way I am doing things (working with w,g,tau) the q defined by (w,tau) changes implicitly when tau changes. This is OK: the new tau is optimizing O(q,g,tau) for the current implicit q (=q(current_w,current_tau) assumed fixed), and as long as we go on to update w, it will find (implicitly) the new q that optimizes O(q,g,tau) this new tau. But the potential for non-monotonic behavior of the obective (after a tau update, before w update) is something to be aware of. And we should probably add a warning if the function is called to update tau with w fixed.

```{r}
set.seed(2)
fit_verbose = ebproj_init(U, b = 1)
objs = numeric(50)
objs_primal = numeric(50)
for (iter in 1:50) {
  fit_verbose = ebproj_update_tau(fit_verbose)
  objs[iter]  = ebproj_objective(fit_verbose)
  objs_primal[iter] = ebproj_primal_objective(fit_verbose)
}
plot(objs, type = "l", xlab = "iteration", ylab = "objective",
     main = "EBProj objective over iterations")
plot(objs_primal, type = "l", xlab = "iteration", ylab = "objective",
     main = "EBProj objective over iterations")
```

### fastica init

This is the trace of the objective initializing with fastica 

```{r}
# plot objective over iterations to check convergence
set.seed(2)
fit_verbose = ebproj_init_fastica(U, b = 1)
objs = numeric(50)
for (iter in 1:50) {
  fit_verbose = ebproj_update_tau(fit_verbose)
  fit_verbose = ebproj_update_g(fit_verbose)
  fit_verbose = ebproj_update_w(fit_verbose)
  objs[iter]  = ebproj_objective(fit_verbose)
}
plot(objs, type = "l", xlab = "iteration", ylab = "objective",
     main = "EBProj objective over iterations")
```

## Simple test 1b (p=0.1)

Generate data from a single asymmetric binary source with $p = 0.1$.
```{r}
set.seed(1)
n = 200; p_dim = 1000; k = 10

# true source: binary with p = 0.1
p_true = 0.1
s_raw  = as.numeric(runif(n) < p_true)
s_true = (s_raw - p_true) / sqrt(p_true * (1 - p_true))   # zero-mean, unit-var

# mix into high-dimensional data
a = rnorm(p_dim)
a = a / sqrt(sum(a^2))
X = outer(a, s_true) + matrix(rnorm(p_dim * n, 0, 0.1), p_dim, n)
```

The data matrix $X$ is $p \times n$.
To whiten $X$ we center it and take the first $k$ right singular vectors ($U$ is $n \times k$).
```{r}
Xcenter = X - rowMeans(X)
sv = svd(Xcenter, nu = 0, nv = k)
U  = sv$v              # n x k
```

Now fit the model. I initialize at fastica solution (run fastica by setting sigma2=c and
updating only w); then set $sigma2=c=mean(x tanh(x))$ and check that the solution does not change when I update c (fix tau and g0). fastica finds the true source in this instance, but it is actually a local minimum rather than local maximum of the fastica objective. 
```{r}
fit = ebproj_init_fastica(U, b = k/2)
fit = ebproj_fit(fit, max_iter = 200, fix_g = TRUE, fix_tau = TRUE, verbose = TRUE)
fit = ebproj_fit(fit, max_iter = 200, fix_g0 = TRUE, fix_tau = TRUE, verbose = TRUE)
# check recovery: x = UDw gives the n-vector of projections
x_hat = ebproj_calc_x(fit)
cat(sprintf("Correlation with true source: %.4f\n", abs(cor(x_hat, s_true))))
```


Now I run from the fastica solution. It kind of converges (slowly) to the null solution (prior y0,y1 moving to 0). Even though the correlation with the true source remains high,
it is not really finding a good solution in my opinion; what I actually wanted it to do was to realize that the prior should be a 0.1,0.9 binary mixture and move there. I think the objective should be much higher at that point. 
```{r}
fit = ebproj_fit(fit, max_iter = 200, verbose = TRUE)
cat(sprintf("Converged in %d iterations\n", fit$iter))
cat(sprintf("Estimated p = %.3f (true = %.3f)\n", fit$p, p_true))
cat(sprintf("y_0 = %.3f, y_1 = %.3f\n", fit$y_0, fit$y_1))

# check recovery: x = UDw gives the n-vector of projections
x_hat = ebproj_calc_x(fit)
cat(sprintf("Correlation with true source: %.4f\n", abs(cor(x_hat, s_true))))
```

What is going on is that at the fastICA solution the 50-50 prior is actually very bad and delicately balanced (after all the fastICA solution is a local minimum). So we need to give it a better initialization. Here I illustrate by manually initializing $y_0,y_1$ to the minimum and maximum of x_hat. It immediately converges to a much better objective value from this fit. 
```{r}
hist(x_hat)
fit$y_0 = min(x_hat)
fit$y_1 = max(x_hat)
fit = ebproj_fit(fit, max_iter = 200, verbose = TRUE)
cat(sprintf("Converged in %d iterations\n", fit$iter))
cat(sprintf("Estimated p = %.3f (true = %.3f)\n", fit$p, p_true))
cat(sprintf("y_0 = %.3f, y_1 = %.3f\n", fit$y_0, fit$y_1))
cat(sprintf("Correlation with true source: %.4f\n", abs(cor(x_hat, s_true))))
```


Try a random start, and plot objective over iterations. Interestingly we see the non-monotonic behavior here, and it does not converge to the true source. 
```{r}
set.seed(2)
fit_verbose = ebproj_init(U, b = 1)
objs = numeric(50)
for (iter in 1:50) {
  fit_verbose = ebproj_update_tau(fit_verbose)
  fit_verbose = ebproj_update_g(fit_verbose)
  fit_verbose = ebproj_update_w(fit_verbose)
  objs[iter]  = ebproj_objective(fit_verbose)
}
plot(objs, type = "l", xlab = "iteration", ylab = "objective",
     main = "EBProj objective over iterations")
x_hat = ebproj_calc_x(fit_verbose)
cat(sprintf("Correlation with true source: %.4f\n", abs(cor(x_hat, s_true))))
hist(x_hat)
```


### Initialization with tau fixed

Try with tau fixed: it converges to a null-like solution:
```{r}
set.seed(2)
fit = ebproj_init(U, b = 1, tau = 1)
fit = ebproj_fit(fit, max_iter = 200, fix_tau=TRUE, verbose = TRUE)
```


```{r}
set.seed(2)
fit = ebproj_init(U, b = 1, tau = 1)
fit = ebproj_fit(fit, max_iter = 200, fix_tau=TRUE, verbose = TRUE)
```

