---
title: "EB FastICA with Asymmetric Prior"
author: "Matthew Stephens (using Claude)" 
date: "2026-07-29"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

Standard fastICA maximizes the log-cosh contrast function, which is equivalent
to assuming a symmetric binary (Rademacher) prior on the independent components.
For sparse or skewed sources (e.g. indicator variables where the "on" fraction
$p \ll 0.5$), the expected log-cosh contrast falls **below** the Gaussian
baseline, causing fastICA to actively avoid the true source direction.

This analysis generalizes the framework to an **asymmetric** Rademacher prior
parameterized by its up-probability $p$. We alternate between:

1. Updating the projection direction $w$ via a Newton-like fixed-point step
   (for fixed $p$).
2. Updating $p$ by maximizing the marginal log-likelihood (for fixed $w$).

At $p = 0.5$ the algorithm exactly reduces to standard log-cosh fastICA.
Two scale anchors are compared:

- **M-estimator** ($c = s$): robust but slightly over-inflated variance.
- **Golden-ratio** ($c \approx 0.618\,s$): satisfies $c^2 + cs = s^2$, so the
  assumed generative variance equals the empirical data variance.

## Helpers

```{r}
prewhiten = function(X, n.comp) {
  X = X - rowMeans(X)
  sqrt(ncol(X)) * t(svd(X)$v[, 1:n.comp])
}
```

## Implementation

With $\sigma^2 = cs$, the score function and its derivative are:

$$M(x) = \frac{y_1}{cs}\,\pi(x) + \frac{y_0}{cs}\,(1-\pi(x)), \qquad
  M'(x) = \frac{\pi(x)(1-\pi(x))}{s^2\,p(1-p)}$$

where $\pi(x) = \sigma(\Delta(x))$ is the logistic sigmoid of

$$\Delta(x) = \frac{1}{\sqrt{p(1-p)}}\!\left(\frac{x}{s}
  - \frac{c\,(1-2p)}{2s\,\sqrt{p(1-p)}}\right) + \log\frac{p}{1-p}$$

and $y_1 = c\sqrt{(1-p)/p} > 0$, $y_0 = -c\sqrt{p/(1-p)} < 0$ are the prior
support points.

```{r}
asym_score = function(x, p, c, s) {
  y1     = c * sqrt((1-p)/p)
  y0     = -c * sqrt(p/(1-p))
  kappa3 = (1 - 2*p) / sqrt(p*(1-p))
  Delta  = (x/s - c*kappa3/(2*s)) / sqrt(p*(1-p)) + log(p/(1-p))
  pi_x   = plogis(Delta)
  list(
    M  = (y1/(c*s)) * pi_x + (y0/(c*s)) * (1 - pi_x),
    Mp = pi_x * (1 - pi_x) / (s^2 * p * (1-p))
  )
}
```

The marginal log-likelihood in $p$ for fixed projections $x = Yw$:

$$J(p) = \frac{1}{n}\sum_{i=1}^n \log\!\left(
  p\,e^{\,x_i y_1/(cs)\,-\,y_1^2/(2cs)}
  + (1-p)\,e^{\,x_i y_0/(cs)\,-\,y_0^2/(2cs)}\right)$$

```{r}
asym_obj_p = function(p, x, c, s) {
  y1  = c * sqrt((1-p)/p)
  y0  = -c * sqrt(p/(1-p))
  a1  = x * y1/(c*s) - y1^2/(2*c*s)
  a0  = x * y0/(c*s) - y0^2/(2*c*s)
  lp  = log(p); l1p = log(1-p)
  m   = pmax(lp + a1, l1p + a0)
  mean(m + log(exp(lp + a1 - m) + exp(l1p + a0 - m)))
}
```

Two diagonal Hessian approximations are supported, following the notation
in [ebproj_newton](ebproj_newton.html):

- **fastICA**: $\tilde H = \overline{M'(x)}\,\mathbf{I}$ — uniform sample weights.
- **trace**: $\tilde H = c_\text{trace}\,\mathbf{I}$ where
  $c_\text{trace} = \tfrac{1}{k}\sum_i M'(x_i)\,S_{ii}$ and
  $S_{ii} = \tfrac{1}{n}\|Y_{:i}\|^2$ is the squared distance of sample $i$
  from the origin (with $\sum_i S_{ii} = k$ for whitened data).

Both reduce to the same Newton fixed-point update structure:
$$w \leftarrow \tfrac{1}{n}Y M(x) - \tilde H\,w, \qquad w \leftarrow w / \|w\|$$

```{r}
fastica_asym_r1 = function(Y, s = 1, anchor = c("M", "golden"),
                            hess = c("fastICA", "trace"),
                            tol = 1e-6, max_iter = 500, eps = 0.01,
                            w_init = NULL) {
  anchor = match.arg(anchor)
  hess   = match.arg(hess)
  c      = if (anchor == "M") s else s * (sqrt(5)-1)/2
  m = nrow(Y); n = ncol(Y)
  S_diag = colSums(Y^2) / n   # S_ii = ||Y[:,i]||^2 / n  (sum = m)
  w = if (is.null(w_init)) rnorm(m) else w_init
  w = w / sqrt(sum(w^2))
  p = 0.5
  for (iter in seq_len(max_iter)) {
    w_old = w; p_old = p
    x  = as.vector(t(Y) %*% w)
    sc = asym_score(x, p, c, s)
    h  = if (hess == "trace") sum(sc$Mp * S_diag) / m else mean(sc$Mp)
    w  = as.vector(Y %*% sc$M) / n - h * w
    w  = w / sqrt(sum(w^2))
    x   = as.vector(t(Y) %*% w)
    opt = optimize(\(pp) asym_obj_p(pp, x, c, s), c(eps, 1-eps), maximum = TRUE)
    p   = opt$maximum
    if (1 - abs(sum(w * w_old)) < tol && abs(p - p_old) < tol) break
  }
  list(w = w, p = p, iter = iter, c = c)
}
```

Standard log-cosh fastICA for comparison:

```{r}
fastica_r1update = function(X, w) {
  w = w / sqrt(sum(w^2))
  P  = t(X) %*% w
  G  = tanh(P); G2 = 1 - tanh(P)^2
  w  = X %*% G - mean(G2) * ncol(X) * w
  w / sqrt(sum(w^2))
}

fastica_r1update_trace = function(X, w, S_diag) {
  w = w / sqrt(sum(w^2))
  P  = t(X) %*% w
  G  = tanh(P); G2 = 1 - tanh(P)^2
  h  = sum(G2 * S_diag) / nrow(X)   # trace Hessian: Σ G2_i S_ii / k
  w  = as.vector(X %*% G) / ncol(X) - h * w
  w / sqrt(sum(w^2))
}

run_seeds_lc = function(Y, S_true, hess = "fastICA", n_seeds = 100, n_iter = 200) {
  maxcor = numeric(n_seeds)
  objs   = numeric(n_seeds)
  S_diag = colSums(Y^2) / ncol(Y)
  for (seed in seq_len(n_seeds)) {
    set.seed(seed)
    w = rnorm(nrow(Y))
    if (hess == "trace") {
      for (i in seq_len(n_iter)) w = fastica_r1update_trace(Y, w, S_diag)
    } else {
      for (i in seq_len(n_iter)) w = fastica_r1update(Y, w)
    }
    maxcor[seed] = max(abs(cor(t(S_true), t(Y) %*% w)))
    objs[seed]   = mean(log(cosh(as.vector(t(Y) %*% w))))
  }
  structure(maxcor, obj = objs)   # vector; objs accessible via attr(., "obj")
}

run_seeds_asym = function(Y, S_true, anchor, hess = "fastICA", n_seeds = 100) {
  maxcor = numeric(n_seeds); ps = numeric(n_seeds); objs = numeric(n_seeds)
  for (seed in seq_len(n_seeds)) {
    set.seed(seed)
    res = fastica_asym_r1(Y, anchor = anchor, hess = hess, w_init = rnorm(nrow(Y)))
    maxcor[seed] = max(abs(cor(t(S_true), t(Y) %*% res$w)))
    ps[seed]     = min(res$p, 1 - res$p)
    objs[seed]   = asym_obj_p(res$p, as.vector(t(Y) %*% res$w), res$c, 1)
  }
  list(maxcor = maxcor, p = ps, obj = objs)
}

# Warm-start variant: run log-cosh to convergence, then hand off to asymmetric
run_seeds_asym_warm = function(Y, S_true, anchor, hess = "fastICA", n_seeds = 100,
                                n_iter_lc = 200) {
  maxcor = numeric(n_seeds); ps = numeric(n_seeds); objs = numeric(n_seeds)
  for (seed in seq_len(n_seeds)) {
    set.seed(seed)
    w = rnorm(nrow(Y))
    for (i in seq_len(n_iter_lc)) w = fastica_r1update(Y, w)
    res = fastica_asym_r1(Y, anchor = anchor, hess = hess, w_init = w)
    maxcor[seed] = max(abs(cor(t(S_true), t(Y) %*% res$w)))
    ps[seed]     = min(res$p, 1 - res$p)
    objs[seed]   = asym_obj_p(res$p, as.vector(t(Y) %*% res$w), res$c, 1)
  }
  list(maxcor = maxcor, p = ps, obj = objs)
}

# Sort by |obj − Gaussian baseline| (most extreme first), colour by success
plot_obj_lc = function(lc_result, title, lc_gauss) {
  objs = attr(lc_result, "obj")
  succ = lc_result > 0.9
  ord  = order(abs(objs - lc_gauss), decreasing = TRUE)
  cols = ifelse(succ[ord], "steelblue", "tomato")
  plot(seq_along(objs), objs[ord],
       pch = 19, cex = 0.7, col = cols,
       xlab = "rank (1 = most extreme log-cosh)",
       ylab = "log-cosh objective",
       main = paste0(title, "\n(blue = found true source, red = not)"))
  abline(h = lc_gauss, lty = 2, col = "grey50")
  legend("topright", c("true source", "other", "Gaussian baseline"),
         pch = c(19, 19, NA), lty = c(NA, NA, 2),
         col = c("steelblue", "tomato", "grey50"), bty = "n", cex = 0.8)
}

plot_obj_asym = function(asym_result, title) {
  objs = asym_result$obj
  succ = asym_result$maxcor > 0.9
  ord  = order(objs, decreasing = TRUE)
  cols = ifelse(succ[ord], "steelblue", "tomato")
  plot(seq_along(objs), objs[ord],
       pch = 19, cex = 0.7, col = cols,
       xlab = "rank (1 = highest asymmetric objective)",
       ylab = "asymmetric log-likelihood",
       main = paste0(title, "\n(blue = found true source, red = not)"))
  legend("topright", c("true source", "other"),
         pch = 19, col = c("steelblue", "tomato"), bty = "n", cex = 0.8)
}
```

## Sanity check: $p = 0.5$ recovers log-cosh

At $p = 0.5$, $s = 1$: $\Delta(x) = 2x$, $\pi(x) = (1 + \tanh x)/2$, and
$M(x) = \tanh(x)$ — the standard fastICA score.

```{r}
x_grid = seq(-3, 3, length.out = 300)
sc05   = asym_score(x_grid, p = 0.5, c = 1, s = 1)

plot(x_grid, sc05$M, type = "l", col = "steelblue", lwd = 2,
     xlab = "x", ylab = "M(x)",
     main = "Score function at p = 0.5 vs tanh(x)")
lines(x_grid, tanh(x_grid), col = "tomato", lty = 2, lwd = 2)
legend("topleft", c("asymmetric M(x), p = 0.5", "tanh(x)"),
       col = c("steelblue", "tomato"), lty = c(1, 2), lwd = 2, bty = "n")
```

The curves are numerically identical.

## Theoretical motivation

When does log-cosh fail? For a standardised binary source $(\text{Bernoulli}(p)$,
zero mean, unit variance), the expected log-cosh contrast is:

```{r}
expected_logcosh = function(p) {
  z1 =  sqrt((1-p)/p); z0 = -sqrt(p/(1-p))
  p * log(cosh(z1)) + (1-p) * log(cosh(z0))
}
lc_stable = function(z) abs(z) + log1p(exp(-2*abs(z))) - log(2)
lc_gauss  = integrate(\(z) lc_stable(z) * dnorm(z), -Inf, Inf)$value

pvec = seq(0.01, 0.99, by = 0.01)
lc   = sapply(pvec, expected_logcosh)

plot(pvec, lc, type = "l", col = "steelblue", lwd = 2,
     xlab = "p  (probability of positive state)",
     ylab = "E[log cosh(x)]",
     main = "Log-cosh contrast vs Gaussian baseline")
abline(h = lc_gauss, lty = 2, col = "grey40")
legend("top", c("E[log cosh], binary source", "Gaussian baseline"),
       col = c("steelblue", "grey40"), lty = c(1, 2), lwd = 2, bty = "n")
```

Standard fastICA with log-cosh **actively avoids** sources with $p < $ about
$0.25$ or $p > 0.75$: the contrast falls below the Gaussian baseline, so the
global maximum of $E[\log\cosh]$ on the sphere is not at the true source.
However, the true source direction can still be a **local minimum**, and the
fixed-point iteration can converge there — especially when $p$ is small enough
to create a deep, distinctive basin (see the $p=0.1$ nine-groups result below).

### Asymmetric objective for $p_\text{true} = 0.2$

Unlike log-cosh, the asymmetric objective can be evaluated at the optimal test
$p$ matching the true source, placing it **above** the Gaussian baseline.

```{r}
set.seed(42)
n_pop  = 100000
c_gr   = (sqrt(5) - 1) / 2   # golden-ratio anchor
z_02   = ifelse(runif(n_pop) < 0.2, sqrt(0.8/0.2), -sqrt(0.2/0.8))
z_N    = rnorm(n_pop)

p_grid      = seq(0.02, 0.98, by = 0.01)
obj_asym_02 = sapply(p_grid, \(p) asym_obj_p(p, z_02, c_gr, 1))
obj_asym_N  = sapply(p_grid, \(p) asym_obj_p(p, z_N,  c_gr, 1))

plot(p_grid, obj_asym_02, type = "l", col = "steelblue", lwd = 2,
     xlab = "test p  (optimization variable)",
     ylab = "E[asymmetric log-likelihood]",
     main = "Asymmetric objective: source p_true = 0.2  (golden-ratio anchor)")
lines(p_grid, obj_asym_N, col = "grey40", lty = 2, lwd = 2)
abline(v = 0.2, lty = 3, col = "tomato", lwd = 1.5)
legend("topright",
       c("source (p_true = 0.2)", "Gaussian source (varying p_test)", "p = 0.2"),
       col = c("steelblue", "grey40", "tomato"),
       lty = c(1, 2, 3), lwd = c(2, 2, 1.5), bty = "n")
```

The grey dashed curve shows the same asymmetric objective evaluated on a
**Gaussian** source as $p_\text{test}$ varies — it serves as a reference, not a
constant horizontal baseline.  The blue curve (true $p = 0.2$ source) peaks
near $p_\text{test} = 0.2$ and lies **above** the Gaussian reference curve in
that region — the optimizer can successfully identify the source by maximising
over $p$.  At $p_\text{test} = 0.5$ the asymmetric objective reduces to the
log-cosh contrast, and the blue curve dips **below** the grey reference,
explaining why standard fastICA fails for this source.

## Score functions for varying $p$

As $p$ decreases below 0.5 the score shifts and steepens, penalising the
positive tail more heavily — appropriate for sources that are rarely "on".

```{r}
pvec2 = c(0.05, 0.1, 0.2, 0.3, 0.5)
cols  = c("purple", "tomato", "darkorange", "steelblue", "black")
plot(NULL, xlim = c(-3, 3), ylim = c(-2.5, 2.5),
     xlab = "x", ylab = "M(x)",
     main = "Asymmetric score functions  (c = s = 1)")
for (i in seq_along(pvec2))
  lines(x_grid, asym_score(x_grid, pvec2[i], 1, 1)$M, col = cols[i], lwd = 2)
legend("topleft", paste0("p = ", pvec2), col = cols, lwd = 2, bty = "n")
abline(h = 0, lty = 3, col = "grey60")
```

## 9 overlapping groups ($k = 9$)

Nine sparse binary sources, each active in some fraction of 100 samples,
whitened to $k = 9$.

### $p \approx 0.2$ (20/100 active)

```{r}
set.seed(1)
n = 100; p_dim = 1000; K = 9
L  = matrix(0, nrow = n, ncol = K)
for (i in 1:K) L[sample(n, 20), i] = 1
FF = matrix(rnorm(p_dim * K), nrow = p_dim)
X9 = t(L %*% t(FF) + matrix(rnorm(n * p_dim, 0, 0.1), nrow = n))
Z9 = prewhiten(X9, K)
S9 = t(L)
```

```{r}
mc9_lc      = run_seeds_lc(Z9,       S9, hess = "fastICA", n_seeds = 100)
mc9_lc_tr   = run_seeds_lc(Z9,       S9, hess = "trace",   n_seeds = 100)
mc9_M       = run_seeds_asym(Z9,      S9, "M",      hess = "fastICA", n_seeds = 100)
mc9_gr      = run_seeds_asym(Z9,      S9, "golden", hess = "fastICA", n_seeds = 100)
mc9_gr_tr   = run_seeds_asym(Z9,      S9, "golden", hess = "trace",   n_seeds = 100)
mc9_grw     = run_seeds_asym_warm(Z9, S9, "golden", hess = "fastICA", n_seeds = 100)
mc9_grw_tr  = run_seeds_asym_warm(Z9, S9, "golden", hess = "trace",   n_seeds = 100)

cat("9-groups (p ~ 0.2, k = 9, n_seeds = 100):\n")
cat(sprintf("  log-cosh (fastICA)            mean = %.3f   frac > 0.9 = %.2f\n",
    mean(mc9_lc), mean(mc9_lc > 0.9)))
cat(sprintf("  log-cosh (trace)              mean = %.3f   frac > 0.9 = %.2f\n",
    mean(mc9_lc_tr), mean(mc9_lc_tr > 0.9)))
cat(sprintf("  asym M-est (random, fastICA)  mean = %.3f   frac > 0.9 = %.2f   mean_p = %.3f\n",
    mean(mc9_M$maxcor),      mean(mc9_M$maxcor      > 0.9), mean(mc9_M$p)))
cat(sprintf("  asym golden (random, fastICA) mean = %.3f   frac > 0.9 = %.2f   mean_p = %.3f\n",
    mean(mc9_gr$maxcor),     mean(mc9_gr$maxcor     > 0.9), mean(mc9_gr$p)))
cat(sprintf("  asym golden (random, trace)   mean = %.3f   frac > 0.9 = %.2f   mean_p = %.3f\n",
    mean(mc9_gr_tr$maxcor),  mean(mc9_gr_tr$maxcor  > 0.9), mean(mc9_gr_tr$p)))
cat(sprintf("  asym golden (warm, fastICA)   mean = %.3f   frac > 0.9 = %.2f   mean_p = %.3f\n",
    mean(mc9_grw$maxcor),    mean(mc9_grw$maxcor    > 0.9), mean(mc9_grw$p)))
cat(sprintf("  asym golden (warm, trace)     mean = %.3f   frac > 0.9 = %.2f   mean_p = %.3f\n",
    mean(mc9_grw_tr$maxcor), mean(mc9_grw_tr$maxcor > 0.9), mean(mc9_grw_tr$p)))
```

Sorting runs by their objective reveals whether the objective alone can select
true sources — making the raw success rate less critical:

```{r fig.width=9, fig.height=4}
par(mfrow = c(1, 2))
plot_obj_lc(mc9_lc,  "log-cosh, 9-groups p≈0.2",  lc_gauss)
plot_obj_asym(mc9_gr, "asymmetric, 9-groups p≈0.2")
par(mfrow = c(1, 1))
```

### $p \approx 0.1$ (10/100 active)

```{r}
set.seed(2)
L_01  = matrix(0, nrow = n, ncol = K)
for (i in 1:K) L_01[sample(n, 10), i] = 1
FF_01 = matrix(rnorm(p_dim * K), nrow = p_dim)
X9_01 = t(L_01 %*% t(FF_01) + matrix(rnorm(n * p_dim, 0, 0.1), nrow = n))
Z9_01 = prewhiten(X9_01, K)
S9_01 = t(L_01)
```

```{r}
mc9_01_lc     = run_seeds_lc(Z9_01,       S9_01, hess = "fastICA", n_seeds = 100)
mc9_01_lc_tr  = run_seeds_lc(Z9_01,       S9_01, hess = "trace",   n_seeds = 100)
mc9_01_gr     = run_seeds_asym(Z9_01,      S9_01, "golden", hess = "fastICA", n_seeds = 100)
mc9_01_gr_tr  = run_seeds_asym(Z9_01,      S9_01, "golden", hess = "trace",   n_seeds = 100)
mc9_01_grw    = run_seeds_asym_warm(Z9_01, S9_01, "golden", hess = "fastICA", n_seeds = 100)
mc9_01_grw_tr = run_seeds_asym_warm(Z9_01, S9_01, "golden", hess = "trace",   n_seeds = 100)

cat("9-groups (p ~ 0.1, k = 9, n_seeds = 100):\n")
cat(sprintf("  log-cosh (fastICA)            mean = %.3f   frac > 0.9 = %.2f\n",
    mean(mc9_01_lc), mean(mc9_01_lc > 0.9)))
cat(sprintf("  log-cosh (trace)              mean = %.3f   frac > 0.9 = %.2f\n",
    mean(mc9_01_lc_tr), mean(mc9_01_lc_tr > 0.9)))
cat(sprintf("  asym golden (random, fastICA) mean = %.3f   frac > 0.9 = %.2f   mean_p = %.3f\n",
    mean(mc9_01_gr$maxcor),     mean(mc9_01_gr$maxcor     > 0.9), mean(mc9_01_gr$p)))
cat(sprintf("  asym golden (random, trace)   mean = %.3f   frac > 0.9 = %.2f   mean_p = %.3f\n",
    mean(mc9_01_gr_tr$maxcor),  mean(mc9_01_gr_tr$maxcor  > 0.9), mean(mc9_01_gr_tr$p)))
cat(sprintf("  asym golden (warm, fastICA)   mean = %.3f   frac > 0.9 = %.2f   mean_p = %.3f\n",
    mean(mc9_01_grw$maxcor),    mean(mc9_01_grw$maxcor    > 0.9), mean(mc9_01_grw$p)))
cat(sprintf("  asym golden (warm, trace)     mean = %.3f   frac > 0.9 = %.2f   mean_p = %.3f\n",
    mean(mc9_01_grw_tr$maxcor), mean(mc9_01_grw_tr$maxcor > 0.9), mean(mc9_01_grw_tr$p)))
```

Log-cosh fastICA succeeds 82% of the time at $p \approx 0.1$ despite having
$E[\log\cosh(z)] < $ Gaussian baseline.  Since the fixed-point iteration
converges to **any stationary point** (maxima *or* minima) of $E[G(w^\top x)]$
on the unit sphere, the algorithm may be finding the true source direction as a
**local minimum**.  We check this directly.

```{r}
lc_obj_01 = attr(mc9_01_lc, "obj")   # already computed inside run_seeds_lc
succ = mc9_01_lc > 0.9
cat(sprintf("Gaussian baseline: %.4f\n", lc_gauss))
cat(sprintf("Successful seeds (n=%d): mean log-cosh obj = %.4f  [%s baseline]\n",
    sum(succ),  mean(lc_obj_01[succ]),
    ifelse(mean(lc_obj_01[succ])  < lc_gauss, "BELOW", "ABOVE")))
cat(sprintf("Failed seeds    (n=%d): mean log-cosh obj = %.4f  [%s baseline]\n",
    sum(!succ), mean(lc_obj_01[!succ]),
    ifelse(mean(lc_obj_01[!succ]) < lc_gauss, "BELOW", "ABOVE")))
```

Both successful and failed seeds land **below** the Gaussian baseline,
confirming that the fixed-point iteration finds local *minima* in all cases for
this dataset.  With 9 very sparse sources ($p = 0.1$), every direction in the
whitened space has $E[\log\cosh] < $ Gaussian baseline — the sparse structure
dominates everywhere.  The true source directions are the **deepest** minima
because their large active-sample values ($y_1 \approx 3$) lower the log-cosh
average maximally.

The practical question is whether the **objective value alone can select true
sources**, making the raw success rate less critical:

```{r fig.width=9, fig.height=4}
par(mfrow = c(1, 2))
plot_obj_lc(mc9_01_lc,  "log-cosh, 9-groups p≈0.1",  lc_gauss)
plot_obj_asym(mc9_01_gr, "asymmetric, 9-groups p≈0.1")
par(mfrow = c(1, 1))
```

### Why does the golden-ratio anchor do better?

The two anchors produce different score functions via the $c$-dependent bias
term in $\Delta(x)$. The golden-ratio anchor ($c = 0.618s$) satisfies
$c^2 + cs = s^2$, so the total assumed generative variance matches the empirical
variance $s^2$. The M-estimator ($c = s$) over-inflates the assumed variance
to $2s^2$, shifting the logistic midpoint and softening the asymmetry penalty.

We can visualise this: at $p = 0.2$, $s = 1$, the two anchors produce
noticeably different score functions:

```{r}
x_grid2 = seq(-4, 4, length.out = 400)
sc_M  = asym_score(x_grid2, p = 0.2, c = 1,              s = 1)
sc_gr = asym_score(x_grid2, p = 0.2, c = (sqrt(5)-1)/2,  s = 1)

plot(x_grid2, sc_M$M,  type = "l", col = "steelblue", lwd = 2,
     xlab = "x", ylab = "M(x)",
     main = "Score functions at p = 0.2: M-estimator vs golden-ratio")
lines(x_grid2, sc_gr$M, col = "tomato", lwd = 2)
lines(x_grid2, tanh(x_grid2), col = "grey50", lty = 2, lwd = 1.5)
legend("topleft",
       c("M-estimator (c = s)", "golden-ratio (c = 0.618s)", "tanh  (p = 0.5)"),
       col = c("steelblue", "tomato", "grey50"),
       lty = c(1, 1, 2), lwd = c(2, 2, 1.5), bty = "n")
```

## Sanity check: symmetric Rademacher ($p = 0.5$, $k = 9$)

Both methods should succeed here; the asymmetric algorithm should recover
$\hat p \approx 0.5$ automatically.  We compare three variants:

- **log-cosh**: standard fastICA.
- **asym golden (random start)**: alternating optimization from a random $w$.
- **asym golden (warm start)**: log-cosh run to convergence first, then hand
  the resulting $w$ to the asymmetric optimizer.

The warm start tests whether the degradation seen with random starts is
purely an initialization issue.

```{r}
set.seed(2)
S_sym = matrix(sample(c(-1, 1), K * n, replace = TRUE), nrow = K)
X_sym = t(S_sym) %*% t(FF) + matrix(rnorm(n * p_dim, 0, 0.1), nrow = n)
Z_sym = prewhiten(t(X_sym), K)

mc_s_lc      = run_seeds_lc(Z_sym,       S_sym, hess = "fastICA", n_seeds = 100)
mc_s_lc_tr   = run_seeds_lc(Z_sym,       S_sym, hess = "trace",   n_seeds = 100)
mc_s_gr      = run_seeds_asym(Z_sym,      S_sym, "golden", hess = "fastICA", n_seeds = 100)
mc_s_gr_tr   = run_seeds_asym(Z_sym,      S_sym, "golden", hess = "trace",   n_seeds = 100)
mc_s_grw     = run_seeds_asym_warm(Z_sym, S_sym, "golden", hess = "fastICA", n_seeds = 100)
mc_s_grw_tr  = run_seeds_asym_warm(Z_sym, S_sym, "golden", hess = "trace",   n_seeds = 100)

cat("Symmetric Rademacher (p_true = 0.5, k = 9):\n")
cat(sprintf("  log-cosh (fastICA)            mean = %.3f   frac > 0.9 = %.2f\n",
    mean(mc_s_lc), mean(mc_s_lc > 0.9)))
cat(sprintf("  log-cosh (trace)              mean = %.3f   frac > 0.9 = %.2f\n",
    mean(mc_s_lc_tr), mean(mc_s_lc_tr > 0.9)))
cat(sprintf("  asym golden (random, fastICA) mean = %.3f   frac > 0.9 = %.2f   mean_p = %.3f\n",
    mean(mc_s_gr$maxcor),     mean(mc_s_gr$maxcor     > 0.9), mean(mc_s_gr$p)))
cat(sprintf("  asym golden (random, trace)   mean = %.3f   frac > 0.9 = %.2f   mean_p = %.3f\n",
    mean(mc_s_gr_tr$maxcor),  mean(mc_s_gr_tr$maxcor  > 0.9), mean(mc_s_gr_tr$p)))
cat(sprintf("  asym golden (warm, fastICA)   mean = %.3f   frac > 0.9 = %.2f   mean_p = %.3f\n",
    mean(mc_s_grw$maxcor),    mean(mc_s_grw$maxcor    > 0.9), mean(mc_s_grw$p)))
cat(sprintf("  asym golden (warm, trace)     mean = %.3f   frac > 0.9 = %.2f   mean_p = %.3f\n",
    mean(mc_s_grw_tr$maxcor), mean(mc_s_grw_tr$maxcor > 0.9), mean(mc_s_grw_tr$p)))
```

```{r fig.width=9, fig.height=4}
par(mfrow = c(1, 2))
plot_obj_lc(mc_s_lc,  "log-cosh, 9-groups p=0.5",  lc_gauss)
plot_obj_asym(mc_s_gr, "asymmetric, 9-groups p=0.5")
par(mfrow = c(1, 1))
```

**Are the two objectives comparable?**  At $p = 0.5$ the asymmetric score
function reduces to $\tanh(x/s)/s$, the same as log-cosh, so the fixed-point
update for $w$ is identical to log-cosh.  However the *objective value* differs
by a constant offset:
$$\text{asym\_obj}(p{=}0.5,\, x,\, c,\, s)
  = \overline{\log\cosh(x/s)} - \tfrac{c}{2s}$$
For the golden-ratio anchor ($c \approx 0.618$, $s = 1$) the offset is
$-c/2 \approx -0.309$.

```{r}
succ_s   = mc_s_gr$maxcor > 0.9
c_gr     = (sqrt(5) - 1) / 2
lc_obj_s = attr(mc_s_lc, "obj")

cat(sprintf("Successful seeds (n=%d): mean p = %.3f   mean asym_obj = %.4f\n",
    sum(succ_s),  mean(mc_s_gr$p[succ_s]),  mean(mc_s_gr$obj[succ_s])))
cat(sprintf("Failed seeds    (n=%d): mean p = %.3f   mean asym_obj = %.4f\n",
    sum(!succ_s), mean(mc_s_gr$p[!succ_s]), mean(mc_s_gr$obj[!succ_s])))

# Fraction of top-K runs (by asym_obj) that found a true source
for (K_top in c(10, 20, 50)) {
  top_K = order(mc_s_gr$obj, decreasing = TRUE)[1:K_top]
  cat(sprintf("Fraction of top-%d asym_obj runs that are true sources: %.2f\n",
      K_top, mean(succ_s[top_K])))
}

# Theoretical offset at p = 0.5: asym_obj = lc_obj - c/2
cat(sprintf("\nTheoretical offset (asym at p=0.5) - lc_obj = -c/2 = %.4f\n", -c_gr/2))
cat(sprintf("Observed offset for successful seeds: %.4f  (p ≠ 0.5, so not exact)\n",
    mean(mc_s_gr$obj[succ_s] - lc_obj_s[succ_s])))
```

**What is happening:**

- Successful seeds do *not* converge to exactly $p = 0.5$.  With finite data
  and approximate symmetry, the optimizer finds $p \approx 0.4$–$0.5$.  At
  exactly $p = 0.5$ the asymmetric score equals log-cosh, but the joint $(w,p)$
  iteration drifts $p$ slightly away.
- Failed seeds converge to $p \approx 0.15$–$0.2$ — they found a noise direction
  that the optimizer interpreted as a sparse source.
- The mean asymmetric objective IS higher for successful seeds, so the objective
  has selection power, but some failed seeds (with inflated objectives from their
  small $p$) rank among the top runs.  This makes the asymmetric objective a
  slightly noisier selector than log-cosh for a symmetric true source.

## Single source, over-complete whitening ($k = 20$)

In all tests so far the whitening dimension $k$ matched the number of true
sources.  Here we use $k = 20$ whitened components to represent a **single**
source — the "over-complete" regime from [ebproj_newton](ebproj_newton.html).

With $k = 20$ the true source direction occupies only one of the 20 whitened
dimensions.  The samples where the source is "on" have larger
$S_{ii} = \|Y_{:i}\|^2/n$ than "off" samples (their projection onto the
leading singular vector is large), while those same "on" samples have near-zero
$M'(x_i)$ (the posterior is saturated).  The **trace** Hessian therefore
down-weights "on" samples relative to "off" samples, which may give a
different curvature estimate than the **fastICA** (isotropic) version.

The $p = 0.5$ simulation matches [ebproj_newton](ebproj_newton.html) exactly
(`set.seed(10)`, $n = 200$, $p = 1000$, single mixing vector, Rademacher source,
$k = 20$ whitened components).  The $p = 0.1$ simulation reuses the same mixing
vector and dimensions with a sparse binary source.

```{r}
# Exactly as in ebproj_newton Test 1
set.seed(10)
n_k20 = 200; p_k20 = 1000; k20 = 20
A_k20    = matrix(rnorm(p_k20), nrow = p_k20)
S_k20_05 = matrix(sample(c(-1, 1), n_k20, replace = TRUE), nrow = 1)
X_k20_05 = A_k20 %*% S_k20_05 + matrix(rnorm(p_k20 * n_k20, 0, 0.1), nrow = p_k20)
Z_k20_05 = prewhiten(X_k20_05, k20)

# Same mixing vector, sparse binary sources (p = 0.2 and p = 0.1)
set.seed(11)
S_k20_02_raw = matrix(as.numeric(runif(n_k20) < 0.2), nrow = 1)
S_k20_02     = (S_k20_02_raw - 0.2) / sqrt(0.2 * 0.8)
X_k20_02 = A_k20 %*% S_k20_02 + matrix(rnorm(p_k20 * n_k20, 0, 0.1), nrow = p_k20)
Z_k20_02 = prewhiten(X_k20_02, k20)

set.seed(12)
S_k20_01_raw = matrix(as.numeric(runif(n_k20) < 0.1), nrow = 1)
S_k20_01     = (S_k20_01_raw - 0.1) / sqrt(0.1 * 0.9)
X_k20_01 = A_k20 %*% S_k20_01 + matrix(rnorm(p_k20 * n_k20, 0, 0.1), nrow = p_k20)
Z_k20_01 = prewhiten(X_k20_01, k20)
```

### $p = 0.5$, $k = 20$

```{r}
mc_k20_05_lc     = run_seeds_lc(Z_k20_05,       S_k20_05, hess = "fastICA", n_seeds = 100)
mc_k20_05_lc_tr  = run_seeds_lc(Z_k20_05,       S_k20_05, hess = "trace",   n_seeds = 100)
mc_k20_05_gr     = run_seeds_asym(Z_k20_05,      S_k20_05, "golden", hess = "fastICA", n_seeds = 100)
mc_k20_05_gr_tr  = run_seeds_asym(Z_k20_05,      S_k20_05, "golden", hess = "trace",   n_seeds = 100)
mc_k20_05_grw    = run_seeds_asym_warm(Z_k20_05, S_k20_05, "golden", hess = "fastICA", n_seeds = 100)
mc_k20_05_grw_tr = run_seeds_asym_warm(Z_k20_05, S_k20_05, "golden", hess = "trace",   n_seeds = 100)

cat("Single source, k=20 whitening, p=0.5 (n_seeds=100):\n")
cat(sprintf("  log-cosh (fastICA)            mean = %.3f   frac > 0.9 = %.2f\n",
    mean(mc_k20_05_lc), mean(mc_k20_05_lc > 0.9)))
cat(sprintf("  log-cosh (trace)              mean = %.3f   frac > 0.9 = %.2f\n",
    mean(mc_k20_05_lc_tr), mean(mc_k20_05_lc_tr > 0.9)))
cat(sprintf("  asym golden (random, fastICA) mean = %.3f   frac > 0.9 = %.2f   mean_p = %.3f\n",
    mean(mc_k20_05_gr$maxcor),     mean(mc_k20_05_gr$maxcor     > 0.9), mean(mc_k20_05_gr$p)))
cat(sprintf("  asym golden (random, trace)   mean = %.3f   frac > 0.9 = %.2f   mean_p = %.3f\n",
    mean(mc_k20_05_gr_tr$maxcor),  mean(mc_k20_05_gr_tr$maxcor  > 0.9), mean(mc_k20_05_gr_tr$p)))
cat(sprintf("  asym golden (warm, fastICA)   mean = %.3f   frac > 0.9 = %.2f   mean_p = %.3f\n",
    mean(mc_k20_05_grw$maxcor),    mean(mc_k20_05_grw$maxcor    > 0.9), mean(mc_k20_05_grw$p)))
cat(sprintf("  asym golden (warm, trace)     mean = %.3f   frac > 0.9 = %.2f   mean_p = %.3f\n",
    mean(mc_k20_05_grw_tr$maxcor), mean(mc_k20_05_grw_tr$maxcor > 0.9), mean(mc_k20_05_grw_tr$p)))
```

```{r fig.width=9, fig.height=4}
par(mfrow = c(1, 2))
plot_obj_lc(mc_k20_05_lc,  "log-cosh, 1-source k=20 p=0.5",  lc_gauss)
plot_obj_asym(mc_k20_05_gr, "asymmetric, 1-source k=20 p=0.5")
par(mfrow = c(1, 1))
```

All methods fail on this problem: with $n = 200$ and $k = 20$ whitened
components, the signal-to-noise ratio in any single direction is low and a
random $w$ starts nearly orthogonal to the true source.  The asymmetric
optimizer also drifts $\hat p$ far from 0.5 (mean $\hat p \approx 0.07$–$0.09$)
before $w$ has converged, so it misidentifies the source as highly sparse even
when it is symmetric.

### $p = 0.2$, $k = 20$

```{r}
mc_k20_02_lc     = run_seeds_lc(Z_k20_02,       S_k20_02, hess = "fastICA", n_seeds = 100)
mc_k20_02_lc_tr  = run_seeds_lc(Z_k20_02,       S_k20_02, hess = "trace",   n_seeds = 100)
mc_k20_02_gr     = run_seeds_asym(Z_k20_02,      S_k20_02, "golden", hess = "fastICA", n_seeds = 100)
mc_k20_02_gr_tr  = run_seeds_asym(Z_k20_02,      S_k20_02, "golden", hess = "trace",   n_seeds = 100)
mc_k20_02_grw    = run_seeds_asym_warm(Z_k20_02, S_k20_02, "golden", hess = "fastICA", n_seeds = 100)
mc_k20_02_grw_tr = run_seeds_asym_warm(Z_k20_02, S_k20_02, "golden", hess = "trace",   n_seeds = 100)

cat("Single source, k=20 whitening, p=0.2 (n_seeds=100):\n")
cat(sprintf("  log-cosh (fastICA)            mean = %.3f   frac > 0.9 = %.2f\n",
    mean(mc_k20_02_lc), mean(mc_k20_02_lc > 0.9)))
cat(sprintf("  log-cosh (trace)              mean = %.3f   frac > 0.9 = %.2f\n",
    mean(mc_k20_02_lc_tr), mean(mc_k20_02_lc_tr > 0.9)))
cat(sprintf("  asym golden (random, fastICA) mean = %.3f   frac > 0.9 = %.2f   mean_p = %.3f\n",
    mean(mc_k20_02_gr$maxcor),     mean(mc_k20_02_gr$maxcor     > 0.9), mean(mc_k20_02_gr$p)))
cat(sprintf("  asym golden (random, trace)   mean = %.3f   frac > 0.9 = %.2f   mean_p = %.3f\n",
    mean(mc_k20_02_gr_tr$maxcor),  mean(mc_k20_02_gr_tr$maxcor  > 0.9), mean(mc_k20_02_gr_tr$p)))
cat(sprintf("  asym golden (warm, fastICA)   mean = %.3f   frac > 0.9 = %.2f   mean_p = %.3f\n",
    mean(mc_k20_02_grw$maxcor),    mean(mc_k20_02_grw$maxcor    > 0.9), mean(mc_k20_02_grw$p)))
cat(sprintf("  asym golden (warm, trace)     mean = %.3f   frac > 0.9 = %.2f   mean_p = %.3f\n",
    mean(mc_k20_02_grw_tr$maxcor), mean(mc_k20_02_grw_tr$maxcor > 0.9), mean(mc_k20_02_grw_tr$p)))
```

```{r fig.width=9, fig.height=4}
par(mfrow = c(1, 2))
plot_obj_lc(mc_k20_02_lc,  "log-cosh, 1-source k=20 p=0.2",  lc_gauss)
plot_obj_asym(mc_k20_02_gr, "asymmetric, 1-source k=20 p=0.2")
par(mfrow = c(1, 1))
```

All methods struggle at $p = 0.2$ in the $k = 20$ setting.  Log-cosh achieves 0%
even though it succeeds at $p = 0.1$.  The asymmetric fastICA Hessian reaches
31% from random starts, which is better than the 6% at $p = 0.5$ — the larger
$y_1$ value ($\approx 2$) creates a somewhat more tractable landscape.

### $p = 0.1$, $k = 20$

```{r}
mc_k20_01_lc     = run_seeds_lc(Z_k20_01,       S_k20_01, hess = "fastICA", n_seeds = 100)
mc_k20_01_lc_tr  = run_seeds_lc(Z_k20_01,       S_k20_01, hess = "trace",   n_seeds = 100)
mc_k20_01_gr     = run_seeds_asym(Z_k20_01,      S_k20_01, "golden", hess = "fastICA", n_seeds = 100)
mc_k20_01_gr_tr  = run_seeds_asym(Z_k20_01,      S_k20_01, "golden", hess = "trace",   n_seeds = 100)
mc_k20_01_grw    = run_seeds_asym_warm(Z_k20_01, S_k20_01, "golden", hess = "fastICA", n_seeds = 100)
mc_k20_01_grw_tr = run_seeds_asym_warm(Z_k20_01, S_k20_01, "golden", hess = "trace",   n_seeds = 100)

cat("Single source, k=20 whitening, p=0.1 (n_seeds=100):\n")
cat(sprintf("  log-cosh (fastICA)            mean = %.3f   frac > 0.9 = %.2f\n",
    mean(mc_k20_01_lc), mean(mc_k20_01_lc > 0.9)))
cat(sprintf("  log-cosh (trace)              mean = %.3f   frac > 0.9 = %.2f\n",
    mean(mc_k20_01_lc_tr), mean(mc_k20_01_lc_tr > 0.9)))
cat(sprintf("  asym golden (random, fastICA) mean = %.3f   frac > 0.9 = %.2f   mean_p = %.3f\n",
    mean(mc_k20_01_gr$maxcor),     mean(mc_k20_01_gr$maxcor     > 0.9), mean(mc_k20_01_gr$p)))
cat(sprintf("  asym golden (random, trace)   mean = %.3f   frac > 0.9 = %.2f   mean_p = %.3f\n",
    mean(mc_k20_01_gr_tr$maxcor),  mean(mc_k20_01_gr_tr$maxcor  > 0.9), mean(mc_k20_01_gr_tr$p)))
cat(sprintf("  asym golden (warm, fastICA)   mean = %.3f   frac > 0.9 = %.2f   mean_p = %.3f\n",
    mean(mc_k20_01_grw$maxcor),    mean(mc_k20_01_grw$maxcor    > 0.9), mean(mc_k20_01_grw$p)))
cat(sprintf("  asym golden (warm, trace)     mean = %.3f   frac > 0.9 = %.2f   mean_p = %.3f\n",
    mean(mc_k20_01_grw_tr$maxcor), mean(mc_k20_01_grw_tr$maxcor > 0.9), mean(mc_k20_01_grw_tr$p)))
```

```{r fig.width=9, fig.height=4}
par(mfrow = c(1, 2))
plot_obj_lc(mc_k20_01_lc,  "log-cosh, 1-source k=20 p=0.1",  lc_gauss)
plot_obj_asym(mc_k20_01_gr, "asymmetric, 1-source k=20 p=0.1")
par(mfrow = c(1, 1))
```

Log-cosh is the best method here on random starts (83%), while the asymmetric
fastICA Hessian trails (71%) and trace Hessian is worst (61%).  Warm-starting
from log-cosh lifts both asymmetric variants to 85%, matching log-cosh.

The trace Hessian gap arises because
$$c_{\text{trace}} = \overline{M'(x)} + \frac{n}{k}\,\mathrm{Cov}(M'(x_i),\, S_{ii})$$
The "on" samples ($p = 0.1$, $\approx 20$ out of $n=200$) have small
$M'(x_i) \approx 0$ (posterior saturated) but large $S_{ii}$ (they lie far from
the origin along the source direction), giving $\text{Cov}(M', S) < 0$.  With
$n/k = 200/20 = 10$ this is amplified 10-fold, making $c_{\text{trace}}$
substantially smaller than $\overline{M'(x)}$ and destabilising Newton steps from
random starts.  Warm-starting largely fixes this.

## Asymmetry parameter recovery ($k = 1$)

Using $k = 1$ whitening (which isolates each source exactly), we verify that
the estimated $\hat p$ tracks the true sparse fraction. Because of sign
ambiguity in the ICA direction, we report $\min(\hat p,\, 1-\hat p)$, i.e.
the probability of the rare state.

```{r}
set.seed(99)
n_rec = 500; p_dim_rec = 1000
A_rec = matrix(rnorm(p_dim_rec), nrow = p_dim_rec)

p_trues  = c(0.05, 0.10, 0.15, 0.20, 0.30, 0.40, 0.50)
p_hat_M  = numeric(length(p_trues))
p_hat_gr = numeric(length(p_trues))

for (j in seq_along(p_trues)) {
  pt = p_trues[j]
  S  = matrix(as.numeric(runif(n_rec) < pt), nrow = 1)
  S  = (S - pt) / sqrt(pt * (1-pt))
  X  = A_rec %*% S + matrix(rnorm(p_dim_rec * n_rec, 0, 0.1), nrow = p_dim_rec)
  Z  = prewhiten(X, 1)
  set.seed(1)
  r_M  = fastica_asym_r1(Z, anchor = "M",      w_init = rnorm(1))
  r_gr = fastica_asym_r1(Z, anchor = "golden",  w_init = rnorm(1))
  p_hat_M[j]  = min(r_M$p,  1 - r_M$p)
  p_hat_gr[j] = min(r_gr$p, 1 - r_gr$p)
}

plot(p_trues, p_hat_M, pch = 19, col = "steelblue",
     xlim = c(0, 0.52), ylim = c(0, 0.52),
     xlab = "true p  (sparse fraction)",
     ylab = "estimated p  (rare-state probability)",
     main = "Asymmetry parameter recovery  (k = 1 whitening)")
points(p_trues, p_hat_gr, pch = 17, col = "tomato")
abline(0, 1, lty = 2, col = "grey50")
legend("topleft", c("M-estimator", "golden-ratio"),
       col = c("steelblue","tomato"), pch = c(19,17), bty = "n")
```

Both anchors track the true sparse fraction closely across $p \in [0.05, 0.5]$.

## Summary

The asymmetric fastICA algorithm alternates between a Newton-like fixed-point
update for $w$ (identical to standard fastICA at $p = 0.5$) and 1D
optimization of $p$.  Two Hessian approximations are compared: **fastICA**
(isotropic, $\bar{M'(x)}\,\mathbf{I}$) and **trace** (weighted by
$S_{ii} = \|Y_{:i}\|^2/n$). Key findings from 100 random seeds each:

Fraction of seeds achieving max $|\text{cor}| > 0.9$ (100 seeds):

| Setting | lc-fastICA | lc-trace | asym random fastICA | asym random trace | asym warm fastICA | asym warm trace |
|---------|------------|----------|---------------------|-------------------|-------------------|-----------------|
| Sym ($p=0.5$, $k=9$) | 0.99 | **1.00** | 0.77 | 0.74 | 0.99 | 0.99 |
| 9 groups ($p\approx 0.2$, $k=9$) | 0.00 | 0.00 | **0.95** | 0.93 | 0.65 | 0.66 |
| 9 groups ($p\approx 0.1$, $k=9$) | 0.82 | 0.16 | **0.98** | 0.94 | **1.00** | **1.00** |
| 1 source ($p=0.5$, $k=20$, $n=200$) | 0.06 | **0.19** | 0.03 | 0.04 | 0.06 | 0.06 |
| 1 source ($p=0.2$, $k=20$, $n=200$) | 0.00 | 0.00 | **0.31** | 0.24 | 0.17 | 0.17 |
| 1 source ($p=0.1$, $k=20$, $n=200$) | 0.76 | 0.29 | **0.78** | 0.68 | **0.78** | **0.78** |

- **Log-cosh fails completely** for the 9-groups case because sparse sources
  have $E[\log\cosh] < $ Gaussian baseline, causing the algorithm to prefer
  noise directions.
- **The golden-ratio anchor** ($c \approx 0.618s$) dramatically outperforms
  the M-estimator anchor.  Its assumed generative variance matches the empirical
  variance ($c^2 + cs = s^2$), giving a better-calibrated asymmetry penalty.
- **Warm-starting has opposite effects depending on source type**:
  - *Symmetric sources*: a random start lets $p$ drift from 0.5 before $w$
    has converged, degrading performance.  Warm-starting from log-cosh
    avoids this and fully recovers high success rates.
  - *Asymmetric sources*: log-cosh converges to a wrong direction (noise PC),
    and the asymmetric method then inherits that bad start.  A random start
    performs better because it can reach the sparse source from a neutral position.
- **Log-cosh finds sparse sources as local minima**.  $E[\log\cosh]$ falls
  below the Gaussian baseline for $p \lesssim 0.25$, so the true source is a
  local *minimum* (not maximum) of the log-cosh contrast on the sphere.
  The fixed-point iteration converges to stationary points of either sign; for
  $p = 0.1$ ($y_1 \approx 3$, very large active-sample values) the minimum is
  deep enough that random starts land in its basin 82% (9 groups) or 76%
  (single source) of the time.  For $p = 0.2$ ($y_1 = 2$) the basin is too
  shallow and success drops to 0%.  The asymmetric method avoids this
  threshold entirely by adapting $p$ to match the source.
- **Trace Hessian: helps symmetric, hurts sparse**.  For symmetric sources
  ($p = 0.5$) the trace correction consistently helps (100% vs 99% at $k=9$;
  19% vs 6% at $k=20$).  But for sparse sources it collapses log-cosh from
  82% to 16% (9 groups, $p=0.1$) and from 76% to 29% (single source, $k=20$,
  $p=0.1$).  The mechanism is $n/k$ amplification:
  $c_{\text{trace}} = \overline{M'} + (n/k)\,\text{Cov}(M', S)$.
  For symmetric sources $\text{Cov}(M', S) \approx 0$; for sparse sources the
  strong negative covariance (active samples: small $M'$, large $S_{ii}$) is
  amplified by $n/k$, shrinking the correction and destabilising Newton steps.
- **Asymmetry recovery** ($k=1$ whitening): $\hat p = \min(p, 1-p)$ correctly
  tracks the true sparse fraction over the range $[0.05, 0.5]$.
