---
title: "fastica_asymmetric_04"
author: "Matthew Stephens"
date: "2026-08-22"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

This file implements simplified single-update versions of the asymmetric fastICA
Newton and gradient steps from `fastica_asymmetric_03.Rmd`.

The prior $g_0$ is binary with probabilities $(1-p, p)$ and atoms $(y_0, y_1)$
chosen so that $g_0$ has mean 0 and variance 1:
$$y_1 = \sqrt{\frac{1-p}{p}}, \qquad y_0 = -\sqrt{\frac{p}{1-p}}.$$
The effective atoms are $cy_0$ and $cy_1$.  The default scale is $c = c_\text{golden} = 2/(1+\sqrt{5})$
and the default residual variance is $\sigma^2 = c$.

When $p = 0.5$ and $c = \sigma^2$ both functions reduce to standard fastICA
(with $g(t) = \log\cosh$): the posterior mean becomes $c\tanh(x)$ and the
posterior derivative becomes $c(1-\tanh^2(x))$, and the c factors have no effect under
the final $\ell_2$ normalisation.


## Objective Function

The per-observation contribution to the objective is
$$\log Z(x;\, p, c, \sigma^2) = \log\!\left[(1-p)\exp\!\left(\frac{2x\,cy_0 - (cy_0)^2}{2\sigma^2}\right) + p\exp\!\left(\frac{2x\,cy_1 - (cy_1)^2}{2\sigma^2}\right)\right]$$
with $y_1 = \sqrt{(1-p)/p}$, $y_0 = -\sqrt{p/(1-p)}$, and the full objective is $J(w) = \sum_i \log Z(x_i)$.

A natural question is: what is the best expected value of $\log Z(x)$ achievable by a binary source with parameter $p_\text{source}$, when we include an intercept (allow the source mean $\mu$ to be non-zero)?  We constrain the second moment to 1, so the source has mean $\mu$ and variance $1 - \mu^2$, giving atoms
$$a_0 = \mu - p_s\,d, \quad a_1 = \mu + (1-p_s)\,d, \quad d = \sqrt{\frac{1-\mu^2}{p_s(1-p_s)}}.$$
We then maximize over $\mu \in (-1, 1)$:
$$\max_\mu\, E_{p_s,\mu}[\log Z(x)] = \max_\mu\left[(1-p_s)\log Z(a_0) + p_s\log Z(a_1)\right].$$
The Gaussian baseline is similarly $\max_\mu\, E_{N(\mu,\,1-\mu^2)}[\log Z(x)]$.

We plot this as a function of $p_\text{source}$ for several values of $p$ used in the objective (solid lines), with the Gaussian baseline shown as a dashed horizontal line for each $p_\text{objective}$.

```{r}
c_val   <- 2 / (1 + sqrt(5))
s2_val  <- c_val
p_objs  <- c(0.01, 0.05, 0.1, 0.2, 0.3, 0.4, 0.5)
cols    <- palette.colors(length(p_objs), palette = "Okabe-Ito")

log_Z_vec <- function(x, p, c, s2) {
  y1  <- sqrt((1 - p) / p);  y0 <- -sqrt(p / (1 - p))
  ey0 <- c * y0;  ey1 <- c * y1
  lp0 <- log(1 - p) + (2*x*ey0 - ey0^2) / (2*s2)
  lp1 <- log(p)     + (2*x*ey1 - ey1^2) / (2*s2)
  m   <- pmax(lp0, lp1)
  log(exp(lp0 - m) + exp(lp1 - m)) + m
}

# Expected log Z for a binary source with mean mu, second moment 1
# (so variance = 1 - mu^2); atoms are mu +/- scaled y0/y1.
E_logZ_binary_shifted <- function(p_source, mu, p_obj, c, s2) {
  d  <- sqrt((1 - mu^2) / (p_source * (1 - p_source)))
  a0 <- mu - p_source       * d
  a1 <- mu + (1 - p_source) * d
  (1 - p_source) * log_Z_vec(a0, p_obj, c, s2) +
       p_source  * log_Z_vec(a1, p_obj, c, s2)
}

# Max over mu in (-1, 1) of the expected log Z for a binary source.
# Uses a grid search (not optimize()) because the objective can be non-unimodal
# in mu: the degenerate solution near the boundary competes with interior maxima
# and golden-section search would get trapped in local optima.
E_logZ_binary_maxmu <- function(p_source, p_obj, c, s2, n_grid = 201) {
  mu_grid <- seq(-1 + 1e-4, 1 - 1e-4, length.out = n_grid)
  max(sapply(mu_grid, function(mu)
    E_logZ_binary_shifted(p_source, mu, p_obj, c, s2)))
}

# Max over mu of E[log Z] for a Gaussian with mean mu, second moment 1.
E_logZ_gauss_maxmu <- function(p_obj, c, s2, n_grid = 201) {
  mu_grid <- seq(-1 + 1e-4, 1 - 1e-4, length.out = n_grid)
  max(sapply(mu_grid, function(mu)
    integrate(function(x) log_Z_vec(x, p_obj, c, s2) * dnorm(x, mu, sqrt(1 - mu^2)),
              -8, 8)$value))
}

x_grid    <- seq(-5, 5, length.out = 500)
logZ_mat  <- sapply(p_objs, function(p) log_Z_vec(x_grid, p, c_val, s2_val))
matplot(x_grid, logZ_mat, type = "l", lty = 1, col = cols, lwd = 2,
        xlab = "x", ylab = "log Z(x)",
        main = "log Z(x) by p_objective")
legend("top", legend = paste0("p_obj = ", p_objs), col = cols, lty = 1, lwd = 2)

p_sources <- seq(0.01, 0.99, by = 0.01)

# mu = 0 (zero-mean source)
E_mat_mu0 <- sapply(p_objs, function(po)
  sapply(p_sources, function(ps) E_logZ_binary_shifted(ps, 0, po, c_val, s2_val)))

E_gauss_mu0 <- sapply(p_objs, function(po)
  integrate(function(x) log_Z_vec(x, po, c_val, s2_val) * dnorm(x), -8, 8)$value)

# max over mu
E_mat_maxmu <- sapply(p_objs, function(po)
  sapply(p_sources, function(ps) E_logZ_binary_maxmu(ps, po, c_val, s2_val)))

E_gauss_maxmu <- sapply(p_objs, E_logZ_gauss_maxmu, c = c_val, s2 = s2_val)

par(mfrow = c(1, 2))
matplot(p_sources, E_mat_mu0, type = "l", lty = 1, col = cols, lwd = 2,
        xlab = expression(p[source]), ylab = expression(E[source]*"[log Z(x)]"),
        main = expression(mu == 0))
for (i in seq_along(p_objs))
  abline(h = E_gauss_mu0[i], col = cols[i], lty = 2, lwd = 1.5)
legend("topright",
       legend = c(paste0("p_obj = ", p_objs), "Gaussian (dashed)"),
       col = c(cols, "grey50"), lty = c(rep(1, length(p_objs)), 2), lwd = 2,
       cex = 0.8)

matplot(p_sources, E_mat_maxmu, type = "l", lty = 1, col = cols, lwd = 2,
        xlab = expression(p[source]), ylab = "",
        main = expression(max[mu]))
for (i in seq_along(p_objs))
  abline(h = E_gauss_maxmu[i], col = cols[i], lty = 2, lwd = 1.5)

par(mfrow = c(1, 1))
```

### Effect of the intercept for a fixed $p_\text{obj}$

To see directly whether the intercept helps find sources with $p_\text{source} \ne p_\text{obj}$, we fix $p_\text{obj}$ and plot $E[\log Z]$ as a function of $\mu$ for several values of $p_\text{source}$.  A curve that peaks above its Gaussian baseline is a source detectable by this objective; how far the peak sits from $\mu = 0$ shows how much the intercept is contributing.

Here are plots for p=0.5 to 0.05. We see that p=0.5 (fastICA) the sparse binary sources are below the Gaussian baseline at mu=0 whereas the balanced binary sources are above the Gaussian baseline. While the sparse binary source curves go above the baseline near mu=-1, this is a rather restricted part of the space that we would probably have to deliberately search (eg by initializing the intercept near that location) to find it. 

In contrast, for p=0.25, almost all the sources (except the vary rarest ones) are above the Gaussian baseline at mu=0 and even more so if you allow mu to be estimated away from 0. 
Obviously for finding truly sparse sources it would pay to use a smaller p. 

However, by the time you go down to $p=0.01$ the ability to find more common sources, even 0.1, might be quite restricted. 

Based on these curves I am inclined to go with running the method for a range of $p$ such as $p=0.25,0.05,0.01$.


```{r}
p_srcs_plot   <- c(0.02, 0.05, 0.1, 0.2, 0.3, 0.4, 0.5)
cols_src      <- palette.colors(length(p_srcs_plot), palette = "Okabe-Ito")
mu_grid_fine  <- seq(-0.99, 0.99, length.out = 201)

for(p_obj_focus in c(0.5,0.4,0.33,0.25,0.1,0.05,0.01)){
  E_vs_mu <- sapply(p_srcs_plot, function(ps)
    sapply(mu_grid_fine, function(mu)
      E_logZ_binary_shifted(ps, mu, p_obj_focus, c_val, s2_val)))

  E_gauss_vs_mu <- sapply(mu_grid_fine, function(mu)
    integrate(function(x) log_Z_vec(x, p_obj_focus, c_val, s2_val) *
                dnorm(x, mu, sqrt(1 - mu^2)), -8, 8)$value)

  matplot(mu_grid_fine, E_vs_mu, type = "l", lty = 1, col = cols_src, lwd = 2,
        xlab = expression(mu), ylab = "E[log Z(x)]",
        main = sprintf("E[log Z] vs intercept  (p_obj = %.2f)", p_obj_focus))
  lines(mu_grid_fine, E_gauss_vs_mu, lty = 2, col = "grey30", lwd = 2)
  abline(v = 0, lty = 3, col = "grey60")
  legend("bottomleft",
       legend = c(paste0("p_src = ", p_srcs_plot), "Gaussian"),
       col = c(cols_src, "grey30"), lty = c(rep(1, length(p_srcs_plot)), 2),
       lwd = 2, cex = 0.8)
}
```


## Update functions

```{r}
# Newton update (ica-hessian approximation).
# U is (n.comp) x n (whitened data).
# W is (n.comp) x n_starts (one weight vector per column).
# P = t(U) %*% W is n x n_starts (source estimates).
# When p = 0.5 and c = s2 this is identical to standard fastica_update.
binary_newton_update <- function(U, W, p = 0.5, c = 2 / (1 + sqrt(5)), s2 = c) {
  y1  <- sqrt((1 - p) / p)
  y0  <- -sqrt(p / (1 - p))
  ey0 <- c * y0
  ey1 <- c * y1

  P   <- t(U) %*% W                              # n x n_starts

  lp0 <- log(1 - p) + (2*P*ey0 - ey0^2) / (2*s2)
  lp1 <- log(p)     + (2*P*ey1 - ey1^2) / (2*s2)
  m   <- pmax(lp0, lp1)
  e0  <- exp(lp0 - m);  e1 <- exp(lp1 - m)
  pi1 <- e1 / (e0 + e1)                          # n x n_starts

  Mx  <- (1 - pi1) * ey0 + pi1 * ey1             # posterior mean
  Mpx <- pi1 * (1 - pi1) * (ey1 - ey0)^2 / s2   # dM/dx

  W <- U %*% Mx - sweep(W, 2, colSums(Mpx), "*")
  sweep(W, 2, sqrt(colSums(W^2)) + 1e-15, "/")
}

# Gradient update (no Hessian correction; equivalent to hessian = "none" in ebproj).
binary_gradient_update <- function(U, W, p = 0.5, c = 2 / (1 + sqrt(5)), s2 = c) {
  y1  <- sqrt((1 - p) / p)
  y0  <- -sqrt(p / (1 - p))
  ey0 <- c * y0
  ey1 <- c * y1

  P   <- t(U) %*% W                              # n x n_starts

  lp0 <- log(1 - p) + (2*P*ey0 - ey0^2) / (2*s2)
  lp1 <- log(p)     + (2*P*ey1 - ey1^2) / (2*s2)
  m   <- pmax(lp0, lp1)
  e0  <- exp(lp0 - m);  e1 <- exp(lp1 - m)
  pi1 <- e1 / (e0 + e1)                          # n x n_starts

  Mx  <- (1 - pi1) * ey0 + pi1 * ey1             # posterior mean

  W <- U %*% Mx
  sweep(W, 2, sqrt(colSums(W^2)) + 1e-15, "/")
}

# Objective J(W) = sum_i log Z(x_i, g, s2), returned as a vector of length n_starts.
binary_objective <- function(U, W, p = 0.5, c = 2 / (1 + sqrt(5)), s2 = c) {
  y1  <- sqrt((1 - p) / p)
  y0  <- -sqrt(p / (1 - p))
  ey0 <- c * y0
  ey1 <- c * y1

  P   <- t(U) %*% W                              # n x n_starts

  lp0 <- log(1 - p) + (2*P*ey0 - ey0^2) / (2*s2)
  lp1 <- log(p)     + (2*P*ey1 - ey1^2) / (2*s2)
  m   <- pmax(lp0, lp1)
  colSums(log(exp(lp0 - m) + exp(lp1 - m)) + m)  # length n_starts
}

# matrix version of r1 fastica
# U is (n.comp+1) x n (whitened data plus intercept).  
# W is (n.comp+1) x n_starts (one weight vector per column).
# P = t(U) %*% W is n x n_starts (source estimates).
fastica_update = function(U, W) {
  P  <- t(U) %*% W          # n x n_starts: source estimates
  G  <- tanh(P)
  G2 <- 1 - G^2
  W  <- U %*% G - sweep(W, 2, colSums(G2), "*")
  # Add epsilon (1e-15) to prevent 0/0
  sweep(W, 2, sqrt(colSums(W^2)) + 1e-15, "/")
}

```

## Simulation

```{r}
### $p \approx 0.2$ (20/100 active)
set.seed(1)
n = 100; p_dim = 1000; K = 9
L  = matrix(0, nrow = n, ncol = K)
for (i in 1:K) L[sample(n, 20), i] = 1
FF = matrix(rnorm(p_dim * K), nrow = p_dim)
Y = L %*% t(FF) + matrix(rnorm(n * p_dim, 0, 0.1), nrow = n)
```

I use U_aug (U augmented with an intercept) because it works well for fastICA and maybe should not hurt the binary version? We will see.

First I run ICA:
```{r}
n.comp = 9
Y     <- scale(Y, scale = FALSE)  # center columns; also densifies for SVD
U     <- sqrt(nrow(Y)) * t(svd(Y)$u[, 1:n.comp])
U_aug <- rbind(rep(1, ncol(U)), U)

n_starts = 1000
n_iter = 50  
set.seed(1)
W <- matrix(rnorm(nrow(U_aug) * n_starts), nrow(U_aug), n_starts)
W <- sweep(W, 2, sqrt(colSums(W^2)), "/")
for (i in seq_len(n_iter))
  W <- fastica_update(U_aug, W)
  #W <- binary_newton_update(U_aug, W)

Lhat <- t(U_aug) %*% W  # n x n_starts
plot(colMeans(log(cosh(Lhat))))
```

Now I run the binary update. Since p=0.5 it should give the same results. Note that the binary objective function is sum(log(cosh))+nc/2 so the axis is different but results are equivalent. 
```{r}
set.seed(1)
W <- matrix(rnorm(nrow(U_aug) * n_starts), nrow(U_aug), n_starts)
W <- sweep(W, 2, sqrt(colSums(W^2)), "/")
for (i in seq_len(n_iter))
  W <- binary_newton_update(U_aug, W)
plot(binary_objective(U_aug,W))
```

Now I try optimizing with p=0.1. As one might hope, the maxima look like true sources. 
The minima look like the intercept (although numerically there is some tiny variation since the correlation is computable, and interestingly they all have the same correlation for reasons I don't understand). Also, every one of the nine sources is found by some start.
```{r}
set.seed(1)
p = 0.1
W <- matrix(rnorm(nrow(U_aug) * n_starts), nrow(U_aug), n_starts)
W <- sweep(W, 2, sqrt(colSums(W^2)), "/")
for (i in seq_len(n_iter))
  W <- binary_newton_update(U_aug, W, p=p)
plot(binary_objective(U_aug,W,p=p))

Lhat <- t(U_aug) %*% W  # n x n_starts
id_max = which.max(binary_objective(U_aug,W,p=p))
plot(Lhat[,id_max])
id_min = which.min(binary_objective(U_aug,W,p=p))
plot(Lhat[,id_min])

bestcor = apply(abs(cor(Lhat,L)),1,max)
plot(bestcor, binary_objective(U_aug,W,p=p))

apply(abs(cor(Lhat,L)),2,max)
```

Now I try optimizing with p=0.25. Something unexpected (to me) happened: the maxima are now sometimes true sources, but sometimes the intercept. If you go back to the plot for p=0.25 you can see the degenerate solution happens to have an expected value just around the same as the maximum for a binary 0.2 source (actually slightly higher). So maybe I need to pay some attention to this and choose slightly different p than I suggested above. (This said, it does still find the true sources; but we would maybe prefer the maximum not to occur at a degenerate solution? These are easily removed though so does it matter much?) 

```{r}
set.seed(1)
p = 0.25
W <- matrix(rnorm(nrow(U_aug) * n_starts), nrow(U_aug), n_starts)
W <- sweep(W, 2, sqrt(colSums(W^2)), "/")
for (i in seq_len(n_iter))
  W <- binary_newton_update(U_aug, W, p=p)
plot(binary_objective(U_aug,W,p=p))
plot(binary_objective(U_aug,W,p=p), ylim=c(20.05,20.15), main = "degenerate solution is just slightly higher")

Lhat <- t(U_aug) %*% W  # n x n_starts
id_max = which.max(binary_objective(U_aug,W,p=p))
plot(Lhat[,id_max])

bestcor = apply(abs(cor(Lhat,L)),1,max)
plot(bestcor,col=1+(binary_objective(U_aug,W,p=p)>20.105), main="red=degenerate solution")

apply(abs(cor(Lhat,L)),2,max)
```

