---
title: "flash_f1_size"
author: "Matthew Stephens"
date: "2021-10-21"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

Following up on some results from Joonsuk Kang, 
I wanted to investigate how the "size" (PVE) of the first factor 
affects ability to detect a second factor in a simple situation.

## Simulation

I'll simulate two factors, both random normal:

```{r}
library("flashr")
n = 100
p = 100
k = 2
LL = matrix(rnorm(p*k), nrow=n, ncol=k)
FF = matrix(rnorm(p*k), nrow=p, ncol=k)
E = matrix(rnorm(n*p),nrow=n,ncol=p)
Y = LL %*% t(FF) + E
svd(Y)$d[1:5] # singular values
fit1 = flashr::flash(Y,verbose=FALSE)
cor(flash_get_ldf(fit1)$l,LL)
```

Actually you see the non-identifiability here: because both factors
are a similar size they have similar eigenvalues so you get the rotation-invariance issue.

Now make first factor stronger. This makes the two factors identifiable and results are more accurate:
```{r}
FF[,1] = FF[,1]* 10
Y = LL %*% t(FF) + E
svd(Y)$d[1:5] # singular values
fit1 = flashr::flash(Y, verbose=FALSE)
cor(flash_get_ldf(fit1)$l,LL)
```

Now make first factor much stronger. It still seems to work....
```{r}
FF[,1] = FF[,1]* 1000
Y = LL %*% t(FF) + E
svd(Y)$d[1:5] # singular values
fit1 = flashr::flash(Y,verbose=F)
cor(flash_get_ldf(fit1)$l,LL)
```
