---
title: "flashier_pbmc_purified"
author: "Matthew Stephens"
date: "2024-04-04"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

I run flashier (sparse EBNMF) on the purified pbmcs. I use some code from Eric Weine to plot the results.


```{r}
library(dplyr)
library(tidyr)
library(purrr)
library(stringr)
library(Matrix)
library(flashier)
library(ggplot2)
```

Load the data (in a sparse matrix called counts)
```{r}
load("../data/pbmc_purified.RData")
```

```{r}
lcounts = counts
lcounts@x = log(counts@x + 1)
```

Run flashier
```{r}
fit1 = flash(lcounts,S = 0.01, ebnm_fn = ebnm_point_exponential, var_type = 2)
saveRDS(fit1,file='../output/fit.nn.pbmc.purified.rds')
```

Note that the first factor is "intercept-like", capturing library size and some kind of background expression,
```{r}
plot(fit1$L_pm[,1],rowSums(lcounts),main="loadings for factor 1 vs rowSums of lcounts")
```


Here I am using Eric's code to plot a heatmap of the results, ordered by cell type
```{r}
FF_log1p = fit1$L_pm
FF_log1p <- scale(FF_log1p, center = FALSE, scale = apply(FF_log1p, 2, max))
colnames(FF_log1p) <- paste0("k", 1:ncol(FF_log1p))

cell.type <- samples$celltype

# Downsample the number of cells and sort them using tSNE.
set.seed(8675309)
cell.idx <- numeric(0)
cell.types <- levels(cell.type)
for (i in 1:length(cell.types)) {
  which.idx <- which(cell.type == cell.types[i])
  # Downsample common cell types.
  if (length(which.idx) > 1250) {
    which.idx <- sample(which.idx, 1250)
  }
  # Don't include rare cell types.
  if (length(which.idx) > 20) {
    # Sort using tsne.
    tsne.res <- Rtsne::Rtsne(
      FF_log1p[which.idx, ],
      dims = 1,
      pca = FALSE,
      normalize = FALSE,
      perplexity = min(100, floor((length(which.idx) - 1) / 3) - 1),
      theta = 0.1,
      max_iter = 1000,
      eta = 200,
      check_duplicates = FALSE
    )$Y[, 1]
    which.idx <- which.idx[order(tsne.res)]
    cell.idx <- c(cell.idx, which.idx)
  }
}

cell.type <- cell.type[cell.idx]
cell.type <- droplevels(cell.type)

FF_log1p <- FF_log1p[cell.idx, ]

make.heatmap.tib <- function(FF) {
  tib <- as_tibble(scale(FF, center = FALSE, scale = apply(FF, 2, max))) %>%
    mutate(Cell.type = cell.type) %>%
    arrange(Cell.type) %>%
    mutate(Cell.idx = row_number())
  
  tib <- tib %>%
    pivot_longer(
      -c(Cell.idx, Cell.type),
      names_to = "Factor",
      values_to = "Loading",
      values_drop_na = TRUE
    ) %>%
    mutate(Factor = as.numeric(str_extract(Factor, "[0-9]+")))
  
  return(tib)
}

log1p_tib <- make.heatmap.tib(FF_log1p)
heatmap.tib = log1p_tib

tib <- heatmap.tib %>%
  group_by(Cell.type, Cell.idx) %>%
  summarize()

cell_type_breaks <- c(1, which(tib$Cell.type[-1] != tib$Cell.type[-nrow(tib)]))
label_pos <- cell_type_breaks / 2 + c(cell_type_breaks[-1], nrow(tib)) / 2

plt <- ggplot(heatmap.tib, aes(x = Factor, y = -Cell.idx, fill = Loading)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "firebrick") +
  labs(y = "") +
  scale_y_continuous(breaks = -label_pos,
                     minor_breaks = NULL,
                     labels = levels(cell.type)) +
  scale_x_continuous(breaks = seq(0, 30, 5)) +
  theme_minimal() +
  geom_hline(yintercept = -cell_type_breaks, size = 0.1) +
  theme(legend.position = "none", 
        strip.text = element_text(size = 16)) 

plt
```

List driving genes:
```{r}
for(k in 1:18){
  print(paste0("factor ",k))
  print(genes$symbol[order(fit1$F_pm[,k],decreasing =TRUE)[1:50]])
}
```

I tried backfitting but it gave a memory error. We should look into that.
```{r}
#fit2 = flash_backfit(fit1)
```


