---
title: "flashier_stocks"
author: "Matthew Stephens"
date: "2021-05-22"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

In preparing homework for my class I ran
flashier on some stock data.
(Note that flash errored out here, and in any case flashier
was quite a bit faster...)
It got me thinking about appropriate application of flash/flashier
for time series, so I'm reporting the results here.

The data were downloaded as in https://stephens999.github.io/stat34800/stocks.html
We probably could do with looking at a bigger dataset if
we want to take this seriously, but I do that for now.

```{r}
# AAPL: Apple
# NFLX: Netflix
# AMZN: Amazon
# MMM: 3M
# K: Kellogs
# O: Realty Income Corp
# NSRGY: Nestle
# LDSVF: Lindt
# JPM: JP Morgan Chase
# JNJ: Johnson and Johnson
# TSLA: Tesla
# V: Visa
# PFE: Pfizer
```



```{r}
prices = read.csv("../data/prices.csv")
log_prices = log(prices)
log_returns = apply(log_prices,2, diff)
```

You can see some structure in the correlation matrix: the tech companys are correlated, as are
the PFE/JNJ and the financial companies (V,JPM).

```{r}
S = cor(log_returns)
heatmap(S, xlab = names(prices), symm=TRUE)
```


### Flashier on the raw data

Note I tried to use column-specific residual variances
but it errored out. So I switched to constant variances.
Backfitting seemed to clean up the factors so I did that.

```{r}
#library("flashr")
library("ebnm")
library("flashier")
#fit.f = flashr::flash(as.matrix(log_returns),ebnm_fn = "ebnm_pl") ## errors out
#fit.f = flashier::flash(as.matrix(log_returns),prior.family = prior.point.laplace(), var.type = 2) # this produces an error
fit.f = flashier::flash(as.matrix(log_returns),prior.family = prior.point.laplace(), var.type = 0, backfit=TRUE)
```

Plot the factors. 

```{r}
for(i in 1:11){
  barplot(fit.f$loadings.pm[[2]][,i], names.arg=names(prices), horiz=TRUE,las=2, main=paste0("Factor ",i))
}

for(i in 1:11){
  plot(fit.f$loadings.pm[[1]][,i], main=paste0("Factor ",i))
}
```



### Flashier on the correlation matrix

Try factor analysis on the correlation matrix. We see
a lot fewer factors. I am interested why this is. I didn't really
expect such a big difference when the factors are dense, as they
seem to be here (in time space).

One possibility is that although the inferred factors are dense, they are long tailed and the inferred factors are driven by the
outliers. It might be worth doing some simulations with
dense long-tailed factors and comparing inference from correlation vs
full data.

Another possibility is that many of the raw data results are
driven by stock-specific factors. Here stock-specific factors
are not represented because I remove the diagonal. And the
2 factors identified could be thought of as 3 factors if we went non-negative... in that case the differences between the results
do not look so stark. 

I did try running this without removing diagonal, but the stock-specific
factors don't get picked up; I think this may be due to problems
converging to them (eg svd would not initialize near them...)
That may suggest it could also be difficult to find other sparse factors
of course (eg involving pairs of stocks). 


```{r}
Smiss = S
diag(Smiss) <- NA
S.f = flashier::flash(Smiss,prior.family = prior.point.laplace(), var.type = 0, backfit=TRUE)
for(i in 1:2){
  barplot(S.f$loadings.pm[[2]][,i], names.arg=names(prices), horiz=TRUE,las=2, main=paste0("Factor ",i))
}
```

### Flashier on the covariance matrix

Try factor analysis on the covariance matrix. We 
still see a lot fewer factors. 


```{r}
S2 = cov(log_returns)
S2.f = flashier::flash(S2,prior.family = prior.point.laplace(), var.type = 0, backfit = TRUE)
for(i in 1:4){
  barplot(S2.f$loadings.pm[[2]][,i], names.arg=names(prices), horiz=TRUE,las=2, main=paste0("Factor ",i))
}
```

### Flashier on the standardized raw data

Here I tried rerunning on the standardized raw data to see how that affects things. 

```{r}
fit.f3 = flashier::flash(scale(log_returns),prior.family = prior.point.laplace(), var.type = 0, backfit = TRUE)
for(i in 1:13){
  barplot(fit.f3$loadings.pm[[2]][,i], names.arg=names(prices), horiz=TRUE,las=2, main=paste0("Factor ",i))
}

for(i in 1:13){
  plot(fit.f3$loadings.pm[[1]][,i], main=paste0("Factor ",i))
}
```


### Wavelet transformed data

Since these are time series it would
be nice to try wavelet transforming them before applying flashier.
Here I explore some of these ideas.

First define haar transform functions
```{r}
haar = function(x,scale= sqrt(2)){
  if(length(x)==1){
    return(x)
  }
  else{
    x = matrix(x,nrow=2)
    diff = (x[1,]-x[2,])/scale
    sum = (x[1,]+x[2,])/scale
    return(c(diff, haar(sum)))
  }
}

haar_inv = function(x,scale=sqrt(2)){
  n=length(x)
  if(n==1){
    return(x)
  }
  x = matrix(scale*x,nrow=2,byrow=TRUE)
  smoothed = haar_inv(x[2,]) 
  return(as.vector(rbind(smoothed+x[1,], smoothed-x[1,]))/2)
}
```

Now I plot the above fitted factors after haar transform. There isn't an
obvious decrease in sparsity (not surpising since there was not an obvious spatiol component.)
```{r}
for(i in 1:13){
  plot(haar(fit.f3$loadings.pm[[1]][1:2048,i]), main=paste0("Factor ",i, " (transformed space)"))
}
```

Compute the haar transform of log returns:
```{r}
lp.h = log_returns[1:2048,]
# do haar wavelet decomposition on log-returns and save in lp.h
for(i in 1:ncol(log_prices)){
  lp.h[,i] = haar(log_returns[1:2048,i])
}
```

Quick look at correlations of the transformed data.
```{r}
S.h = cor(lp.h)
heatmap(S.h, xlab = names(prices), symm=TRUE)
```

Maybe it makes sense just to do the higher scales?
```{r}
low_res = 2048-(0:255)
S.h = cor(lp.h[low_res,])
heatmap(S.h, xlab = names(prices), symm=TRUE)
```


I'm not quite sure of the right way to proceed here... I'm just going
to apply flash to the wavelet transformed data, even though that does not really seem quite right (the iid prior on wavelet coefficients at different scales does not really seem sensible.)
```{r}
lp.h.f = flashier::flash(lp.h, prior.family=prior.point.laplace(), backfit=TRUE)
for(i in 1:8){
  barplot(lp.h.f$loadings.pm[[2]][,i], names.arg=names(prices), horiz=TRUE,las=2, main=paste0("Factor ",i))
}

for(i in 1:8){
  plot(lp.h.f$loadings.pm[[1]][,i], main=paste0("Factor ",i," (transformed space)"))
}

for(i in 1:8){
  plot(haar_inv(lp.h.f$loadings.pm[[1]][,i]), main=paste0("Factor ",i),type="l")
}
```


