---
title: "Simple illustration of issue with conservative behaviour of variational method"
author: "Matthew Stephens"
date: 2017-10-12
output: workflowr::wflow_html
---




## Simple illustration

Here's an example where flashr2 produces a rank 0 result, even when the first
eigenvalue is clearly outstanding.

```{r}
library("flashr2")
n = 100
p = 300

set.seed(7)
# seed 5 also produces an example of same behaviour

Y = matrix(rnorm(n*p),nrow=n,ncol=p)

l= rnorm(n)
f= rnorm(p,0,2/sqrt(p))
Y1 = Y + outer(l,f)
Y1.svd = svd(Y1,1,1)
f1 = flash_r1(Y1,verbose=TRUE,var_type = "constant")
plot(Y1.svd$d)
```

My thesis is that although in this case the variational lower bound
favors the null (rank 0) solution, the actual likelihood - if we could
compute it - would favor the rank 1 solution.

For investigating this it may be helpful to run flash without the null check at the end, to see what the optimimum is that it found. This is done as follows:
```{r}
f2 = flash_r1(Y1,verbose=TRUE,var_type = "constant",nullcheck = FALSE)
```





