---
title: "hierinf_toyexample"
author: "Matthew Stephens"
date: "2019-07-15"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

The aim here is to illustrate in a very simple example why the hierarchical inference approach,
e.g. as  implemented in `hierinf`, may select too many variables.

Suppose we have three highly-correlated variables $x_1,x_2,x_3$, and that $x_2,x_3$ are most highly correlated. Then the hierchical clustering
will put $x_2,x_3$ together first, and then join with $x_1$. As a consequence
the hierarchical approach cannot select just $x_1$ and $x_2$ say, even
if the data would support that selection (e.g.  because $x_1$ is the effect
variable and $x_2$ is sufficiently correlated with $x_1$ to be difficult
to distinguish from it, whereas $x_3$ is not).
The hierarchical approach can  only choose to select $x_1$ alone or all 3. On the other hand SuSiE can select
any pair of variables that is appropriate.


# Illustration

Here we provide a simple simulation that illustrates the idea.
(Note that this is not deterministic.. so a different seed may
produce a different result.)

```{r}
library("hierinf")
library("mvtnorm")
```

First simulate some data with $(x_2,x_3)$ most highly correlated, and
with $x_1$ being most correlated with $x_2$. Note that $x_1$ is the effect variable.
```{r}
set.seed(4)
n = 10000 # chosen large to reduce the variation  among simulations, although different seeds can give different answers
rho12 = 0.97
rho13 = 0.92
rho23 = 0.98
Sigma = cbind(c(1,rho12,rho13),c(rho12,1,rho23),c(rho13,rho23,1))

x = mvtnorm::rmvnorm(n, c(0,0,0),Sigma)
b = c(0.07,0,0) # this was chosen such that there is usually some  uncertainty in the variable selection
y = drop(x %*% b + rnorm(n))
```

Run `susie`:  
```{r}
s = susieR::susie(x,y)
susieR::susie_get_cs(s)
susieR::susie_get_pip(s)
```

Note that susie reports a CS with  just
variables $(x_1,x_2)$. Essentially this is because, although $x_3$ is
correlated with  $x_1$, it is sufficiently independent of $x_1$
to not be confused with it ($x_3$ has very small PIP).

This result is impossible for `hierinf` because it constrains
itself by the hierarchical  structure (which does not "really" exist here - it
is imposed by the method.) Here `hierinf` reports a significant cluster containing all 3  variables:
```{r}
colnames(x)<-1:3
h = hierinf::test_hierarchy(x,y,alpha=0.01,dendr=cluster_var(x))
h
```

