---
title: "fit Lasso by EM"
author: "Matthew Stephens"
date: "2020-06-02"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

The idea here is to implement an EM algorithm for the Lasso.
My implementation is based on [these lecture notes](http://citeseerx.ist.psu.edu/viewdoc/summary?doi=10.1.1.330.69) from M Figuerdo.


```{r}
# see objective computation for scaling of eta and residual variance s2
lasso_em = function(y,X,b.init,s2=1,eta=1,niter=100){
  n = nrow(X)
  p = ncol(X)
  b = b.init # initiolize
  XtX = t(X) %*% X
  Xty = t(X) %*% y
  obj = rep(0,niter)
  for(i in 1:niter){
    W =  diag(as.vector((1/abs(b)) * sqrt(2/eta)))
    V = chol2inv(chol(s2*W + XtX))
    b = V %*% Xty
    
    # the computation here was intended to avoid 
    # infinities in the weights by working with Winv
    # could improve this...
    # Winv_sqrt = diag(as.vector(sqrt(abs(b) * sqrt(eta/2))))
    # V = chol2inv(chol(diag(s2,p) + Winv_sqrt %*% XtX %*% Winv_sqrt))
    # b = Winv_sqrt %*% V %*% Winv_sqrt %*% Xty
    
    r =  (y - X %*% b)
    obj[i] = (1/(2*s2)) * (sum(r^2)) + sqrt(2/eta)*(sum(abs(b)))
  }
  return(list(bhat=b, obj=obj))
}
```

Here I try it out on a trend filtering example. I run twice from two different random
initializations

```{r}
set.seed(100)
sd = 1
n = 100
p = n
X = matrix(0,nrow=n,ncol=n)
for(i in 1:n){
  X[i:n,i] = 1:(n-i+1)
}
#X = X %*% diag(1/sqrt(colSums(X^2)))

btrue = rep(0,n)
btrue[40] = 8
btrue[41] = -8
y = X %*% btrue + sd*rnorm(n)

plot(y)
lines(X %*% btrue)

y.em1 = lasso_em(y,X,b.init = rnorm(100),1,1,100)
lines(X %*% y.em1$bhat,col=2)

y.em2 = lasso_em(y,X,b.init = rnorm(100),1,1,100)
lines(X %*% y.em2$bhat,col=3)


```

And check the objective is decreasing:
```{r}
plot(y.em1$obj,type="l")
lines(y.em2$obj,col=2)
```


## Bayesian Lasso

We should be able to do the same thing to fit the ``variational bayesian lasso", by which here I mean compute the (approximate) posterior mean under the double-exponential prior.

```{r}
# see objective computation for scaling of eta and residual variance s2
# if compute_mode=TRUE we have the regular LASSO
blasso_em = function(y,X,b.init,s2=1,eta=1,niter=100,compute_mode=FALSE){
  n = nrow(X)
  p = ncol(X)
  b = b.init # initiolize
  XtX = t(X) %*% X
  Xty = t(X) %*% y
  obj = rep(0,niter)
  EB = rep(1,p)
  EB2 = rep(1,p)
  
  for(i in 1:niter){
    W = as.vector(1/sqrt(EB2 + EB^2) * sqrt(2/eta))
     
    V = chol2inv(chol(XtX+ diag(s2*W))) 
    Sigma1 = s2*V  # posterior variance of b
    EB2 = diag(Sigma1)
    if(compute_mode){
      EB2 = rep(0,p)
    }
    mu1 = as.vector(V %*% Xty) # posterior mean of b
    EB = mu1
    
  }
  return(list(bhat=EB))
}
```

```{r}
y.em3 = blasso_em(y,X,b.init = rnorm(100),1,1,100)
plot(y,main="red=blasso; green = lasso")
lines(X %*% y.em3$bhat,col=2)
lines(X %*% y.em2$bhat,col=3)
plot(y.em3$bhat,y.em2$bhat, xlab="blasso",ylab="lasso")
```

Check the posterior mode is same as my Lasso implementation 
```{r}
y.em4 = blasso_em(y,X,b.init = rnorm(100),1,1,100,compute_mode=TRUE)
plot(y)
lines(X %*% y.em4$bhat,col=2)
lines(X %*% y.em2$bhat,col=3)
plot(y.em4$bhat,y.em2$bhat)
```

