---
title: "miximash"
author: "Matthew Stephens"
date: "2019-03-09"
output: workflowr::wflow_html
---

## Introduction

This is my first attempt to code an EM algorithm for
the miximash model using ash.

NOTES:
 - should edit ash to allow vloglik to be output directly

```{r}
library(ashr)
```

```{r}
#' fix point function for miximash em algorithm
#' @param w an N by K matrix of weights w
#' @param pi a K vector of mixture proportions
#' @param Bhat an N by R matrix of observations
#' @param Shat an N by R matrix of standard errors
miximash.fixpt = function(w,pi,Bhat,Shat){
  K = ncol(w)
  N = nrow(w)
  R = ncol(Bhat)
  loglik = array(0,dim=c(N,K,R)) #matrix of loglikelihoods
  
  #initialize glist
  glist = vector("list",K)
  for(k in 1:K){
    glist[[k]] = vector("list",R)
  }
      
  for(k in 1:K){
    for(r in 1:R){
      a = ashr::ash(Bhat[,r],Shat[,r],weights = w[,k],outputlevel = 1,method="shrink",mixcompdist="normal",gridmult=2^0.25)
      glist[[k]][[r]] = ashr::get_fitted_g(a)
      loglik[,k,r] = ashr::calc_vloglik(glist[[k]][[r]],ashr::set_data(Bhat[,r],Shat[,r],alpha=0))
    }
  }
  
  # compute the log-likelihood for each observation in each component
  loglik.sum = apply(loglik,c(1,2),sum) # sums across 3rd dimension, which is R
  loglik.max = apply(loglik.sum,1,max)
  loglik.sum = loglik.sum - loglik.max# avoid overflow 
  
  w = t(pi*exp(t(loglik.sum))) # an N by K matrix of weights
  w.sum = rowSums(w) # an N vector
  w = w/w.sum 
  loglik.total = sum( log(w.sum) + loglik.max)
  pi = colMeans(w)
  return(list(w=w,pi=pi,loglik=loglik.total,glist=glist))
}


#' @param loglik_array an N by K by R array
#' @param pi a K vectro
miximash.loglik = function(loglik_array,pi){
  loglik.sum = apply(loglik_array,c(1,2),sum) # sums across 3rd dimension, which is R
  offset = apply(loglik.sum,1,max)
  loglik.sum= loglik.sum - offset
  lik = exp(loglik.sum)
  sum(log(colSums(pi*t(lik)))+offset)
}

miximash = function(Bhat,Shat,w=NULL,tol=0.001,max_iter = 100){
  N = nrow(Bhat)
  R = ncol(Bhat)
  if(is.null(w)){
    w= array(runif(N*K), dim=c(N,K)) #initialize weights
  }
  pi= rep(1/K,K)

  loglik=-Inf

  state = list(w=w,pi=pi)
  for(i in 1:max_iter){
    state = miximash.fixpt(state$w,state$pi,Bhat,Shat)
    loglik = c(loglik,state$loglik)
    if( (loglik[i+1]-loglik[i]) < tol ) break;
  }
  return(list(state=state,loglik=loglik))
}



```

Simulate some data with 5 conditions.
The first group (1-50) have effects in 1-4.
The second group (51-75) have effects in 1,5.
The third group is null.

```{r}
K = 3
N = 100
R = 5
sd = 16

set.seed(1)
B = matrix(0,nrow=N,ncol=R) 
B[1:(N/2),1] = sd*rnorm((N/2))
B[1:(N/2),2] = sd*rnorm((N/2))
B[1:(N/2),3] = sd*rnorm((N/2))
B[1:(N/2),4] = sd*rnorm((N/2))

B[((N/2)+1):(3*N/4),1] = sd*rnorm((N/4))
B[((N/2)+1):(3*N/4),5] = sd*rnorm((N/4))
heatmap(B^2,Rowv=NA,Colv=NA)

Bhat = B + rnorm(N*R)
Shat = matrix(1,nrow=N,ncol=R)
```


Set up glist containing true effects for testing
```{r}
glist = vector("list",K)
for(k in 1:K){
  glist[[k]] = vector("list",R)
}

# set up true values
glist[[1]][[1]] = normalmix(c(1),c(0),c(16)) 
glist[[1]][[2]] = normalmix(c(1),c(0),c(16))
glist[[1]][[3]] = normalmix(c(1),c(0),c(16))
glist[[1]][[4]] = normalmix(c(1),c(0),c(16))
glist[[1]][[5]] = normalmix(c(1),c(0),c(0))

glist[[2]][[1]] = normalmix(c(1),c(0),c(16)) 
glist[[2]][[2]] = normalmix(c(1),c(0),c(0))
glist[[2]][[3]] = normalmix(c(1),c(0),c(0))
glist[[2]][[4]] = normalmix(c(1),c(0),c(0))
glist[[2]][[5]] = normalmix(c(1),c(0),c(16))

glist[[3]][[1]] = normalmix(c(1),c(0),c(0)) 
glist[[3]][[2]] = normalmix(c(1),c(0),c(0))
glist[[3]][[3]] = normalmix(c(1),c(0),c(0))
glist[[3]][[4]] = normalmix(c(1),c(0),c(0))
glist[[3]][[5]] = normalmix(c(1),c(0),c(0))

loglik = array(0,dim=c(N,K,R)) #matrix of loglikelihoods
  

for(k in 1:K){
    for(r in 1:R){
      loglik[,k,r] = ashr::calc_vloglik(glist[[k]][[r]],ashr::set_data(Bhat[,r],Shat[,r],alpha=0))
  }
}

loglik_oracle = miximash.loglik(loglik,c(0.5,0.25,0.25))
loglik_oracle
w_true = array(0,dim=c(N,K))
w_true[1:50,1]=1
w_true[51:75,2]=1
w_true[76:100,3]=1
```

Run miximash (twice from random; third time from truth)
```{r}
s1 = miximash(Bhat,Shat)
s2 = miximash(Bhat,Shat)
s3 = miximash(Bhat,Shat,w=w_true) #initalize from truth
```

```{r}
par(mfrow=c(1,1))
par(mai=.6*c(1,1,1,1))
plot(s3$loglik,type="l",ylim=c(min(s1$loglik[-1]),max(s3$loglik)),xlim=c(1,20))

lines(s1$loglik,col="green")
lines(s2$loglik,col="green")

abline(h=loglik_oracle,col="red")

x = seq(-10,10,length=100)
par(mfrow=c(3,5))
par(mai=.1*c(1,1,1,1))
for(k in 1:K){
  for(r in 1:R){
    plot(x,ashr::dens(s3$state$glist[[k]][[r]],x),type="l",ylim=c(0,0.4))
  }
}

```

# Test loglikelihoods

This was an initial test of the log-likelihood computations - could
be useful as a unit test...

```{r}
N= 8
K=2
R=3
glist = vector("list",K)
for(k in 1:K){
  glist[[k]] = vector("list",R)
}

# set up N(0,0) and N(0,4)
glist[[1]][[1]] = normalmix(c(1),c(0),c(0)) 
glist[[1]][[2]] = normalmix(c(1),c(0),c(0))
glist[[1]][[3]] = normalmix(c(1),c(0),c(0))

glist[[2]][[1]] = normalmix(c(1),c(0),c(4))
glist[[2]][[2]] = normalmix(c(1),c(0),c(4))
glist[[2]][[3]] = normalmix(c(1),c(0),c(4))

loglik = array(0,dim=c(N,K,R)) #matrix of loglikelihoods
  

set.seed(1)
B = matrix(0,nrow=N,ncol=R) 
B[1:(N/2),] = rnorm((N/2)*R,0,4) 

Bhat = B + rnorm(N*R)

Shat = matrix(1,nrow=N,ncol=R)

for(k in 1:K){
    for(r in 1:R){
      loglik[,k,r] = ashr::calc_vloglik(glist[[k]][[r]],ashr::set_data(Bhat[,r],Shat[,r],alpha=0))
  }
}

loglik[,1,]==dnorm(Bhat,0,1,log=TRUE)
loglik[,2,]==dnorm(Bhat,0,sqrt(17),log=TRUE)
```

