---
title: "mr.ash.changepoint"
author: "Matthew Stephens"
date: "2019-11-01"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r}
library("mr.ash.alpha")
library("glmnet")
library("genlasso")
```

## Introduction

I'm going to try the current version of `mr.ash.alpha` on a changepoint problem for my own interest.

## Single changepoint

First simulate data:
```{r}
set.seed(100)
n = 100
p = n
X = matrix(0,nrow=n,ncol=n)
for(i in 1:n){
  X[i:n,i] = 1
}

btrue = rep(0,n)
btrue[50] = 8 
Y = X %*% btrue + rnorm(n)
plot(Y)
lines(X %*% btrue)
```

This works great out of the box:
```{r}
fit.ma = mr.ash(X,Y)
plot(Y)
lines(X %*% btrue,lwd=1)
lines(predict(fit.ma,X),col=2,lwd=2)

fit.glmnet = glmnet(X,Y)
fit.glmnet.cv = glmnet::cv.glmnet(X,Y)
lines(predict(fit.glmnet.cv,X),col=3,lwd=2)

fit.genlasso = trendfilter(Y,ord=0)
fit.genlasso.cv = cv.trendfilter(fit.genlasso)
lines(fit.genlasso$fit[,which(fit.genlasso.cv$lambda==fit.genlasso.cv$lambda.min)],lwd=2,col=4)
```




## Double changepoint

Now we do a harder case, similar to the Susie paper:
```{r}
btrue[52] = -8 
Y = X %*% btrue + rnorm(n,0,0.1)
plot(Y)
lines(X %*% btrue)
```

```{r}
fit.ma = mr.ash(X,Y)
plot(Y)
lines(predict(fit.ma,X),col=2)

fit.glmnet = glmnet(X,Y)
fit.glmnet.cv = glmnet::cv.glmnet(X,Y)
lines(predict(fit.glmnet.cv,X),col=3,lwd=2)

fit.genlasso = trendfilter(Y,ord=0)
fit.genlasso.cv = cv.trendfilter(fit.genlasso)
lines(fit.genlasso$fit[,which(fit.genlasso.cv$lambda==fit.genlasso.cv$lambda.min)],lwd=2,col=4)
```

It is interesting that glmnet and genlasso give quite different answers. Clearly `glmnet` is entirely missing the changepoint. If you zoom in you will see that `genlasso` is overfitting - it is not just getting the single changepoint but fitting the data everywhere! 

Also it looks like in this case `mr.ash` is converging to local optima. (the change is a bit too early...)
Here we try warmstarts of `mr.ash` from genlasso solution, and from truth.
However, both give NaN as answers. Not sure why.


```{r}
b.genlasso  = fit.genlasso$beta[,which(fit.genlasso.cv$lambda==fit.genlasso.cv$lambda.min)]

plot(Y)
lines(b.genlasso)

fit.ma.warm = mr.ash(X,Y,beta.init = b.genlasso)
fit.ma.warm2 = mr.ash(X,Y,beta.init = btrue)


predict(fit.ma.warm,X)
predict(fit.ma.warm2,X)
```






