---
title: "mr_missing_data"
author: "Matthew Stephens"
date: "2020-03-06"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

The idea here is to look at multiple regression with missing data.

## Pseudo-data approach

Youxin Zou suggested we take a pseudo-data approach to fitting 
multivariate multiple regression (MMR). This approach unifies the full-data and
summary data approaches, and could also deal with missing data. So I wanted
to start by checking it out.

The basic MMR model can be written 
$$Y \sim MN(Xb, I, V)$$
where $MN$ denotes the matrix normal distribution.

It follows that:
$$X'Y \sim MN(X'Xb, X'X, V)$$
Now assume $X$ has SVD $X=UDV'$, so $R=X'X$ has eigendecomposition $R=UD^2U'$.
Define transform $T = D^{-1} U'$.
Then 

$$TX'Y = D^{-1}U'UDV'Y = V'Y$$
$$TX'Xb = D^{-1}U'UD^2U'b = DU'b$$
$$TX'XT' = D^{-1}U'UD^2U'UD^{-1} = I$$

So $$V'Y \sim MN(DU'b,I,V)$$

So we can fit the MMR model by using a regression model to outcome 
$$\tilde{Y} := V'Y$$
and covariate
$$\tilde{X} :=  DU'$$

With no missing data $X'Y$ is sufficient for $b$,so we should get exactly the same results (even from a Bayesian approach, because the likelihood for $b$ is the same in each case). 

Let's try this out via simulation in the univariate case (no missingness).

```{r}
library(susieR)
n= 100
p = 100
X = matrix(rnorm(n*p),nrow=n,ncol=p)
b = rep(0,p)
b[1] = 2
b[3] = 2
Y = X %*% b + rnorm(n)
Y.s = susieR::susie(X,Y,10,standardize = FALSE,intercept=FALSE)

## pseduo data
X.svd = svd(t(X))
U = X.svd$u
D = X.svd$d 
V = X.svd$v
Ytilde = t(V) %*% Y
Xtilde = diag(D) %*% t(U)
Y.stilde = susieR::susie(Xtilde,Ytilde,10,standardize = FALSE, intercept =FALSE)

all.equal(susie_get_pip(Y.s),susie_get_pip(Y.stilde))
```

Try the singular case:
```{r}
set.seed(1)
n= 20
p = 100
X = matrix(rnorm(n*p),nrow=n,ncol=p)
b = rep(0,p)
b[1] = 2
b[3] = 2
Y = X %*% b + rnorm(n)
Y.s = susieR::susie(X,Y,10,standardize = FALSE,intercept=FALSE)

# Y.s = susieR::susie(X,Y,10,standardize = FALSE,intercept=FALSE,estimate_prior_variance=FALSE, scaled_prior_variance = 1/sd(Y)^2)

## pseduo data
X.svd = svd(t(X))
U = X.svd$u
D = X.svd$d 
V = X.svd$v
Ytilde = t(V) %*% Y
Xtilde = diag(D) %*% t(U)
Y.stilde = susieR::susie(Xtilde,Ytilde,10,standardize = FALSE, intercept =FALSE)

all.equal(susie_get_pip(Y.s),susie_get_pip(Y.stilde))


```
