---
title: "mtd"
author: "Matthew Stephens"
date: "2019-04-22"
output: workflowr::wflow_html
---

## Introduction

Here we do a small simulation study to 
compare with the results of Mandozzi and Buhlmann (2016,JASA).

We consider particularly the 
"high correlation within large blocks" design which is
in some way most similar to genetic data, and also a scenario where the
existing methods do poorly. See Table 4 of MB.

```{r}
library("susieR")
```


```{r}
# simulate with 10 blocks of 20 correlated variables
simdata_hcl = function(n=100,snr=8,rho=0.99){
  x = NULL
  b = vector()
  for(i in 1:10){ 
    p=20
    z = rnorm(n,sd=sqrt(rho))
    e = matrix(rnorm(n*p,sd=sqrt(1-rho)),nrow=n)
    x = cbind(x,z + e)
    index = sample(1:p,1)
    beta = rep(0,p)
    beta[index] = (-1)^(runif(1)<0.5)
    b = c(b,beta)
  }
  
  xb = x %*% cbind(b)
  sigma = sqrt(sum(xb^2)/ (n * snr^2))
  y = xb + rnorm(n,sd=sigma)
  return(list(x=x,y=y,b=b))
}
```

## Run susie

Run susie with $L=20$ (so not cheating by telling it the true value).
Note that these simulations have very small residual variance (high PVE)
and regular susie, initialized from the null, converges to poor local optima in
such cases, so we first run susie with fixed large pve (0.99) and then use that
to initialize.

```{r}

sim_study_hcl = function(n_iter=100,rho){

  out_susie = list(n_iter=n_iter, rho=rho, nselect = rep(NA,n_iter), select_true = rep(NA,n_iter),mtd_size = NULL)

  for(i in 1:n_iter){
    dat = simdata_hcl(rho=rho)
    fit.s0 = susie(dat$x,dat$y,L=20,residual_variance = 0.01*var(dat$y),estimate_residual_variance=FALSE)
    fit.s = susie(dat$x,dat$y,L=20,s_init = fit.s0)
    cs.s = fit.s$sets$cs # susie cs's
    cs.nsig = lapply(cs.s, FUN=function(x){return(length(intersect(which(dat$b !=0),x)))}) # number of signals in each cs
    out_susie$nselect[i] = length(cs.nsig)
    out_susie$select_true[i] = sum(cs.nsig>0)
    
    mtd = cs.s[which(cs.nsig>0)]
    out_susie$mtd_size = c(out_susie$mtd_size, unlist(lapply(mtd,length)))
  }
  
  return(out_susie)
}


summarize_sim  = function(out_susie){
  hist(out_susie$nselect,breaks=seq(-0.5,20.5,length=22),main="number of CSs selected")
  
# FWER
  fwer= mean((out_susie$nselect-out_susie$select_true)>0)

  #number of mtd and sizes
  mtd = mean(out_susie$select_true)
  mtd1 = sum(out_susie$mtd_size==1)/out_susie$n_iter # number of singletons
  mtd2 = sum(out_susie$mtd_size==2)/out_susie$n_iter # number of doubletons
  mtd3.10  = sum(out_susie$mtd_size>=3 & out_susie$mtd_size<=10 )/out_susie$n_iter # number of 3-10
  mtd10.20  = sum(out_susie$mtd_size>=10 & out_susie$mtd_size<=20)/out_susie$n_iter # number of 3-10
  
  return(list(fwer= fwer, mtd = mtd,mtd1 = mtd1,  mtd2 = mtd2, mtd3.10= mtd3.10, mtd10.20 = mtd10.20))
}
```


```{r}
set.seed(1)
out_susie.99 = sim_study_hcl(n=100,rho=0.99)
s.99 = summarize_sim(out_susie.99)
```

```{r}
set.seed(1)
out_susie.95 = sim_study_hcl(n=100,rho=0.95)
s.95 = summarize_sim(out_susie.95)
```


```{r}
set.seed(1)
out_susie.9 = sim_study_hcl(n=100,rho=0.9)
s.9 = summarize_sim(out_susie.9)
```


```{r}
set.seed(1)
out_susie.0 = sim_study_hcl(n=100,rho=0)
s.0 = summarize_sim(out_susie.0)
```

## Summarize results

Here we compare with results for method "C" from Mandozzi and Buhlmann (2016,JASA), Table 4.
Note that susie does not try to control fwer so it is difficult to compare
the results in many ways: susie has more mtd and more singletons (mtd1)
but generally at cost of higher fwer. 

However,for rho=0.99 susie get better fwer and better detection.

Note also that very occassionally susie reports overlapping CSs, which is why
in some cases the number of true detections is very slightly above 10.
```{r}
mb.0 = list(fwer=0 , mtd = 10,mtd1 = 10,  mtd2 = 0, mtd3.10= 0, mtd10.20 = 0)
mb.9 = list(fwer=0 , mtd = 6.62,mtd1 = 0.06,  mtd2 = 0, mtd3.10= 0.2, mtd10.20 = 6.1)
mb.95 = list(fwer=0.05 , mtd = 9.94,mtd1 = 1.25,  mtd2 = 0, mtd3.10= 0.77, mtd10.20 = 7.92)
mb.99 = list(fwer=0.88 , mtd = 8.55,mtd1 = 3.26,  mtd2 = 0.14, mtd3.10= 2.8, mtd10.20 = 2.35)
knitr::kable(rbind(s.99,mb.99))
knitr::kable(rbind(s.95,mb.95))
knitr::kable(rbind(s.9,mb.9))
knitr::kable(rbind(s.0,mb.0))
```

