---
title: "poisson_log1p"
author: "Matthew Stephens"
date: "2024-07-02"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

I wanted to look at normal approximation to the Poisson log-likelihood with the log1p link.

The model is:
$$Y \sim Poisson(\lambda)$$

$$\log(1+\lambda) = b, (b>0)$$
So 
$$Y | b \sim Poisson(\exp(b)-1)$$
The log-likelihood is 
$$l(b) = Y \log(\exp(b)-1) - (\exp(b)-1)$$
The derivative of the log-likelihood is
$$l'(b) = Y \frac{\exp(b)}{\exp(b)-1} - \exp(b)$$
The second derivative of the log-likelihood is
$$l''(b) = Y \frac{-exp(b)}{(\exp(b)-1)^2} - \exp(b) = -exp(b)(Y/(exp(b)-1)^2 + 1)$$

Writing the RHS in terms of $\lambda = exp(b)-1$ we have
$$l''(b) = -(\lambda+1)[Y/\lambda^2 + 1]$$
Also the mle for b is $\hat{b} = log(1+Y)$

This suggests the normal approximation to the likelihood of
$l(b) \approx N(b; \hat{b}, s^2)$
where $$s^2 = -1/l''(\hat{b}) = 1/(\hat{\lambda}+1)[Y/\hat{\lambda}^2 + 1] = 1/(Y+1)(1/Y+1)=Y/(1+Y)^2 \quad (Y>0)$$
and $s^2 = 1$ for $Y=0$.


# Plots

First we plot the log-likelihood for $Y=1$ as a function of $b$.
Then we overlay the normal approximation to the log-likelihood for $Y=1$.

```{r}
b = seq(0,3,length=100)
loglik = function(b,Y) {
  if(Y==0){
    return(-(exp(b)-1))} 
  else 
    return(Y*log(exp(b)-1) - (exp(b)-1))
}

y = 1
ll = loglik(b,y)
plot(b,ll-max(ll),type="l",xlab="b",ylab="log-likelihood",main="log-likelihood for Y=1")
s2 = y/(1+y)^2
bhat = log(y+1)
ll_approx = dnorm(b,bhat,sqrt(s2),log=TRUE)
lines(b,ll_approx-max(ll_approx),col="red")
```

Do the same for $y=2$:
```{r}
y = 2
ll = loglik(b,y)
plot(b,ll-max(ll),type="l",xlab="b",ylab="log-likelihood",main="log-likelihood for Y=2")
s2 = y/(1+y)^2
bhat = log(y+1)
ll_approx = dnorm(b,bhat,sqrt(s2),log=TRUE)
lines(b,ll_approx-max(ll_approx),col="red")
```

For $Y=0$ we see a problem: the derivative of the log-likelihood does not vanish at $b=0$ (it is -1). So we need to modify the normal approximation to be $N(-1,1)$ rather than $N(0,1)$.

```{r}
b = seq(0,1,length=100)
y = 0
ll = loglik(b,y)
plot(b,ll-max(ll),type="l",xlab="b",ylab="log-likelihood",main="log-likelihood for Y=0")
s2 = 1
bhat = -1
ll_approx = dnorm(b,bhat,sqrt(s2),log=TRUE)

lines(b,ll_approx-max(ll_approx),col="red")
```




