---
title: "prec_vs_cov"
author: "Matthew Stephens"
date: "2021-01-22"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

Here I look at the difference between the precision and covariance
for ridge regression.

With suitable scaling, the precision matrix (Omega)
is $(X'X+I)$ and covariance (Sigma) is inverse of this.
I'm going to look at this when the elements of $X$ are iid N(0,1).

First I am going to check I have got the right formulae to compute the diagonal elements of Omega and Sigma from an SVD of X, by comparing with direct inversion:
```{r}
n = 10
p = 20
X = matrix(rnorm(n*p),nrow=n)

Omega = t(X) %*% X + diag(p)
Sigma = chol2inv(chol(Omega))

compute_diags = function(X){
  udv = svd(X)
  d2 = udv$d^2
  v = udv$v
  Sigma_diag = 1 - colSums((d2/(1+d2))*t(v^2))
  Omega_diag = 1 + colSums(d2*t(v^2))
  return(list(Sigma_diag=Sigma_diag,Omega_diag = Omega_diag, inv_Omega_diag = 1/Omega_diag))
}

res = compute_diags(X)

res$Sigma_diag - diag(Sigma)
res$Omega_diag - diag(Omega)
```


# Low dimension case (n>p)

If n>>p then diagonal of Sigma and inverse of diagonal of Omega are similar, and look like 1/n, independent of p.
```{r}
compare_means = function(n,p){
  X = matrix(rnorm(n*p),nrow=n)
  res = compute_diags(X)
  lapply(res,mean)
}


compare_means(1000,100)
compare_means(10000,100)


compare_means(100,20)
compare_means(1000,20)
compare_means(10000,20)
```

# High dimension case (n<p)

If p>n then the inverse of Omega_diag continues to look like 1/n
but Sigma looks like 1-(n/p). 
```{r}
compare_means(100,100)
compare_means(100,1000)
compare_means(100,10000)
compare_means(100,100000)

```


# For n=p

For n=p inv_Omega looks like 1/n, Sigma looks like 1/sqrt(p) [= 1/sqrt(n)].
```{r}
compare_means(10,10)
compare_means(20,20)
compare_means(50,50)
compare_means(100,100)
compare_means(200,200)
compare_means(500,500)
compare_means(1000,1000)

pvec = c(10,20,50,100,200,500,1000)
res = rep(0,length(pvec)) 
for(i in 1:length(pvec)){
  res[i] = compare_means(pvec[i],pvec[i])$Sigma_diag
}
plot(pvec,res)
plot(log(pvec),log(res))
lm(log(res)~log(pvec))
plot(pvec^(-0.5),res)
abline(a=0,b=1)
```
