---
title: "primepca"
author: "Matthew Stephens"
date: "2019-10-05"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

I briefly experiment with primePCA package for PCA with missing data
and compare its results with those from softImpute. To make the two comparable
I run both with no centering (set `center=FALSE` in primePCA).

```{r}
library("primePCA")
library("softImpute")
```


This first try is 50% missingness in every row, a rank 1 matrix:
```{r}
set.seed(123)
n = 100
p = 200
missprob = rep(0.5,100) #make every row have 50% missing
u = rnorm(n)
v = rnorm(p)
X = u %*% t(v) + rnorm(n*p)

for(i in 1:n){
  for(j in 1:p){
    if(runif(1)<missprob[i]){X[i,j]=NA}
  }
}
res.p = primePCA(X, 1,trace.it=TRUE,center=FALSE)
res.s = softImpute(X,1)
plot(res.s$v,res.p$V_cur,main="v from prime vs softimpute")
abline(a=0,b=1,col=2)
```

This is the same but missingness varies by row (uniform on 0,1).
```{r}
set.seed(123)
n = 100
p = 200
missprob = runif(n) #
u = rnorm(n)
v = rnorm(p)
X = u %*% t(v) + rnorm(n*p)

for(i in 1:n){
  for(j in 1:p){
    if(runif(1)<missprob[i]){X[i,j]=NA}
  }
}
res.p = primePCA(X, 1,trace.it=TRUE,center=FALSE)
res.s = softImpute(X,1)
plot(res.s$v,res.p$V_cur,main="v from prime vs softimpute")
abline(a=0,b=-1,col=2)
```


This example has higher missingness (0.8,1)
```{r}
set.seed(123)
n = 100
p = 200
missprob = 0.8+ 0.2*runif(n) #at least 80% missing
u = rnorm(n)
v = rnorm(p)
X = u %*% t(v) + rnorm(n*p)

for(i in 1:n){
  for(j in 1:p){
    if(runif(1)<missprob[i]){X[i,j]=NA}
  }
}
res.p = primePCA(X, 1,trace.it=TRUE,center=FALSE)
res.s = softImpute(X,1)
plot(res.s$v,res.p$V_cur,main="v from prime vs softimpute")
abline(a=0,b=-1,col=2)
cor(cbind(v,res.p$V_cur,res.s$v))
```


...and higher missingness again, (0.9,1). (I increased n so that every column has sufficient non-missing entries):
```{r}
set.seed(123)
n = 1000
p = 200
missprob = 0.9+ 0.1*runif(n) #at least 90% missing
u = rnorm(n)
v = rnorm(p)
X = u %*% t(v) + rnorm(n*p)

for(i in 1:n){
  for(j in 1:p){
    if(runif(1)<missprob[i]){X[i,j]=NA}
  }
}
res.p = primePCA(X, 1,trace.it=TRUE,center=FALSE)
res.s = softImpute(X,1)
plot(res.s$v,res.p$V_cur,main="v from prime vs softimpute")
abline(a=0,b=-1,col=2)
cor(cbind(v,res.p$V_cur,res.s$v))
```

Interestingly, the results from `trace.it=TRUE` in primePCA suggest it is maybe entering an infinite loop in this case. I guess that maybe this is probably because of changes in the rows selected, and indeed was able to avoid it by setting very large `thresh_sigma=1e100`. (In this case it appears to just filter out the rows with only one entry; in the other case it sometimes filters out one additional row).
