---
title: "Untitled"
author: "First Last"
date: YYYY-MM-DD
output: workflowr::wflow_html
---




Here we try fitting mash with all the UDI (Unassociated, Directly associated, indirectly associated) models. The whole thing would take a while,
so as a short cut we begin by just fitting the top signals to learn which
of those matrices are most useful.

First read in the data
```{r read}
z.old = read.table("../data/bmass.HaemgenRBC2016.Vs2.PreviousSNPs.ZScores.hclust.vs1.txt",header=TRUE,stringsAsFactors = FALSE)
z.new = read.table("../data/bmass.HaemgenRBC2016.Vs2.NewSNPs.ZScores.hclust.vs1.txt",header=TRUE,stringsAsFactors = FALSE)
z.random = read.table("../data/bmass.HaemgenRBC2016.MAF01.Vs2.MergedDataSources.200kRanSubset.ChrBPMAFMarkerZScores.vs1.txt.gz",header=TRUE,stringsAsFactors = FALSE)
z.old.m = as.matrix(z.old[,2:7])
z.new.m = as.matrix(z.new[,2:7])
z.random.m = as.matrix(z.random[,4:9])

```

Set up the data and covariance matrices for mash.
Notice we don't use the singleton matrices in the canonical ones
here because cov_udi includes these.

```{r mash_setup}
library("mashr")
library("ashr")
data = set_mash_data(z.random.m,1)
V = mashr:::estimate_null_correlation(data)
data = set_mash_data(z.random.m,1,V)

# use the old hits to set up the data-driven covariances
data.old = set_mash_data(z.old.m,1,V)
data.new = set_mash_data(z.new.m,1,V)

U.pca = cov_pca(data.old,3,subset=NULL)
U.d = cov_ed(data.old,U.pca,subset=NULL)
U.c = cov_canonical(data,cov_methods = c("identity","simple_het","equal_effects"))
U.udi = cov_udi(data)
```

```

```{r}
m.cdu.old = mashr::mash(data.old,c(U.c,U.d,U.udi))
saveRDS(m.cdu.old,file="../data/m.cdu.old.RDS")
subset.U = (get_estimated_pi(m.cdu.old)[-1]>1e-10)
Ulist = c(U.c,U.d,U.udi)[subset.U]
m.cdu = mashr::mash(data,Ulist)
```



