---
title: "Ridge hyperparameter estimation: convergence issues"
author: "Matthew Stephens"
date: "2021-02-17"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

During various simulations I found examples where EB ridge seemed to 
converge to different hyperparameters depending on initialization.
This suggested the log-likelihood might be multi-modal.
Here I investigate this.

First I experimented to find a small simulation that exhibited this behaviour. In this example the difference in final elbo is not large, but it is noticeable nonetheless about (0.15 log-likelihood unit). Increasing n and p can give examples with more extreme differences.

```{r}
  library(ebmr.alpha)
  set.seed(12)
  n <- 100
  p <- 200
  p_causal <- 100 # number of causal variables (simulated effects N(0,1))
  pve <- 0.99
  sim=list()
    sim$X =  matrix(rnorm(n*p,sd=1),nrow=n)
    B <- rep(0,p)
    causal_variables <- sample(x=(1:p), size=p_causal)
    B[causal_variables] <- rnorm(n=p_causal, mean=0, sd=1)

    sim$B = B
    sim$Y = sim$X %*% sim$B
    sigma2 = ((1-pve)/(pve))*sd(sim$Y)^2
    E = rnorm(n,sd = sqrt(sigma2))
    sim$Y = sim$Y + E
    fit.init = ebmr.init(sim$X,sim$Y,sb2 = 1, residual_variance=1 )
    fit.ebr = ebmr.update(fit.init, maxiter = 20, ebnv_fn = ebnv.pm)
    
    fit.init2 = ebmr.init(sim$X,sim$Y,sb2 = 0.1, residual_variance=0.01 )
    fit.ebr2 = ebmr.update(fit.init2, maxiter = 20, ebnv_fn = ebnv.pm)
    fit.ebr$elbo
    fit.ebr2$elbo
```



To get a better idea of what is going on I try running the EM steps directly. The code uses SVD of X to turn the problem of estimating
hyperparameters in ridge regression
 into a different problem: estimating he hyperparmaters
 in the independent measumrements model
 $$ytilde_j \sim N(0, s2 * (1+sb2 d2_j))$$
 The code in`ridge_indep_em3` fits this by EM, 
as given in more details [here](ridge_em_svd.html).

```{r}
  Xtilde = sim$X
  Xtilde.svd = svd(Xtilde)

  # maximum likelihood estimation
  ytilde = drop(t(Xtilde.svd$u) %*% sim$Y)
  df = length(sim$Y) - length(ytilde)
  ss = sum(sim$Y^2) - sum(ytilde^2)
  d2 = Xtilde.svd$d^2
  
  yt.em1 = ebmr.alpha:::ridge_indep_em3(ytilde, d2, ss, df, tol=1e-8, maxiter=10000, s2.init = 1, sb2.init = 1,  update_s2=TRUE)
  
  yt.em2 = ebmr.alpha:::ridge_indep_em3(ytilde, d2, ss, df, tol=1e-8, maxiter=10000, s2.init = .01, sb2.init = .1,  update_s2=TRUE)
  
  
  sum(dnorm(ytilde, 0, sd = sqrt(yt.em1$s2* (1+d2*yt.em1$sb2)), log=TRUE))
  sum(dnorm(ytilde, 0, sd = sqrt(yt.em2$s2* (1+d2*yt.em2$sb2)), log=TRUE))
```

## Contour of log-likelihood

Let's look at this log-likelihood surface more carefully. Here the parameters are plotted on the log scale.
```{r}
loglik = function(s2,sb2){
  sum(dnorm(ytilde, 0, sd = sqrt(s2* (1+d2*sb2)), log=TRUE))
}

n1=n2=100
ll = matrix(0,nrow=n1,ncol=n2)
ls2.seq = seq(-4.6,1.2,length=n1)
lsb2.seq = seq(-2,4.1,length=n2)
for(i in 1:n1){
  for(j in 1:n2)
    ll[i,j] = loglik(exp(ls2.seq[i]),exp(lsb2.seq[j]))
}
contour(ls2.seq,lsb2.seq,ll,levels = seq(-364.2,-364.5,length=10))

```

Since there is such a strong ridge I look at the sum and difference.
The red point indicate the solutions the 2 EM runs found.
```{r}
n1=500
n2=500
ll = matrix(0,nrow=n1,ncol=n2)
lsum = seq(-.85,-0.4,length=n1)
ldiff = seq(-6,10,length=n2)
for(i in 1:n1){
  for(j in 1:n2)
    ll[i,j] = loglik(exp(0.5*(lsum[i]-ldiff[j])),exp(0.5*(lsum[i]+ldiff[j])))
}
contour(lsum,ldiff,ll,levels = seq(-364.2,-364.5,length=30),xlab="log(sb2)+log(s2)",ylab="log(sb2)-log(s2)")

points(log(yt.em1$s2)+log(yt.em1$sb2),log(yt.em1$sb2)-log(yt.em1$s2),col=2)
points(log(yt.em2$s2)+log(yt.em2$sb2),log(yt.em2$sb2)-log(yt.em2$s2),col=2)

```

## Longer runs fix the problem

So, at least in this log(parameters) space the log-likelihood does not
appear to be multimodal. This prompted me to try running the EM longer with more stringent threshold. I found it eventually converges to the same solution, but it takes many iterations:
```{r}
yt.em1.long = ebmr.alpha:::ridge_indep_em3(ytilde, d2, ss, df, tol=1e-16, maxiter=100000, s2.init = 1, sb2.init = 1,  update_s2=TRUE)
  
yt.em2.long = ebmr.alpha:::ridge_indep_em3(ytilde, d2, ss, df, tol=1e-16, maxiter=100000, s2.init = .01, sb2.init = .1,  update_s2=TRUE)

plot(yt.em2.long$loglik[-1],xlab="iteration", ylab="loglik")
points(yt.em1.long$loglik[-1],col=2)
```



