---
title: "Experiment with simple fine mapping approach"
author: "Matthew Stephens"
date: 2017-10-17
output: html_document
---

<!-- The file analysis/chunks.R contains chunks that define default settings
shared across the workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

# Simulate some data

We will simulate data with some "LD structure". b=3 blocks of p=10 variables,
each correlated through some latent Z.
The phenotype model is $Y = X_1+...+X_b + e$ where
$X_1,\dots,X_b$ are each a member of a block, and $e$ is $N(0,sd)$.

```{r}
set.seed(1)
simulate = function(n=100,p=10,b=3,sd=1){
  Z = list()
  X = list()
  Y = list()
  for(i in 1:b) Z[[i]] = rnorm(n)
  for(i in 1:b) X[[i]] = Z[[i]] + matrix(rnorm(n*p),nrow=n)
  for(i in 1:b) Y[[i]] = X[[i]][,1] 
  
  X = do.call(cbind,X) # bind columns of X and Y
  Y = do.call(cbind,Y)
  Y = rowSums(Y) # each of the betas is 1
  
  Y = Y + rnorm(n,sd = sd)
  return(list(X=X,Y=Y))
}
d = simulate()
```

Note how the first (causal) variable in each block is more correlated than others.
```{r}
plot(cor(d$Y,d$X)[1,],main= "correlation of Y with each variable")
```


```{r}
#to run this chunk had to install varbvs
#library(devtools)
#install_github("pcarbo/varbvs",subdir = "varbvs-R")
library("varbvs")
sd.grid <- c(0,10^seq(-2,1,length.out = 19))
res = varbvsmix(d$X,Z=NULL,d$Y,sd.grid)
plot(res$lfsr)
```


Try adding three variables that are exact duplicates of the causal ones.
Notice it picks just one of the causal ones and zeros out the others.
```{r}
res2 = varbvsmix(cbind(d$X[,1],d$X[,11],d$X[,21],d$X),Z=NULL,d$Y,sd.grid)
plot(res2$lfsr)
```



## Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
