---
title: "simple_transform_simulation"
author: "Matthew Stephens"
date: "2019-05-10"
output: workflowr::wflow_html
---

## Introduction

Let's try a simple simulation to test t test on log-tranformed count data.

Specifically I simulate $X_i | s_i \sim Poi(s_i \lambda_i)$ 
for two groups where the library size $s_i$ differs between the groups
(by a factor of 10, so quite extreme)
but distribution of $\lambda_i$ is the same in each group.

Indeed, here i just fix the $\lambda_i$ to be all equal, 
to a value such that the data
have mean 1 in one group and mean 10 in the other group.

Then do transform $Y_i = \log(X_i/(s_i/median(s_i)) +  1)$


```{r}
set.seed(1)
n = 100
s = c(rep(10^5,n), rep(10^4,n))
l = rep(1/10^4,2*n)

niter = 1000
pval = rep(0,niter)
for(i in 1:niter){
  x = rpois(2*n, s*l)
  y = log(x/(s/median(s))+1)
  pval[i] = t.test(y[1:100],y[101:200])$p.value
}
hist(pval)

```

So we see the t test p values are very non-uniform.
One can see why one might worry about this....

Plot one example:
```{r}
plot(y)
mean(y[1:100])
mean(y[101:200])
```


## smaller difference in library size

Try the same thing but with only a factor 2 in library size

```{r}
set.seed(1)
n = 100
s = c(rep(10^5,n), rep(0.5*10^5,n))
l = rep(1/(0.5*10^5),2*n)

niter = 1000
pval = rep(0,niter)
for(i in 1:niter){
  x = rpois(2*n, s*l)
  y = log(x/(s/median(s))+1)
  pval[i] = t.test(y[1:100],y[101:200])$p.value
}
hist(pval)

```

## Briefly explore bias correction

(Not sure this is 100% correct... needs checking)

According to Taylor series expansion https://users.rcc.uchicago.edu/~aksarkar/singlecell-modes/transforms.html the bias should be V(x)/2(E(x) + 1)^2 where in the first
simulation x is x/(1.8) or x/0.18 in the two groups (because this is s/median(s))


```{r}
(10/1.8^2) /(2*(10/1.8+1)^2) -  (1/0.18^2) / (2*(1/0.18+1)^2)
```

So if that is right, to second order, the difference in mean of y between two groups should be 0.32 in our first simulation...so try correcting for this....

```{r}
set.seed(1)
n = 100
s = c(rep(10^5,n), rep(10^4,n))
l = rep(1/10^4,2*n)

niter = 1000
pval = rep(0,niter)
for(i in 1:niter){
  x = rpois(2*n, s*l)
  y = log(x/(s/median(s))+1)
  pval[i] = t.test(y[1:100],y[101:200]+0.32)$p.value
}
hist(pval)

```



