---
title: "Experiment with simple fine mapping approach"
author: "Matthew Stephens"
date: 2017-10-17
output: workflowr::wflow_html
---




# Simulate some data

We will simulate data with some "LD structure". b=3 blocks of p=10 variables,
each correlated through some latent Z.
The phenotype model is $Y = X_1+...+X_b + e$ where
$X_1,\dots,X_b$ are each a member of a block, and $e$ is $N(0,sd)$.

```{r}
set.seed(1)
simulate = function(n=100,p=10,b=3,sd=1){
  Z = list()
  X = list()
  Y = list()
  for(i in 1:b) Z[[i]] = rnorm(n)
  for(i in 1:b) X[[i]] = Z[[i]] + matrix(rnorm(n*p),nrow=n)
  for(i in 1:b) Y[[i]] = X[[i]][,1] 
  
  X = do.call(cbind,X) # bind columns of X and Y
  Y = do.call(cbind,Y)
  Y = rowSums(Y) # each of the betas is 1
  
  Y = Y + rnorm(n,sd = sd)
  return(list(X=X,Y=Y))
}
d = simulate()
```

Note how the first (causal) variable in each block is more correlated than others.
```{r}
plot(cor(d$Y,d$X)[1,],main= "correlation of Y with each variable")
```


```{r}
#to run this chunk had to install varbvs
#library(devtools)
#install_github("pcarbo/varbvs",subdir = "varbvs-R")
library("varbvs")
sd.grid <- c(0,10^seq(-2,1,length.out = 19))
res = varbvsmix(d$X,Z=NULL,d$Y,sd.grid)
plot(res$lfsr)
```


Try adding three variables that are exact duplicates of the causal ones.
Notice it picks just one of the causal ones and zeros out the others.
```{r}
res2 = varbvsmix(cbind(d$X[,1],d$X[,11],d$X[,21],d$X),Z=NULL,d$Y,sd.grid)
plot(res2$lfsr)
```

# Real Data

Now try the ACTN example from GTEx. These
data were received from Gao:

  - data: `/project/compbio/external_private_supp/GTExV8_HDF5`, 
  - code: http://stephenslab.github.io/gtex-eqtls/analysis/20171017_Extract_Full_Tissue_Data.html

```{r}
d = readRDS("../data/Muscle_Skeletal.ACTN3.1Mb.RDS")
storage.mode(d$X)<- "double"
d.v = varbvsmix(as.matrix(d$X),d$Z,d$y,sd.grid)
plot(d.v$lfsr)

0.739233, 0.022280
```

remove that top SNP and try again to find next snp:
```{r}
d.v2 = varbvsmix(as.matrix(d$X[,-which.min(d.v$lfsr)]),d$Z,d$y,sd.grid)
which.min(d.v2$lfsr)
```

```{r}
summary(lm(d$y ~ d$X[,2654] + d$Z))
summary(lm(d$y ~ d$X[,2656] + d$Z))
```

```{r}
logABF = function(betahat,sebetahat,sigmaa){
T = betahat/sebetahat
lambda = sebetahat^2/(sebetahat^2+sigmaa^2)
return(0.5*log(lambda) + 0.5*T^2 *(1-lambda))
}


log10ABF = function(b,s,sa){log10(exp(logABF(b,s,sa)))}

log10ABF(-0.739233,0.022280,1)
log10ABF( 0.74047 ,0.02251,1)

log10ABF(-0.739233,0.022280,.3)
log10ABF( 0.74047 ,0.02251,.3)

0.74047/0.02251
-0.739233/0.022280
```




