---
title: "Try SSVD on GTEX Z scores"
author: "Matthew Stephens"
date: 2017-12-06
output: html_document
---

<!-- The file analysis/chunks.R contains chunks that define default settings
shared across the workflowr files. -->
```{r read-chunk, include=FALSE, cache=FALSE}
knitr::read_chunk("chunks.R")
```

<!-- Update knitr chunk options -->
```{r knitr-opts-chunk, include=FALSE}
```

<!-- Insert the date the file was last updated -->
```{r last-updated, echo=FALSE, results='asis'}
```

<!-- Insert the code version (Git commit SHA1) if Git repository exists and R
 package git2r is installed -->
```{r code-version, echo=FALSE, results='asis'}
```

<!-- Add your analysis here -->

# Analyse GTEX Z scores with SSVD

Read in the top Z scores:
```{r}
Zscores = readRDS("../data/MatrixEQTLSumStats.Portable.Z.rds")
Ztop = Zscores$test.z
dim(Ztop)
```

Fit 8 and 16 factors:
```{r}
z.ssvd.8 = ssvd::ssvd(Ztop,r=8,method="theory")
z.ssvd.16 = ssvd::ssvd(Ztop,r=16,method="theory")
```

Plot results
```{r}
par(mfcol=c(2,4))
for(k in 1:8){
  barplot(z.ssvd.8$v[,k],main="SSVD (K=8)")
}

for(k in 1:16){
  barplot(z.ssvd.16$v[,k],main="SSVD (K=16)")
}
```

# Notes

- When fitting 16 factors the first 8 are the same as with 8. That
is the procedure is "greedy". 

- My first impression is that for K=16 the second set of 8 factors is
not quite as "clean" (or sparse) as the first set, or as the flash results.
I wonder if this is a consequence of imposing orthogonality?



## Session information

<!-- Insert the session information into the document -->
```{r session-info}
```
