---
title: "susie_flash"
author: "Matthew Stephens"
date: "2021-05-23"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

My idea here is to look at using flash, with
susie to solve the ebnm problem.
Susie is very sparse, so the idea is that this might
provide a good solution for very sparse factors. This is potentially
helpful because in some applications we have problems where the prior
and factors start off dense (from PCA initialization say),
and then it can't move away from that solution.

```{r}
library(susieR)
library(flashier)
```

### Try flashier

I coded up an ebnm function based on susie (which assumes only
a small number of values are non-zero).

```{r}
ebnm_susie = function(x,s, g_init,fix_g, output){
  z = x/s
  fit.susie = susieR::susie_rss(z, R = diag(length(z)))
  posterior <- list(mean = drop(s*susie_get_posterior_mean(fit.susie)), sd = drop(s*susie_get_posterior_sd(fit.susie)))
  posterior$second_moment = posterior$mean^2 + posterior$sd^2
  return(list(posterior=posterior, log_likelihood = susie_get_objective(fit.susie), fitted_g = "susie"))
}
```

Now I run it on data with 10 independent variables. So the
covariance will be close to the identity. My idea is that
it should pick up the 10 variables as sparse "factors".
(This is a bit weird compared with regular factor analysis, but
I think it should work that way in principle.)
However, it stopped after fitting one factor for reasons I don't understand yet.


```{r}
n = 100
p = 10
x = matrix(rnorm(n*p),nrow=n,ncol=p)
S = cov(x)
S.f = flashier::flash(S, prior.family = as.prior(ebnm_susie),  backfit=FALSE, var.type=2)
S.f$loadings.pm
```


### Try flashr

Here I tried `flashr` instead (which requires a different output format
for the ebnm function, unfortunately). It seems to kind of work.
```{r}
ebnm_susie2 = function(x,s, ebnm_param, output=NULL){
  z = x/s
  fit.susie = susieR::susie_rss(drop(z), R = diag(length(drop(z))))
  posterior <- list(mean = s*susie_get_posterior_mean(fit.susie), sd = s*susie_get_posterior_sd(fit.susie))
  posterior$second_moment = posterior$mean^2 + posterior$sd^2
  return(list(postmean = drop(posterior$mean), postmean2 = drop(posterior$second_moment), penloglik = drop(susie_get_objective(fit.susie)), fitted.g = NULL))
}
fit.flash = flashr::flash(S,ebnm_fn = "ebnm_susie2")
image(fit.flash$ldf$l)
image(fit.flash$ldf$f)
```


## Block structure

Now I simulate data with a strong block structure.
It finds the blocks, but it fits each block using multiple factors, presumably
because it is "over-shrinking" each factor so it needs
to add more factors to compensate.

```{r}
x1 = rnorm(n)
x2 = rnorm(n)
x = cbind(x1,x1,x1,x2,x2,x2,x2,x2)
S = cov(x)
image(S)
fit.flash = flashr::flash(S,ebnm_fn = "ebnm_susie2")
fit.flash$ldf$f

```


Try looking at rank 1 solution.. indeed it is underfitting... 
Maybe because the residual variance is too big because the signal
has not yet been removed? Maybe adding a diagonal
element could help with this? (ie Y = LDF' + E where D is a 
diagonal matrix estimated as a free parameter?)

```{r}
fit.flash = flashr::flash(S,ebnm_fn = "ebnm_susie2", Kmax=1)
flashr::flash_get_fitted_values(fit.flash)
fit.flash$ldf$l
```

