---
title: "tree_pca"
author: "Matthew Stephens"
date: "2020-07-20"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

I want to take a look at the simulated data from 
[here](http://jhmarcus.com/drift-workflow/random_init.html#introduction) 
where there is a tree structure plus admixture event,
to see how simple things like PCA behave.

Here is the code used there to simulate the data.
```{r}
set.seed(666)

n_per_pop <- 60
p <- 10000

a <- rnorm(p)
b <- rnorm(p)
c <- rnorm(p)
d <- rnorm(p, sd = 0.5)
e <- rnorm(p, sd = 0.5)
f <- rnorm(p, sd = 0.5)
g <- rnorm(p, sd = 0.5)

popA <- c(rep(1, n_per_pop), rep(0, 4 * n_per_pop))
popB <- c(rep(0, n_per_pop), rep(1, n_per_pop), rep(0, 3 * n_per_pop))
popC <- c(rep(0, 2 * n_per_pop), rep(1, n_per_pop), rep(0, 2 * n_per_pop))
popD <- c(rep(0, 3 * n_per_pop), rep(1, n_per_pop), rep(0, n_per_pop))
popE <- c(rep(0, 4 * n_per_pop), rep(1, n_per_pop))

E.factor <- (a + b + e) / 2 + (a + c + f) / 3 + (a + c + g) / 6

Y <- cbind(popA, popB, popC, popD, popE) %*% 
  rbind(a + b + d, a + b + e, a + c + f, a + c + g, E.factor)
Y <- Y + rnorm(5 * n_per_pop * p, sd = 0.1)
rownames(Y) <- c(rep("A",n_per_pop),rep("B",n_per_pop),rep("C",n_per_pop),rep("D",n_per_pop),
                 rep("E",n_per_pop))

```



## Results of SVD 

Here I do an SVD of `Y` and
plot the first two left singular vectors (which I will call the PCs for now):

```{r}
Y.svd = svd(Y)
plot(Y.svd$u[,1], main="PC1")
plot(Y.svd$u[,2], main="PC2")
```


So we can see that the second PC captures the deepest split of the tree (A,B vs C,D). 
This should be expected, at least in hindsight, as the first PC captures the mean term.

Check this by looking at the right singular vectors:
PC2 should correspond to the drift event `c-b`:
```{r}
plot(Y.svd$v[,2],c-b, main="PC2 vs c-b")
```

And PC1 should be the mean
```{r}
plot(Y.svd$v[,1],colMeans(Y), main="PC1 vs mean")
```


## Iterate SVD

Here I repeat that process hierarchically to see how it goes...

First I split on the second PC
```{r}
split =Y.svd$u[,2]>0
Y.0 = Y[!split,]
Y.1 = Y[split,]
```

Now apply svd to left and right splits, and plot. 

The left split contains only one admixed individual (E), so the second
PC nicely splits A vs B:
```{r}
Y.0.svd = svd(Y.0)
plot(Y.0.svd$u[,2],type="n")
text(Y.0.svd$u[,2],rownames(Y.0))
```

However, the right split contains most of the admixed individuals and these throw
off the PCA from splitting on C vs D (maybe not suprisingly):
```{r}
Y.1.svd = svd(Y.1)
plot(Y.1.svd$u[,2],type="n")
text(Y.1.svd$u[,2],rownames(Y.1))
```

In fact here PC3 is closer to the split we want:
```{r}
plot(Y.1.svd$u[,3],type="n")
text(Y.1.svd$u[,3],rownames(Y.1))
```

