---
title: "tree_ebcd"
author: "Matthew Stephens"
date: "2024-05-31"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

I want to understand what the EBCD solution for $Z$ looks like in the case of a tree (where factors in $L$ are linearly dependent), and compare it with the "ridgeless regression" solution (ridge regression with a very small penalty). I also look at orthogonalizing the ridgeless regression solution. It turns out that the orthogonal estimates are considerably further from the true values than the ridgeless solution.

It makes me want to reassess the possibility of doing a version of EBMF with sparse prior on L and normal prior on Z, integrating out $Z$ (jointly across all factors) by exploiting the fact that p(Z|L,X) is analytically available for a normal prior on Z.
That is, something like a variational approximation of the form q(L,Z) = q(l1)...q(lk) q(Z). I think Jason Willwerscheid 



## Background

The EBCD model is $X=ZL' + E$ where $X$ is $p \times n$, $Z$ is $p \times k$, and $L$ is $n \times k$, and $Z'Z=I_k$ and the error matrix $E$ has constant error terms.

Given $L$ the EBCD solution is $\hat{Z} = Polar.U(XL)$. 

Also, given $L$ the ridgeless regression solution is $\hat{Z} = (XL)(L'L + \lambda I)^{-1}$ with $\lambda \to 0$.

Here I compute these estimates on simulated data.

## Simulate data

```{r}
# set up L to be a tree with 4 tips and 7 branches (including top shared branch)
set.seed(1)
L = cbind(c(1,1,1,1),c(1,1,0,0),c(0,0,1,1),diag(4))
p = 1000
Z = matrix(rnorm(7*p)/sqrt(p),ncol=7)
X = Z %*% t(L)
```

Now let's compute the EBCD solution for $Z$ given $L$.
```{r}
Polar.U = function(X) {
  svdX = svd(X)
  U = svdX$u
  V = svdX$v
  Z = U %*% V
  return(Z)
}

Zhat_EBCD = Polar.U(X %*% L)
```

And the ridgeless regression solution for $Z$ given $L$. I use $lambda=1e-6$.
```{r}
Zhat_ridgeless = X %*% L %*% solve(t(L) %*% L + 1e-6*diag(7)) 
```

Now let's compare the two solutions.
```{r}
plot(Zhat_EBCD,Zhat_ridgeless)
```

We can also orthogonalize the ridgeless solution:
```{r}
plot(Zhat_EBCD,Polar.U(Zhat_ridgeless))
```

## Comparison with true $Z$

Compare the estimates with the truth, it is clear that the orthogonal solutions are not as close to the truth as the (non-orthogonal) ridgeless solution.
```{r}
plot(Z,Zhat_EBCD)
plot(Z,Zhat_ridgeless)
plot(Z,Polar.U(Zhat_ridgeless))
```

## Estimates of $L$ from estimated $Z$

The maximum likelihood estimate for $L$ is $X'Z(Z'Z)^-1$, or just $X'Z$ for the case where $Z$ is orthogonal. We can see that the estimates for $L$ from these estimated $Z$s are very different from the true $L$. 
```{r}
t(X) %*% Z # use true Z for comparison
t(X) %*% Zhat_EBCD
t(X) %*% Polar.U(Zhat_ridgeless)
```

The use of Zhat_ridgeless directly gives a good estimate of $L$ (the MLE from Zhat_ridgeless gives essentially the exact answer, presumably because it is just inverting the process that gave Zhat_ridgeless).
```{r}
t(X) %*% Zhat_ridgeless 
t(X) %*% Zhat_ridgeless %*% solve(t(Zhat_ridgeless) %*% Zhat_ridgeless + 1e-6*diag(7))
```


