---
title: "tree_pca_02"
author: "Matthew Stephens"
date: "2020-07-29"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

Following up on a [previous analysis](tree_pca.html) to look at how
PCA behaves on some tree-structure covariance matrices.


## Tree-structure covariance - 4 tips

I look at the simple 4-tip tree where the root first splits into 2 branches,
and then each branch splits into two tips.


This function produces the covariance matrix of this tree, with 
branch lengths given by `b`.
The first two elements of `b` give 
the left and right branches of the first split, etc
(There are 6 branches in all.)
```{r}
Sigma4_fun = function(b){
  Sigma4 = matrix(0,nrow=4, ncol=4) # shared branch
  Sigma4[1:2,1:2] = Sigma4[1:2,1:2] + b[1] # left
  Sigma4[3:4,3:4] = Sigma4[3:4,3:4] + b[2] # right split
  for(i in 1:4)
    Sigma4[i,i] = Sigma4[i,i] + b[2+i]
  Sigma4
}
```


### Even trees

Here we assume all the branch lengths are the same. We see that
the PCA (eigenvectors) reproduce the tree. 
```{r}
Sigma4_even = Sigma4_fun(rep(1,6))
Sigma4_even
eigen(Sigma4_even)
```

However, note that the equal eigenvalues (3,3) and (1,1)
mean that there is ambiguity about these eigenvectors. Any
linear cobination of these last two eigenvectors will also be
an eigenvector. For example:
```{r}
v=c(-1,1,-1,1)
Sigma4_even %*% v
```

This means that if the truth is a tree like this,
the representation learned could be quite prior dependent.
It makes it more attractive to make
assumptions like sparsity,  or even a hierarchical structure
on the loadings. (One long-term question is whether a sparsity
assumption is enough, or whether we really need to put in
a hierarchical assumption to recover trees.)


### The effects of centering

We can also center the covariance matrix before eigen-decomposition,
which is analogous to including a constant factor. This way each
eigenvector represents a "split" of the samples in the tree (rather
than the top two branches being put in two different eigen vectors).
```{r}
Sigma4_even_centered = Sigma4_even-mean(Sigma4_even)
eigen(Sigma4_even_centered)
```


### Balanced trees

By a (locally-)balanced tree I mean one where every split is balanced (although
not all splits are the same). We see similar results in this case to the even tree
(but this is not true for the locally-balanced 8 tip tree!)

```{r}
Sigma4_balanced = Sigma4_fun(c(1,1,2,2,3,3))
Sigma4_balanced
eigen(Sigma4_balanced)
eigen(Sigma4_balanced-mean(Sigma4_balanced))
```

### Un-balanced trees

With arbitrary branch lengths we don't see the nice exact results.
For example, the first eigen-vector, the loadings are not equal for all
the samples that split the same way.
However the general patterns are somewhat similar. 

```{r}
Sigma4_unbalanced = Sigma4_fun(c(1,2,3,4,5,6))
Sigma4_unbalanced
eigen(Sigma4_unbalanced)
eigen(Sigma4_unbalanced-mean(Sigma4_unbalanced))
```

### Un-balanced trees: correlation matrix

Part of the issue is that with arbitrary branch lengths the different
nodes do not have the same variance. Taking the correlation matrix
instead of the covariance matrix seemms to make the results for irregular trees
closer to the "regular" trees. 

```{r}
eigen(cov2cor(Sigma4_unbalanced)-mean(cov2cor(Sigma4_unbalanced)))
```

## Tree-structure covariance - 8 tips


Here I create an 8-tip tree by joining two 4-tip trees with a top splitting branch
whose branch lengths are given by `b_top`. 
```{r}
Sigma8_fun = function(b_top,b_left,b_right){
  Sigma8 = matrix(0,nrow=8,ncol=8)
  Sigma8[1:4,1:4] = b_top[1] + Sigma4_fun(b_left) # left
  Sigma8[5:8,5:8] = b_top[2] + Sigma4_fun(b_right) # right split
  Sigma8
}
```


### Balanced trees

Note that for the 8-tip case, the (locally-)balanced trees don't give such clean
results as the 4-tip case. In particular, the first eigenvector varies
among samples that split the same way. However if you take the strategy of splitting
each eigenvector at its largest gap, then I think every eigenvector splits the tree according to 
some branch...
```{r}
Sigma8_balanced = Sigma8_fun(c(1,1),c(2,2,3,3,4,4),c(5,5,6,6,7,7))
Sigma8_balanced
eigen(Sigma8_balanced)
eigen(Sigma8_balanced-mean(Sigma8_balanced))
```

Using the correlation matrix instead helps make the results cleaner and more "tree-like":
```{r}
eigen(cov2cor(Sigma8_balanced)-mean(cov2cor(Sigma8_balanced)))
```


### Add an outgroup

Here I add a 9th "outgroup" to an even 8-tip tree to see what happens.
I try eigen-decomposition of correlation matrix 
both with and without centering. In both cases
the first eigen-vector has no weight on the 9th population. This might not
get in the way of partition-based approaches because in both cases the first 
eigenvector partitions the samples 
consistent with the unrooted tree. However, it may complicate
the issue of taking account of admixed populations; how to distinguish
admixed vs an outgroup?

```{r}
Sigma8_even = Sigma8_fun(c(1,1),rep(1,6),rep(1,6))
Sigma9 = rbind(cbind(1+Sigma8_even, rep(0,8)),rep(0,9))
Sigma9[9,9] = 1
Sigma9
Sigma9_cor = cov2cor(Sigma9)
Sigma9_cor
```


```{r}
eigen(Sigma9_cor)
eigen(Sigma9_cor-mean(Sigma9_cor))
```


## Try admixture 

Here I add an admixed population that I intended to represent $X=qX_1 + (1-q)X_2$,
from which one obtains $cov(X,X_k) = q^2 cov(X_1,X_k) + (1-q)^2 cov(X_2,X_k)$.
```{r}
q = 0.5
Sigma8_even = Sigma8_fun(c(1,1),rep(1,6),rep(1,6))
Sigma9 = rbind(Sigma8_even, q^2*Sigma8_even[1,]+(1-q)^2*Sigma8_even[2,])
Sigma9 = cbind(Sigma9, c(q^2*Sigma8_even[1,]+(1-q)^2*Sigma8_even[2,], 0))
Sigma9[9,9] = q^2*Sigma8_even[1,1]+(1-q)^2*Sigma8_even[1,1] + 2*q*(1-q)*Sigma8_even[1,2]
Sigma9
cov2cor(Sigma9)

eigen(cov2cor(Sigma9)-mean(cov2cor(Sigma9)))
```

These results do not look promising. While the first 3 eigenvectors make some sense,
the fourth eigenvector picks out the admixed
population, and in fact contrasts it with the populations it is admixed with, which is not what we want.

Trying it without centering. In this case the 5th eigenvector is contrasting admixed population.
```{r}
eigen(cov2cor(Sigma9))
```

And try without taking correlation.. it still happens.
```{r}
eigen(Sigma9)
```


## Graph Laplacian

The graph laplacian for a graph with edge weights (distances) $w$ is $L=D-W$ where $D$ is a diagonal
matrix with $d_{ii} = \sum_j w_{ij}$. Cutting a graph into two can be acheived by taking the top eigenvector
of the Graph Laplacian, as [here](https://www-liebertpub-com.proxy.uchicago.edu/doi/10.1089/cmb.2009.0028)
for example.

Here we have functions to compute distance matrices for the 4-tip and 8-tip trees as above.

```{r}
D4_fun = function(b){
  D4 = matrix(0,nrow=4, ncol=4) 
  D4[1:2,3:4] = b[1] + b[2] # top split
  D4[1,-1] = D4[1,-1] + b[3]
  D4[2,-2] = D4[2,-2] + b[4]
  D4[3,-3] = D4[3,-3] + b[5]
  D4[4,-4] = D4[4,-4] + b[6]
  D4 + t(D4)
}

D8_fun = function(b_top, b_left, b_right){
  D8 = matrix(0,nrow=8, ncol=8) 
  D8[1:4,-(1:4)] = b_top[1] 
  D8[5:8,-(5:8)] = b_top[2] # top split
  
  D8[1:2,-(1:2)] = D8[1:2,-(1:2)] + b_left[1]
  D8[3:4,-(3:4)] = D8[3:4,-(3:4)] + b_left[2]
  D8[1,-1] = D8[1,-1] + b_left[3]
  D8[2,-2] = D8[2,-2] + b_left[4]
  D8[3,-3] = D8[3,-3] + b_left[5]
  D8[4,-4] = D8[4,-4] + b_left[6]
  
  D8[5:6,-(5:6)] = D8[5:6,-(5:6)] + b_right[1]
  D8[7:8,-(7:8)] = D8[7:8,-(7:8)] + b_right[2]
  D8[5,-5] = D8[5,-5] + b_right[3]
  D8[6,-6] = D8[6,-6] + b_right[4]
  D8[7,-7] = D8[7,-7] + b_right[5]
  D8[8,-8] = D8[8,-8] + b_right[6]
  
  D8 + t(D8)
  
}
```

Interestingly, in the case with even branches not even the top eigenvector is unique -- there are three equal eigenvalues. (However, all the top eigenvectors would produce a "valid" split
of the tree if you applied the strategy of splitting them at the largest gap. The second and third eigenvectors
would split into 2 vs 6 rather than 4 vs 4. Would all linear combinations of eigenvectors produce a valid split? Not sure...) 
```{r}
D8_even = D8_fun(c(1,1),rep(1,6),rep(1,6))
D8_even
L_even = diag(rowSums(D8_even)) - D8_even 
L_even
eigen(L_even)
```


### Balanced tree

```{r}
D8_balanced = D8_fun(c(1,1),c(2,2,3,3,4,4),c(5,5,6,6,7,7))
D8_balanced
eigen(diag(rowSums(D8_balanced))-D8_balanced)
```


### Outgroup

Add an outgroup to the even tree. The top eigenvector splits 5 vs 4.
```{r}
D9 = rbind(cbind(D8_even, rep(5,8)),rep(5,9))
D9[9,9] = 0
L9 = diag(rowSums(D9)) - D9 
eigen(L9)
```

Notice here that, unlike the correlation case, the length of the outgroup will matter. If we make it long
enough then the first eigenvector will split 8 vs 1.
```{r}
D9 = rbind(cbind(D8_even, rep(50,8)),rep(50,9))
D9[9,9] = 0
L9 = diag(rowSums(D9)) - D9 
eigen(L9)
```


## Relationship between distance and covariance

This section is a bit of a diversion and may be ignored...
I just wanted to remind myself about the relationship between the pairwise
distances and covariances.

Basically if $D$ is a distance matrix, and $L$ is the centering
matrix (so premultiplying by $L$ subtracts the) then
$W=-LDL'$ is symmetric positive semi-definite, and closely related to the covariance matrix. See Petkova et al for more discussion....

```{r}
D4_fun = function(b){
  D4 = matrix(0,nrow=4, ncol=4) 
  D4[1:2,3:4] = b[1] + b[2] # top split
  D4[1,-1] = D4[1,-1] + b[3]
  D4[2,-2] = D4[2,-2] + b[4]
  D4[3,-3] = D4[3,-3] + b[5]
  D4[4,-4] = D4[4,-4] + b[6]
  D4 + t(D4)
}
```

```{r}
D4_even = D4_fun(rep(1,6))
D4_even

L = diag(4) - matrix(1/4,nrow=4,ncol=4)
W = -L %*% D4_even %*% t(L)
W
0.5*W - Sigma4_even
eigen(W)
```



