---
title: "explore vicar package; t distribution"
author: "Matthew Stephens"
date: 2017-04-14
output: workflowr::wflow_html
---




# Background

I wanted to try out David's vicar package, particularly `mouthwash_secondstep`, and
see how the $t$ distribution vs normal distribution works, particularly with the variance inflation option.

# Simulate data

I start with "null" data, but with t errors.

```{r}
set.seed(1)
n=1000
bhat = rt(n,df=4) # t with 4 df
shat = rep(1,n)
library(ashr)
bhat.ash.t4 = ash(bhat,shat,df = 4)
bhat.ash.norm = ash(bhat,shat)
get_pi0(bhat.ash.norm)
get_pi0(bhat.ash.t4)
min(get_lfsr(bhat.ash.norm))
min(get_lfsr(bhat.ash.t4))
```
So we see that use of a normal likelihood with $t$ data creates "false positives",
not suprisingly.

Now I wanted to check if this also occurs in mouthwash - the idea was that
maybe the use of a variance inflation parameter in mouthwash would avoid this behaviour.
However, it appeared not to.
```{r}
library("vicar")
a = matrix(rep(1,n),nrow = n, ncol=1) # just put in an "intercept" confounder with no effect
a_seq = bhat.ash.norm$fitted_g$a
b_seq = bhat.ash.norm$fitted_g$b
lambda_seq = rep(1,length(a_seq))
lambda_seq[1] = 10
bhat.m.norm = mouthwash_second_step(bhat,shat,a,lambda_seq = lambda_seq,a_seq = a_seq, b_seq=b_seq,mixing_dist = "uniform", likelihood="normal", scale_var = FALSE, degrees_freedom = 100000) #I had to set very high df to get it to run
bhat.m.t4 = mouthwash_second_step(bhat,shat,a,lambda_seq = lambda_seq,a_seq = a_seq, b_seq=b_seq,mixing_dist = "uniform", likelihood="t", scale_var = FALSE, degrees_freedom = 4)
bhat.m.norm.c = mouthwash_second_step(bhat,shat,a,lambda_seq = lambda_seq,a_seq = a_seq, b_seq=b_seq,mixing_dist = "uniform", likelihood="normal", scale_var = TRUE, degrees_freedom = 100000) #I had to set very high df to get it to run
bhat.m.t4.c = mouthwash_second_step(bhat,shat,a,lambda_seq = lambda_seq,a_seq = a_seq, b_seq=b_seq,mixing_dist = "uniform", likelihood="t", scale_var = TRUE, degrees_freedom = 4)

bhat.m.norm$pi0
bhat.m.t4$pi0
bhat.m.norm.c$pi0
bhat.m.t4.c$pi0

min(bhat.m.norm$result$lfsr)
min(bhat.m.t4$result$lfsr)
min(bhat.m.norm.c$result$lfsr)
min(bhat.m.t4.c$result$lfsr)


```

So actually the scaling seems to make things worse here.
Actually this seems to be because the scaling parameter is estimated to be <1.

```{r}
bhat.m.norm.c$xi
bhat.m.t4.c$xi
```

What about trying the inflation penalty to help avoid it
```{r}
bhat.m.norm.c.pen = mouthwash_second_step(bhat,shat,a,lambda_seq = lambda_seq,a_seq = a_seq, b_seq=b_seq,mixing_dist = "uniform", likelihood="normal", scale_var = TRUE, var_inflate_pen = 1)
bhat.m.norm.c.pen$xi
```



